### 2021-08-29 documents/cluster_neural_population.tex
reorganize to get the dir

% load packages
\documentclass[]{article}
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{lmodern}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{url}
\usepackage{parskip}
\usepackage{graphicx}
\usepackage{subfigure}

\title{Model-based Clustering for Neural Populations}
\author{}
\date{}
\begin{document}
\maketitle

The goal for this research is to do model-based clustering for  neural populations, by making use of features for each counting process observation.

\section{Notations}

Assume we can observe neural activities for \(N\) neurons, with counting observation up to \(T\) steps. Therefore, the observation is a \textsl{N-by-T} matrix, \(\mathbf{Y} \in \mathbb{Z}_{\geq 0}^{N \times T}\) ,with each row represents the recording from single neuron. Denote the recording for neuron \(i\) as
\(\mathbf{y}_{i} = (y_{i1},\ldots,y_{\text{iT}})'\), \(i = 1,\ldots N.\), with the cluster index for neuron \(i\) as
\(z_{i} \in \{ 1,\ldots\}\). The number of neurons in cluster \(j\) is
\(n_{j} = \sum_{i = 1}^{N}{I(z_{i} = j)}\), and
\(\sum_{j = 1,2,\ldots}^{}n_{j} = N\). The proportion/ probability in
cluster \(z_{i}\) is \(\rho_{z_{i}}\).

\section{Clustering Wrapper}
The model-based clustering problem can be transformed into fitting the mixture model (MM). The likelihood for each cluster depends on how we model the counting observation, but fitting strategies for MM are the same for all models. Here, I choose to fit the MM by Gibbs sampler. Depending on whether the number of cluster is finite or not, there are two versions: finite mixture model (FMM) and Dirichlet process mixture model (DPMM).

\subsection{Finite Mixture Model}	
Assume the number of cluster is \(J\). The full likelihood for these \(N\) neurons is
\[L = \prod_{i = 1}^{N}{\rho_{z_{i}}f\left( \mathbf{y}_{i}|\mathbf{\Theta}_{z_{i}} \right)} = \prod_{j = 1}^{J}{\rho_{j}^{n_{j}}\left\lbrack \prod_{i:z_{i} = j}^{}{f\left( \mathbf{y}_{i}|\mathbf{\Theta}_{j} \right)} \right\rbrack}\]
, where \(\mathbf{\Theta}_{j}\) contains all parameters in cluster \(j\) defined by the specific model. Therefore, the parameters need to update are:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Cluster indicator: \(\left\{ z_{i} \right\}_{i = 1}^{N}\)
	\item
	Cluster proportion: \(\mathbf{\rho} = (\rho_{1},\ldots\rho_{J})'\)
	\item
	Model parameters: \(\mathbf{\Theta}_{j}\)
\end{enumerate}
The (conditional) priors for clustering-related parameters:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Cluster indicator \(\left\{ z_{i} \right\}_{i = 1}^{N}\):
	\(P\left( z_{i} = j \right) = \rho_{j}\)
	\item
	Cluster proportion \(\mathbf{\rho} = (\rho_{1},\ldots\rho_{J})'\):
	\[\mathbf{\rho} \sim Dir(\delta_{1},\ldots\delta_{J})\]
	, where \(\delta_{1} = \ldots = \delta_{J} = 1\)
\end{enumerate}

So, the MCMC( Gibbs sampler) iteration for FMM is:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Update \(\left\{ z_{i} \right\}_{i = 1}^{N}\):
	\[P\left( z_{i} = j \middle| \mathbf{y}_{i},\left\{ \mathbf{\Theta}_{j} \right\}_{j = 1}^{J} \right) \propto \rho_{j}f\left( \mathbf{y}_{i}|\mathbf{\Theta}_{j} \right)\]
	\item
	Update \(\mathbf{\rho} = \left( \rho_{1},\ldots\rho_{J} \right)^{'}\):
	\[\mathbf{\rho}|\ \left\{ \mathbf{y}_{i} \right\}_{i = 1}^{N},\ \left\{ z_{i} \right\}_{i = 1}^{N},\left\{ \mathbf{\Theta}_{j} \right\}_{j = 1}^{J} \sim Dir(\delta_{1} + n_{1},\ldots\delta_{J} + n_{J})\]
	\item
	Update \(\mathbf{\Theta}_{j}\): this is defined by the specific model. When there's no
	\(z_{i} = j\), just sample \(\mathbf{\Theta}_{j}\) from priors or by other observation-independent ways.
\end{enumerate}

\subsection{Dirichlet Process Mixture Model}

Since calculation of posterior predictive distribution can be hard or even impossible for complicated models, instead of using the popular CRP representation of DP (Neal, 2020), I choose to use the slice sampler (\href{https://www.tandfonline.com/doi/full/10.1080/03610910601096262}{Walker, 2007}).

Use the ''stick-breaking'' construction for cluster proportion, i.e.
\[\rho_{1} = \eta_{1}\]
\[\rho_{j} = \left( 1 - \eta_{1} \right) \cdot \ldots \cdot \left( 1 - \eta_{j - 1} \right)\eta_{j}\]
\[\eta_{j} \sim Beta(1,\alpha)\]

In the slice sampler for DPMM, the parameters need to update are:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	``stick-breaking'' elements: \(\eta_{j}\)
	\item
	Augment latent variable: \(\left\{ u_{i} \right\}_{i = 1}^{N}\)
	\item
	Model parameters: \(\mathbf{\Theta}_{j}\)
	\item
	Cluster indicator: \(\left\{ z_{i} \right\}_{i = 1}^{N}\)
\end{enumerate}

So, the MCMC( Gibbs sampler) iteration for DPMM is:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	update \(\eta_{j}\), for
	\(j = 1,\ldots,{{z^{*} = max}\left\{ z_{i} \right\}}_{i = 1}^{N}\) as
	\[\eta_{j}|\left\{ z_{i} \right\}_{i = 1}^{N},\ldots \sim Beta(n_{j} + 1,\ N - \sum_{l = 1}^{j}n_{l} + \alpha)\]
	\item
	update \(\left\{ u_{i} \right\}_{i = 1}^{N}\):
	\[u_{i}|\mathbf{\rho,\ldots} \sim U(0,\ \rho_{z_{i}})\]
	\item
	update \(\eta_{j}\), for \(j = z^{*} + 1,\ldots,\ s^{*}\). \(s^{*}\)
	is the smallest value, s.t.
	\(\sum_{j = 1}^{s^{*}}\rho_{j} > 1 - \min{\{ u_{1},\ldots,u_{N}\}}\)
	\[\eta_{j} \sim Beta(1,\alpha)\]
	\item
	Update \(\mathbf{\Theta}_{j}\): this is defined by the specific model. When there's no
	\(z_{i} = j\), just sample \(\mathbf{\Theta}_{j}\) from priors or by other observation-independent ways.
	\item
	Update \(\left\{ z_{i} \right\}_{i = 1}^{N}\)
	\[P\left( z_{i} = j \middle| \mathbf{y}_{i},\left\{ \mathbf{\Theta}_{j} \right\},\mathbf{\rho,}\left\{ u_{i} \right\}_{i = 1}^{N} \right) = \frac{f\left( \mathbf{y}_{i}|\mathbf{\Theta}_{j} \right)}{\sum_{j:\rho_{j} > u_{i}}^{}{f\left( \mathbf{y}_{i}|\mathbf{\Theta}_{j} \right)}}\]
\end{enumerate}

\section{linear Dynamical System Model}
Here, I model the observations by a linear dynamical system (LDS) model.

LDS models the multi-dimensional time series using a lower dimensional latent representation of the system, which evolves over time according to linear dynamics. By specifying the linear dynamics and process noise covariance, we can also handle the interactions between different neural populations (clusters).

\subsection{Model Details}
Denote the latent vector in cluster \(j\) as
\(\mathbf{x}_{t}^{(j)} \in \mathbb{R}^{p_{j}}\). For simplicity, assume all \(p_{j} = p\). Each observation follows a Poisson distribution:
\[\log\lambda_{\text{it}} = d_{i} + \mathbf{c'}_{i}\mathbf{x}_{t}^{(z_{i})}\]
\[y_{\text{it}} \sim Poisson(\lambda_{\text{it}})\]
, where \(\mathbf{c}_{i} \in \mathbb{R}^{p}\) and
\(\mathbf{x}_{t}^{(z_{i})} \in \mathbb{R}^{p}\).

Although the loading (\(d_{i}\) and \(\mathbf{c}_{i}\)) is determined by neuron index \(i\), the distribution is also cluster-dependent. That is,
\[\left( d_{i},\mathbf{c}'_{i} \right)' \sim N(\bm{\mu}_{\text{dc}}^{\left( z_{i} \right)},\mathbf{\Sigma}_{\text{dc}}^{(z_{i})})\]
By doing this, the loading within each cluster is also correlated.

Denote all latent states as \(\mathbf{x}_{t} = \left( {\mathbf{x'}_{t}^{(1)}},{\mathbf{x'}_{t}^{(2)}},\ldots \right)'\) and they evolve linearly with a Gaussian noise:
\[\mathbf{x}_{1} \sim N(\mathbf{x}_{0},\mathbf{Q}_{0})\]
\[\mathbf{x}_{t + 1}|\mathbf{x}_{t} \sim N(\mathbf{A}\mathbf{x}_{t} + \mathbf{b},\mathbf{Q})\]
For simplicity, assume \(\mathbf{Q}_{0}\) is known (e.g.
\(\mathbf{Q}_{0} = \mathbf{I} \times 10^{-2}\)).

If we assume process noise covariance is block diagonal (\href{https://papers.nips.cc/paper/2020/hash/aa1f5f73327ba40d47ebce155e785aaf-Abstract.html}{Joshua et al., 2020}), we can write things as:
\[\mathbf{x}_{t + 1}^{(j)}|\mathbf{x}_{t}^{(1)},\mathbf{x}_{t}^{(2)},\ldots \sim N(\sum_{l = 1,\ldots}^{}\mathbf{A}_{j \leftarrow l}\mathbf{x}_{t}^{(l)} + \mathbf{b}_{j},\mathbf{Q}^{(j)})\]

Notice \(\left\{ \mathbf{A}_{j \leftarrow l} \right\}\) forms the full transition matrix as:
\[\mathbf{A} = \ \begin{pmatrix}
	\mathbf{A}_{1 \leftarrow 1} & \mathbf{A}_{1 \leftarrow 2} & \ldots \\
	\mathbf{A}_{2 \leftarrow 1} & \mathbf{A}_{2 \leftarrow 2} & \ldots \\
	\ldots\  & \ldots & \ldots \\
\end{pmatrix}\]
Denote the \(j^{\text{th}}\) row block of \(\mathbf{A}\) as
\(\mathbf{A}_{j} = \begin{pmatrix}
	\mathbf{A}_{j \leftarrow 1} & \mathbf{A}_{j \leftarrow 2} & \ldots \\
\end{pmatrix}\). Then,
\(\sum_{l = 1,\ldots}^{}\mathbf{A}_{j \leftarrow l}\mathbf{x}_{t}^{(l)} + \mathbf{b}_{j}\mathbf{=}\mathbf{A}_{j}\mathbf{x}_{t} + \mathbf{b}_{j}\).

If we further let \(\mathbf{Q}\) be diagonal, with the
\(k^{\text{th}}\) row of \(\mathbf{x}_{t}\), \(\mathbf{A}\),
\(\mathbf{b}\) denoted as \(x_{\text{kt}}\), \(\mathbf{a}_{k}\), \(b_{k}\). The corresponding process noise variance is \(q_{k}\). Then:
\[x_{k,t + 1}|x_{\text{kt}} \sim N\left( \mathbf{a}_{k}^{'}\mathbf{x}_{t} + b_{k},q_{k} \right)\]

For completeness, Gibbs samplers for all three versions (no constraint, block diagonal and diagonal \(\mathbf{Q}\)) are shown below.

Since the progress noise is independent at each step,
\(f\left( \mathbf{y}_{i}|\mathbf{\Theta}_{j} \right) = \prod_{t = 1}^{T}{P(y_{\text{it}}|\mathbf{\Theta}_{j})}\),
where \(P( \cdot )\) is the Poisson density and \(\mathbf{\Theta}_{j}\) contains all parameters in cluster \(j\).

\subsection{Conditional Priors for Parameters}
The parameters need to estimate:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Latent vectors: \(\left\{ \mathbf{x}_{t} \right\}_{t=1}^T\)
	\item
	Initials: \(\mathbf{x}_{0}\)
	\item
	Linear mapping (loading) for latent vectors:
	\(\left\{ d_{i} \right\}_{i = 1}^{N}\) and
	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\)
	\item
	Mean and covariance for loading in each cluster:
	\(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and
	\(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\)
	\item
	Linear dynamics for latent vectors: \(\mathbf{A}\) and \(\mathbf{b}\)
	\item
	Process noise: \(\mathbf{Q}\)
\end{enumerate}

The conditional priors for these parameters:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Latent vectors \(\left\{ \mathbf{x}_{t} \right\}_{t=1}^T\): the conditional prior is
	defined by
	\[\mathbf{x}_{1} \sim N(\mathbf{x}_{0},\mathbf{Q}_{0})\]
	\[\mathbf{x}_{t + 1}|\mathbf{x}_{t} \sim N(\mathbf{A}\mathbf{x}_{t} + \mathbf{b},\mathbf{Q})\]
	\item
	Initials \(\mathbf{x}_{0}\): assume there are \(J\) clusters,
	\[\mathbf{x}_{0} \sim N(\bm{\mu}_{\mathbf{x}_{00}},\ \mathbf{\Sigma}_{\mathbf{x}_{00}})\]
	, where \(\bm{\mu}_{\mathbf{x}_{00}} = \mathbf{0}_{\text{Jp}}\) and \(\mathbf{\Sigma}_{\mathbf{x}_{00}} = \mathbf{I}_{\text{Jp}}\)
	\item
	Linear mapping (loading) for latent vectors
	\(\left\{ d_{i} \right\}_{i = 1}^{N}\) and
	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\):
	\[\left( d_{i},\mathbf{c}'_{i} \right)' \sim N(\bm{\mu}_{\text{dc}}^{\left( z_{i} \right)},\mathbf{\Sigma}_{\text{dc}}^{(z_{i})})\]
	\item
	Mean and covariance for loading in each cluster
	\(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and
	\(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\):
	\[\bm{\mu}_{\text{dc}}^{(j)} \sim N(\mathbf{\delta}_{dc0},\mathbf{T}_{dc0})\]
	, where \(\bm{\delta}_{dc0} = \mathbf{0}_{p + 1}\) and
	\(\mathbf{T}_{dc0} = \mathbf{I}_{p + 1}\)

	\[\mathbf{\Sigma}_{\text{dc}}^{(j)} \sim W^{- 1}\left( \Psi_{dc0},\nu_{dc0} \right)\]
	, where \(\nu_{dc0} = p + 1 + 2\) and
	\(\Psi_{dc0} = \mathbf{I}_{p + 1} \times 10^{-4}\)
	\item
	Linear dynamics for latent vectors \(\mathbf{A}\) and \(\mathbf{b}\): if the number of cluster is \(J\)
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		No constraints on \(\mathbf{Q}\): denote \(\widetilde{\mathbf{A}} = vec(\mathbf{A})\)
		\[\left( \mathbf{b}',\widetilde{\mathbf{A}} \right)' \sim N\left( \bm{\mu}_{BA_{0}},\mathbf{\Sigma}_{BA_{0}} \right)\]
		, where \(\bm{\mu}_{BA_{0}} = \left( \mathbf{0}_{\text{Jp}}',vec\left( \mathbf{I}_{\text{Jp}} \right)' \right)'\)
		and \(\mathbf{\Sigma}_{BA_{0}} = 0.25\mathbf{I}_{Jp(1 + Jp)}\)
		\item
		Assume \(\mathbf{Q}\) is block-diagonal: denote
		\({\widetilde{\mathbf{A}}}_{j} = vec\left( \mathbf{A}_{j} \right)\)
		\[\left( \mathbf{b}'_{j}, {\widetilde{\mathbf{A}}}'_{j} \right)' \sim N\left( \bm{\mu}_{bA_{0}},\mathbf{\Sigma}_{bA_{0}} \right)\]
		, where \(\bm{\mu}_{bA_{0}} = \left( \mathbf{0}'_{p + (j - 1)p^{2}}, vec\left( \mathbf{I}_{p} \right)', \mathbf{0}_{(J - j)p^{2}}' \right)'\)
		and \(\mathbf{\Sigma}_{bA_{0}} = 0.25\mathbf{I}_{p(1 + pJ)}\)
		\item
		Assume \(\mathbf{Q}\) is diagonal:
		\[\left( b_{k},\mathbf{a}'_{k} \right)' \sim N\left( \mathbf{\mu}_{ba_{k}0},\mathbf{\Sigma}_{ba0} \right)\]
		, where \(\mathbf{\mu}_{ba_{k}0} = \left( 0,\mathbf{e}'_{k} \right)'\) and \(\mathbf{\Sigma}_{ba0} = 0.25\mathbf{I}_{Jp + 1}\)	
	\end{enumerate}
	\item
	Process noise \(\mathbf{Q}\): if the number of cluster is \(J\)
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		No constraints on \(\mathbf{Q}\):
		\[\mathbf{Q} \sim W^{- 1}(\Psi_{\mathbf{Q}_{0}},\nu_{\mathbf{Q}_{0}})\]
		, where \(\nu_{\mathbf{Q}_{0}} = Jp + 2\) and
		\(\Psi_{\mathbf{Q}_{0}} = \mathbf{I}_{\text{Jp}} \times 10^{- 4}\). (To make the mean of \(\mathbf{Q}\) loosely centered around
		\(\mathbf{I}_{Jp} \times 10^{- 4}\))
		\item
		Assume \(\mathbf{Q}\) is block-diagonal:
		\[\mathbf{Q}^{(j)} \sim W^{- 1}\left( \Psi_{0},\nu_{0} \right)\]
		, where \(\nu_{0} = p + 2\) and
		\(\Psi_{0} = \mathbf{I}_{p} \times 10^{- 4}\). (To make the mean of \(\mathbf{Q}^{(j)}\) loosely centered around
		\(\mathbf{I}_{p} \times 10^{- 4}\))
		\item
		Assume \(\mathbf{Q}\) is diagonal:
		\[q_{k} \sim IG(\frac{\nu_{0}}{2},\frac{\nu_{0}\sigma_{0}^{2}}{2})\]
		, where \(\nu_{0} = 4\) and \(\sigma_{0}^{2} = 10^{- 4}\).
	\end{enumerate}
\end{enumerate}

\subsection{MCMC (Gibbs Sampler)}

\subsubsection{Update \(\left\{ \mathbf{x}_{t} \right\}_{t=1}^T\)}
Use Laplace approximation and make use of the block tri-diagonal Hessian.

Denote \(t^{\text{th}}\) column of mean firing rate and observation as \({\widetilde{\bm{\lambda}}}_{t} = \left( \lambda_{1t},\ldots,\lambda_{\text{Nt}} \right)'\)
and \({\widetilde{\mathbf{y}}}_{t} = (y_{1t},\ldots,y_{\text{Nt}})'\).
The linear mapping matrix for all observations is \(\mathbf{C}\), such that
\(\log{\widetilde{\bm{\lambda}}}_{t} = \mathbf{d} + \mathbf{C}\mathbf{x}_{t}\).
Let \(\mathbf{x} = \left( \mathbf{x'}_{1},\ldots,\mathbf{x'}_{T} \right)'\)and
\(f\left( \mathbf{x} \right) = \log{P(\mathbf{x}|\left\{ \mathbf{y}_{i} \right\}_{i = 1}^{N},\mathbf{C},\mathbf{Q}_{0},\mathbf{A},\mathbf{b},\mathbf{Q},\ldots)}\)
The first and second derivative with respect to \(\mathbf{x}\), for \(t=2, \ldots, T-1\):
\begin{align*}
	\frac{\partial f}{\partial\mathbf{x}_{1}} &= \mathbf{C'}\left( {\widetilde{\mathbf{y}}}_{1} - {\widetilde{\bm{\lambda}}}_{1} \right) - \mathbf{Q}_{0}^{- 1}\left( \mathbf{x}_{1} - \mathbf{x}_{0} \right) + \mathbf{A}'\mathbf{Q}^{- 1}\mathbf{(}\mathbf{x}_{2} - \mathbf{A}\mathbf{x}_{1} - \mathbf{b)} \\
	\frac{\partial f}{\partial\mathbf{x}_{t}} &= \mathbf{C'}\left( {\widetilde{\mathbf{y}}}_{t} - {\widetilde{\bm{\lambda}}}_{t} \right) - \mathbf{Q}^{- 1}\left( \mathbf{x}_{t} - \mathbf{A}\mathbf{x}_{t - 1} - \mathbf{b} \right) + \mathbf{A}'\mathbf{Q}^{- 1}(\mathbf{x}_{t + 1} - \mathbf{A}\mathbf{x}_{t} -\mathbf{b)}\\
	\frac{\partial f}{\partial\mathbf{x}_{T}} &= \mathbf{C'}\left( {\widetilde{\mathbf{y}}}_{T} - {\widetilde{\bm{\lambda}}}_{T} \right) - \mathbf{Q}^{- 1}\left( \mathbf{x}_{T} - \mathbf{A}\mathbf{x}_{T - 1} - \mathbf{b} \right)\\
	\frac{\partial^{2}f}{\partial\mathbf{x}_{1}\partial\mathbf{x}'_{1}} &= - \mathbf{C'}\text{Diag}\left( {\widetilde{\bm{\lambda}}}_{1} \right)\mathbf{C} - \mathbf{Q}_{0}^{- 1} - \mathbf{A}'\mathbf{Q}^{- 1}\mathbf{A}\\
	\frac{\partial^{2}f}{\partial\mathbf{x}_{t}\partial\mathbf{x}'_{t}} &= - \mathbf{C'}\text{Diag}\left( {\widetilde{\bm{\lambda}}}_{t} \right)\mathbf{C} - \mathbf{Q}^{- 1} - \mathbf{A}'\mathbf{Q}^{- 1}\mathbf{A}\\
	\frac{\partial^{2}f}{\partial\mathbf{x}_{T}\partial\mathbf{x}'_{T}} &= - \mathbf{C'}\text{Diag}\left( {\widetilde{\bm{\lambda}}}_{T} \right)\mathbf{C} - \mathbf{Q}^{- 1}\\
	\frac{\partial^{2}f}{\partial\mathbf{x}_{1}\partial\mathbf{x}'_{2}} &= \frac{\partial^{2}f}{\partial\mathbf{x}_{t}\partial\mathbf{x}'_{t + 1}} = \mathbf{A}'\mathbf{Q}^{- 1}
	&&\frac{\partial^{2}f}{\partial\mathbf{x}_{t}\partial\mathbf{x}'_{t - 1}} = \mathbf{Q}^{- 1}\mathbf{A}
\end{align*}

So, the gradient is:

\[\nabla = \frac{\partial f}{\partial\mathbf{x}} = \left( \left( \frac{\partial f}{\partial\mathbf{x}_{1}} \right)',\ \ldots,\left( \frac{\partial f}{\partial\mathbf{x}_{T}} \right)' \right)'\]

And the block tri-diagonal Hessian:

\[H = \frac{\partial^{2}f}{\partial\mathbf{x}\partial\mathbf{x}^{'}} = \begin{pmatrix}
	\frac{\partial^{2}f}{\partial\mathbf{x}_{1}\partial\mathbf{x}'_{1}} & \mathbf{A'}\mathbf{Q}^{- 1} & 0 & \cdots & 0 \\
	\mathbf{Q}^{- 1}\mathbf{A} & \frac{\partial^{2}f}{\partial\mathbf{x}_{2}\partial\mathbf{x}'_{2}} & \mathbf{A'}\mathbf{Q}^{- 1} & \cdots & \vdots \\
	0 & \mathbf{Q}^{- 1}\mathbf{A} & \frac{\partial^{2}f}{\partial\mathbf{x}_{3}\partial\mathbf{x}'_{3}} & \cdots & \vdots \\
	\vdots & \vdots & \vdots & \ddots & \vdots \\
	0 & \cdots & \cdots & \cdots & \frac{\partial^{2}f}{\partial\mathbf{x}_{T}\partial\mathbf{x}'_{T}} \\
\end{pmatrix}\]

Use Newton-Raphson to find
\(\bm{\mu}_{\mathbf{x}} =\text{argmax}_{\mathbf{x}}\left( f\left( \mathbf{x} \right) \right)\)
and \(\mathbf{\Sigma}_{\mathbf{x}}= -\left\lbrack \frac{\partial^{2}f}{\partial\mathbf{x}\partial\mathbf{x}'}\left. \ \mathbf{} \right|_{\mathbf{X} = \bm{\mu}_{\mathbf{X}}} \right\rbrack^{\mathbf{- 1}}\), such that \((P(\mathbf{x}|\left\{ \mathbf{y}_{i} \right\}_{i = 1}^{N},\ldots)\ \approx N\left(\bm{\mu}_{\mathbf{x}}, \mathbf{\Sigma}_{\mathbf{x}}\right)\).
When using Newton-Raphson (NR), \(H\backslash\nabla\) in MATLAB will make use
of block tri-diagonal structure automatically.

However, NR is not robust to bad initials. At the first few iterations, simply using fitting from previous step may lead to infinite Hessian. When the initial from previous step fails, use the approximation at recursive priors, , i.e. the adaptive smoother estimates, as the initial. The adaptive smoother estimates are from backward RTS smoother from adaptive filter, and the details about Poisson adaptive filter can be found in  \href{http://www.stat.columbia.edu/~liam/teaching/neurostat-spr11/papers/brown-et-al/eden2004.pdf}{Eden et al., 2004}.

To sample efficiently and make best use of sparse covariance, use Cholesky decomposition of
\(\mathbf{\Sigma}_{\mathbf{x}}^{- 1} = \mathbf{R}\mathbf{R}'\):
sample
\(\mathbf{Z} \sim N(\mathbf{R}'\mathbf{\mu}_{\mathbf{x}},\mathbf{I})\),
then
\(\mathbf{x} = \left( \mathbf{R}' \right)^{- 1}\mathbf{Z} \sim N(\bm{\mu}_{\mathbf{x}},\mathbf{\Sigma}_{\mathbf{x}})\).

\subsubsection{Update \(\mathbf{x}_{0}\)}
\[P\left( \mathbf{x}_{0}|\mathbf{x}_{1},\ \mathbf{Q}_{0}\ldots \right) \propto N(\mathbf{x}_{1}|\mathbf{x}_{0},\ \mathbf{Q}_{0})N(\mathbf{x}_{0}|\bm{\mu}_{\mathbf{x}_{00}},\ \mathbf{\Sigma}_{\mathbf{x}_{00}})\]
By conjugacy, \(\mathbf{x}_{0}|\mathbf{x}_{1},\ \mathbf{Q}_{0}\ldots \sim N(\bm{\mu}_{\mathbf{x}_{0}},\ \mathbf{\Sigma}_{\mathbf{x}_{0}})\)
\begin{align*}
	\mathbf{\Sigma}_{\mathbf{x}_{0}} &= \left\lbrack \mathbf{\Sigma}_{\mathbf{x}_{00}}^{- 1} + \mathbf{Q}_{0}^{- 1} \right\rbrack^{- 1}\\
	\bm{\mu}_{\mathbf{x}_{0}} &= \mathbf{\Sigma}_{\mathbf{x}_{0}}\left( \mathbf{\Sigma}_{\mathbf{x}_{00}}^{- 1}\bm{\mu}_{\mathbf{x}_{00}} + \mathbf{Q}_{0}^{- 1}\mathbf{x}_{1} \right)
\end{align*}

\subsubsection{Update \(\left\{ d_{i} \right\}_{i = 1}^{N}\) and
	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\)}
To update efficiently, use Laplace approximation again. Denote
\(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\)
and \(\left( 1,{\mathbf{x}_{t}'^{\left( z_{i} \right)}} \right) = {{\widetilde{\mathbf{x}}}_{t}'^{\left( z_{i} \right)}}\).
\begin{align*}
	P\left( \bm{\zeta}_{i} \middle| \mathbf{y}_{i},\ \left\{ \mathbf{x}_{t}^{\left( z_{i} \right)} \right\}_{t = 1}^{T},\ldots \right) &= \exp{f\left( \bm{\zeta}_{i} \right)} \approx N\left( \bm{\zeta}_{i}|\bm{\mu}_{\bm{\zeta}_{i}},\mathbf{\Sigma}_{\bm{\zeta}_{i}}\  \right)\\
	\frac{\partial f}{\partial\bm{\zeta}_{i}} &= \frac{\partial l}{\partial\bm{\zeta}_{i}} - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(z_{i})} \right) = \left\lbrack \sum_{t = 1}^{T}{\widetilde{\mathbf{x}}}_{t}^{\left( z_{i} \right)}\left( y_{\text{it}} - \lambda_{\text{it}} \right) \right\rbrack - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(z_{i})} \right)\\
	\frac{\partial^{2}f}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}} &= \frac{\partial^{2}l}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}} - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1} = -\left\lbrack \sum_{t = 1}^{T}{\lambda_{\text{it}}{\widetilde{\mathbf{x}}}_{t}^{\left( z_{i} \right)}{{\widetilde{\mathbf{x}}}_{t}'^{\left( z_{i} \right)}}} \right\rbrack - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}
\end{align*}
, where \(l\) is Poisson log-likelihood. Then use Newton-Raphson to find
\(\bm{\mu}_{\bm{\zeta}_{i}} = \text{argmax}_{\bm{\zeta}_{i}}\left( f\left( \bm{\zeta}_{i} \right) \right)\)
and \(\mathbf{\Sigma}_{\bm{\zeta}_{i}} = -\left\lbrack \frac{\partial^{2}f}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}}\left.  \right|_{\bm{\zeta}_{i} = \bm{\mu}_{\bm{\zeta}_{i}}} \right\rbrack^{-1}\). If initial as the previous step fits fails, simply use prior mean of \(\bm{\mu}_{\bm{\zeta}_{i}}\), i.e. \(\bm{\delta}_{dc0}\).

\subsubsection{Update \(\left\{ \mathbf{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and \(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\)}
As above, denote \(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\).
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Mean \(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\): by
	conjugacy, \(\bm{\mu}_{\text{dc}}^{(j)} \sim N\left( \bm{\delta}_{\text{dc}},\mathbf{T}_{\text{dc}}\  \right)\)
	\begin{align*}
		\mathbf{T}_{\text{dc}}^{- 1} &= \left( \mathbf{T}_{dc0}^{- 1} + n_{j}{\mathbf{\Sigma}_{\text{dc}}^{(j)}}^{- 1} \right)^{- 1}\\
		\bm{\delta}_{\text{dc}} &= \mathbf{T}_{\text{dc}}\left( \mathbf{T}_{dc0}^{- 1}\bm{\delta}_{dc0} + {\mathbf{\Sigma}_{\text{dc}}^{(j)}}^{- 1}\sum_{i:z_{i} = j}^{}\bm{\zeta}_{i} \right)
	\end{align*}
	\item
	Covariance \(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\): by conjugacy, \(\mathbf{\Sigma}_{\text{dc}}^{(j)} \sim W^{- 1}\left( \Psi_{\text{dc}},\nu_{\text{dc}} \right)\)
	\begin{align*}
		\nu_{\text{dc}} &= n_{j} + \nu_{dc0}\\
		\Psi_{\text{dc}} &= \Psi_{dc0} + \sum_{i:z_{i} = j}^{}{\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(j)} \right)\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(j)} \right)'}
	\end{align*}
\end{enumerate}

\subsubsection{Update \(\mathbf{A}\) and \(\mathbf{b}\)}
Assume there are \(J\) clusters.
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	No constraints on \(\mathbf{Q}\):
	\[P\left( \left( \mathbf{b}',{\widetilde{\mathbf{A}}'} \right)' \middle| \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T},\mathbf{Q},\ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}|\mathbf{A}\mathbf{x}_{t - 1} + \mathbf{b},\mathbf{Q} \right)} \right\rbrack P\left( \left( \mathbf{b}',{\widetilde{\mathbf{A}}'} \right)' \right)\]
	Let
	\(\left( \mathbf{b'},{\widetilde{\mathbf{A}}'} \right)' = \bm{\gamma} \in \mathbb{R}^{Jp(1 + Jp)}\),
	\(\left( 1,\ {\mathbf{x}'_{t - 1}} \right)\bigotimes\mathbf{I}_{\text{Jp}}\mathbf{=}{\widetilde{\mathbf{G}}}_{t - 1} \in \mathbb{R}^{Jp \times Jp(1 + pJ)}\)
	and
	\(\widetilde{\mathbf{G}} = \left( {\widetilde{\mathbf{G}}'}_{1},\ldots,{\widetilde{\mathbf{G}}'}_{T - 1} \right)'\).
	By conjugacy, \(\bm{\gamma}|\left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T},\mathbf{Q},\ldots \sim N(\bm{\mu}_{\bm{\gamma}},\mathbf{\Sigma}_{\bm{\gamma}})\)
	\begin{align*}
		\mathbf{\Sigma}_{\bm{\gamma}} &= \left\lbrack \mathbf{\Sigma}_{BA_{0}}^{- 1} + {\widetilde{\mathbf{G}}}^{\mathbf{'}}\mathbf{I}_{T - 1}\bigotimes\left( \mathbf{Q} \right)^{- 1}\widetilde{\mathbf{G}} \right\rbrack^{- 1}\\
		\bm{\mu}_{\bm{\gamma}} &= \mathbf{\Sigma}_{\bm{\gamma}}\left( \mathbf{\Sigma}_{BA_{0}}^{- 1}\bm{\mu}_{BA_{0}} + {\widetilde{\mathbf{G}}}^{'}\mathbf{I}_{T - 1}\bigotimes\left( \mathbf{Q} \right)^{- 1}\left( \mathbf{x}_{2}^{'},\ldots,\mathbf{x}_{T}^{'} \right)' \right)
	\end{align*}

	\item
	Assume \(\mathbf{Q}\) is block-diagonal:
	\[P\left( \left( \mathbf{b}'_{j}, {\widetilde{\mathbf{A}}'}_{j}\ \right)' \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\mathbf{Q}^{(j)},\ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}^{(j)}|\mathbf{A}_{j}\mathbf{x}_{t - 1} + \mathbf{b}_{j},\mathbf{Q}^{(j)} \right)} \right\rbrack P\left( \left( \mathbf{b}'_{j}, {\widetilde{\mathbf{A}}'}_{j} \right)' \right)\]
	Notice that \(\mathbf{A}_{j}\mathbf{x}_{t - 1} + \mathbf{b}_{j}\mathbf{=}\left( \left( 1,\ {\mathbf{x}'_{t - 1}} \right)\bigotimes\mathbf{I}_{p} \right)\left( \mathbf{b}_{j}^{\mathbf{'}}\mathbf{,}{\widetilde{\mathbf{A}}'}_{j} \right)'\), then the problem is reduced to Bayesian linear regression. Denote \(\left( 1,\ {\mathbf{x}'_{t - 1}} \right)\bigotimes\mathbf{I}_{p}\mathbf{=}{\widetilde{\mathbf{X}}}_{t - 1} \in \mathbb{R}^{p \times p(1 + pJ)}\),
	\(\widetilde{\mathbf{X}} = \left({\widetilde{\mathbf{X}}'}_{1},\ldots,{\widetilde{\mathbf{X}}'}_{T - 1} \right)'\)
	and\(\left( \mathbf{b}'_{j}, {\widetilde{\mathbf{A}}'}_{j} \right)' = \mathbf{\gamma}_{j} \in \mathbb{R}^{p(1 + pJ)}\).
	Then
	\begin{align*}
		\prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}^{(j)}|\mathbf{A}_{j}\mathbf{x}_{t - 1} + \mathbf{b}_{j},\mathbf{Q}^{(j)} \right)} &= \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}^{(j)}|{\widetilde{\mathbf{X}}}_{t - 1}\mathbf{\gamma}_{j},\mathbf{Q}^{(j)} \right)}\\
		&= N\left( \left( {\mathbf{x}_{2}'^{(j)}},\ldots,{\mathbf{x}_{T}'^{(j)}} \right)'|\widetilde{\mathbf{X}}\bm{\gamma}_{j},\ \mathbf{I}_{T - 1}\bigotimes\mathbf{Q}^{(j)} \right)
	\end{align*}
	By conjugacy, \(P\left( \bm{\gamma}_{j} \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\mathbf{Q}^{(j)},\ldots \right) = N\left(\bm{\gamma}_{j}|\bm{\mu}_{\bm{\gamma}_{j}},\mathbf{\Sigma}_{\bm{\gamma}_{j}} \right)\)
	\begin{align*}
		\mathbf{\Sigma}_{\bm{\gamma}_{j}} &= \left\lbrack \mathbf{\Sigma}_{bA_{0}}^{- 1} + {\widetilde{\mathbf{X}}'}\mathbf{I}_{T - 1}\bigotimes\left( \mathbf{Q}^{(j)} \right)^{- 1}\widetilde{\mathbf{X}} \right\rbrack^{- 1}\\
		\bm{\mu}_{\bm{\gamma}_{j}} &= \mathbf{\Sigma}_{\bm{\gamma}_{j}}\left( \mathbf{\Sigma}_{bA_{0}}^{- 1}\bm{\mu}_{bA_{0}} + {\widetilde{\mathbf{X}}'}\mathbf{I}_{T - 1}\bigotimes\left( \mathbf{Q}^{(j)} \right)^{- 1}\left( {\mathbf{x}_{2}'^{(j)}},\ldots,{\mathbf{x}_{T}'^{(j)}} \right)' \right)
	\end{align*}

	\item
	Assume \(\mathbf{Q}\) is diagonal:
	\[P\left( \left( b_{k},\mathbf{a}'_{k} \right)'|\left\{ x_{kt} \right\}_{t = 1}^{T},q_{k},\ \ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( x_{\text{kt}}|\mathbf{a}'_{k}\mathbf{x}_{t - 1} + b_{k},q_{k} \right)} \right\rbrack P\left( \left( b_{k},\mathbf{a}'_{k} \right)' \right)\]
	Rewrite \(\mathbf{a}'_{k}\mathbf{x}_{t - 1} + b_{k} = \left( 1,\mathbf{x}'_{t - 1} \right)\left( b_{k},\mathbf{a}'_{k} \right)'\). Then the problem is reduced to Bayesian linear regression. Denote \(\left( b_{k},\mathbf{a}'_{k} \right)' = \bm{\gamma}_{k} \in \mathbb{R}^{Jp + 1}\)
	and
	\(\widetilde{\mathbf{X}} = \left( \left( 1,\mathbf{x}'_{1} \right)',\ldots,\left( 1,\mathbf{x}'_{T - 1} \right)' \right)^{'}\). Then by conjugacy, \(P\left( \bm{\gamma}_{k}|\left\{ x_{\text{kt}} \right\}_{t = 1}^{T},q_{k},\ \ldots \right) = N\left( \bm{\gamma}_{k}|\bm{\mu}_{\bm{\gamma}_{k}},\mathbf{\Sigma}_{\bm{\gamma}_{k}} \right)\)
	\begin{align*}
		\mathbf{\Sigma}_{\bm{\gamma}_{k}} &= \left\lbrack \mathbf{\Sigma}_{ba0}^{- 1} + \frac{{\widetilde{\mathbf{X}}'}\widetilde{\mathbf{X}}}{q_{k}} \right\rbrack^{- 1}\\
		\bm{\mu}_{\bm{\gamma}_{k}} &= \mathbf{\Sigma}_{\mathbf{\gamma}_{k}}\left( \mathbf{\Sigma}_{ba0}^{- 1}\mathbf{\mu}_{ba0} + {\widetilde{\mathbf{X}}'}\left( x_{k2},\ldots,x_{\text{kT}} \right)'/q_{k} \right)
	\end{align*}	
\end{enumerate}

\subsubsection{Update \(\mathbf{Q}\)}
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	No constraints on \(\mathbf{Q}\):\\
	Let \(\mathbf{A}\mathbf{x}_{t - 1} + \mathbf{b} = \bm{\mu}_{x_{t}}\)
	\[P\left( \mathbf{Q} \middle| \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T},\mathbf{A},\mathbf{b},\ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}|\bm{\mu}_{\mathbf{x}_{t}},\ \mathbf{Q} \right)} \right\rbrack W^{- 1}(\mathbf{Q}|\Psi_{\mathbf{Q}_{0}},\nu_{\mathbf{Q}_{0}})\]
	By conjugacy,
	\[P\left( \mathbf{Q} \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\ \mathbf{A},\mathbf{b},\ldots \right) = W^{- 1}\left( \mathbf{Q}|\Psi_{0} + \sum_{t = 2}^{T}{\left( \mathbf{x}_{t} - \bm{\mu}_{\mathbf{x}_{t}} \right)\left( \mathbf{x}_{t} - \bm{\mu}_{\mathbf{x}_{t}} \right)^{'}},\ T - 1 + \nu_{0} \right)\]

	\item
	Assume \(\mathbf{Q}\) is block-diagonal:\\
	Let \(\mathbf{A}_{j}\mathbf{x}_{t - 1} + \mathbf{b}_{j}\mathbf{=}\bm{\mu}_{\mathbf{x}_{t}^{(j)}}\),
	\[P\left( \mathbf{Q}^{(j)} \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\ \mathbf{A}_{j},\mathbf{b}_{j},\ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}^{(j)}|\bm{\mu}_{\mathbf{x}_{t}^{(j)}},\mathbf{Q}^{(j)} \right)} \right\rbrack W^{- 1}\left( \mathbf{Q}^{(j)}|\Psi_{0},\nu_{0} \right)\]
	By conjugacy,
	\[P\left( \mathbf{Q}^{(j)} \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\ \mathbf{A}_{j},\mathbf{b}_{j},\ldots \right) = W^{- 1}\left( \mathbf{Q}^{(j)}|\Psi_{0} + \sum_{t = 2}^{T}{\left( \mathbf{x}_{t}^{(j)} - \bm{\mu}_{\mathbf{x}_{t}^{(j)}} \right)\left( \mathbf{x}_{t}^{(j)} - \bm{\mu}_{\mathbf{x}_{t}^{(j)}} \right)^{'}},\ T - 1 + \nu_{0} \right)\]

	\item
	Assume \(\mathbf{Q}\) is diagonal:\\
	Let \(\mathbf{a}_{k}^{'}\mathbf{x}_{t - 1} + b_{k} = \mu_{x_{\text{kt}}}\),
	\[P\left( q_{k}|\left\{ x_{\text{kt}} \right\}_{t = 1}^{T},b_{k},\mathbf{a}_{k}^{'},\ \ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( x_{\text{kt}}|\mu_{x_{\text{kt}}},q_{k} \right)} \right\rbrack\text{IG}\left( q_{k} \middle| \frac{\nu_{0}}{2},\frac{\nu_{0}\sigma_{0}^{2}}{2} \right)\]
	By conjugacy,
	\[P\left( q_{k}|\left\{ x_{\text{kt}} \right\}_{t = 1}^{T},b_{k},\mathbf{a}_{k}^{'},\ \ldots \right) = IG\left( q_{k}|\frac{\nu_{0} + T - 1}{2},\frac{\nu_{0}\sigma_{0}^{2} + \sum_{t = 2}^{T}\left( x_{\text{kt}} - \mu_{x_{\text{kt}}} \right)^{2}}{2} \right)\]
\end{enumerate}

\section{Simulations}
There are two set of simulation examples. The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics. In all of the following results, the covariance of process noise \(\mathbf{Q}\) is assumed to be block diagonal. The code for non-constrained and diagonal \(\mathbf{Q}\) can be found in XXX and XXX. The fitting results for all these three are similar, because the underlying true \(\mathbf{Q}\) is diagonal.

Before doing the clustering, I first assume the cluster labels are known to see the fitting performance. All the fitting results with labels are shown in means from iteration 50 to 100.

\subsection{Simulation 1: Generate from LDS Directly}
In this simulation, there are 3 clusters with 10 neurons in each cluster. The dimension of latents in each cluster is 2. In the linear dynamics, the bias term \(\mathbf{b}\) is zero. There's no within-population interaction but has some weak between-population interactions. In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal. The details of simulation can be found in the first section of XXX.

\subsubsection{Labeled Data: No Clustering}
The code for fitting is XXX, and some results are in Figure \ref{fig:LDS labeled}.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image001.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image002.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image003.png}}%
	}
	\caption{LDS sample with labels}
	\label{fig:LDS labeled}
\end{figure}

The fitting for overall mean firing rate is perfect and the dynamics pattern is mostly recovered. The fitting of latents is OK up to scale. In other words, the model captures the oscillating pattern of latents.

But the latent patterns of these three clusters seems quite similar. That means the differences in the mean firing rates may be purely explained by the loading, i.e. \(\mathbf{d}\) and \(\mathbf{C}\). To see that, I set the number of cluster to be 1 and the results are shown in Figure \ref{fig:LDS labeled one cluster}.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image004.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image005.png}}%
	}
	\caption{LDS sample with labels, one cluster}
	\label{fig:LDS labeled one cluster}
\end{figure}

Again, the fitting for overall mean firing rate is perfect. This is why it's important to make the loading, \(\mathbf{d}\) and \(\mathbf{C}\), also be cluster-dependent (loading within clusters are correlated). If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.

\subsubsection{Unlabeled Data: Clustering}
To give the full path of clustering, I show results in GIFs. The code for FMM and DPMM can be found in XXX and XXX.

There are three results:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Fit by FMM with \(\mathbf{J} = 3\), starting from random cluster assignments: 
	\item
	Fit by FMM with \(\mathbf{J} = 3\), starting from all neurons in single cluster: 
	\item
	Fit by DPMM (\(\alpha = X\)), starting from each neuron forms its own cluster:
\end{enumerate}
I didn't show the DPMM starting from single cluster. Since in my current implementation, the cluster generation is not efficient. In other words, the newly generated cluster will usually not be sampled and the algorithm usually gets stuck in 1 or 2 clusters.

This is a big problem... That means if we occasionally combine two clusters into one, we may never correct the mistake. Further, when the data grows, the number of cluster will be hard to grow. Fix that later.

\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
In this simulation, there are again 3 clusters with 2 latents in each. Besides set 10 neurons in each cluster, I further simulate 50 neurons in each cluster to see performance in a larger scale. Now, the latents are generated directly without specifying the underlying linear dynamics of latents. However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal. The details of simulation can be found in the first section of XXX.

\subsubsection{Labeled Data: No Clustering}
As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster. The code can be found in XXX.
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Smaller dataset (10 neurons each):\\
	The results with true cluster assignment is shown in Figure \ref{fig:no A labeled 10}
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image006.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image007.png}}%
			\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image008.png}}%
		}
		\caption{unspecified dynamics with labels, 10 neurons each}
		\label{fig:no A labeled 10}
	\end{figure}
	And results by forcing all neurons to be in single cluster (Figure \ref{fig:no A labeled one cluster 10})
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image009.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image010.png}}%
		}
		\caption{unspecified dynamics with labels, one cluster, 10 neurons each}
		\label{fig:no A labeled one cluster 10}
	\end{figure}

	Again, the fittings for overall mean firing rate are perfect in both cases and the dynamics is roughly diagonal. The oscillating pattern of fitted latents with true cluster index is not significant, because of plot scaling. The pattern is easier to see in the second fitting.

	\item
	Larger dataset (50 neurons each):\\
	The results with true cluster assignment is shown in Figure \ref{fig:no A labeled 50}
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image011.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image012.png}}%
			\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image013.png}}%
		}
		\caption{unspecified dynamics with labels, 50 neurons each}
		\label{fig:no A labeled 50}
	\end{figure}
	And results by forcing all neurons to be in single cluster (Figure \ref{fig:no A labeled one cluster 50})
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image014.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image015.png}}%
		}
		\caption{unspecified dynamics with labels, one cluster, 50 neurons each}
		\label{fig:no A labeled one cluster 50}
	\end{figure}

	Now we can see the pattern more clearly.
\end{enumerate}

\subsubsection{Unlabeled Data: Clustering}
As previous, in each example, I show results from (1) FMM starting from random cluster assignment \(\left(J = 3\right) \), (2) FMM starting from single cluster and (3) DPMM starting from multiple clusters. The DPMM starting from single cluster need to be fixed later.

\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Smaller dataset (10 neurons each):\\
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from random cluster assignments:
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from single cluster:
		\item
		Fit by DP (\(\alpha = X\)), start from assuming each neuron forms its own cluster:
	\end{enumerate}

	\item
	Larger dataset (50 neurons each):\\
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from random cluster assignments:
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from single cluster:
		\item
		Fit by DP (\(\alpha = X\)), start from random assignment for 10 clusters:
	\end{enumerate}
\end{enumerate}

The performance is not bad. The neuron at top of each cluster cannot be clustered correctly by its nature: the signals are too weak to be clearly clustered.

\section{TODO}
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
	\item
	improve DPMM
\end{enumerate}

\end{document}

### 2021-08-29 LDS/sim_example/recycle/fast_test.m

@@ -1,257 +1,0 @@
-rng(2)
-n = 10;
-nClus = 3;
-N = n*nClus;
-p = 2;
-T = 1000;
-
-Lab = repelem(1:nClus, n);
-% d = ones(n*nClus,1)*0;
-% C_all = reshape(normrnd(0.08,1e-3,n*nClus*p,1), [], p);
-
-d = randn(n*nClus,1)/5;
-C_trans = zeros(n*nClus, p*nClus);
-for k = 1:length(Lab)
-%     C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C_all(k,:);
-    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k));
-end
-
-X = zeros(p*nClus, T);
-x0 = [1.2 1.2 0.5 0.5 1 1]*10*0;
-Q0 = eye(nClus*p)*1e-2;
-X(:,1) = mvnrnd(x0, Q0)';
-
-b1 = ones(p,1)*0.01;
-b2 = ones(p,1)*0;
-b3 = ones(p,1)*-0.03;
-b = [b1;b2;b3];
-b = b*0;
-
-Q1 = 1e-3*eye(p);
-Q2 = 1e-3*eye(p);
-Q3 = 1e-3*eye(p);
-Q = blkdiag(Q1, Q2, Q3);
-
-% Generate X offline (A unspecified)
-for i=1:size(Q,1)
-    k = ceil(rand()*20)+5;
-    X(i,:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
-end
-
-% let's generate lambda
-logLam = zeros(n*nClus, T);
-logLam(:,1) = d + C_trans*X(:,1);
-
-for t=2:T
-    logLam(:, t) = d + C_trans*X(:,t);
-end
-
-Y = poissrnd(exp(logLam));
-clusterPlot(Y, Lab)
-
-subplot(1,3,1)
-plot(X(1:p,:)')
-subplot(1,3,2)
-plot(X(p+1:2*p,:)')
-subplot(1,3,3)
-plot(X(2*p+1:3*p,:)')
-
-%%
-% nClus = 2;
-% Lab = [ones(1,2*n) 2*ones(1,n)];
-rng(3)
-ng = 100;
-
-% pre-allocation
[…]

### 2021-08-29 LDS/sim_example/recycle/fast_test_cluster.m

@@ -1,188 +1,0 @@
-% rng(123)
-rng(2)
-n = 10;
-nClus = 3;
-N = n*nClus;
-p = 2;
-T = 1000;
-
-Lab = repelem(1:nClus, n);
-
-d = randn(N,1)/5;
-C_trans = zeros(n*nClus, p*nClus);
-for k = 1:length(Lab)
-%     C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C_all(k,:);
-    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k));
-end
-
-X = zeros(p*nClus, T);
-x0 = [1.2 1.2 0.5 0.5 1 1]*10*0;
-Q0 = eye(nClus*p)*1e-2;
-X(:,1) = mvnrnd(x0, Q0)';
-
-b1 = ones(p,1)*0.01;
-b2 = ones(p,1)*0;
-b3 = ones(p,1)*-0.03;
-b = [b1;b2;b3];
-b = b*0;
-
-Q1 = 1e-3*eye(p);
-Q2 = 1e-3*eye(p);
-Q3 = 1e-3*eye(p);
-Q = blkdiag(Q1, Q2, Q3);
-
-% Generate X offline (A unspecified)
-for i=1:size(Q,1)
-    k = ceil(rand()*20)+5;
-    X(i,:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
-end
-
-% let's generate lambda
-logLam = zeros(n*nClus, T);
-logLam(:,1) = d + C_trans*X(:,1);
-
-for t=2:T
-    logLam(:, t) = d + C_trans*X(:,t);
-end
-
-Y = poissrnd(exp(logLam));
-figure(11)
-clusterPlot(Y, Lab)
-
-figure(12)
-subplot(1,3,1)
-plot(X(1:p,:)')
-subplot(1,3,2)
-plot(X(p+1:2*p,:)')
-subplot(1,3,3)
-plot(X(2*p+1:3*p,:)')
-%%
-rng(2)
-ng = 20;
-kMM = 4;
-
-% pre-allocation
-Z_fit = zeros(N, ng);
-RHO_fit = zeros(kMM, ng);
-d_fit = zeros(N, ng);
-C_fit = zeros(N, p, ng);
[…]

### 2021-08-29 test/newtonTest.m

@@ -1,135 +1,0 @@
-addpath(genpath('D:\github\state-space-clustering'));
-
-%% simulation 1: simple example
-rng(1)
-C = 1;
-d = 0;
-T = 1000;
-X_true = [1*ones(1, round(T/2)) 2*ones(1, T - round(T/2))];
-lam_true = exp(X_true);
-n = poissrnd(lam_true);
-
-% try different methods
-A = 1;
-b=0;
-x0 = 1;
-W0 = 1e-1;
-Q = 1e-4;
-
-% filtering
-X_filt = ppasmoo_poissexp_v2(n,C,d,x0,W0,A,b,Q);
-
-% direct Newton-Raphson
-tic;
-der = @(vecX) derX(vecX, d, C, x0, W0, Q, A, b, n);
-hess = @(vecX) hessX(vecX, d, C, W0, Q, A, n);
-% [X_newton,fx,niSigX,~] = newton(der,hess,ones(T,1)*x0,1e-10,1000);
-[X_newton,fx,niSigX,~] = newton(der,hess,X_filt(:),1e-10,1000);
-toc;
-
-
-
-figure(1)
-hold on
-plot(X_true, 'k')
-plot(X_filt, 'r')
-plot(X_newton, 'b')
-legend('true', 'filter', 'newton')
-hold off
-
-%% simulation 2: ~ LDS
-clear all;close all;clc;
-rng(2)
-nCell = 10;
-nClus = 3;
-N = nCell*nClus;
-p = 2;
-T = 1000;
-
-Lab = repelem(1:nClus, nCell);
-d1 = ones(nCell,1)*0;
-d2 = ones(nCell,1)*1;
-d3 = ones(nCell,1)*-5;
-d = [d1;d2;d3];
-C_all1 = reshape(normrnd(0.08,1e-4,nCell*p,1), [], p);
-C_all2 = reshape(normrnd(-0.02,1e-4,nCell*p,1), [], p);
-C_all3 = reshape(normrnd(-0.18,1e-4,nCell*p,1), [], p);
-C_all = [C_all1; C_all2; C_all3];
-C_trans = zeros(nCell*nClus, p*nClus);
-for k = 1:length(Lab)
-    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C_all(k,:);
-end
-
-X = zeros(p*nClus, T);
[…]

### 2021-08-29 documents/cluster_neural_population.tex
document draft 1 complete

@@
 In all of the following results, the covariance of process noise \(\mathbf{Q}\) is assumed to be block diagonal.
-The code for non-constrained and diagonal \(\mathbf{Q}\) can be found in XXX and XXX.
+For reference, see the code for \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/old_example/sim_LDS_noConstraint.m}{non-constrained} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/old_example/sim_LDS_diag.m}{diagonal} \(\mathbf{Q}\).
+But these two legacy versions assuming the loading is independent on cluster assignments.
+If we need to use any of one of them, make sure to replace the updates for loading to be cluster-related.
 The fitting results for all these three are similar, because the underlying true \(\mathbf{Q}\) is diagonal.
@@
 In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal.
-The details of simulation can be found in the first section of XXX.
+The details of simulation can be found in the first section of the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample.m}{simulation 1}.
 
 \subsubsection{Labeled Data: No Clustering}
-The code for fitting is XXX, and some results are in Figure \ref{fig:LDS labeled}.
+The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample.m}{here}, and some results are in Figure \ref{fig:LDS labeled}.
 
@@
 \subsubsection{Unlabeled Data: Clustering}
-To give the full path of clustering, I show results in GIFs. The code for FMM and DPMM can be found in XXX and XXX.
+To give the full path of clustering, I show results in GIFs.
+Click to see the code for \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample_MM.m}{FMM} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample_DP.m}{DPMM}.
 
@@
 	\item
-	Fit by FMM with \(\mathbf{J} = 3\), starting from random cluster assignments: 
+	Fit by FMM with \(J = 3\), starting from random cluster assignments:
+	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_MM_above.gif}{GIF result 1}
 	\item
-	Fit by FMM with \(\mathbf{J} = 3\), starting from all neurons in single cluster: 
+	Fit by FMM with \(J = 3\), starting from all neurons in single cluster: 
+	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_MM_below.gif}{GIF result 2}
 	\item
-	Fit by DPMM (\(\alpha = X\)), starting from each neuron forms its own cluster:
+	Fit by DPMM (\(\alpha = 10\)), starting from each neuron forms its own cluster:
+	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_DP_above_10.gif}{GIF result 3}
+
 \end{enumerate}
@@
 However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal.
-The details of simulation can be found in the first section of XXX.
+The details of simulation can be found in the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample.m}{code}.
 
 \subsubsection{Labeled Data: No Clustering}
-As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster. The code can be found in XXX.
+As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster.
+The code can be found \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample.m}{here}.
 \begin{enumerate}
@@
 The DPMM starting from single cluster need to be fixed later.
+These are code for
+\href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample_MM.m}{FMM} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample_DP.m}{DPMM}.
 
@@
 		\item
-		Fit by FMM with \(\mathbf{J} = 3\), start from random cluster assignments:
+		Fit by FMM with \(J = 3\), start from random cluster assignments:
+		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_MM_above.gif}{GIF result 4}
 		\item
-		Fit by FMM with \(\mathbf{J} = 3\), start from single cluster:
+		Fit by FMM with \(J = 3\), start from single cluster:
+		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_MM_below.gif}{GIF result 5}
 		\item
-		Fit by DP (\(\alpha = X\)), start from assuming each neuron forms its own cluster:
+		Fit by DP (\(\alpha = 5\)), start from assuming each neuron forms its own cluster:
+		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_DP_above_5.gif}{GIF result 6}
 	\end{enumerate}
@@
 		Fit by FMM with \(\mathbf{J} = 3\), start from random cluster assignments:
+		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_above.gif}{GIF result 7}
 		\item
 		Fit by FMM with \(\mathbf{J} = 3\), start from single cluster:
+		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_below.gif}{GIF result 8}
 		\item
-		Fit by DP (\(\alpha = X\)), start from random assignment for 10 clusters:
+		Fit by DP (\(\alpha = 1\)), start from random assignment for 10 clusters:
+		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_DP_above1.gif}{GIF result 9}
 	\end{enumerate}

### 2021-08-30 documents/cluster_neural_population.tex
modify documents
add some reading lists for DPMM

@@
 		\item
-		Fit by FMM with \(\mathbf{J} = 3\), start from random cluster assignments:
+		Fit by FMM with \(J = 3\), start from random cluster assignments:
 		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_above.gif}{GIF result 7}
 		\item
-		Fit by FMM with \(\mathbf{J} = 3\), start from single cluster:
+		Fit by FMM with \(J = 3\), start from single cluster:
 		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_below.gif}{GIF result 8}
@@
 	\item
-	improve DPMM
+	improve DPMM.
+In current brute force implementation, the number of potential clusters can even go beyond number of neurons (\(N\)).
+There are several improvements, e.g.
+\href{https://link.springer.com/article/10.1007/s11222-009-9150-y}{Kalli et al., 2011}, \href{http://proceedings.mlr.press/v37/gea15.html}{Ge et al., 2015} and \href{https://link.springer.com/article/10.1007/s11222-014-9471-3}{Hastie et al, 2015}.
+Check them later.
 \end{enumerate}

### 2021-09-03 documents/cluster_neural_population.tex
problem detect

@@
 
-\subsubsection{Update \(\left\{ \mathbf{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and \(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\)}
+\subsubsection{Update \(\left\{ \mathbf{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and \(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\)} \label{loading prior}
 As above, denote \(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\).
@@
 
-\subsubsection{Update \(\mathbf{A}\) and \(\mathbf{b}\)}
-Assume there are \(J\) clusters.
+\subsubsection{Update \(\mathbf{A}\) and \(\mathbf{b}\)} \label{dynamics update}
+
 \begin{enumerate}
@@
 \section{Simulations}
+\textcolor{red}{\textbf{Currently, the latent \(x_t\) estimation is not accurate enough. This may come from bad simulation/ slow convergence or other sources. See details in section \ref{problems}}}\\
+
 There are two set of simulation examples.
@@
 
-Before doing the clustering, I first assume the cluster labels are known to see the fitting performance. All the fitting results with labels are shown in means from iteration 50 to 100.
+Before doing the clustering, I first assume the cluster labels are known to see the fitting performance.
+All the fitting results with labels are shown in means from iteration 50 to 100 (\textcolor{red}{\textbf{The iteration number is not enough to achieve stationary distribution}}).
 
@@
 The fitting for overall mean firing rate is perfect and the dynamics pattern is mostly recovered.
-The fitting of latents is OK up to scale.
+The fitting of latents is OK (\textcolor{red}{\textbf{may not be accurate enough}}\\) up to scale.
 In other words, the model captures the oscillating pattern of latents.
@@
 
+\section{PROBLEMS} \label{problems}
+This section shows what cause the inaccuracy in the latent estimation.
+
+\subsection{Estimation of \(\mathbf{d}\) and \(\mathbf{C}\) alone}
+At first, as guessed by Ian, this may be caused by bad estimation of loading (\(\mathbf{d}\) and \(\mathbf{C}\)).
+So I first turn off all others except for loading (related) parameters and set them to be true values.
+There are 2 versions: (1) update priors as in subsection \ref{loading prior} and (2) no update of priors, and just set the prior for each as the standard normal.
+For the estimation with prior update, the results are mean from iteration 50 to 100 (Figure \ref{loading alone}).
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[]{\includegraphics[width=0.6\textwidth]{image016.png}}%
+		\subfigure[]{\includegraphics[width=0.6\textwidth]{image017.png}}%
+	}
+	\caption{estimation of loading}
+	\label{loading alone}
+\end{figure}
+
+The mixing of convergence for loading priors (in subsection \ref{loading prior}) are fine.
+The update of priors don't influence results a lot, but this will help clustering a lot.
+It seems estimation of loading is fine.
+Let's see what happens when estimation of latents \(\mathbf{x}_t\) is on at the same time.
+
+\subsection{Estimation of \(\mathbf{d}\), \(\mathbf{C}\) and \(\mathbf{x}_{t}\)}
+When the estimation of \(\mathbf{x}_t\) is on, we need more iterations. The results are mean from iteration 500 to 1000 (Figure \ref{loading and latent}).
+
+The estimation of loading:
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[]{\includegraphics[width=0.6\textwidth]{image018.png}}%
+		\subfigure[]{\includegraphics[width=0.6\textwidth]{image020.png}}%
+	}
+	\caption{estimation of loading, with latents on}
+	\label{loading and latent}
+\end{figure}
+
+And the estimation of corresponding latents (Figure \ref{loading and latents}):
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[no update of prior]{\includegraphics[width=0.6\textwidth]{image019.png}}%
+		\subfigure[with update of prior]{\includegraphics[width=0.6\textwidth]{image021.png}}%
+	}
+	\caption{estimation of loading}
+	\label{loading and latents}
+\end{figure}
+
+Things are still fine. The remaining parts may influence latent estimations are dynamics (\(\mathbf{b}\), \(\mathbf{A}\)) and prior process noise \(\mathbf{Q}\).
+
+I've checked update of (1) dynamics alone, (2) \(\mathbf{Q}\) and (3) \(\mathbf{Q}\) and \(\mathbf{x}_{t}\).
+All of them are fine.
+\textbf{So the problem must come from estimations of dynamics and \(\mathbf{x}_t\)}.
+
+\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\) and \(\mathbf{x}_t\)}
+I first estimate dynamics by assuming block diagonal \(\mathbf{Q}\) (as in subsection \ref{dynamics update}).
+When turning on (1) latents \(\mathbf{x}_t\) and (2) dynamics (\(\mathbf{b}\) and \(\mathbf{A}\)), the trace of \(\mathbf{b}\) and \(\mathbf{A}\) are (Figure \ref{trace of dynamics, blk}):\\
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width = .7\textwidth]{image022.png}
+	\caption{trace of dynamics, block diagonal}
+	\label{trace of dynamics, blk}
+\end{figure}
+
+The mixing is not good, and we may haven't achieve stationary distribution yet. Anyway, I average results from iteration 5000 to 10000, and the results are shown in Figure \ref{lat and dyn, blk}).
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[latent]{\includegraphics[width=0.6\textwidth]{image023.png}}%
+		\subfigure[dynamics]{\includegraphics[width=0.6\textwidth]{image024.png}}%
+	}
+	\caption{estimation of latents and dynamics, block diagonal}
+	\label{lat and dyn, blk}
+\end{figure}
+
+OK, things are not perfect now.
+What if I don't assume block diagonal \(\mathbf{Q}\)?
+For the non-constraint version, the trace plot of \(\mathbf{b}\) and \(\mathbf{A}\) are in Figure \ref{trace of dynamics, full}.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width = .7\textwidth]{image025.png}
+	\caption{trace of dynamics, no constraint}
+	\label{trace of dynamics, full}
+\end{figure}\\
+
+Again, the average from iteration 5000 to 10000 is in Figure \ref{lat and dyn, full}).\\
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[latent]{\includegraphics[width=0.6\textwidth]{image026.png}}%
+		\subfigure[dynamics]{\includegraphics[width=0.6\textwidth]{image027.png}}%
+	}
+	\caption{estimation of latents and dynamics, no constraint}
+	\label{lat and dyn, full}
+\end{figure}
+
+Based on trace plot and latent, things look a bit better. But it's still not good enough. The dynamics is even worse, although the latent is better.
+
+\textbf{Another Observation}:
+Even with these inaccurate latent, the overall mean firing rate still looks perfect (with true loading \(\mathbf{d}\) and \(\mathbf{C}\)), because of small loading. See Figure \ref{mfr, blk and full}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[block diagonal]{\includegraphics[width=0.6\textwidth]{image028.png}}%
+		\subfigure[no constraint]{\includegraphics[width=0.6\textwidth]{image029.png}}%
+	}
+	\caption{overall mean firing rate, block diagonal and no constraint}
+	\label{mfr, blk and full}
+\end{figure}
+
+In PLDS, they use EM, so there's no noise. But here I tackle things by MCMC: if the simulation is not very identifiable, then there will be a lot of noise by its nature.
+
+Anyway, the clustering is based on the overall mean firing rate pattern. Although the estimation of latent may not be accurate enough, this should not influence the clustering result.
+
+Maybe the next step is to test the current algorithm by a more identifiable simulation. For example, use a larger loading to make the latent effect become more significant.
+
+\subsection{Simulation 1 Revisit}
+The results above remind me that the 100 iterations is not enough to achieve stationary distribution. I need to check the sampling trace.
+Here, I rerun the MCMC for simulation 1, but use the no-constraint \(\mathbf{Q}\) version. Since no constraint version is better than block-diagonal one.
+
+First, I show the trace of sampling (Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\) as in Figure \ref{norms}.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width = .8\textwidth]{image030.png}
+	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
+	\label{norms}
+\end{figure}\\
+
+It seems we may have not achieved stationary yet. And I show the average from iteration 8000 to 10000 as in Figure \ref{fig:LDS labeled revisit}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image031.png}}%
+		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image032.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image033.png}}%
+	}
+	\caption{LDS sample with labels}
+	\label{fig:LDS labeled revisit}
+\end{figure}
+
+Things look much better! That means we need to improve the mixing of \(\mathbf{b}\) and \(\mathbf{A}\) later. (Also see if the mixing will be better for a more identifiable simulation)
+
+\clearpage
 \section{TODO}
@@
 	\def\labelenumi{(\arabic{enumi})}
+	\item
+	Use a more identifiable simulation, e.g. larger loading, to see the performance
+	\item
+	Currently, the mixing of \(\mathbf{b}\) and \(\mathbf{A}\) is terrible, and I need find a way to break the dependence between dynamics (\(\mathbf{b}\) and \(\mathbf{A}\)) and latent (\(\mathbf{x}_t\)).
+For example, see if it's possible to do collapsed Gibbs sampler?
+Or switch to EM as in PLDS?
+If I switch to EM, then the DPMM should also be done in EM.
+(That would be more challenging).
+\\
+
+	Basically the question is
+	\[y_{\text{it}} \sim Poisson(\lambda_{\text{it}})\]
+	\[\log\lambda_{\text{it}} = d_{i} + \mathbf{c'}_{i}\mathbf{x}_{t}^{(z_{i})}\]
+	\[\mathbf{x}_{1} \sim N(\mathbf{x}_{0},\mathbf{Q}_{0})\]
+	\[\mathbf{x}_{t + 1}|\mathbf{x}_{t} \sim N(\mathbf{A}\mathbf{x}_{t} + \mathbf{b},\mathbf{Q})\]
+	And we need to sample \(\mathbf{x}_t\), \(\mathbf{b}\), \(\mathbf{A}\) and \(\mathbf{Q}\) efficiently. (Assume loading is known now).
+
+	Or even a more basic question: how to efficiently sample state-space and transition matrix at the same time in Kalman filter?
+The dependency between transition matrix and state vectors causes bad mixing.
+
+	\item
+	Check sampling trace and mixing.
 	\item

### 2021-09-03 LDS/debug/debug_bA.m

@@ -1,0 +1,114 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(N,1);
+C_trans = zeros(N, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(N, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+ng = 2;
+idx = 2;
+
+%% fitting: MCMC
+rng(3)
+X_fit = repmat(X, 1, 1, ng); % true
+x0_fit = repmat(x0,1,ng); %true
+d_fit = repmat(d,1,ng); % true
+C_fit = repmat([sum(C_trans(:,1:p:end), 2) sum(C_trans(:,2:p:end), 2)],...
+    1,1,ng); % true
[…]

### 2021-09-03 LDS/debug/debug_bA_X.m

@@ -1,0 +1,207 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(N,1);
+C_trans = zeros(N, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(N, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+ng = 10000;
+
+
+%% fitting: MCMC
+rng(3)
+X_fit = zeros(nClus*p, T, ng);
+x0_fit = zeros(nClus*p, ng);
+d_fit = repmat(d,1,ng); % true
+C_fit = repmat([sum(C_trans(:,1:p:end), 2) sum(C_trans(:,2:p:end), 2)],...
[…]

### 2021-09-03 LDS/debug/debug_bA_X_full.m

@@ -1,0 +1,252 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+
+%% simulation
+rng(1)
+n = 40;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(N,1);
+C_trans = zeros(N, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(N, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+ng = 1000;
+% idx = 1000:ng;
+
+%% fitting: MCMC
+rng(3)
+X_fit = zeros(nClus*p, T, ng);
+x0_fit = zeros(nClus*p, ng);
+d_fit = repmat(d,1,ng); % true
[…]

### 2021-09-03 LDS/debug/debug_dC.m

@@ -54,8 +54,12 @@ for t=2:T
 end
 
 Y = poissrnd(exp(logLam));
-ng = 50;
-idx = 30:ng;
+imagesc(exp(logLam))
+colorbar()
+% clusterPlot(Y, Lab)
+
+ng = 100; % 40
+idx = 50:ng; % 20:ng
 
 %% fitting: MCMC, no update prior
 
@@ -99,27 +103,30 @@ end
 figure
 subplot(1,3,1)
 plot(d,mean(d_fit(:,idx), 2),'rx');
-title('true vs estimated d')
+ylabel('estimate')
+title('d')
 C_fit_mean = mean(C_fit(:,:,idx), 3);
 subplot(1,3,2)
 plot(sum(C_trans(:,1:p:end), 2),C_fit_mean(:,1),'rx');
-title('true vs estimated 1st column of C')
+title('1st column of C')
+xlabel('true')
 subplot(1,3,3)
 plot(sum(C_trans(:,2:p:end), 2),C_fit_mean(:,2),'rx');
-title('true vs estimated 2nd column of C')
+title('2nd column of C')
+sgtitle('No Update of Prior')
 
 %% fitting: MCMC, update priors
-X_fit = repmat(X, 1, 1, ng); % true
-x0_fit = repmat(x0',1,ng); %true
-A_fit = repmat(A,1,1,ng); % true
-b_fit = repmat(b,1,ng); % true
-Q_fit = repmat(Q,1,1,ng); % true
+X_fit2 = repmat(X, 1, 1, ng); % true
+x0_fit2 = repmat(x0',1,ng); %true
+A_fit2 = repmat(A,1,1,ng); % true
+b_fit2 = repmat(b,1,ng); % true
+Q_fit2 = repmat(Q,1,1,ng); % true
 
 % pre-allocation
-d_fit = zeros(N, nClus, ng);
-C_fit = zeros(N, nClus*p, ng);
-mudc_fit = zeros(p+1, nClus, ng);
-Sigdc_fit = zeros(p+1, p+1, nClus, ng);
+d_fit2 = zeros(N, nClus, ng);
+C_fit2 = zeros(N, nClus*p, ng);
+mudc_fit2 = zeros(p+1, nClus, ng);
+Sigdc_fit2 = zeros(p+1, p+1, nClus, ng);
 
 % priors
 deltadc0 = zeros(p+1,1);
[…]

### 2021-09-03 LDS/debug/debug_dC_X.m

@@ -14,7 +14,7 @@ Lab = repelem(1:nClus, n);
 pLab = repelem(1:nClus, p);
 
 d = randn(N,1);
-C_trans = zeros(n*nClus, p*nClus);
+C_trans = zeros(N, p*nClus);
 for k = 1:length(Lab)
     C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
 end
@@ -45,7 +45,7 @@ while any(imag(eig(A))==0)
 end
 
 % let's generate lambda
-logLam = zeros(n*nClus, T);
+logLam = zeros(N, T);
 logLam(:,1) = d + C_trans*X(:,1);
 
 for t=2:T
@@ -54,30 +54,30 @@ for t=2:T
 end
 
 Y = poissrnd(exp(logLam));
-ng = 50;
-idx = 30:ng;
+clusterPlot(Y, Lab)
+ng = 1000;
+idx = 900:ng;
 
 %% fitting: MCMC, no update prior
 X_fit = zeros(nClus*p, T, ng);
-d_fit = zeros(n*nClus, ng);
-C_fit = zeros(n*nClus, p, ng);
+d_fit = zeros(N, ng);
+C_fit = zeros(N, p, ng);
 x0_fit = zeros(nClus*p, ng);
 A_fit = repmat(A,1,1,ng); % true
 b_fit = repmat(b,1,ng); % true
 Q_fit = repmat(Q,1,1,ng); % true
 
-
 % priors
 Q0 = eye(nClus*p)*1e-2;
 mux00 = zeros(nClus*p, 1);
-Sigx00 = eye(nClus*p)*1e2;
+Sigx00 = eye(nClus*p);
 mudc0 = zeros(p+1,1);
 Sigdc0 = eye(p+1);
 
 % initials
 % initial for d_fit: 0
-C_fit(:,:,1) = reshape(normrnd(0,1e-2,n*nClus*p,1), [], p);
-C_trans_tmp = zeros(n*nClus, p*nClus);
+C_fit(:,:,1) = reshape(normrnd(0,1e-2,N*p,1), [], p);
+C_trans_tmp = zeros(N, p*nClus);
 for k = 1:length(Lab)
     C_trans_tmp(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C_fit(k,:,1);
 end
@@ -91,7 +91,7 @@ for g = 2:ng
     
[…]

### 2021-09-03 LDS/lds_sample_noConstraint.m

@@ -1,0 +1,372 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+clear all;close all;clc;
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+% 
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+	A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+
+figure(1)
+imagesc(A)
+colorbar()
+xlabel('sending')
+ylabel('receiving')
+
+figure(2)
+imagesc(exp(logLam))
+colorbar()
+
+figure(3)
+clusterPlot(Y, Lab)
+
+figure(4)
+subplot(1,3,1)
+plot(X(1:p,:)')
[…]

### 2021-09-06 documents/cluster_neural_population.tex
ignore mv-glm, fix that...

@@
 	Linear mapping (loading) for latent vectors
-	\(\left\{ d_{i} \right\}_{i = 1}^{N}\) and
-	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\):
-	\[\left( d_{i},\mathbf{c}'_{i} \right)' \sim N(\bm{\mu}_{\text{dc}}^{\left( z_{i} \right)},\mathbf{\Sigma}_{\text{dc}}^{(z_{i})})\]
+
+	\textcolor{red}{\textbf{I just ignore that it is just multivariate GLM problem. Do it later...}}\\
+
 	\item
 	Mean and covariance for loading in each cluster
-	\(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and
-	\(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\):
-	\[\bm{\mu}_{\text{dc}}^{(j)} \sim N(\mathbf{\delta}_{dc0},\mathbf{T}_{dc0})\]
-	, where \(\bm{\delta}_{dc0} = \mathbf{0}_{p + 1}\) and
-	\(\mathbf{T}_{dc0} = \mathbf{I}_{p + 1}\)
 
-	\[\mathbf{\Sigma}_{\text{dc}}^{(j)} \sim W^{- 1}\left( \Psi_{dc0},\nu_{dc0} \right)\]
-	, where \(\nu_{dc0} = p + 1 + 2\) and
-	\(\Psi_{dc0} = \mathbf{I}_{p + 1} \times 10^{-4}\)
+	\textcolor{red}{\textbf{MV-GLM...}}\\
+
 	\item
 	Linear dynamics for latent vectors \(\mathbf{A}\) and \(\mathbf{b}\): if the number of cluster is \(J\)
-	\begin{enumerate}
-		\def\labelenumi{\alph{enumi}.}
-		\item
-		No constraints on \(\mathbf{Q}\): denote \(\widetilde{\mathbf{A}} = vec(\mathbf{A})\)
-		\[\left( \mathbf{b}',\widetilde{\mathbf{A}} \right)' \sim N\left( \bm{\mu}_{BA_{0}},\mathbf{\Sigma}_{BA_{0}} \right)\]
-		, where \(\bm{\mu}_{BA_{0}} = \left( \mathbf{0}_{\text{Jp}}',vec\left( \mathbf{I}_{\text{Jp}} \right)' \right)'\)
-		and \(\mathbf{\Sigma}_{BA_{0}} = 0.25\mathbf{I}_{Jp(1 + Jp)}\)
-		\item
-		Assume \(\mathbf{Q}\) is block-diagonal: denote
-		\({\widetilde{\mathbf{A}}}_{j} = vec\left( \mathbf{A}_{j} \right)\)
-		\[\left( \mathbf{b}'_{j}, {\widetilde{\mathbf{A}}}'_{j} \right)' \sim N\left( \bm{\mu}_{bA_{0}},\mathbf{\Sigma}_{bA_{0}} \right)\]
-		, where \(\bm{\mu}_{bA_{0}} = \left( \mathbf{0}'_{p + (j - 1)p^{2}}, vec\left( \mathbf{I}_{p} \right)', \mathbf{0}_{(J - j)p^{2}}' \right)'\)
-		and \(\mathbf{\Sigma}_{bA_{0}} = 0.25\mathbf{I}_{p(1 + pJ)}\)
-		\item
-		Assume \(\mathbf{Q}\) is diagonal:
-		\[\left( b_{k},\mathbf{a}'_{k} \right)' \sim N\left( \mathbf{\mu}_{ba_{k}0},\mathbf{\Sigma}_{ba0} \right)\]
-		, where \(\mathbf{\mu}_{ba_{k}0} = \left( 0,\mathbf{e}'_{k} \right)'\) and \(\mathbf{\Sigma}_{ba0} = 0.25\mathbf{I}_{Jp + 1}\)	
-	\end{enumerate}
+
+	\textcolor{red}{\textbf{MV-GLM...}}\\
+
 	\item
 	Process noise \(\mathbf{Q}\): if the number of cluster is \(J\)
-	\begin{enumerate}
-		\def\labelenumi{\alph{enumi}.}
-		\item
-		No constraints on \(\mathbf{Q}\):
-		\[\mathbf{Q} \sim W^{- 1}(\Psi_{\mathbf{Q}_{0}},\nu_{\mathbf{Q}_{0}})\]
-		, where \(\nu_{\mathbf{Q}_{0}} = Jp + 2\) and
-		\(\Psi_{\mathbf{Q}_{0}} = \mathbf{I}_{\text{Jp}} \times 10^{- 4}\). (To make the mean of \(\mathbf{Q}\) loosely centered around
-		\(\mathbf{I}_{Jp} \times 10^{- 4}\))
-		\item
-		Assume \(\mathbf{Q}\) is block-diagonal:
-		\[\mathbf{Q}^{(j)} \sim W^{- 1}\left( \Psi_{0},\nu_{0} \right)\]
-		, where \(\nu_{0} = p + 2\) and
-		\(\Psi_{0} = \mathbf{I}_{p} \times 10^{- 4}\). (To make the mean of \(\mathbf{Q}^{(j)}\) loosely centered around
-		\(\mathbf{I}_{p} \times 10^{- 4}\))
-		\item
-		Assume \(\mathbf{Q}\) is diagonal:
-		\[q_{k} \sim IG(\frac{\nu_{0}}{2},\frac{\nu_{0}\sigma_{0}^{2}}{2})\]
-		, where \(\nu_{0} = 4\) and \(\sigma_{0}^{2} = 10^{- 4}\).
-	\end{enumerate}
+
+	\textcolor{red}{\textbf{MV-GLM...}}\\
+
 \end{enumerate}
@@
 To update efficiently, use Laplace approximation again. Denote
-\(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\)
-and \(\left( 1,{\mathbf{x}_{t}'^{\left( z_{i} \right)}} \right) = {{\widetilde{\mathbf{x}}}_{t}'^{\left( z_{i} \right)}}\).
-\begin{align*}
-	P\left( \bm{\zeta}_{i} \middle| \mathbf{y}_{i},\ \left\{ \mathbf{x}_{t}^{\left( z_{i} \right)} \right\}_{t = 1}^{T},\ldots \right) &= \exp{f\left( \bm{\zeta}_{i} \right)} \approx N\left( \bm{\zeta}_{i}|\bm{\mu}_{\bm{\zeta}_{i}},\mathbf{\Sigma}_{\bm{\zeta}_{i}}\  \right)\\
-	\frac{\partial f}{\partial\bm{\zeta}_{i}} &= \frac{\partial l}{\partial\bm{\zeta}_{i}} - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(z_{i})} \right) = \left\lbrack \sum_{t = 1}^{T}{\widetilde{\mathbf{x}}}_{t}^{\left( z_{i} \right)}\left( y_{\text{it}} - \lambda_{\text{it}} \right) \right\rbrack - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(z_{i})} \right)\\
-	\frac{\partial^{2}f}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}} &= \frac{\partial^{2}l}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}} - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1} = -\left\lbrack \sum_{t = 1}^{T}{\lambda_{\text{it}}{\widetilde{\mathbf{x}}}_{t}^{\left( z_{i} \right)}{{\widetilde{\mathbf{x}}}_{t}'^{\left( z_{i} \right)}}} \right\rbrack - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}
-\end{align*}
-, where \(l\) is Poisson log-likelihood. Then use Newton-Raphson to find
-\(\bm{\mu}_{\bm{\zeta}_{i}} = \text{argmax}_{\bm{\zeta}_{i}}\left( f\left( \bm{\zeta}_{i} \right) \right)\)
-and \(\mathbf{\Sigma}_{\bm{\zeta}_{i}} = -\left\lbrack \frac{\partial^{2}f}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}}\left.
-\right|_{\bm{\zeta}_{i} = \bm{\mu}_{\bm{\zeta}_{i}}} \right\rbrack^{-1}\).
-If initial as the previous step fits fails, simply use prior mean of \(\bm{\mu}_{\bm{\zeta}_{i}}\), i.e.
-\(\bm{\delta}_{dc0}\).
+
+\textcolor{red}{\textbf{MV-GLM...}}\\
 
 \subsubsection{Update \(\left\{ \mathbf{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and \(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\)} \label{loading prior}
-As above, denote \(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\).
-\begin{enumerate}
-	\def\labelenumi{(\arabic{enumi})}
-	\item
-	Mean \(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\): by
-	conjugacy, \(\bm{\mu}_{\text{dc}}^{(j)} \sim N\left( \bm{\delta}_{\text{dc}},\mathbf{T}_{\text{dc}}\  \right)\)
-	\begin{align*}
-		\mathbf{T}_{\text{dc}}^{- 1} &= \left( \mathbf{T}_{dc0}^{- 1} + n_{j}{\mathbf{\Sigma}_{\text{dc}}^{(j)}}^{- 1} \right)^{- 1}\\
-		\bm{\delta}_{\text{dc}} &= \mathbf{T}_{\text{dc}}\left( \mathbf{T}_{dc0}^{- 1}\bm{\delta}_{dc0} + {\mathbf{\Sigma}_{\text{dc}}^{(j)}}^{- 1}\sum_{i:z_{i} = j}^{}\bm{\zeta}_{i} \right)
-	\end{align*}
-	\item
-	Covariance \(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\): by conjugacy, \(\mathbf{\Sigma}_{\text{dc}}^{(j)} \sim W^{- 1}\left( \Psi_{\text{dc}},\nu_{\text{dc}} \right)\)
-	\begin{align*}
-		\nu_{\text{dc}} &= n_{j} + \nu_{dc0}\\
-		\Psi_{\text{dc}} &= \Psi_{dc0} + \sum_{i:z_{i} = j}^{}{\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(j)} \right)\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(j)} \right)'}
-	\end{align*}
-\end{enumerate}
+
+\textcolor{red}{\textbf{MV-GLM...}}\\
 
@@
 
-\begin{enumerate}
-	\def\labelenumi{(\arabic{enumi})}
-	\item
-	No constraints on \(\mathbf{Q}\):
-	\[P\left( \left( \mathbf{b}',{\widetilde{\mathbf{A}}'} \right)' \middle| \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T},\mathbf{Q},\ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}|\mathbf{A}\mathbf{x}_{t - 1} + \mathbf{b},\mathbf{Q} \right)} \right\rbrack P\left( \left( \mathbf{b}',{\widetilde{\mathbf{A}}'} \right)' \right)\]
-	Let
-	\(\left( \mathbf{b'},{\widetilde{\mathbf{A}}'} \right)' = \bm{\gamma} \in \mathbb{R}^{Jp(1 + Jp)}\),
-	\(\left( 1,\ {\mathbf{x}'_{t - 1}} \right)\bigotimes\mathbf{I}_{\text{Jp}}\mathbf{=}{\widetilde{\mathbf{G}}}_{t - 1} \in \mathbb{R}^{Jp \times Jp(1 + pJ)}\)
-	and
-	\(\widetilde{\mathbf{G}} = \left( {\widetilde{\mathbf{G}}'}_{1},\ldots,{\widetilde{\mathbf{G}}'}_{T - 1} \right)'\).
-	By conjugacy, \(\bm{\gamma}|\left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T},\mathbf{Q},\ldots \sim N(\bm{\mu}_{\bm{\gamma}},\mathbf{\Sigma}_{\bm{\gamma}})\)
-	\begin{align*}
-		\mathbf{\Sigma}_{\bm{\gamma}} &= \left\lbrack \mathbf{\Sigma}_{BA_{0}}^{- 1} + {\widetilde{\mathbf{G}}}^{\mathbf{'}}\mathbf{I}_{T - 1}\bigotimes\left( \mathbf{Q} \right)^{- 1}\widetilde{\mathbf{G}} \right\rbrack^{- 1}\\
-		\bm{\mu}_{\bm{\gamma}} &= \mathbf{\Sigma}_{\bm{\gamma}}\left( \mathbf{\Sigma}_{BA_{0}}^{- 1}\bm{\mu}_{BA_{0}} + {\widetilde{\mathbf{G}}}^{'}\mathbf{I}_{T - 1}\bigotimes\left( \mathbf{Q} \right)^{- 1}\left( \mathbf{x}_{2}^{'},\ldots,\mathbf{x}_{T}^{'} \right)' \right)
-	\end{align*}
-
-	\item
-	Assume \(\mathbf{Q}\) is block-diagonal:
-	\[P\left( \left( \mathbf{b}'_{j}, {\widetilde{\mathbf{A}}'}_{j}\ \right)' \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\mathbf{Q}^{(j)},\ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}^{(j)}|\mathbf{A}_{j}\mathbf{x}_{t - 1} + \mathbf{b}_{j},\mathbf{Q}^{(j)} \right)} \right\rbrack P\left( \left( \mathbf{b}'_{j}, {\widetilde{\mathbf{A}}'}_{j} \right)' \right)\]
-	Notice that \(\mathbf{A}_{j}\mathbf{x}_{t - 1} + \mathbf{b}_{j}\mathbf{=}\left( \left( 1,\ {\mathbf{x}'_{t - 1}} \right)\bigotimes\mathbf{I}_{p} \right)\left( \mathbf{b}_{j}^{\mathbf{'}}\mathbf{,}{\widetilde{\mathbf{A}}'}_{j} \right)'\), then the problem is reduced to Bayesian linear regression.
-Denote \(\left( 1,\ {\mathbf{x}'_{t - 1}} \right)\bigotimes\mathbf{I}_{p}\mathbf{=}{\widetilde{\mathbf{X}}}_{t - 1} \in \mathbb{R}^{p \times p(1 + pJ)}\),
-	\(\widetilde{\mathbf{X}} = \left({\widetilde{\mathbf{X}}'}_{1},\ldots,{\widetilde{\mathbf{X}}'}_{T - 1} \right)'\)
-	and\(\left( \mathbf{b}'_{j}, {\widetilde{\mathbf{A}}'}_{j} \right)' = \mathbf{\gamma}_{j} \in \mathbb{R}^{p(1 + pJ)}\).
-	Then
-	\begin{align*}
-		\prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}^{(j)}|\mathbf{A}_{j}\mathbf{x}_{t - 1} + \mathbf{b}_{j},\mathbf{Q}^{(j)} \right)} &= \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}^{(j)}|{\widetilde{\mathbf{X}}}_{t - 1}\mathbf{\gamma}_{j},\mathbf{Q}^{(j)} \right)}\\
-		&= N\left( \left( {\mathbf{x}_{2}'^{(j)}},\ldots,{\mathbf{x}_{T}'^{(j)}} \right)'|\widetilde{\mathbf{X}}\bm{\gamma}_{j},\ \mathbf{I}_{T - 1}\bigotimes\mathbf{Q}^{(j)} \right)
-	\end{align*}
-	By conjugacy, \(P\left( \bm{\gamma}_{j} \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\mathbf{Q}^{(j)},\ldots \right) = N\left(\bm{\gamma}_{j}|\bm{\mu}_{\bm{\gamma}_{j}},\mathbf{\Sigma}_{\bm{\gamma}_{j}} \right)\)
-	\begin{align*}
-		\mathbf{\Sigma}_{\bm{\gamma}_{j}} &= \left\lbrack \mathbf{\Sigma}_{bA_{0}}^{- 1} + {\widetilde{\mathbf{X}}'}\mathbf{I}_{T - 1}\bigotimes\left( \mathbf{Q}^{(j)} \right)^{- 1}\widetilde{\mathbf{X}} \right\rbrack^{- 1}\\
-		\bm{\mu}_{\bm{\gamma}_{j}} &= \mathbf{\Sigma}_{\bm{\gamma}_{j}}\left( \mathbf{\Sigma}_{bA_{0}}^{- 1}\bm{\mu}_{bA_{0}} + {\widetilde{\mathbf{X}}'}\mathbf{I}_{T - 1}\bigotimes\left( \mathbf{Q}^{(j)} \right)^{- 1}\left( {\mathbf{x}_{2}'^{(j)}},\ldots,{\mathbf{x}_{T}'^{(j)}} \right)' \right)
-	\end{align*}
-
-	\item
-	Assume \(\mathbf{Q}\) is diagonal:
-	\[P\left( \left( b_{k},\mathbf{a}'_{k} \right)'|\left\{ x_{kt} \right\}_{t = 1}^{T},q_{k},\ \ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( x_{\text{kt}}|\mathbf{a}'_{k}\mathbf{x}_{t - 1} + b_{k},q_{k} \right)} \right\rbrack P\left( \left( b_{k},\mathbf{a}'_{k} \right)' \right)\]
-	Rewrite \(\mathbf{a}'_{k}\mathbf{x}_{t - 1} + b_{k} = \left( 1,\mathbf{x}'_{t - 1} \right)\left( b_{k},\mathbf{a}'_{k} \right)'\).
-Then the problem is reduced to Bayesian linear regression.
-Denote \(\left( b_{k},\mathbf{a}'_{k} \right)' = \bm{\gamma}_{k} \in \mathbb{R}^{Jp + 1}\)
-	and
-	\(\widetilde{\mathbf{X}} = \left( \left( 1,\mathbf{x}'_{1} \right)',\ldots,\left( 1,\mathbf{x}'_{T - 1} \right)' \right)^{'}\).
-Then by conjugacy, \(P\left( \bm{\gamma}_{k}|\left\{ x_{\text{kt}} \right\}_{t = 1}^{T},q_{k},\ \ldots \right) = N\left( \bm{\gamma}_{k}|\bm{\mu}_{\bm{\gamma}_{k}},\mathbf{\Sigma}_{\bm{\gamma}_{k}} \right)\)
-	\begin{align*}
-		\mathbf{\Sigma}_{\bm{\gamma}_{k}} &= \left\lbrack \mathbf{\Sigma}_{ba0}^{- 1} + \frac{{\widetilde{\mathbf{X}}'}\widetilde{\mathbf{X}}}{q_{k}} \right\rbrack^{- 1}\\
-		\bm{\mu}_{\bm{\gamma}_{k}} &= \mathbf{\Sigma}_{\mathbf{\gamma}_{k}}\left( \mathbf{\Sigma}_{ba0}^{- 1}\mathbf{\mu}_{ba0} + {\widetilde{\mathbf{X}}'}\left( x_{k2},\ldots,x_{\text{kT}} \right)'/q_{k} \right)
-	\end{align*}	
-\end{enumerate}
+\textcolor{red}{\textbf{MV-GLM...}}\\
 
 \subsubsection{Update \(\mathbf{Q}\)}
-\begin{enumerate}
-	\def\labelenumi{(\arabic{enumi})}
-	\item
-	No constraints on \(\mathbf{Q}\):\\
-	Let \(\mathbf{A}\mathbf{x}_{t - 1} + \mathbf{b} = \bm{\mu}_{x_{t}}\)
-	\[P\left( \mathbf{Q} \middle| \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T},\mathbf{A},\mathbf{b},\ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}|\bm{\mu}_{\mathbf{x}_{t}},\ \mathbf{Q} \right)} \right\rbrack W^{- 1}(\mathbf{Q}|\Psi_{\mathbf{Q}_{0}},\nu_{\mathbf{Q}_{0}})\]
-	By conjugacy,
-	\[P\left( \mathbf{Q} \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\ \mathbf{A},\mathbf{b},\ldots \right) = W^{- 1}\left( \mathbf{Q}|\Psi_{0} + \sum_{t = 2}^{T}{\left( \mathbf{x}_{t} - \bm{\mu}_{\mathbf{x}_{t}} \right)\left( \mathbf{x}_{t} - \bm{\mu}_{\mathbf{x}_{t}} \right)^{'}},\ T - 1 + \nu_{0} \right)\]
 
-	\item
-	Assume \(\mathbf{Q}\) is block-diagonal:\\
-	Let \(\mathbf{A}_{j}\mathbf{x}_{t - 1} + \mathbf{b}_{j}\mathbf{=}\bm{\mu}_{\mathbf{x}_{t}^{(j)}}\),
-	\[P\left( \mathbf{Q}^{(j)} \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\ \mathbf{A}_{j},\mathbf{b}_{j},\ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( \mathbf{x}_{t}^{(j)}|\bm{\mu}_{\mathbf{x}_{t}^{(j)}},\mathbf{Q}^{(j)} \right)} \right\rbrack W^{- 1}\left( \mathbf{Q}^{(j)}|\Psi_{0},\nu_{0} \right)\]
-	By conjugacy,
-	\[P\left( \mathbf{Q}^{(j)} \middle| \left\{ \mathbf{x}_{t}^{(j)} \right\}_{t = 1}^{T},\ \mathbf{A}_{j},\mathbf{b}_{j},\ldots \right) = W^{- 1}\left( \mathbf{Q}^{(j)}|\Psi_{0} + \sum_{t = 2}^{T}{\left( \mathbf{x}_{t}^{(j)} - \bm{\mu}_{\mathbf{x}_{t}^{(j)}} \right)\left( \mathbf{x}_{t}^{(j)} - \bm{\mu}_{\mathbf{x}_{t}^{(j)}} \right)^{'}},\ T - 1 + \nu_{0} \right)\]
-
-	\item
-	Assume \(\mathbf{Q}\) is diagonal:\\
-	Let \(\mathbf{a}_{k}^{'}\mathbf{x}_{t - 1} + b_{k} = \mu_{x_{\text{kt}}}\),
-	\[P\left( q_{k}|\left\{ x_{\text{kt}} \right\}_{t = 1}^{T},b_{k},\mathbf{a}_{k}^{'},\ \ldots \right) \propto \left\lbrack \prod_{t = 2}^{T}{N\left( x_{\text{kt}}|\mu_{x_{\text{kt}}},q_{k} \right)} \right\rbrack\text{IG}\left( q_{k} \middle| \frac{\nu_{0}}{2},\frac{\nu_{0}\sigma_{0}^{2}}{2} \right)\]
-	By conjugacy,
-	\[P\left( q_{k}|\left\{ x_{\text{kt}} \right\}_{t = 1}^{T},b_{k},\mathbf{a}_{k}^{'},\ \ldots \right) = IG\left( q_{k}|\frac{\nu_{0} + T - 1}{2},\frac{\nu_{0}\sigma_{0}^{2} + \sum_{t = 2}^{T}\left( x_{\text{kt}} - \mu_{x_{\text{kt}}} \right)^{2}}{2} \right)\]
-\end{enumerate}
+\textcolor{red}{\textbf{MV-GLM...}}\\
 
 \section{Simulations}
+
+\textcolor{red}{\textbf{Old results}}\\
+
 \textcolor{red}{\textbf{Currently, the latent \(x_t\) estimation is not accurate enough. This may come from bad simulation/ slow convergence or other sources. See details in section \ref{problems}}}\\
@@
 	\item
-	Use a more identifiable simulation, e.g. larger loading, to see the performance
-	\item
-	Currently, the mixing of \(\mathbf{b}\) and \(\mathbf{A}\) is terrible, and I need find a way to break the dependence between dynamics (\(\mathbf{b}\) and \(\mathbf{A}\)) and latent (\(\mathbf{x}_t\)).
-For example, see if it's possible to do collapsed Gibbs sampler?
-Or switch to EM as in PLDS?
-If I switch to EM, then the DPMM should also be done in EM.
-(That would be more challenging).
-\\
-
-	Basically the question is
-	\[y_{\text{it}} \sim Poisson(\lambda_{\text{it}})\]
-	\[\log\lambda_{\text{it}} = d_{i} + \mathbf{c'}_{i}\mathbf{x}_{t}^{(z_{i})}\]
-	\[\mathbf{x}_{1} \sim N(\mathbf{x}_{0},\mathbf{Q}_{0})\]
-	\[\mathbf{x}_{t + 1}|\mathbf{x}_{t} \sim N(\mathbf{A}\mathbf{x}_{t} + \mathbf{b},\mathbf{Q})\]
-	And we need to sample \(\mathbf{x}_t\), \(\mathbf{b}\), \(\mathbf{A}\) and \(\mathbf{Q}\) efficiently. (Assume loading is known now).
-
-	Or even a more basic question: how to efficiently sample state-space and transition matrix at the same time in Kalman filter?
-The dependency between transition matrix and state vectors causes bad mixing.
-
+	\textcolor{red}{\textbf{Do MV-GLM}}\\
 	\item
@@
 Check them later.
+	\item
+	After all of them are resolved, switch to GP version
+
 \end{enumerate}

### 2021-09-06 PLDS/examples/PLDSExample.m

@@ -3,6 +3,7 @@
 % Lars Buesing, Jakob H Macke, 2014
 %
 addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering\pop_spike_dyn'));
+% addpath(genpath('D:\github\state-space-clustering'))
 %%
 clear all
 close all
@@ -13,7 +14,7 @@ close all
 xDim    = 5;												% latent dimensiom
 yDim    = 100;											    	% observed dimension = no of neurons
 T       = 100;												% no of time bins per trial; here a time step is approx 10ms 
-Trials  = 1;		    										% no trials
+Trials  = 2;		    										% no trials
 maxIter = 100;		    										% max no of EM iterations for fitting the model
 
 

### 2021-09-06 LDS/debug/debug_bA_Q_X_blkDiag.m
transfer for simulation

@@ -1,7 +1,6 @@
 addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
 % addpath(genpath('D:\github\state-space-clustering'));
 
-
 %% simulation
 rng(1)
 n = 40;
@@ -55,11 +54,11 @@ end
 
 Y = poissrnd(exp(logLam));
 clusterPlot(Y, Lab)
-ng = 1000;
-% idx = 1000:ng;
+
 
 %% fitting: MCMC
 rng(3)
+ng = 100;
 X_fit = zeros(nClus*p, T, ng);
 x0_fit = zeros(nClus*p, ng);
 d_fit = repmat(d,1,ng); % true
@@ -67,28 +66,22 @@ C_fit = repmat([sum(C_trans(:,1:p:end), 2) sum(C_trans(:,2:p:end), 2)],...
     1,1,ng); % true
 A_fit = zeros(nClus*p, nClus*p, ng);
 b_fit = zeros(nClus*p, ng);
-
-% A_fit = repmat(A,1,1,ng); % true
-% b_fit = repmat(b,1,ng); % true
-
-Q_fit = repmat(Q,1,1,ng); % true
-% Q_fit = zeros(nClus*p, nClus*p, ng);
+Q_fit = zeros(nClus*p, nClus*p, ng);
 
 
 % priors
 Q0 = eye(nClus*p)*1e-2;
 mux00 = zeros(nClus*p, 1);
 Sigx00 = eye(nClus*p);
-muBA0_mat = [zeros(nClus*p,1) eye(nClus*p)];
-muBA0 = muBA0_mat(:);
-SigBA0 = sparse(eye(length(muBA0))*0.1);
-% Psi0 = eye(p)*1e-4;
-% nu0 = p+2;
+BA0_all = [zeros(nClus*p,1) eye(nClus*p)]';
+Lamb0 = eye(nClus*p + 1);
+Psi0 = eye(p)*1e-4;
+nu0 = p+2;
 
 % initials
 % initial for b_fit: 0
 A_fit(:,:,1) = eye(nClus*p);
-% Q_fit(:,:,1) = eye(nClus*p)*1e-4;
+Q_fit(:,:,1) = eye(nClus*p)*1e-4;
 
 C_trans_tmp = zeros(N, p*nClus);
 for k = 1:length(Lab)
@@ -143,34 +136,27 @@ for g = 2:ng
     mux0 = Sigx0*(Sigx00\mux00 + Q0\X_fit(:,1,g));
[…]

### 2021-09-06 LDS/debug/debug_bA_Q_blkDiag.m

@@ -65,40 +65,48 @@ x0_fit = repmat(x0,1,ng); %true
 d_fit = repmat(d,1,ng); % true
 C_fit = repmat([sum(C_trans(:,1:p:end), 2) sum(C_trans(:,2:p:end), 2)],...
     1,1,ng); % true
+
+% need to estimate
 A_fit = zeros(nClus*p, nClus*p, ng);
 b_fit = zeros(nClus*p, ng);
-Q_fit = repmat(Q,1,1,ng); % true
+Q_fit = zeros(nClus*p, nClus*p, ng);
 
 % priors
-mubA0_mat = [zeros(nClus*p,1) eye(nClus*p)];
-SigbA0 = eye(p*(1+p*nClus));
+BA0_all = [zeros(nClus*p,1) eye(nClus*p)]';
+Lamb0 = eye(nClus*p + 1);
+Psi0 = eye(p)*1e-4;
+nu0 = p+2;
 
 % initials
 % initial for b_fit: 0
 A_fit(:,:,1) = eye(nClus*p);
+Q_fit(:,:,1) = eye(nClus*p)*1e-4;
 
 for g = 2:ng
     
     disp(g)
-    % (4) update b_fit & A_fit
-    for l = unique(Lab)
-        
-        latentId = ((l-1)*p+1):(l*p);
-        mubA0 = mubA0_mat(latentId, :);
-        mubA0 = mubA0(:);
-        Z_tmp = X_fit(latentId,2:T,g);
-        Z_tmp2 = Z_tmp(:);
-        
-        X_tmp = kron([ones(1,T-1); X_fit(:, 1:(T-1), g)]', eye(p));
-        SigbA_tmp = inv(inv(SigbA0) + X_tmp'*kron(eye(T-1), inv(Q_fit(latentId,latentId,g-1)))*X_tmp);
-        SigbA_tmp = (SigbA_tmp + SigbA_tmp')/2;
-        mubA_tmp = SigbA_tmp*(inv(SigbA0)*mubA0 +...
-            X_tmp'*kron(eye(T-1), inv(Q_fit(latentId,latentId,g-1)))*Z_tmp2);
-        bAtmp = reshape(mvnrnd(mubA_tmp, SigbA_tmp)', [], 1+nClus*p);
-        b_fit(latentId,g) = bAtmp(:,1);
-        A_fit(latentId,:,g) = bAtmp(:,2:end);
-    end
     
+    for l = unique(Lab)
[…]

### 2021-09-06 LDS/debug/debug_bA_Q_full.m

@@ -1,0 +1,120 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(N,1);
+C_trans = zeros(N, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(N, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+ng = 2;
+idx = 2;
+
+%% fitting: MCMC
+rng(3)
+X_fit = repmat(X, 1, 1, ng); % true
+x0_fit = repmat(x0,1,ng); %true
+d_fit = repmat(d,1,ng); % true
[…]

### 2021-09-06 LDS/debug/debug_dC_ind.m

@@ -1,7 +1,6 @@
 addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
 % addpath(genpath('D:\github\state-space-clustering'));
 
-
 %% simulation
 rng(1)
 n = 10;

### 2021-09-06 LDS/lds_sample.m

@@ -106,9 +106,8 @@ Taudc0 = eye(p+1);
 Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
-mubA0_mat = [zeros(nClus*p,1) eye(nClus*p)];
-SigbA0 = eye(p*(1+p*nClus))*0.25;
-
+BA0_all = [zeros(nClus*p,1) eye(nClus*p)]';
+Lamb0 = eye(nClus*p + 1);
 Psi0 = eye(p)*1e-4;
 nu0 = p+2;
 
@@ -244,39 +243,26 @@ for g = 2:ng
 %     mudc_fit(:,:,g)
 %     Sigdc_fit(:,:,:,g)
     
-    % (5) update b_fit & A_fit
     for l = unique(Lab)
-        
         latentId = id2id(l,p);
-        mubA0 = mubA0_mat(latentId, :);
-        mubA0 = mubA0(:);
-        Z_tmp = X_fit(latentId,2:T,g);
-        Z_tmp2 = Z_tmp(:);
         
-        X_tmp = kron([ones(1,T-1); X_fit(:, 1:(T-1), g)]', eye(p));
-        SigbA_tmp = inv(inv(SigbA0) + X_tmp'*kron(eye(T-1), inv(Q_fit(latentId,latentId,g-1)))*X_tmp);
-        SigbA_tmp = (SigbA_tmp + SigbA_tmp')/2;
-        mubA_tmp = SigbA_tmp*(inv(SigbA0)*mubA0 +...
-            X_tmp'*kron(eye(T-1), inv(Q_fit(latentId,latentId,g-1)))*Z_tmp2);
-        bAtmp = reshape(mvnrnd(mubA_tmp, SigbA_tmp)', [], 1+nClus*p);
-        b_fit(latentId,g) = bAtmp(:,1);
-        A_fit(latentId,:,g) = bAtmp(:,2:end);
-    end
-    
-    %     [b_fit(:,g) b]
-    %     imagesc(abs(A_fit(:,:,g) - A))
-    %     colorbar()
-    
-    % (6) update Q_fit
-    for l = unique(Lab)
-        latentId = ((l-1)*p+1):(l*p);
-        mux = A_fit(latentId,:,g)*X_fit(:,1:(T-1),g) +...
-            repmat(b_fit(latentId, g), 1, T-1);
-        xq = X_fit(latentId,2:T,g) - mux;
+        % (4)update Q
[…]

### 2021-09-08 LDS/blkDiag/lds_sample_DP_blkDiag.m
new results

@@ -59,7 +59,7 @@ clusterPlot(Y, Lab)
 %% MCMC setting
 rng(3)
 alphaDP = 10;
-ng = 50;
+ng = 100;
 
 % pre-allocation
 Z_fit = zeros(N, ng);
@@ -73,7 +73,7 @@ Sigx00_f = @(nClus) eye(nClus*p);
 deltadc0 = zeros(p+1,1);
 Taudc0 = eye(p+1);
 
-Psidc0 = eye(p+1)*1e-2;
+Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 BA0_all_f = @(nClus) [zeros(nClus*p,1) eye(nClus*p)]';
@@ -86,7 +86,7 @@ nu0 = p+2;
 % Z_fit(:,1) = ones(N, 1);
 % N cluster
 Z_fit(:,1) = 1:N;
-
+% Z_fit(:,1) = randsample(2, N, true);
 
 % reorder Y by labels
 clusMax = max(Z_fit(:,1));

### 2021-09-08 LDS/blkDiag/lds_sample_MM_blkDiag.m

@@ -56,9 +56,9 @@ Y = poissrnd(exp(logLam));
 clusterPlot(Y, Lab)
 
 %%
-rng(3)
+rng(6)
 ng = 50;
-kMM = 3;
+kMM = 30;
 
 % pre-allocation
 Z_fit = zeros(N, ng);
@@ -83,7 +83,7 @@ Sigx00 = eye(kMM*p);
 deltadc0 = zeros(p+1,1);
 Taudc0 = eye(p+1);
 
-Psidc0 = eye(p+1)*1e-2;
+Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 BA0_all = [zeros(kMM*p,1) eye(kMM*p)]';
@@ -93,8 +93,8 @@ nu0 = p+2;
 
 
 % initials
-% Z_fit(:,1) = ones(1, N);
-Z_fit(:,1) = randsample(kMM, N, true);
+Z_fit(:,1) = ones(1, N);
+% Z_fit(:,1) = randsample(kMM, N, true);
 RHO_fit(:,1) = ones(kMM,1)/kMM;
 
 mudc_fit(:,:,1) = zeros(p+1, kMM);
@@ -165,6 +165,7 @@ for g = 2:ng
     logp_tmp = repmat(log(RHO_fit(:,g-1)'), N, 1) + LLHD;
     clus_tmp = mnrnd(ones(N, 1), softmax(logp_tmp')');
     [Z_fit(:,g), ~] = find(clus_tmp');
+    
     
     % (2) update RHO_fit
     nClus = histc(Z_fit(:,g),1:kMM);

### 2021-09-08 LDS/blkDiag/lds_sample_blkDiag.m

@@ -107,7 +107,7 @@ Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 BA0_all = [zeros(nClus*p,1) eye(nClus*p)]';
-Lamb0 = eye(nClus*p + 1);
+Lamb0 = eye(nClus*p + 1)*.25;
 Psi0 = eye(p)*1e-4;
 nu0 = p+2;
 

### 2021-09-08 LDS/blkDiag/unspecifiedA_sample_DP_blkDiag.m

@@ -50,8 +50,8 @@ clusterPlot(Y, Lab)
 
 %% MCMC setting
 rng(3)
-alphaDP = 1;
-ng = 50;
+alphaDP = 10;
+ng = 100;
 
 % pre-allocation
 Z_fit = zeros(N, ng);
@@ -77,7 +77,8 @@ nu0 = p+2;
 % single cluster
 % Z_fit(:,1) = ones(N, 1);
 % N cluster
-Z_fit(:,1) = 1:N;
+Z_fit(:,1) = randsample(2, N, true);
+% Z_fit(:,1) = 1:N;
 
 
 % reorder Y by labels

### 2021-09-08 LDS/blkDiag/unspecifiedA_sample_MM_blkDiag.m

@@ -1,6 +1,6 @@
 %% simulation
 rng(123)
-n = 10;
+n = 50;
 nClus = 3;
 N = n*nClus;
 p = 2;
@@ -51,8 +51,8 @@ clusterPlot(Y, Lab)
 
 %% MCMC setting
 rng(3)
-ng = 50;
-kMM = 3;
+ng = 100;
+kMM = 30;
 
 % pre-allocation
 Z_fit = zeros(N, ng);
@@ -77,7 +77,7 @@ Sigx00 = eye(kMM*p);
 deltadc0 = zeros(p+1,1);
 Taudc0 = eye(p+1);
 
-Psidc0 = eye(p+1)*1e-2;
+Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 BA0_all = [zeros(kMM*p,1) eye(kMM*p)]';

### 2021-09-08 LDS/blkDiag/unspecifiedA_sample_blkDiag.m

@@ -87,7 +87,7 @@ Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 BA0_all = [zeros(nClus*p,1) eye(nClus*p)]';
-Lamb0 = eye(nClus*p + 1);
+Lamb0 = eye(nClus*p + 1)*.25;
 Psi0 = eye(p)*1e-4;
 nu0 = p+2;
 
@@ -276,7 +276,7 @@ plot(A_norm_fro)
 title('Frobenius norm of A')
 
 %%
-idx = 1500:ng;
+idx = 1800:2000;
 
 subplot(1,2,1)
 imagesc(exp(C_trans*X + d))

### 2021-09-08 LDS/clusPlot.m

@@ -1,12 +1,13 @@
 addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
 %%
 clear all;close all;clc;
-load('C:\Users\gaw19004\Desktop\LDS_backup\unspecifiedA_50_DP_above.mat')
+load('C:\Users\gaw19004\Desktop\LDS_backup\new\noA_full_MM_below_4.mat')
 
-plotFolder = "C:\Users\gaw19004\Desktop\LDS_backup\ClusPlots\noA_50_DP_above_1\";
+plotFolder = "C:\Users\gaw19004\Desktop\LDS_backup\new\plot\noA_full_MM_3\";
 for k = 1:ng
     clus = figure;
     clusterPlot(Y, Z_fit(:,k)')
-    title("k = " + k);
+    title("MM, Max Clust = " + kMM + ", k = " + k); % for MM
+%     title("DP, \alpha = " + alphaDP + ", k = " + k); % for DP
     saveas(clus, plotFolder+k+".png")
 end

### 2021-09-08 LDS/debug/debug_bA_Q_X_blkDiag.m

@@ -3,7 +3,7 @@ addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
 
 %% simulation
 rng(1)
-n = 40;
+n = 10;
 nClus = 3;
 N = n*nClus;
 p = 2;
@@ -58,7 +58,7 @@ clusterPlot(Y, Lab)
 
 %% fitting: MCMC
 rng(3)
-ng = 100;
+ng = 5000;
 X_fit = zeros(nClus*p, T, ng);
 x0_fit = zeros(nClus*p, ng);
 d_fit = repmat(d,1,ng); % true
@@ -172,6 +172,8 @@ for g = 2:ng
     plot(A_fit_fro(1:g))
 end
 
+ng = 1000;
+
 figure
 subplot(1,2,1)
 plot(b_fit_norm(1:ng))
@@ -181,7 +183,7 @@ plot(A_fit_fro(1:ng))
 title('Frobenius norm of A')
 % idx = 50:100;
 % idx = 500:1000;
-% idx = 5000:ng;
+% idx = 4000:ng;
 
 figure
 subplot(3,2,1)

### 2021-09-08 LDS/debug/debug_bA_Q_X_full.m

@@ -3,7 +3,7 @@ addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
 
 %% simulation
 rng(1)
-n = 40;
+n = 10;
 nClus = 3;
 N = n*nClus;
 p = 2;
@@ -58,7 +58,7 @@ clusterPlot(Y, Lab)
 
 %% fitting: MCMC
 rng(3)
-ng = 1000;
+ng = 5000;
 X_fit = zeros(nClus*p, T, ng);
 x0_fit = zeros(nClus*p, ng);
 d_fit = repmat(d,1,ng); % true
@@ -175,8 +175,8 @@ subplot(1,2,2)
 plot(A_fit_fro(1:ng))
 title('Frobenius norm of A')
 % idx = 800:1000;
-% idx = 500:1000;
-% idx = 5000:ng;
+% idx = 1000:2000;
+% idx = 4000:ng;
 
 figure
 subplot(3,2,1)

### 2021-09-08 LDS/full/full_gibbsLoop_DP_v2.m

@@ -94,6 +94,10 @@ if(sum(isnan(muXvec)) ~= 0)
 end
 
 % use Cholesky decomposition to sample efficiently
+
+% [L,D,P] = ldl(-hess_tmp+ 1e-13*sparse(eye(size(hess_tmp))));
+% R = P*L*sqrt(D);
+
 R = chol(-hess_tmp,'lower'); % sparse
 z = randn(length(muXvec), 1) + R'*muXvec;
 Xsamp = R'\z;

### 2021-09-08 LDS/full/lds_sample_DP_full.m

@@ -57,8 +57,8 @@ figure(1)
 clusterPlot(Y, Lab)
 
 %% MCMC setting
-rng(3)
-alphaDP = 1;
+rng(10)
+alphaDP = 15;
 ng = 50;
 
 % pre-allocation
@@ -73,7 +73,7 @@ Sigx00_f = @(nClus) eye(nClus*p);
 deltadc0 = zeros(p+1,1);
 Taudc0 = eye(p+1);
 
-Psidc0 = eye(p+1)*1e-2;
+Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 
@@ -86,8 +86,8 @@ nu0_f = @(nClus) nClus*p+2;
 % single cluster
 % Z_fit(:,1) = ones(N, 1);
 % N cluster
-Z_fit(:,1) = 1:N;
-
+% Z_fit(:,1) = 1:N;
+Z_fit(:,1) = randsample(5, N, true);
 
 % reorder Y by labels
 clusMax = max(Z_fit(:,1));

### 2021-09-08 LDS/full/lds_sample_full.m

@@ -108,7 +108,7 @@ Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 BA0 = [zeros(nClus*p,1) eye(nClus*p)]';
-Lamb0 = eye(nClus*p + 1);
+Lamb0 = eye(nClus*p + 1)*.25;
 Psi0 = eye(nClus*p)*1e-4;
 nu0 = nClus*p+2;
 
@@ -293,7 +293,7 @@ title('Frobenius norm of A')
 
 
 %%
-idx = 1500:ng;
+idx = 500:1000;
 
 subplot(1,2,1)
 imagesc(exp(C_trans*X + d))

### 2021-09-08 LDS/full/unspecifiedA_sample_DP_full.m

@@ -53,8 +53,8 @@ clusterPlot(Y, Lab)
 
 %% MCMC setting
 rng(3)
-alphaDP = 1;
-ng = 50;
+alphaDP = 10;
+ng = 100;
 
 % pre-allocation
 Z_fit = zeros(N, ng);
@@ -68,7 +68,7 @@ Sigx00_f = @(nClus) eye(nClus*p);
 deltadc0 = zeros(p+1,1);
 Taudc0 = eye(p+1);
 
-Psidc0 = eye(p+1)*1e-2;
+Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 
@@ -81,8 +81,8 @@ nu0_f = @(nClus) nClus*p+2;
 % single cluster
 % Z_fit(:,1) = ones(N, 1);
 % N cluster
-Z_fit(:,1) = 1:N;
-
+% Z_fit(:,1) = 1:N;
+Z_fit(:,1) = randsample(2, N, true);
 
 % reorder Y by labels
 clusMax = max(Z_fit(:,1));

### 2021-09-08 LDS/full/unspecifiedA_sample_MM_full.m

@@ -1,6 +1,6 @@
 %% simulation
 rng(123)
-n = 10;
+n = 50;
 nClus = 3;
 N = n*nClus;
 p = 2;
@@ -51,8 +51,8 @@ clusterPlot(Y, Lab)
 
 %% MCMC setting
 rng(3)
-ng = 50;
-kMM = 3;
+ng = 100;
+kMM = 4;
 
 % pre-allocation
 Z_fit = zeros(N, ng);

### 2021-09-08 LDS/full/unspecifiedA_sample_full.m

@@ -58,7 +58,7 @@ clusterPlot(Y, Lab)
 % nClus = 1;
 % Lab = ones(1,N);
 rng(3)
-ng = 2000;
+ng = 1000;
 
 % pre-allocation
 X_fit = zeros(nClus*p, T, ng);
@@ -87,7 +87,7 @@ Psidc0 = eye(p+1)*1e-4;
 nudc0 = p+1+2;
 
 BA0 = [zeros(nClus*p,1) eye(nClus*p)]';
-Lamb0 = eye(nClus*p + 1);
+Lamb0 = eye(nClus*p + 1)*.25;
 Psi0 = eye(nClus*p)*1e-4;
 nu0 = nClus*p+2;
 
@@ -137,14 +137,20 @@ for g = 2:ng
     A_tmp = A_fit(:,:,g-1);
     b_tmp = b_fit(:,g-1);
     Q_tmp = Q_fit(:,:,g-1);
-%     X_tmp = X_fit(:,:,g-1);
-    X_tmp = ppasmoo_poissexp_v2(Y,C_tmp,d_tmp,x0_tmp,Q0,A_tmp,b_tmp,Q_tmp);
-%     hess_tmp = hessX(muX(:), d_tmp, C_tmp, Q0, Q_tmp, A_tmp, Y);
+    X_tmp = X_fit(:,:,g-1);
     
-    % if use Newton?
     gradHess = @(vecX) gradHessX(vecX, d_tmp, C_tmp, x0_tmp, Q0, Q_tmp, A_tmp, b_tmp, Y);
     [muXvec,~,hess_tmp,~] = newtonGH(gradHess,X_tmp(:),1e-10,1000);
     muX = reshape(muXvec, [], T);
+    
+    % X_tmp = ppasmoo_poissexp_v2(Y,C_tmp,d_tmp,x0_tmp,Q0,A_tmp,b_tmp,Q_tmp);
+    % hess_tmp = hessX(muX(:), d_tmp, C_tmp, Q0, Q_tmp, A_tmp, Y);
+    
+    if(sum(isnan(muXvec)) ~= 0)
+        disp('use adaptive smoother initial')
+        X_tmp = ppasmoo_poissexp_v2(Y,C_tmp,d_tmp,x0_tmp,Q0,A_tmp,b_tmp,Q_tmp);
+        [muXvec,~,hess_tmp,~] = newtonGH(gradHess,X_tmp2(:),1e-6,1000);
+    end
     
     % tic;
[…]

### 2021-09-08 LDS/gradHessX.m

@@ -27,5 +27,7 @@ gradHess{1} = C_trans_tmp'*(Y - lam_tmp) + [-Q0_tmp\(X_all(:,1) - x0_tmp)+...
 gradHess{1} = gradHess{1}(:);
 
 gradHess{2} = blktridiag(hessmed,hessub,hessup);
+% gradHess{2} = (gradHess{2} + gradHess{2}')/2;
+
 
 end

### 2021-09-08 LDS/newtonGH.m

@@ -23,15 +23,6 @@ for k = 1:MaxIter
         break;
     end
     
-%     
-%     if(norm(fx) > 1e4*dhPre{1})
-%        x = nan;
-%        dfdx = nan;
-%        fprintf('skip');
-%        return; 
-%     elseif(norm(fx)<TolFun || norm(dx) < TolX)
-%         break;
-%     end
     
 end
 warning('on');

### 2021-09-08 documents/cluster_neural_population.tex
new documents

@@
 
-For completeness, Gibbs samplers for all three versions (no constraint, block diagonal and diagonal \(\mathbf{Q}\)) are shown below.
+To facilitate derivation in Gibbs sampler, write the linear dynamics of
+\(\mathbf{x}_{t}\) in MV-GLM form, i.e.
+
+\[\mathbf{x}_{t + 1}^{'} = \begin{pmatrix}
+	1 & \mathbf{x}_{t}^{'} \\
+\end{pmatrix}\begin{pmatrix}
+	\mathbf{b}^{'} \\
+	\mathbf{A}^{'} \\
+\end{pmatrix} + \mathbf{\epsilon}_{i}^{'}\]
+
+, where \(\mathbf{\epsilon} \sim N(0,\mathbf{Q})\). Then if we stack the
+latent by row, the problem is reduced to a multivariate general linear model (MV-GLM) problem:
+
+\[\begin{pmatrix}
+	\mathbf{x}_{2}^{'} \\
+	\mathbf{x}_{3}^{'} \\
+	\vdots \\
+	\mathbf{x}_{T}^{'} \\
+\end{pmatrix} = \begin{pmatrix}
+	1 & \mathbf{x}_{1}^{'} \\
+	1 & \mathbf{x}_{2}^{'} \\
+	\vdots & \vdots \\
+	1 & \mathbf{x}_{T - 1}^{'} \\
+\end{pmatrix}\begin{pmatrix}
+	\mathbf{b}^{'} \\
+	\mathbf{A}^{'} \\
+\end{pmatrix} + \mathbf{E}\]
+
+, where
+\(\mathbf{E} = \left( \mathbf{\epsilon}_{1},\ldots,\mathbf{\epsilon}_{T - 1} \right)^{'}\).
 
@@
 	Linear mapping (loading) for latent vectors
+	\(\left\{ d_{i} \right\}_{i = 1}^{N}\) and
+	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\):
+	\[\left( d_{i},\mathbf{c}'_{i} \right)' \sim N(\bm{\mu}_{\text{dc}}^{\left( z_{i} \right)},\mathbf{\Sigma}_{\text{dc}}^{(z_{i})})\]
+	\item
+	Mean and covariance for loading in each cluster
+	Mean and covariance for loading in each cluster
+	\(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\) and
+	\(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\):
+	\[\bm{\mu}_{\text{dc}}^{(j)} \sim N(\mathbf{\delta}_{dc0},\mathbf{T}_{dc0})\]
+	, where \(\bm{\delta}_{dc0} = \mathbf{0}_{p + 1}\) and
+	\(\mathbf{T}_{dc0} = \mathbf{I}_{p + 1}\)
 
-	\textcolor{red}{\textbf{I just ignore that it is just multivariate GLM problem. Do it later...}}\\
+	\[\mathbf{\Sigma}_{\text{dc}}^{(j)} \sim W^{- 1}\left( \Psi_{dc0},\nu_{dc0} \right)\]
+	, where \(\nu_{dc0} = p + 1 + 2\) and
+	\(\Psi_{dc0} = \mathbf{I}_{p + 1} \times 10^{-4}\)
 
 	\item
-	Mean and covariance for loading in each cluster
+	Linear dynamics for latent vectors: \(\mathbf{A}\), \(\mathbf{b}\) and process noise \(\mathbf{Q}\).\\ 
+	Denote \(\mathbf{F} = \begin{pmatrix}
+		\mathbf{b}^{'} \\
+		\mathbf{A}^{'} \\
+	\end{pmatrix}\), \(\mathbf{f} = vec(\mathbf{F})\)
 
-	\textcolor{red}{\textbf{MV-GLM...}}\\
+	\begin{align*}
+		P\left( \mathbf{f},\ \mathbf{Q} \right) &= P(\mathbf{Q})P(\mathbf{f}|\mathbf{Q})\\
+		\mathbf{Q} &\sim W^{- 1}(\Psi_{\mathbf{Q}_{0}},\nu_{\mathbf{Q}_{0}})\\
+		\mathbf{f} &\sim N(\mathbf{f}_0, \mathbf{Q}\bigotimes \bm{\Lambda}_0^{-1})
+	\end{align*}
 
-	\item
-	Linear dynamics for latent vectors \(\mathbf{A}\) and \(\mathbf{b}\): if the number of cluster is \(J\)
-
-	\textcolor{red}{\textbf{MV-GLM...}}\\
-
-	\item
-	Process noise \(\mathbf{Q}\): if the number of cluster is \(J\)
-
-	\textcolor{red}{\textbf{MV-GLM...}}\\
+	, where \(\mathbf{f}_0 = vec(\mathbf{F}_0)\).If the number of cluster is \(J\), \(\nu_{\mathbf{Q}_{0}} = Jp + 2\),
+	\(\Psi_{\mathbf{Q}_{0}} = \mathbf{I}_{\text{Jp}} \times 10^{- 4}\). (To make the mean of \(\mathbf{Q}\) loosely centered around
+	\(\mathbf{I}_{Jp} \times 10^{- 4}\)), \(\mathbf{F}_0 = \begin{pmatrix}
+		\mathbf{0}'_{Jp} \\
+		\mathbf{I}_{\text{Jp}} \\
+	\end{pmatrix}\) and \(\bm{\Lambda}_0 = \mathbf{I}_{Jp + 1}\)
 
@@
 	\(\left\{ \mathbf{c}_{i} \right\}_{i = 1}^{N}\)}
-To update efficiently, use Laplace approximation again. Denote
+To update efficiently, use Laplace approximation and the error is independent for each row (i.e. update things row by row). Denote
+\(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\)
+and \(\left( 1,{\mathbf{x}_{t}'^{\left( z_{i} \right)}} \right) = {{\widetilde{\mathbf{x}}}_{t}'^{\left( z_{i} \right)}}\).
+\begin{align*}
+	P\left( \bm{\zeta}_{i} \middle| \mathbf{y}_{i},\ \left\{ \mathbf{x}_{t}^{\left( z_{i} \right)} \right\}_{t = 1}^{T},\ldots \right) &= \exp{f\left( \bm{\zeta}_{i} \right)} \approx N\left( \bm{\zeta}_{i}|\bm{\mu}_{\bm{\zeta}_{i}},\mathbf{\Sigma}_{\bm{\zeta}_{i}}\  \right)\\
+	\frac{\partial f}{\partial\bm{\zeta}_{i}} &= \frac{\partial l}{\partial\bm{\zeta}_{i}} - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(z_{i})} \right) = \left\lbrack \sum_{t = 1}^{T}{\widetilde{\mathbf{x}}}_{t}^{\left( z_{i} \right)}\left( y_{\text{it}} - \lambda_{\text{it}} \right) \right\rbrack - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(z_{i})} \right)\\
+	\frac{\partial^{2}f}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}} &= \frac{\partial^{2}l}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}} - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1} = -\left\lbrack \sum_{t = 1}^{T}{\lambda_{\text{it}}{\widetilde{\mathbf{x}}}_{t}^{\left( z_{i} \right)}{{\widetilde{\mathbf{x}}}_{t}'^{\left( z_{i} \right)}}} \right\rbrack - {\mathbf{\Sigma}_{\text{dc}}^{\left( z_{i} \right)}}^{- 1}
+\end{align*}
+, where \(l\) is Poisson log-likelihood. Then use Newton-Raphson to find
+\(\bm{\mu}_{\bm{\zeta}_{i}} = \text{argmax}_{\bm{\zeta}_{i}}\left( f\left( \bm{\zeta}_{i} \right) \right)\)
+and \(\mathbf{\Sigma}_{\bm{\zeta}_{i}} = -\left\lbrack \frac{\partial^{2}f}{\partial\bm{\zeta}_{i}\partial\bm{\zeta}'_{i}}\left.
+\right|_{\bm{\zeta}_{i} = \bm{\mu}_{\bm{\zeta}_{i}}} \right\rbrack^{-1}\).
+If initial as the previous step fits fails, simply use prior mean of \(\bm{\mu}_{\bm{\zeta}_{i}}\), i.e.
+\(\bm{\delta}_{dc0}\).
 
-\textcolor{red}{\textbf{MV-GLM...}}\\
+DETOUR: we can write things in the form of MV-(Poisson)-GLM as follows (in cluster j with \(n_j\) observations/ neurons, with corresponding loading be \(\mathbf{d}_{j}\) and \(\mathbf{C}_{j}\)):
+
+\[\log\begin{pmatrix}
+	\lambda_{11} & \cdots & \lambda_{1n_{j}} \\
+	\vdots & \ddots & \vdots \\
+	\lambda_{T1} & \cdots & \lambda_{Tn_{j}} \\
+\end{pmatrix} = \begin{pmatrix}
+	1 & \mathbf{x}_{1}^{'} \\
+	\vdots & \vdots \\
+	1 & \mathbf{x}_{T}^{'} \\
+\end{pmatrix}\begin{pmatrix}
+	\mathbf{d}_{j}^{'} \\
+	\mathbf{C}_{j}^{'} \\
+\end{pmatrix} + \mathbf{E}_j\]
+
+We can further organize things column-wise. let \(\begin{pmatrix}
+	1 & \mathbf{x}_{1}^{'} \\
+	\vdots & \vdots \\
+	1 & \mathbf{x}_{T}^{'} \\
+\end{pmatrix} = \mathbf{G}\), then the design matrix and corresponding log-mean response are \(\begin{pmatrix}
+\mathbf{G} & \mathbf{0} & \cdots & \mathbf{0} \\
+\mathbf{0} & \mathbf{G} & \cdots & \mathbf{0} \\
+\vdots & \vdots & \ddots & \vdots \\
+\mathbf{0} & \cdots & \cdots & \mathbf{G} \\
+\end{pmatrix}\) and \({\log\begin{pmatrix}
+	\lambda_{11}^{(j)} & \cdots & \lambda_{T1}^{(j)} & \lambda_{12}^{(j)} & \cdots & \lambda_{T2}^{(j)} & \cdots & \lambda_{Tn_{j}}^{(j)} \\
+\end{pmatrix}}'\).
+
+However, when I use the Laplace approximation and NR, because of the block-diagonal structure of design matrix, the error is always independent observation by observation.
+This is equivalent to observation-wise update as above.
+
+\textbf{Maybe I did something wrong?}
+The point to say this is to make loading within each cluster depends on each other.
+Check these two later(\href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}).
+
+Now I do it in another way: make the prior cluster-dependent...
 
@@
 
-\textcolor{red}{\textbf{MV-GLM...}}\\
-
-\subsubsection{Update \(\mathbf{A}\) and \(\mathbf{b}\)} \label{dynamics update}
-
-\textcolor{red}{\textbf{MV-GLM...}}\\
-
-\subsubsection{Update \(\mathbf{Q}\)}
-
-\textcolor{red}{\textbf{MV-GLM...}}\\
-
-\section{Simulations}
-
-\textcolor{red}{\textbf{Old results}}\\
-
-\textcolor{red}{\textbf{Currently, the latent \(x_t\) estimation is not accurate enough. This may come from bad simulation/ slow convergence or other sources. See details in section \ref{problems}}}\\
-
-There are two set of simulation examples.
-The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics.
-In all of the following results, the covariance of process noise \(\mathbf{Q}\) is assumed to be block diagonal.
-For reference, see the code for \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/old_example/sim_LDS_noConstraint.m}{non-constrained} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/old_example/sim_LDS_diag.m}{diagonal} \(\mathbf{Q}\).
-But these two legacy versions assuming the loading is independent on cluster assignments.
-If we need to use any of one of them, make sure to replace the updates for loading to be cluster-related.
-The fitting results for all these three are similar, because the underlying true \(\mathbf{Q}\) is diagonal.
-
-Before doing the clustering, I first assume the cluster labels are known to see the fitting performance.
-All the fitting results with labels are shown in means from iteration 50 to 100 (\textcolor{red}{\textbf{The iteration number is not enough to achieve stationary distribution}}).
-
-\subsection{Simulation 1: Generate from LDS Directly}
-In this simulation, there are 3 clusters with 10 neurons in each cluster.
-The dimension of latents in each cluster is 2.
-In the linear dynamics, the bias term \(\mathbf{b}\) is zero.
-There's no within-population interaction but has some weak between-population interactions.
-In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal.
-The details of simulation can be found in the first section of the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample.m}{simulation 1}.
-
-\subsubsection{Labeled Data: No Clustering}
-The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample.m}{here}, and some results are in Figure \ref{fig:LDS labeled}.
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image001.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image002.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image003.png}}%
-	}
-	\caption{LDS sample with labels}
-	\label{fig:LDS labeled}
-\end{figure}
-
-The fitting for overall mean firing rate is perfect and the dynamics pattern is mostly recovered.
-The fitting of latents is OK (\textcolor{red}{\textbf{may not be accurate enough}}\\) up to scale.
-In other words, the model captures the oscillating pattern of latents.
-
-But the latent patterns of these three clusters seems quite similar.
-That means the differences in the mean firing rates may be purely explained by the loading, i.e.
-\(\mathbf{d}\) and \(\mathbf{C}\).
-To see that, I set the number of cluster to be 1 and the results are shown in Figure \ref{fig:LDS labeled one cluster}.
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image004.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image005.png}}%
-	}
-	\caption{LDS sample with labels, one cluster}
-	\label{fig:LDS labeled one cluster}
-\end{figure}
-
-Again, the fitting for overall mean firing rate is perfect.
-This is why it's important to make the loading, \(\mathbf{d}\) and \(\mathbf{C}\), also be cluster-dependent (loading within clusters are correlated).
-If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.
-
-\subsubsection{Unlabeled Data: Clustering}
-To give the full path of clustering, I show results in GIFs.
-Click to see the code for \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample_MM.m}{FMM} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample_DP.m}{DPMM}.
-
-There are three results:
+Purpose: to make loading within each cluster depends on each other, and this will help with clustering.
+As above, denote \(\left( d_{i},\mathbf{c}'_{i} \right)' = \bm{\zeta}_{i} \in \mathbb{R}^{p + 1}\).
 \begin{enumerate}
@@
 	\item
-	Fit by FMM with \(J = 3\), starting from random cluster assignments:
-	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_MM_above.gif}{GIF result 1}
+	Mean \(\left\{ \bm{\mu}_{\text{dc}}^{(j)} \right\}_{j}\): by
+	conjugacy, \(\bm{\mu}_{\text{dc}}^{(j)} \sim N\left( \bm{\delta}_{\text{dc}},\mathbf{T}_{\text{dc}}\  \right)\)
+	\begin{align*}
+		\mathbf{T}_{\text{dc}}^{- 1} &= \left( \mathbf{T}_{dc0}^{- 1} + n_{j}{\mathbf{\Sigma}_{\text{dc}}^{(j)}}^{- 1} \right)^{- 1}\\
+		\bm{\delta}_{\text{dc}} &= \mathbf{T}_{\text{dc}}\left( \mathbf{T}_{dc0}^{- 1}\bm{\delta}_{dc0} + {\mathbf{\Sigma}_{\text{dc}}^{(j)}}^{- 1}\sum_{i:z_{i} = j}^{}\bm{\zeta}_{i} \right)
+	\end{align*}
 	\item
-	Fit by FMM with \(J = 3\), starting from all neurons in single cluster: 
-	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_MM_below.gif}{GIF result 2}
-	\item
-	Fit by DPMM (\(\alpha = 10\)), starting from each neuron forms its own cluster:
-	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_DP_above_10.gif}{GIF result 3}
-
-\end{enumerate}
-I didn't show the DPMM starting from single cluster.
-Since in my current implementation, the cluster generation is not efficient.
-In other words, the newly generated cluster will usually not be sampled and the algorithm usually gets stuck in 1 or 2 clusters.
-
-This is a big problem...
-That means if we occasionally combine two clusters into one, we may never correct the mistake.
-Further, when the data grows, the number of cluster will be hard to grow.
-Fix that later.
-
-\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
-In this simulation, there are again 3 clusters with 2 latents in each.
-Besides set 10 neurons in each cluster, I further simulate 50 neurons in each cluster to see performance in a larger scale.
-Now, the latents are generated directly without specifying the underlying linear dynamics of latents.
-However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal.
-The details of simulation can be found in the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample.m}{code}.
-
-\subsubsection{Labeled Data: No Clustering}
-As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster.
-The code can be found \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample.m}{here}.
-\begin{enumerate}
-	\def\labelenumi{(\arabic{enumi})}
-	\item
-	Smaller dataset (10 neurons each):\\
-	The results with true cluster assignment is shown in Figure \ref{fig:no A labeled 10}
-	\begin{figure}[h!]
-		\makebox[\linewidth][c]{%
-			\centering
-			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image006.png}}%
-			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image007.png}}%
-			\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image008.png}}%
-		}
-		\caption{unspecified dynamics with labels, 10 neurons each}
-		\label{fig:no A labeled 10}
-	\end{figure}
-	And results by forcing all neurons to be in single cluster (Figure \ref{fig:no A labeled one cluster 10})
-	\begin{figure}[h!]
-		\makebox[\linewidth][c]{%
-			\centering
-			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image009.png}}%
-			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image010.png}}%
-		}
-		\caption{unspecified dynamics with labels, one cluster, 10 neurons each}
-		\label{fig:no A labeled one cluster 10}
-	\end{figure}
-
-	Again, the fittings for overall mean firing rate are perfect in both cases and the dynamics is roughly diagonal.
-The oscillating pattern of fitted latents with true cluster index is not significant, because of plot scaling.
-The pattern is easier to see in the second fitting.
-
-	\item
-	Larger dataset (50 neurons each):\\
-	The results with true cluster assignment is shown in Figure \ref{fig:no A labeled 50}
-	\begin{figure}[h!]
-		\makebox[\linewidth][c]{%
-			\centering
-			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image011.png}}%
-			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image012.png}}%
-			\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image013.png}}%
-		}
-		\caption{unspecified dynamics with labels, 50 neurons each}
-		\label{fig:no A labeled 50}
-	\end{figure}
-	And results by forcing all neurons to be in single cluster (Figure \ref{fig:no A labeled one cluster 50})
-	\begin{figure}[h!]
-		\makebox[\linewidth][c]{%
-			\centering
-			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image014.png}}%
-			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image015.png}}%
-		}
-		\caption{unspecified dynamics with labels, one cluster, 50 neurons each}
-		\label{fig:no A labeled one cluster 50}
-	\end{figure}
-
-	Now we can see the pattern more clearly.
+	Covariance \(\left\{ \mathbf{\Sigma}_{\text{dc}}^{(j)} \right\}_{j}\): by conjugacy, \(\mathbf{\Sigma}_{\text{dc}}^{(j)} \sim W^{- 1}\left( \Psi_{\text{dc}},\nu_{\text{dc}} \right)\)
+	\begin{align*}
+		\nu_{\text{dc}} &= n_{j} + \nu_{dc0}\\
+		\Psi_{\text{dc}} &= \Psi_{dc0} + \sum_{i:z_{i} = j}^{}{\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(j)} \right)\left( \bm{\zeta}_{i} - \bm{\mu}_{\text{dc}}^{(j)} \right)'}
+	\end{align*}
 \end{enumerate}
 
-\subsubsection{Unlabeled Data: Clustering}
-As previous, in each example, I show results from (1) FMM starting from random cluster assignment \(\left(J = 3\right) \), (2) FMM starting from single cluster and (3) DPMM starting from multiple clusters.
-The DPMM starting from single cluster need to be fixed later.
-These are code for
-\href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample_MM.m}{FMM} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample_DP.m}{DPMM}.
+\subsubsection{Update \(\mathbf{A}\), \(\mathbf{b}\) and \(\mathbf{Q}\)} 
+Here, I only give the update for full \(\mathbf{Q}\). If we want to assume block-diagonal or diagonal structure of \(\mathbf{Q}\), just update things cluster-by-cluster or latent-by-latent.
 
-\begin{enumerate}
-	\def\labelenumi{(\arabic{enumi})}
-	\item
-	Smaller dataset (10 neurons each):\\
-	\begin{enumerate}
-		\def\labelenumi{\alph{enumi}.}
-		\item
-		Fit by FMM with \(J = 3\), start from random cluster assignments:
-		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_MM_above.gif}{GIF result 4}
-		\item
-		Fit by FMM with \(J = 3\), start from single cluster:
-		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_MM_below.gif}{GIF result 5}
-		\item
-		Fit by DP (\(\alpha = 5\)), start from assuming each neuron forms its own cluster:
-		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_DP_above_5.gif}{GIF result 6}
-	\end{enumerate}
+As shown before, the problem is the usual Bayesian MV-GLM problem. Denote \(\mathbf{F} = \begin{pmatrix}
+	\mathbf{b}^{'} \\
+	\mathbf{A}^{'} \\
+\end{pmatrix}\), \(\mathbf{f} = vec(\mathbf{F})\), \(\mathbf{Y}_\mathbf{bA} = (\mathbf{x}_{2}, \mathbf{x}_{3} \ldots, \mathbf{x}_{T})'\) and \(\mathbf{X}_\mathbf{bA} = \begin{pmatrix}
+1 & \mathbf{x}_{1}^{'} \\
+1 & \mathbf{x}_{2}^{'} \\
+\vdots & \vdots \\
+1 & \mathbf{x}_{T - 1}^{'} \\
+\end{pmatrix}\)
 
-	\item
-	Larger dataset (50 neurons each):\\
-	\begin{enumerate}
-		\def\labelenumi{\alph{enumi}.}
-		\item
-		Fit by FMM with \(J = 3\), start from random cluster assignments:
-		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_above.gif}{GIF result 7}
-		\item
-		Fit by FMM with \(J = 3\), start from single cluster:
-		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_below.gif}{GIF result 8}
-		\item
-		Fit by DP (\(\alpha = 1\)), start from random assignment for 10 clusters:
-		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_DP_above1.gif}{GIF result 9}
-	\end{enumerate}
-\end{enumerate}
+\[\mathbf{Y}_\mathbf{bA} = \mathbf{X}_\mathbf{bA}\mathbf{F}+ \mathbf{E}\]
 
-The performance is not bad. The neuron at top of each cluster cannot be clustered correctly by its nature: the signals are too weak to be clearly clustered.
+, where
+\(\mathbf{E} = \left( \mathbf{\epsilon}_{1},\ldots,\mathbf{\epsilon}_{T - 1} \right)^{'}\) and \(\mathbf{\epsilon} \sim N(0,\mathbf{Q})\). Then posteriors are
+\begin{align*}
+	\mathbf{Q}|\mathbf{Y}_{\mathbf{\text{bA}}},\mathbf{X}_{\mathbf{\text{bA}}},\ldots &\sim W^{-1}(\Psi_{\mathbf{Q}}, \nu_{\mathbf{Q}})\\
+	\mathbf{f}|\mathbf{Y}_{\mathbf{\text{bA}}},\mathbf{X}_{\mathbf{\text{bA}}},\mathbf{Q},\ldots &\sim N(vec(\mathbf{F}_\mathbf{bA}),  \mathbf{Q}\bigotimes \bm{\Lambda}_\mathbf{bA}^{-1})
+\end{align*}
+, with parameters be:
+\begin{align*}
+	\Psi_{\mathbf{Q}} &= \Psi_{\mathbf{Q}_{0}} + (\mathbf{Y}_\mathbf{bA} - \mathbf{X}_\mathbf{bA}\mathbf{F}_\mathbf{bA})'(\mathbf{Y}_\mathbf{bA} - \mathbf{X}_\mathbf{bA}\mathbf{F}_\mathbf{bA}) + (\mathbf{F}_\mathbf{bA} - \mathbf{F}_0)'\bm{\Lambda}_0(\mathbf{F}_\mathbf{bA} - \mathbf{F}_0)\\
+	\nu_{\mathbf{Q}} &= \nu_{\mathbf{Q}_{0}} + T - 1\\
+	\mathbf{F}_\mathbf{bA} &= (\mathbf{X}'_\mathbf{bA}\mathbf{X}_\mathbf{bA} + \bm{\Lambda}_0)^{-1}(\mathbf{X}'_\mathbf{bA}\mathbf{Y}_\mathbf{bA} + \bm{\Lambda}_0\mathbf{F}_0)\\
+	\bm{\Lambda}_\mathbf{bA} &= \mathbf{X}'_\mathbf{bA}\mathbf{X}_\mathbf{bA} + \bm{\Lambda}_0
+\end{align*}
 
-\section{PROBLEMS} \label{problems}
-This section shows what cause the inaccuracy in the latent estimation.
+\section{Model Check [Problems for \(\mathbf{b}\) and \(\mathbf{A}\) partially solved]} \label{problems}
 
 \subsection{Estimation of \(\mathbf{d}\) and \(\mathbf{C}\) alone}
-At first, as guessed by Ian, this may be caused by bad estimation of loading (\(\mathbf{d}\) and \(\mathbf{C}\)).
-So I first turn off all others except for loading (related) parameters and set them to be true values.
+First, turn off all others except for loading (related) parameters and set them to be true values.
 There are 2 versions: (1) update priors as in subsection \ref{loading prior} and (2) no update of priors, and just set the prior for each as the standard normal.
@@
 
-I've checked update of (1) dynamics alone, (2) \(\mathbf{Q}\) and (3) \(\mathbf{Q}\) and \(\mathbf{x}_{t}\).
-All of them are fine.
-\textbf{So the problem must come from estimations of dynamics and \(\mathbf{x}_t\)}.
-
-\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\) and \(\mathbf{x}_t\)}
-I first estimate dynamics by assuming block diagonal \(\mathbf{Q}\) (as in subsection \ref{dynamics update}).
-When turning on (1) latents \(\mathbf{x}_t\) and (2) dynamics (\(\mathbf{b}\) and \(\mathbf{A}\)), the trace of \(\mathbf{b}\) and \(\mathbf{A}\) are (Figure \ref{trace of dynamics, blk}):\\
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width = .7\textwidth]{image022.png}
-	\caption{trace of dynamics, block diagonal}
-	\label{trace of dynamics, blk}
-\end{figure}
-
-The mixing is not good, and we may haven't achieve stationary distribution yet. Anyway, I average results from iteration 5000 to 10000, and the results are shown in Figure \ref{lat and dyn, blk}).
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[latent]{\includegraphics[width=0.6\textwidth]{image023.png}}%
-		\subfigure[dynamics]{\includegraphics[width=0.6\textwidth]{image024.png}}%
-	}
-	\caption{estimation of latents and dynamics, block diagonal}
-	\label{lat and dyn, blk}
-\end{figure}
-
-OK, things are not perfect now.
-What if I don't assume block diagonal \(\mathbf{Q}\)?
-For the non-constraint version, the trace plot of \(\mathbf{b}\) and \(\mathbf{A}\) are in Figure \ref{trace of dynamics, full}.
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width = .7\textwidth]{image025.png}
-	\caption{trace of dynamics, no constraint}
-	\label{trace of dynamics, full}
-\end{figure}\\
-
-Again, the average from iteration 5000 to 10000 is in Figure \ref{lat and dyn, full}).\\
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[latent]{\includegraphics[width=0.6\textwidth]{image026.png}}%
-		\subfigure[dynamics]{\includegraphics[width=0.6\textwidth]{image027.png}}%
-	}
-	\caption{estimation of latents and dynamics, no constraint}
-	\label{lat and dyn, full}
-\end{figure}
-
-Based on trace plot and latent, things look a bit better. But it's still not good enough. The dynamics is even worse, although the latent is better.
-
-\textbf{Another Observation}:
-Even with these inaccurate latent, the overall mean firing rate still looks perfect (with true loading \(\mathbf{d}\) and \(\mathbf{C}\)), because of small loading. See Figure \ref{mfr, blk and full}
+\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\), \(\mathbf{Q}\) and \(\mathbf{x}_t\)}
+I show the results of block-diagonal (Figure \ref{bAQX_blkDiag}) and full (Figure \ref{bAQX_full}) \(\mathbf{Q}\).
+These results are from 2000 MCMC samples.
+The latent and dynamics are from average from 500 to 2000 iterations.
 
@@
 		\centering
-		\subfigure[block diagonal]{\includegraphics[width=0.6\textwidth]{image028.png}}%
-		\subfigure[no constraint]{\includegraphics[width=0.6\textwidth]{image029.png}}%
+		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_trace.png}}%
+		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_latent.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_A.png}}%
 	}
-	\caption{overall mean firing rate, block diagonal and no constraint}
-	\label{mfr, blk and full}
+	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, block-diagonal process noise}
+	\label{bAQX_blkDiag}
 \end{figure}
-
-In PLDS, they use EM, so there's no noise. But here I tackle things by MCMC: if the simulation is not very identifiable, then there will be a lot of noise by its nature.
-
-Anyway, the clustering is based on the overall mean firing rate pattern. Although the estimation of latent may not be accurate enough, this should not influence the clustering result.
-
-Maybe the next step is to test the current algorithm by a more identifiable simulation. For example, use a larger loading to make the latent effect become more significant.
-
-\subsection{Simulation 1 Revisit}
-The results above remind me that the 100 iterations is not enough to achieve stationary distribution. I need to check the sampling trace.
-Here, I rerun the MCMC for simulation 1, but use the no-constraint \(\mathbf{Q}\) version. Since no constraint version is better than block-diagonal one.
-
-First, I show the trace of sampling (Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\) as in Figure \ref{norms}.
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width = .8\textwidth]{image030.png}
-	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
-	\label{norms}
-\end{figure}\\
-
-It seems we may have not achieved stationary yet. And I show the average from iteration 8000 to 10000 as in Figure \ref{fig:LDS labeled revisit}
 
@@
 		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image031.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image032.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image033.png}}%
+		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_full_trace.png}}%
+		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_full_latent.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_full_A.png}}%
+	}
+	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, full process noise}
+	\label{bAQX_full}
+\end{figure}
+
+Both block-diagonal and full looks fine, but the mixing of  full process noise version looks better.
+
+\section{Simulations}
+
+There are two set of simulation examples.
+The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics.
+In all following results, I fit things with both \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{block-diagonal} and \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/full}{full} \(\mathbf{Q}\).
+The fitting results for block-diagonal version is a bit better, because the underlying true \(\mathbf{Q}\) is diagonal.
+In the following part, I only show results from block-diagonal fitting for labeled data.
+For unlabeled data, i.e.
+clustering, all the results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
+
+\subsection{Simulation 1: Generate from LDS Directly}
+In this simulation, there are 3 clusters with 10 neurons in each cluster.
+The dimension of latents in each cluster is 2.
+In the linear dynamics, the bias term \(\mathbf{b}\) is zero.
+There's no within-population interaction but has some weak between-population interactions.
+In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal.
+The details of simulation can be found in the first section of the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{simulation 1}.
+
+\subsubsection{Labeled Data: No Clustering}
+The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{here}.
+After running 2000 iterations, the trace of the sample is shown in Figure \ref{lds norms}.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width = .8\textwidth]{lds_blk_trace.png}
+	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
+	\label{lds norms}
+\end{figure}\\
+
+Some results are in Figure \ref{fig:LDS labeled}. These are means from iteration 1000 to 2000.
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{lds_blk_FR.png}}%
+		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{lds_blk_latent.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{lds_blk_A.png}}%
 	}
 	\caption{LDS sample with labels}
-	\label{fig:LDS labeled revisit}
+	\label{fig:LDS labeled}
 \end{figure}
 
-Things look much better! That means we need to improve the mixing of \(\mathbf{b}\) and \(\mathbf{A}\) later. (Also see if the mixing will be better for a more identifiable simulation)
+Personally, I think the fitting is not bad and the mixing is fine (in previous version, we may need 10000 iterations to get a not bad result. But here 2000 iterations did even better than before).
+
+I also fit the model by assuming one cluster.
+Again, the fitting for overall mean firing rate is perfect.
+This is why it's important to make the loading, \(\mathbf{d}\) and \(\mathbf{C}\), also be cluster-dependent (loading within clusters are correlated).
+
+If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.
+
+\textbf{Think more on it.} Maybe we can do MV-(Poisson)-GLM, to make all loading within cluster correlated.
+
+\subsubsection{Unlabeled Data: Clustering}
+To give the full path of clustering, I show results in GIFs.
+All results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
+The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.
+
+Basically, in my current implementation, algorithms are good to merge clusters.
+However, generating new clusters is very hard...
+In other words, the newly generated cluster will usually not be sampled.
+
+\textbf{This is a big problem}. This makes DPMM loses its power. \textbf{Fix that later}.
+
+\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
+In this simulation, there are again 3 clusters with 2 latents and 10 neurons in each.
+Now, the latents are generated directly without specifying the underlying linear dynamics of latents.
+However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal.
+The details of simulation can be found in the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/unspecifiedA_sample_blkDiag.m}{code}.
+
+\subsubsection{Labeled Data: No Clustering}
+As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster.
+The code can be found \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/unspecifiedA_sample_blkDiag.m}{here}.
+
+This time I ran 1000 iterations. The trace of the samples is shown in Figure \ref{noA norms}.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width = .8\textwidth]{noA_blk_trace.png}
+	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
+	\label{noA norms}
+\end{figure}\\
+
+Some results are in Figure \ref{fig:noA labeled}. These are means from iteration 500 to 1000.
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{noA_blk_FR.png}}%
+		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{noA_blk_latent.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{noA_blk_A.png}}%
+	}
+	\caption{LDS sample with labels}
+	\label{fig:noA labeled}
+\end{figure}
+
+This time the mixing of loading (\(\mathbf{d}\) and \(\mathbf{C}\)) is not fine. That reminds me that \textbf{I may need to make loading within cluster related by MV-(Poisson)-GLM directly.}
+
+\subsubsection{Unlabeled Data: Clustering}
+Again, all results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
+The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.
+
+Besides fitting 3 clusters with 10 neurons each, I further fit a larger scale example (50 neurons for each cluster).
 
 \clearpage
+
 \section{TODO}
@@
 	\item
-	\textcolor{red}{\textbf{Do MV-GLM}}\\
+	Think about MV-(Poisson)-GLM.
+Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
 	\item
-	Check sampling trace and mixing.
+	After resolving (1), see if the MCMC is still trapped in local optima sometimes.
 	\item

### 2021-09-08 LDS/debug/debug_dC_clusCor.m

@@ -106,7 +106,7 @@ for g = 2:ng
         Sigdc0 = sparse(kron(eye(N_clus), Sigdc0_sing));
         mudc0 = repmat(mudc0_sing, N_clus, 1);
         
-        hessdc_all = @(dc) -sparse(X_dc'*diag(lamdc_all(dc))*X_dc - inv(Sigdc0));
+        hessdc_all = @(dc) sparse(-X_dc'*diag(lamdc_all(dc))*X_dc - inv(Sigdc0));
         niSigdc_all = hessdc_all(mudc_all);
         
         % use Cholesky decomposition to sample efficiently

### 2021-09-12 documents/cluster_neural_population.tex
modify TODO
transfer for debug dc MH

@@
 
-Since the progress noise is independent at each step,
-\(f\left( \mathbf{y}_{i}|\mathbf{\Theta}_{j} \right) = \prod_{t = 1}^{T}{P(y_{\text{it}}|\mathbf{\Theta}_{j})}\),
-where \(P( \cdot )\) is the Poisson density and \(\mathbf{\Theta}_{j}\) contains all parameters in cluster \(j\).
+In summary, Let
+\(\mathbf{\Theta} = \left\{ \mathbf{z},\ \mathbf{d},\mathbf{C},\left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T},\mathbf{x}_{0},\mathbf{Q}_{0},\mathbf{A},\ \mathbf{b},\mathbf{Q}\  \right\}\)
+be the set of parameters for LDS. The number of observation/ neuron is
+\(N\), and they can group into \(J\) clusters. In each cluster, there
+are \(p\) latent state vectors. The recording length is \(T\).
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+	\(\mathbf{z} \in \mathbb{Z}^{N}\): cluster index for each neuron/ observation,
+	with \(z_{i} \in \{ 1,\ldots,J\}\) for \(i = 1,\ldots,N\).
+	\item
+	\(\mathbf{d} \in \mathbb{R}^{N}\) and \(\mathbf{C} \in \mathbb{R}^{N \times p}\): baseline
+	\& loading of the latent
+	\item
+	\(\mathbf{x}_{t} \in \mathbb{R}^{\text{Jp}}\): all latent at \(t\).
+	\item
+	\(\mathbf{x}_{t}^{(j)} \in \mathbb{R}^{p}\) means the latent in cluster \(j\) at
+	time \(t\).
+	\item
+	\(\mathbf{x}_{0} \in \mathbb{R}^{\text{Jp}}\) and
+	\(\mathbf{Q}_{0} \in \mathbb{R}^{\text{Jp} \times Jp}\): mean and covariance of
+	\(\mathbf{x}_{1}\). For simplicity, assume \(\mathbf{Q}_{0}\) is known.
+	\item
+	\(\mathbf{A} \in \mathbb{R}^{Jp \times Jp},\ \mathbf{b} \in \mathbb{R}^{Jp},\mathbf{Q} \in \mathbb{R}^{Jp \times Jp}\):
+	linear dynamics of the latent.
+\end{enumerate}
+
+Since the progress noise is independent at each step and the observation is assumed conditional independent, the likelihood is:
+\[ f(\mathbf{y}|\mathbf{\Theta})= \prod_{i = 1}^{N}\prod_{t = 1}^{T}{P(y_{it}|\mathbf{\Theta})} = \prod_{i = 1}^{N}\prod_{t = 1}^{T}{POI(y_{it}|\exp{(d_{i} + \mathbf{c'}_{i}\mathbf{x}_{t}^{(z_{i})})})}
+\]
+, where \(POI(\cdot|\lambda)\) is the density of \(Poisson(\lambda)\).
+
+\textbf{DETOUR}:\\
+Maybe in the future, we may switch to EM.
+To help with this, I also give the likelihood for complete observation \((\mathbf{y}, \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T})\).
+Let \(\mathbf{\Theta}'=\mathbf{\Theta}\backslash \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T}\).
+\[ f(\mathbf{y}, \left\{ \mathbf{x}_{t} \right\}_{t = 1}^{T}|\mathbf{\Theta}')= \prod_{i = 1}^{N}\prod_{t = 1}^{T}{P(y_{it}, \mathbf{x}_{t}|\mathbf{\Theta}')} = \prod_{i = 1}^{N}\prod_{t = 1}^{T}{POI(y_{it}|\exp{(d_{i} + \mathbf{c'}_{i}\mathbf{x}_{t}^{(z_{i})})})}\cdot N(\mathbf{x}_{t}^{(z_{i})} |\mathbf{A}_{z_i}\mathbf{x}_{t} + \mathbf{b}_{z_i}, \mathbf{Q}_{z_i})
+\]
+, where \(\mathbf{A}_j\in \mathbb{R}^{p \times Jp}\) and \(\mathbf{b}_j\in \mathbb{R}^p\)  are the \(j^{\text{th}}\) row block of \(\mathbf{A}\) and \(\mathbf{b}\).
+\(\mathbf{Q}_j \in \mathbb{R}^{p \times p}\) is the \(j^{\text{th}}\) row and column block of \(\mathbf{Q}\).
 
@@
 	After resolving (1), see if the MCMC is still trapped in local optima sometimes.
+Now, the chain seems will get stuck in local mode.
+Maybe updating the loading as a whole will remedy the problem a bit.
 	\item

### 2021-09-12 LDS/debug/debug_dC_ind.m

@@ -57,7 +57,7 @@ imagesc(exp(logLam))
 colorbar()
 % clusterPlot(Y, Lab)
 
-ng = 100; % 40
+ng = 1000; % 40
 idx = 50:ng; % 20:ng
 
 %% fitting: MCMC, no update prior
@@ -74,6 +74,10 @@ A_fit = repmat(A,1,1,ng); % true
 b_fit = repmat(b,1,ng); % true
 Q_fit = repmat(Q,1,1,ng); % true
 
+
+dc = [d_fit(i,1) C_fit(i,:,1)]';
+acs = 0;
+tau = 
 for g = 2:ng
     
     disp(g)
@@ -90,14 +94,32 @@ for g = 2:ng
         [mudc,~,niSigdc,~] = newton(derdc,hessdc,...
             [d_fit(i, g-1) C_fit(i,:,g-1)]',1e-8,1000);
         
-        Sigdc = -inv(niSigdc);
-        Sigdc = (Sigdc + Sigdc')/2;
-        dc = mvnrnd(mudc, Sigdc);
+        % let's step further... MH based on posterior mode
+        % proposal distribution
+        % use Cholesky decomposition
+        R = chol(-niSigdc,'lower'); % sparse
+        z = randn(length(mudc), 1) + R'*(mudc -(dc - mudc));
+        dcStar = R'\z;
+        
+        % lhr
+        lhr = sum(log(poisspdf(Y(i,:)', lamdc(dcStar)))) -...
+            sum(log(poisspdf(Y(i,:)', lamdc(dc)))) +...
+            log(mvnpdf(dcStar, mudc0, Sigdc0)) -...
+            log(mvnpdf(dc, mudc0, Sigdc0));
+        
+        if(log(rand(1)) < lhr)
+            dc = dcStar;
+            acs = acs + 1;
+        end
         
         d_fit(i,g) = dc(1);
         C_fit(i,:,g) = dc(2:end);
     end
 end
+
+acs
+
+
 
 figure
 subplot(1,3,1)

### 2021-09-15 LDS/blkDiag/lds_sample_blkDiag.m
Update lds_sample_blkDiag.m

@@ -33,12 +33,12 @@ Q2 = 1e-3*eye(p);
 Q3 = 1e-3*eye(p);
 Q = blkdiag(Q1, Q2, Q3);
 
-% 
+%
 A = eye(size(Q,1));
 while any(imag(eig(A))==0)
     A= randn(size(Q));
     A = A-diag(diag(A));
-	A(squareform(pdist(pLab'))==0)=0;
+    A(squareform(pdist(pLab'))==0)=0;
     A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
     A = A+eye(size(Q,1))*0.92;
 end
@@ -143,10 +143,12 @@ X_fit(:,:,1) = reshape(muXvec, [], T);
 dc_all = [sum(d_fit(:,:,1), 2) sum(C_fit(:,1:p:end), 2) sum(C_fit(:,2:p:end), 2)]';
 acc = zeros(N,1);
 x_all = reshape(X_fit(:,:,1),[],1);
-
+% taux = sqrt(p*T);
+taux = sqrt(length(x_all))/2.38;
+accx = 0;
 for g = 2:ng
     
-    disp(g)
+    %     disp(g)
     
     % (1) update X_fit
     % adaptive smoothing
@@ -170,38 +172,44 @@ for g = 2:ng
         X_tmp = ppasmoo_poissexp_v2(Y,C_tmp,d_tmp,x0_tmp,Q0,A_tmp,b_tmp,Q_tmp);
         [muXvec,~,hess_tmp,~] = newtonGH(gradHess,X_tmp(:),1e-8,1000);
     end
-        
-    R = chol(-hess_tmp,'lower'); % sparse
-    z = randn(length(muXvec), 1) + R'*muXvec;
-    x_all = R'\z;
     
+    %     R = chol(-hess_tmp,'lower'); % sparse
+    %     z = randn(length(muXvec), 1) + R'*muXvec;
+    %     x_all = R'\z;
+    
+    x_MHStart = round(ng/2);
     % let's use MH again...
-%     if g < 1 % g < round(ng/2)
-%         R = chol(-hess_tmp,'lower'); % sparse
-%         z = randn(length(muXvec), 1) + R'*muXvec;
-%         x_all = R'\z;
-%     else
-%         lamX = @(X) exp(C_tmp*X + d_tmp) ;
-%         
-%         R = chol(-hess_tmp,'lower'); % sparse
-%         z = randn(length(x_all), 1) + R'*x_all;
-%         xStar = R'\z;
-%         
[…]

### 2021-09-15 documents/cluster_neural_population.tex
documentation-Poisson MCMC

@@
 
+\textbf{Step Further:}\\
+Using normal approximation may not be accurate enough, and this may lead to a bad mixing.
+We can step further to use the Metropolis-Hasting, based on the normal approximated variance.
+To be specific, the proposal distribution is \(Q(\mathbf{x}^{*}|\mathbf{x}^{(s)}) \sim N(\mathbf{x}^{(s)}, \alpha\mathbf{\Sigma}_{\mathbf{x}})\), where \(\alpha\) is a scalar to make the acceptance rate to be 0.4 to 0.6.
+According to \href{https://projecteuclid.org/journals/statistical-science/volume-16/issue-4/Optimal-scaling-for-various-Metropolis-Hastings-algorithms/10.1214/ss/1015346320.full}{Roberts and Rosenthal, 2001}.
+For high dimensional MH, the optimal proposal is \(N(x, 2.38^2\Sigma/d)\).
+See details \& experiments in Section \ref{sim}: Simulations.
+
 \subsubsection{Update \(\mathbf{x}_{0}\)}
@@
 
-DETOUR: we can write things in the form of MV-(Poisson)-GLM as follows (in cluster j with \(n_j\) observations/ neurons, with corresponding loading be \(\mathbf{d}_{j}\) and \(\mathbf{C}_{j}\)):
+%DETOUR: we can write things in the form of MV-(Poisson)-GLM as follows (in cluster j with \(n_j\) observations/ neurons, with corresponding loading be \(\mathbf{d}_{j}\) and \(\mathbf{C}_{j}\)):
+%
+%\[\log\begin{pmatrix}
+%	\lambda_{11} & \cdots & \lambda_{1n_{j}} \\
+%	\vdots & \ddots & \vdots \\
+%	\lambda_{T1} & \cdots & \lambda_{Tn_{j}} \\
+%\end{pmatrix} = \begin{pmatrix}
+%	1 & \mathbf{x}_{1}^{'} \\
+%	\vdots & \vdots \\
+%	1 & \mathbf{x}_{T}^{'} \\
+%\end{pmatrix}\begin{pmatrix}
+%	\mathbf{d}_{j}^{'} \\
+%	\mathbf{C}_{j}^{'} \\
+%\end{pmatrix} + \mathbf{E}_j\]
+%
+%We can further organize things column-wise. let \(\begin{pmatrix}
+%	1 & \mathbf{x}_{1}^{'} \\
+%	\vdots & \vdots \\
+%	1 & \mathbf{x}_{T}^{'} \\
+%\end{pmatrix} = \mathbf{G}\), then the design matrix and corresponding log-mean response are \(\begin{pmatrix}
+%\mathbf{G} & \mathbf{0} & \cdots & \mathbf{0} \\
+%\mathbf{0} & \mathbf{G} & \cdots & \mathbf{0} \\
+%\vdots & \vdots & \ddots & \vdots \\
+%\mathbf{0} & \cdots & \cdots & \mathbf{G} \\
+%\end{pmatrix}\) and \({\log\begin{pmatrix}
+%	\lambda_{11}^{(j)} & \cdots & \lambda_{T1}^{(j)} & \lambda_{12}^{(j)} & \cdots & \lambda_{T2}^{(j)} & \cdots & \lambda_{Tn_{j}}^{(j)} \\
+%\end{pmatrix}}'\).
+%
+%However, when I use the Laplace approximation and NR, because of the block-diagonal structure of design matrix, the error is always independent observation by observation.
+This is equivalent to observation-wise update as above.
+%
+%\textbf{Maybe I did something wrong?}
+%The point to say this is to make loading within each cluster depends on each other.
+Check these two later(\href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}).
+%
+%Now I do it in another way: make the prior cluster-dependent...
 
-\[\log\begin{pmatrix}
-	\lambda_{11} & \cdots & \lambda_{1n_{j}} \\
-	\vdots & \ddots & \vdots \\
-	\lambda_{T1} & \cdots & \lambda_{Tn_{j}} \\
-\end{pmatrix} = \begin{pmatrix}
-	1 & \mathbf{x}_{1}^{'} \\
-	\vdots & \vdots \\
-	1 & \mathbf{x}_{T}^{'} \\
-\end{pmatrix}\begin{pmatrix}
-	\mathbf{d}_{j}^{'} \\
-	\mathbf{C}_{j}^{'} \\
-\end{pmatrix} + \mathbf{E}_j\]
-
-We can further organize things column-wise. let \(\begin{pmatrix}
-	1 & \mathbf{x}_{1}^{'} \\
-	\vdots & \vdots \\
-	1 & \mathbf{x}_{T}^{'} \\
-\end{pmatrix} = \mathbf{G}\), then the design matrix and corresponding log-mean response are \(\begin{pmatrix}
-\mathbf{G} & \mathbf{0} & \cdots & \mathbf{0} \\
-\mathbf{0} & \mathbf{G} & \cdots & \mathbf{0} \\
-\vdots & \vdots & \ddots & \vdots \\
-\mathbf{0} & \cdots & \cdots & \mathbf{G} \\
-\end{pmatrix}\) and \({\log\begin{pmatrix}
-	\lambda_{11}^{(j)} & \cdots & \lambda_{T1}^{(j)} & \lambda_{12}^{(j)} & \cdots & \lambda_{T2}^{(j)} & \cdots & \lambda_{Tn_{j}}^{(j)} \\
-\end{pmatrix}}'\).
-
-However, when I use the Laplace approximation and NR, because of the block-diagonal structure of design matrix, the error is always independent observation by observation.
-This is equivalent to observation-wise update as above.
-
-\textbf{Maybe I did something wrong?}
-The point to say this is to make loading within each cluster depends on each other.
-Check these two later(\href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}).
-
-Now I do it in another way: make the prior cluster-dependent...
+\textbf{Step further}:
+Again, use MH to sample the posterior, based on the normal approximated variance.
+Since the dimension of d and C is not large (just d = 3) in the simulation, I didn't use the optimal scalar yet.
+\textbf{Modify it later}.
 
@@
 
-\section{Simulations}
+\section{Simulations}\label{sim}
 
@@
 The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{here}.
-After running 2000 iterations, the trace of the sample is shown in Figure \ref{lds norms}.
+
+Here, I first compare 10k iterations for three methods (consider high dimension of \(\mathbf{x}_t\)): (1) both \(\mathbf{x}_t\) and loading (\(\mathbf{d}\) and \(\mathbf{C}\)) are updated by normal approximation.
+(2) update \(\mathbf{x}_t\) by normal approximation, but update loading by MH.
+(3) update both \(\mathbf{x}_t\) and loading by MH.
+
+The fitting results are average of iteration 5k to 10k.
+
+\textbf{Update \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by normal approximation}
+
+The trace plots are in Figure \ref{double N: trace}
 \begin{figure}[h!]
+	\makebox[\linewidth][c]{%
 		\centering
-	\includegraphics[width = .8\textwidth]{lds_blk_trace.png}
-	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
-	\label{lds norms}
-\end{figure}\\
+		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleN.png}}%
+		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleN.png}}%
+	}
+	\caption{both Normal Approx., trace plots}
+	\label{double N: trace}
+\end{figure}
 
-Some results are in Figure \ref{fig:LDS labeled}. These are means from iteration 1000 to 2000.
+Well, the chain hasn't achieve the stationary distribution.
+So there's no need to see further results.
+But to show the improvement and for comparison, I still show the fitting of overall firing rate, latents and dynamics in Figure \ref{double N: fit}.
 
@@
 		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{lds_blk_FR.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{lds_blk_latent.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{lds_blk_A.png}}%
+		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleN.png}}%
+		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleN.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleN.png}}%
 	}
-	\caption{LDS sample with labels}
-	\label{fig:LDS labeled}
+	\caption{both Normal Approx., fit}
+	\label{double N: fit}
 \end{figure}
 
-Personally, I think the fitting is not bad and the mixing is fine (in previous version, we may need 10000 iterations to get a not bad result. But here 2000 iterations did even better than before).
+\textbf{Update \(\mathbf{x}_t\) by normal approximation and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
+
+By using the normal approximated posterior variance as the proposal variance, the acceptance rates are around 0.45 (Since there are 30 neurons, there are 30 acceptance rates).
+The trace plots are in Figure \ref{dc MH: trace}
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_dcMH.png}}%
+		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_dcMH.png}}%
+	}
+	\caption{latent Normal Approx. + loading MH, trace plots}
+	\label{dc MH: trace}
+\end{figure}
+
+Looks much better, although still not perfect yet. The fitting of overall firing rate, latents and dynamics are in Figure \ref{dc MH: fit}.
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_dcMH.png}}%
+		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_dcMH.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_dcMH.png}}%
+	}
+	\caption{latent Normal Approx. + loading MH, fit}
+	\label{dc MH: fit}
+\end{figure}
+
+See, nearly perfect (especially) for latent.
+
+\textbf{Update both \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
+
+The acceptance rate (by using the optimal scalar) for latent is 0.6227, which is a bit high (but acceptable). Again, the acceptance rates for loading are around 0.45.
+
+The trace plots are in Figure \ref{double MH: trace}
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleMH.png}}%
+		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleMH.png}}%
+	}
+	\caption{both MH, trace plots}
+	\label{double MH: trace}
+\end{figure}
+
+The trace for dynamics is terrible... Maybe this is caused by high dimension of latent (\(d=6000\) in this case).
+
+The fitting of overall firing rate, latents and dynamics are in Figure \ref{double MH: fit}.
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleMH.png}}%
+		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleMH.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleMH.png}}%
+	}
+	\caption{both MH, fit}
+	\label{double MH: fit}
+\end{figure}
+
+OK, the fitting sucks...
+
+One natural idea is to combine MH and normal approximation in latent: at first few steps, use normal approximation at conditional posterior mode to bring things into somewhere close to mode quickly.
+Then use MH to stabilize everything.
+
+In the following, I still sample loading by MH. But for latent, the first half iteration is sampled by normal approximation, while the remainder is sampled from MH.
+
+The red line marks when I switch the sampling method for \(\mathbf{x}_t\).
+
+The trace plots are in Figure \ref{half MH: trace}
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_halfMH.png}}%
+		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_halfMH.png}}%
+	}
+	\caption{half MH for latent, trace plots}
+	\label{half MH: trace}
+\end{figure}
+
+The fitting of overall firing rate, latents and dynamics are in Figure \ref{half MH: fit}.
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_halfMH.png}}%
+		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_halfMH.png}}%
+		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_halfMH.png}}%
+	}
+	\caption{half MH for latent, fit}
+	\label{half MH: fit}
+\end{figure}
+
+Well, everything is stable now. But this implementation is too naive. Maybe use some more advanced adaptive sampling method?
+
+%\begin{figure}[h!]
+%	\centering
+%	\includegraphics[width = .8\textwidth]{lds_blk_trace.png}
+%	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
+%	\label{lds norms}
+%\end{figure}\\
+%Some results are in Figure \ref{fig:LDS labeled}. These are means from iteration 1000 to 2000.
+%
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{lds_blk_FR.png}}%
+%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{lds_blk_latent.png}}%
+%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{lds_blk_A.png}}%
+%	}
+%	\caption{LDS sample with labels}
+%	\label{fig:LDS labeled}
+%\end{figure}
+%
+%Personally, I think the fitting is not bad and the mixing is fine (in previous version, we may need 10000 iterations to get a not bad result. But here 2000 iterations did even better than before).
 
@@
 If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.
-
-\textbf{Think more on it.} Maybe we can do MV-(Poisson)-GLM, to make all loading within cluster correlated.
 
@@
 
-This time I ran 1000 iterations. The trace of the samples is shown in Figure \ref{noA norms}.
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width = .8\textwidth]{noA_blk_trace.png}
-	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
-	\label{noA norms}
-\end{figure}\\
+\textbf{I have deleted the old results to avoid confusion. I will post the new results here later.}
 
-Some results are in Figure \ref{fig:noA labeled}. These are means from iteration 500 to 1000.
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{noA_blk_FR.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{noA_blk_latent.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{noA_blk_A.png}}%
-	}
-	\caption{LDS sample with labels}
-	\label{fig:noA labeled}
-\end{figure}
-
-This time the mixing of loading (\(\mathbf{d}\) and \(\mathbf{C}\)) is not fine. That reminds me that \textbf{I may need to make loading within cluster related by MV-(Poisson)-GLM directly.}
+%This time I ran 1000 iterations. The trace of the samples is shown in Figure \ref{noA norms}.
+%\begin{figure}[h!]
+%	\centering
+%	\includegraphics[width = .8\textwidth]{noA_blk_trace.png}
+%	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
+%	\label{noA norms}
+%\end{figure}\\
+%
+%Some results are in Figure \ref{fig:noA labeled}. These are means from iteration 500 to 1000.
+%
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{noA_blk_FR.png}}%
+%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{noA_blk_latent.png}}%
+%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{noA_blk_A.png}}%
+%	}
+%	\caption{LDS sample with labels}
+%	\label{fig:noA labeled}
+%\end{figure}
+%
+%This time the mixing of loading (\(\mathbf{d}\) and \(\mathbf{C}\)) is not fine. That reminds me that \textbf{I may need to make loading within cluster related by MV-(Poisson)-GLM directly.}
 
 \subsubsection{Unlabeled Data: Clustering}
+\textbf{The clustering results are old. Update when the MFM is implemented}
+
 Again, all results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
@@
 	\item
-	Think about MV-(Poisson)-GLM.
+%	Think about MV-(Poisson)-GLM.
 Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
-	\item
-	After resolving (1), see if the MCMC is still trapped in local optima sometimes.
+	Do MH adaptively.
+%	\item
+%	After resolving (1), see if the MCMC is still trapped in local optima sometimes.
 Now, the chain seems will get stuck in local mode.
@@
 	\item
+	Implement the mixture of finite mixture (MFM) by Jeff Miller.
+	\item
 	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
-	\item
-	improve DPMM.
+%	\item
+%	improve DPMM.
 In current brute force implementation, the number of potential clusters can even go beyond number of neurons (\(N\)).

### 2021-09-15 LDS/debug/debug_X.m

@@ -1,0 +1,188 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(N,1);
+C_trans = zeros(N, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(N, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+%%
+rng(3)
+ng = 1000;
+
+X_fit = zeros(nClus*p, T, ng);
+x0_fit = repmat(x0, 1, ng); % true
+d_fit = repmat(d,1,ng); % true
+C_fit = repmat([sum(C_trans(:,1:p:end), 2) sum(C_trans(:,2:p:end), 2)],...
[…]

### 2021-09-27 documents/cluster_neural_population.tex
documentation

@@
 	\item
-	\(\mathbf{x}_{t} \in \mathbb{R}^{\text{Jp}}\): all latent at \(t\).
+	\(\mathbf{x}_{t} \in \mathbb{R}^{Jp}\): all latent at \(t\).
 	\item
@@
 	\item
-	\(\mathbf{x}_{0} \in \mathbb{R}^{\text{Jp}}\) and
-	\(\mathbf{Q}_{0} \in \mathbb{R}^{\text{Jp} \times Jp}\): mean and covariance of
+	\(\mathbf{x}_{0} \in \mathbb{R}^{Jp}\) and
+	\(\mathbf{Q}_{0} \in \mathbb{R}^{Jp \times Jp}\): mean and covariance of
 	\(\mathbf{x}_{1}\). For simplicity, assume \(\mathbf{Q}_{0}\) is known.
@@
 	\[\mathbf{x}_{0} \sim N(\bm{\mu}_{\mathbf{x}_{00}},\ \mathbf{\Sigma}_{\mathbf{x}_{00}})\]
-	, where \(\bm{\mu}_{\mathbf{x}_{00}} = \mathbf{0}_{\text{Jp}}\) and \(\mathbf{\Sigma}_{\mathbf{x}_{00}} = \mathbf{I}_{\text{Jp}}\)
+	, where \(\bm{\mu}_{\mathbf{x}_{00}} = \mathbf{0}_{Jp}\) and \(\mathbf{\Sigma}_{\mathbf{x}_{00}} = \mathbf{I}_{Jp}\)
 	\item
@@
 	, where \(\mathbf{f}_0 = vec(\mathbf{F}_0)\).If the number of cluster is \(J\), \(\nu_{\mathbf{Q}_{0}} = Jp + 2\),
-	\(\Psi_{\mathbf{Q}_{0}} = \mathbf{I}_{\text{Jp}} \times 10^{- 4}\). (To make the mean of \(\mathbf{Q}\) loosely centered around
+	\(\Psi_{\mathbf{Q}_{0}} = \mathbf{I}_{Jp} \times 10^{- 4}\). (To make the mean of \(\mathbf{Q}\) loosely centered around
 	\(\mathbf{I}_{Jp} \times 10^{- 4}\)), \(\mathbf{F}_0 = \begin{pmatrix}
 		\mathbf{0}'_{Jp} \\
-		\mathbf{I}_{\text{Jp}} \\
+		\mathbf{I}_{Jp} \\
 	\end{pmatrix}\) and \(\bm{\Lambda}_0 = \mathbf{I}_{Jp + 1}\)
@@
 \begin{align*}
-	\mathbf{Q}|\mathbf{Y}_{\mathbf{\text{bA}}},\mathbf{X}_{\mathbf{\text{bA}}},\ldots &\sim W^{-1}(\Psi_{\mathbf{Q}}, \nu_{\mathbf{Q}})\\
-	\mathbf{f}|\mathbf{Y}_{\mathbf{\text{bA}}},\mathbf{X}_{\mathbf{\text{bA}}},\mathbf{Q},\ldots &\sim N(vec(\mathbf{F}_\mathbf{bA}),  \mathbf{Q}\bigotimes \bm{\Lambda}_\mathbf{bA}^{-1})
+	\mathbf{Q}|\mathbf{Y}_{\mathbf{bA}},\mathbf{X}_{\mathbf{bA}},\ldots &\sim W^{-1}(\Psi_{\mathbf{Q}}, \nu_{\mathbf{Q}})\\
+	\mathbf{f}|\mathbf{Y}_{\mathbf{bA}},\mathbf{X}_{\mathbf{bA}},\mathbf{Q},\ldots &\sim N(vec(\mathbf{F}_\mathbf{bA}),  \mathbf{Q}\bigotimes \bm{\Lambda}_\mathbf{bA}^{-1})
 \end{align*}
@@
 
-\section{Model Check [Problems for \(\mathbf{b}\) and \(\mathbf{A}\) partially solved]} \label{problems}
-
-\subsection{Estimation of \(\mathbf{d}\) and \(\mathbf{C}\) alone}
-First, turn off all others except for loading (related) parameters and set them to be true values.
+%\section{Model Check [Problems for \(\mathbf{b}\) and \(\mathbf{A}\) partially solved]} \label{problems}
+%
+%\subsection{Estimation of \(\mathbf{d}\) and \(\mathbf{C}\) alone}
+%First, turn off all others except for loading (related) parameters and set them to be true values.
 There are 2 versions: (1) update priors as in subsection \ref{loading prior} and (2) no update of priors, and just set the prior for each as the standard normal.
-For the estimation with prior update, the results are mean from iteration 50 to 100 (Figure \ref{loading alone}).
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[]{\includegraphics[width=0.6\textwidth]{image016.png}}%
-		\subfigure[]{\includegraphics[width=0.6\textwidth]{image017.png}}%
-	}
-	\caption{estimation of loading}
-	\label{loading alone}
-\end{figure}
-
-The mixing of convergence for loading priors (in subsection \ref{loading prior}) are fine.
+%For the estimation with prior update, the results are mean from iteration 50 to 100 (Figure \ref{loading alone}).
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image016.png}}%
+%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image017.png}}%
+%	}
+%	\caption{estimation of loading}
+%	\label{loading alone}
+%\end{figure}
+%
+%The mixing of convergence for loading priors (in subsection \ref{loading prior}) are fine.
 The update of priors don't influence results a lot, but this will help clustering a lot.
@@
 Let's see what happens when estimation of latents \(\mathbf{x}_t\) is on at the same time.
-
-\subsection{Estimation of \(\mathbf{d}\), \(\mathbf{C}\) and \(\mathbf{x}_{t}\)}
-When the estimation of \(\mathbf{x}_t\) is on, we need more iterations. The results are mean from iteration 500 to 1000 (Figure \ref{loading and latent}).
-
-The estimation of loading:
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[]{\includegraphics[width=0.6\textwidth]{image018.png}}%
-		\subfigure[]{\includegraphics[width=0.6\textwidth]{image020.png}}%
-	}
-	\caption{estimation of loading, with latents on}
-	\label{loading and latent}
-\end{figure}
-
-And the estimation of corresponding latents (Figure \ref{loading and latents}):
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[no update of prior]{\includegraphics[width=0.6\textwidth]{image019.png}}%
-		\subfigure[with update of prior]{\includegraphics[width=0.6\textwidth]{image021.png}}%
-	}
-	\caption{estimation of loading}
-	\label{loading and latents}
-\end{figure}
-
-Things are still fine. The remaining parts may influence latent estimations are dynamics (\(\mathbf{b}\), \(\mathbf{A}\)) and prior process noise \(\mathbf{Q}\).
-
-\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\), \(\mathbf{Q}\) and \(\mathbf{x}_t\)}
-I show the results of block-diagonal (Figure \ref{bAQX_blkDiag}) and full (Figure \ref{bAQX_full}) \(\mathbf{Q}\).
+%
+%\subsection{Estimation of \(\mathbf{d}\), \(\mathbf{C}\) and \(\mathbf{x}_{t}\)}
+%When the estimation of \(\mathbf{x}_t\) is on, we need more iterations. The results are mean from iteration 500 to 1000 (Figure \ref{loading and latent}).
+%
+%The estimation of loading:
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image018.png}}%
+%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image020.png}}%
+%	}
+%	\caption{estimation of loading, with latents on}
+%	\label{loading and latent}
+%\end{figure}
+%
+%And the estimation of corresponding latents (Figure \ref{loading and latents}):
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[no update of prior]{\includegraphics[width=0.6\textwidth]{image019.png}}%
+%		\subfigure[with update of prior]{\includegraphics[width=0.6\textwidth]{image021.png}}%
+%	}
+%	\caption{estimation of loading}
+%	\label{loading and latents}
+%\end{figure}
+%
+%Things are still fine. The remaining parts may influence latent estimations are dynamics (\(\mathbf{b}\), \(\mathbf{A}\)) and prior process noise \(\mathbf{Q}\).
+%
+%\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\), \(\mathbf{Q}\) and \(\mathbf{x}_t\)}
+%I show the results of block-diagonal (Figure \ref{bAQX_blkDiag}) and full (Figure \ref{bAQX_full}) \(\mathbf{Q}\).
 These results are from 2000 MCMC samples.
 The latent and dynamics are from average from 500 to 2000 iterations.
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_trace.png}}%
-		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_latent.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_A.png}}%
-	}
-	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, block-diagonal process noise}
-	\label{bAQX_blkDiag}
-\end{figure}
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_full_trace.png}}%
-		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_full_latent.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_full_A.png}}%
-	}
-	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, full process noise}
-	\label{bAQX_full}
-\end{figure}
-
-Both block-diagonal and full looks fine, but the mixing of  full process noise version looks better.
+%
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_trace.png}}%
+%		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_latent.png}}%
+%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_A.png}}%
+%	}
+%	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, block-diagonal process noise}
+%	\label{bAQX_blkDiag}
+%\end{figure}
+%
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_full_trace.png}}%
+%		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_full_latent.png}}%
+%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_full_A.png}}%
+%	}
+%	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, full process noise}
+%	\label{bAQX_full}
+%\end{figure}
+%
+%Both block-diagonal and full looks fine, but the mixing of  full process noise version looks better.
 
@@
 
-There are two set of simulation examples.
+\textbf{\textcolor{red}{All the simulation results are old and problematic. Update them later.}}
+
+%There are two set of simulation examples.
 The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics.
@@
 clustering, all the results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
-
-\subsection{Simulation 1: Generate from LDS Directly}
-In this simulation, there are 3 clusters with 10 neurons in each cluster.
+%
+%\subsection{Simulation 1: Generate from LDS Directly}
+%In this simulation, there are 3 clusters with 10 neurons in each cluster.
 The dimension of latents in each cluster is 2.
@@
 The details of simulation can be found in the first section of the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{simulation 1}.
-
-\subsubsection{Labeled Data: No Clustering}
-The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{here}.
-
-Here, I first compare 10k iterations for three methods (consider high dimension of \(\mathbf{x}_t\)): (1) both \(\mathbf{x}_t\) and loading (\(\mathbf{d}\) and \(\mathbf{C}\)) are updated by normal approximation.
+%
+%\subsubsection{Labeled Data: No Clustering}
+%The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{here}. 
+%
+%Here, I first compare 10k iterations for three methods (consider high dimension of \(\mathbf{x}_t\)): (1) both \(\mathbf{x}_t\) and loading (\(\mathbf{d}\) and \(\mathbf{C}\)) are updated by normal approximation.
 (2) update \(\mathbf{x}_t\) by normal approximation, but update loading by MH.
 (3) update both \(\mathbf{x}_t\) and loading by MH.
-
-The fitting results are average of iteration 5k to 10k.
-
-\textbf{Update \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by normal approximation}
-
-The trace plots are in Figure \ref{double N: trace}
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleN.png}}%
-		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleN.png}}%
-	}
-	\caption{both Normal Approx., trace plots}
-	\label{double N: trace}
-\end{figure}
-
-Well, the chain hasn't achieve the stationary distribution.
+%
+%The fitting results are average of iteration 5k to 10k.
+%
+%\textbf{Update \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by normal approximation}
+%
+%The trace plots are in Figure \ref{double N: trace}
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleN.png}}%
+%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleN.png}}%
+%	}
+%	\caption{both Normal Approx., trace plots}
+%	\label{double N: trace}
+%\end{figure}
+%
+%Well, the chain hasn't achieve the stationary distribution.
 So there's no need to see further results.
 But to show the improvement and for comparison, I still show the fitting of overall firing rate, latents and dynamics in Figure \ref{double N: fit}.
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleN.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleN.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleN.png}}%
-	}
-	\caption{both Normal Approx., fit}
-	\label{double N: fit}
-\end{figure}
-
-\textbf{Update \(\mathbf{x}_t\) by normal approximation and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
-
-By using the normal approximated posterior variance as the proposal variance, the acceptance rates are around 0.45 (Since there are 30 neurons, there are 30 acceptance rates).
-The trace plots are in Figure \ref{dc MH: trace}
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_dcMH.png}}%
-		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_dcMH.png}}%
-	}
-	\caption{latent Normal Approx. + loading MH, trace plots}
-	\label{dc MH: trace}
-\end{figure}
-
-Looks much better, although still not perfect yet. The fitting of overall firing rate, latents and dynamics are in Figure \ref{dc MH: fit}.
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_dcMH.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_dcMH.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_dcMH.png}}%
-	}
-	\caption{latent Normal Approx. + loading MH, fit}
-	\label{dc MH: fit}
-\end{figure}
-
-See, nearly perfect (especially) for latent.
-
-\textbf{Update both \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
-
-The acceptance rate (by using the optimal scalar) for latent is 0.6227, which is a bit high (but acceptable). Again, the acceptance rates for loading are around 0.45.
-
-The trace plots are in Figure \ref{double MH: trace}
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleMH.png}}%
-		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleMH.png}}%
-	}
-	\caption{both MH, trace plots}
-	\label{double MH: trace}
-\end{figure}
-
-The trace for dynamics is terrible... Maybe this is caused by high dimension of latent (\(d=6000\) in this case).
-
-The fitting of overall firing rate, latents and dynamics are in Figure \ref{double MH: fit}.
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleMH.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleMH.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleMH.png}}%
-	}
-	\caption{both MH, fit}
-	\label{double MH: fit}
-\end{figure}
-
-OK, the fitting sucks...
-
-One natural idea is to combine MH and normal approximation in latent: at first few steps, use normal approximation at conditional posterior mode to bring things into somewhere close to mode quickly.
+%
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleN.png}}%
+%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleN.png}}%
+%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleN.png}}%
+%	}
+%	\caption{both Normal Approx., fit}
+%	\label{double N: fit}
+%\end{figure}
+%
+%\textbf{Update \(\mathbf{x}_t\) by normal approximation and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
+%
+%By using the normal approximated posterior variance as the proposal variance, the acceptance rates are around 0.45 (Since there are 30 neurons, there are 30 acceptance rates).
+%The trace plots are in Figure \ref{dc MH: trace}
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_dcMH.png}}%
+%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_dcMH.png}}%
+%	}
+%	\caption{latent Normal Approx. + loading MH, trace plots}
+%	\label{dc MH: trace}
+%\end{figure}
+%
+%Looks much better, although still not perfect yet. The fitting of overall firing rate, latents and dynamics are in Figure \ref{dc MH: fit}.
+%
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_dcMH.png}}%
+%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_dcMH.png}}%
+%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_dcMH.png}}%
+%	}
+%	\caption{latent Normal Approx. + loading MH, fit}
+%	\label{dc MH: fit}
+%\end{figure}
+%
+%See, nearly perfect (especially) for latent.
+%
+%\textbf{Update both \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
+%
+%The acceptance rate (by using the optimal scalar) for latent is 0.6227, which is a bit high (but acceptable). Again, the acceptance rates for loading are around 0.45.
+%
+%The trace plots are in Figure \ref{double MH: trace}
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleMH.png}}%
+%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleMH.png}}%
+%	}
+%	\caption{both MH, trace plots}
+%	\label{double MH: trace}
+%\end{figure}
+%
+%The trace for dynamics is terrible... Maybe this is caused by high dimension of latent (\(d=6000\) in this case).
+%
+%The fitting of overall firing rate, latents and dynamics are in Figure \ref{double MH: fit}.
+%
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleMH.png}}%
+%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleMH.png}}%
+%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleMH.png}}%
+%	}
+%	\caption{both MH, fit}
+%	\label{double MH: fit}
+%\end{figure}
+%
+%OK, the fitting sucks...
+%
+%One natural idea is to combine MH and normal approximation in latent: at first few steps, use normal approximation at conditional posterior mode to bring things into somewhere close to mode quickly.
 Then use MH to stabilize everything.
-
-In the following, I still sample loading by MH. But for latent, the first half iteration is sampled by normal approximation, while the remainder is sampled from MH.
-
-The red line marks when I switch the sampling method for \(\mathbf{x}_t\).
-
-The trace plots are in Figure \ref{half MH: trace}
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_halfMH.png}}%
-		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_halfMH.png}}%
-	}
-	\caption{half MH for latent, trace plots}
-	\label{half MH: trace}
-\end{figure}
-
-The fitting of overall firing rate, latents and dynamics are in Figure \ref{half MH: fit}.
-
-\begin{figure}[h!]
-	\makebox[\linewidth][c]{%
-		\centering
-		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_halfMH.png}}%
-		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_halfMH.png}}%
-		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_halfMH.png}}%
-	}
-	\caption{half MH for latent, fit}
-	\label{half MH: fit}
-\end{figure}
-
-Well, everything is stable now. But this implementation is too naive. Maybe use some more advanced adaptive sampling method?
+%
+%In the following, I still sample loading by MH. But for latent, the first half iteration is sampled by normal approximation, while the remainder is sampled from MH.
+%
+%The red line marks when I switch the sampling method for \(\mathbf{x}_t\).
+% 
+%The trace plots are in Figure \ref{half MH: trace}
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_halfMH.png}}%
+%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_halfMH.png}}%
+%	}
+%	\caption{half MH for latent, trace plots}
+%	\label{half MH: trace}
+%\end{figure}
+%
+%The fitting of overall firing rate, latents and dynamics are in Figure \ref{half MH: fit}.
+%
+%\begin{figure}[h!]
+%	\makebox[\linewidth][c]{%
+%		\centering
+%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_halfMH.png}}%
+%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_halfMH.png}}%
+%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_halfMH.png}}%
+%	}
+%	\caption{half MH for latent, fit}
+%	\label{half MH: fit}
+%\end{figure}
+%
+%Well, everything is stable now. But this implementation is too naive. Maybe use some more advanced adaptive sampling method?
 
@@
 
-I also fit the model by assuming one cluster.
+%I also fit the model by assuming one cluster.
 Again, the fitting for overall mean firing rate is perfect.
 This is why it's important to make the loading, \(\mathbf{d}\) and \(\mathbf{C}\), also be cluster-dependent (loading within clusters are correlated).
-
-If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.
-
-\subsubsection{Unlabeled Data: Clustering}
-To give the full path of clustering, I show results in GIFs.
+%
+%If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.
+%
+%
+%\subsubsection{Unlabeled Data: Clustering}
+%To give the full path of clustering, I show results in GIFs.
 All results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
 The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.
-
-Basically, in my current implementation, algorithms are good to merge clusters.
+%
+%Basically, in my current implementation, algorithms are good to merge clusters.
 However, generating new clusters is very hard...
 In other words, the newly generated cluster will usually not be sampled.
-
-\textbf{This is a big problem}. This makes DPMM loses its power. \textbf{Fix that later}.
-
-\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
-In this simulation, there are again 3 clusters with 2 latents and 10 neurons in each.
+%
+%\textbf{This is a big problem}. This makes DPMM loses its power. \textbf{Fix that later}.
+%
+%\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
+%In this simulation, there are again 3 clusters with 2 latents and 10 neurons in each.
 Now, the latents are generated directly without specifying the underlying linear dynamics of latents.
@@
 The details of simulation can be found in the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/unspecifiedA_sample_blkDiag.m}{code}.
-
-\subsubsection{Labeled Data: No Clustering}
-As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster.
+%
+%\subsubsection{Labeled Data: No Clustering}
+%As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster.
 The code can be found \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/unspecifiedA_sample_blkDiag.m}{here}.
-
-\textbf{I have deleted the old results to avoid confusion. I will post the new results here later.}
+%
+%
+%\textbf{I have deleted the old results to avoid confusion. I will post the new results here later.}
 
@@
 
-\subsubsection{Unlabeled Data: Clustering}
-\textbf{The clustering results are old. Update when the MFM is implemented}
-
-Again, all results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
+%\subsubsection{Unlabeled Data: Clustering}
+%\textbf{The clustering results are old. Update when the MFM is implemented}
+%
+%Again, all results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
 The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.
-
-Besides fitting 3 clusters with 10 neurons each, I further fit a larger scale example (50 neurons for each cluster).
+%
+%Besides fitting 3 clusters with 10 neurons each, I further fit a larger scale example (50 neurons for each cluster).
+
+\section{Problem from Affine Transformation}
+Let me review the model first.\\
+Assume there are \(J\)
+clusters, with \(n\) neurons in each cluster. Each cluster is governed
+by \(p\) latent vectors. \(\bm{\lambda}_{t}^{(j)} \in \mathbb{R}^{n}\),
+\(\mathbf{x}_{t}^{(j)} \in \mathbb{R}^{p}\), \(\mathbf{d}^{(j)} \in \mathbb{R}^{n}\) and
+\(\mathbf{C}^{(j)} \in \mathbb{R}^{n \times p}\), with \(j = 1,\ldots,J\), be the
+mean firing rate, latent state, intercept and loading (at time $t$) for
+each cluster. Then I just stack all the thing together, and denote
+corresponding things as \(\bm{\lambda}_{t} \in R^{\text{Jn}}\),
+\(\mathbf{x}_{t} \in \mathbb{R}^{Jp}\), \(\mathbf{d} \in \mathbb{R}^{\text{Jn}}\)
+and \(\mathbf{C} \in \mathbb{R}^{Jn \times Jp}\). Notice \(\mathbf{C}\) is block
+diagonal.
+
+\[\log{\begin{pmatrix}
+		\ \bm{\lambda}_{t}^{(1)} \\
+		\vdots \\
+		\ \bm{\lambda}_{t}^{(J)} \\
+	\end{pmatrix} =}\log\bm{\lambda}_{t} = \begin{pmatrix}
+	\ \mathbf{d}^{(1)} \\
+	\vdots \\
+	\ \mathbf{d}^{(J)} \\
+\end{pmatrix} + \begin{pmatrix}
+	\mathbf{C}^{(1)} & \cdots & 0 \\
+	\vdots & \ddots & \vdots \\
+	0 & \cdots & \mathbf{C}^{(J)} \\
+\end{pmatrix}\begin{pmatrix}
+	\ \mathbf{x}_{t}^{(1)} \\
+	\vdots \\
+	\ \mathbf{x}_{t}^{(J)} \\
+\end{pmatrix} = \mathbf{d} + \mathbf{C}\mathbf{x}_{t}\]
+
+The latent states progress linearly with a Gaussian noise:
+
+\[\mathbf{x}_{t + 1}|\mathbf{x}_{t} \sim N(\mathbf{A}\mathbf{x}_{t} + \mathbf{b},\ \mathbf{Q})\]
+
+, where \(\mathbf{A} \in \mathbb{R}^{Jp \times Jp}\),
+\(\mathbf{b} \in \mathbb{R}^{Jp}\) and
+\(\mathbf{Q} \in \mathbb{R}^{Jp \times Jp}\).
+
+Now assume we take an affine transformation of latent states as
+\(\mathbf{x}_{t}^{*} = \mathbf{M}\mathbf{x}_{t} + \mathbf{g}\). To
+preserve the block diagonal structure of \(\mathbf{C}\), \(\mathbf{M}\)
+is also block diagonal. Then,
+\begin{align*}
+	\log\bm{\lambda}_{t} &= \mathbf{d} + \mathbf{C}\mathbf{M}^{- 1}\left( \mathbf{M}\mathbf{x}_{t} + \mathbf{g} \right) - \mathbf{C}\mathbf{M}^{- 1}\mathbf{g}\\
+	&= \left( \mathbf{d} - \mathbf{C}\mathbf{M}^{- 1}\mathbf{g} \right) + \mathbf{C}\mathbf{M}^{- 1}(\mathbf{M}\mathbf{x}_{t} + \mathbf{g})\\
+	\mathbf{M}\mathbf{x}_{t + 1} + \mathbf{g}|\mathbf{x}_{t} &\sim N_{Jp}\left( \mathbf{M}\left( \mathbf{A}\mathbf{x}_{t} + \mathbf{b} \right)+\mathbf{g}, \mathbf{MQM}' \right)\\
+	&= N_{Jp}(\mathbf{MAM}^{- 1}\left( \mathbf{M}\mathbf{x}_{t} + \mathbf{g} \right) + \left( \mathbf{Mb} + \mathbf{g} - \mathbf{MAM}^{- 1}\mathbf{g} \right), \mathbf{MQM}')
+\end{align*}
+
+That means, when \(\left\{ \mathbf{d},\mathbf{C},\mathbf{x}_t,\mathbf{A},\mathbf{b},\mathbf{Q} \right\}\) is a solution,
+\(\left\{ \mathbf{d}^*,\mathbf{C}^*,\mathbf{x}^*_t,\mathbf{A}^*,\mathbf{b}^*,\mathbf{Q}^* \right\}\) is also a
+solution, where
+\(\mathbf{d}^* = \mathbf{d} - \mathbf{C}\mathbf{M}^{- 1}\mathbf{g}\),
+\(\mathbf{C}^* = \mathbf{CM}^{- 1}\), \(\mathbf{x}_{t}^{*} = \mathbf{M}\mathbf{x}_{t} + \mathbf{g}\),
+\(\mathbf{A}^* = \mathbf{MAM}^{- 1}\),
+\(\mathbf{b}^* = \mathbf{Mb} + \mathbf{g} - \mathbf{MAM}^{- 1}\mathbf{g}\)
+and \(\mathbf{Q}^* = \mathbf{MQM}'\), for any block diagonal \(\mathbf{M}\). Therefore, we must add some constraints to ensure the unique solution/ stationary
+posterior!
+
+\subsection{Constraints}
+An affine transformation contains translation and linear transformation.
+
+\subsubsection{Translation}
+There are two strategies: (1) just ignore \(\mathbf{d}\) and let \(\mathbf{Cx}_{t}\)
+absorb all the intercepts; (2) center \(\mathbf{x}_t\) such that mean for each
+row of \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0.
+However, because of the block-diagonal constraints on $\mathbf{M}$, the strategy (1) may be problematic.
+Moreover, when $\mathbf{d}$ is non-zero, the \(\mathbf{x}_t\) will looks ugly.
+Another benefit for (2) is that we can use $\mathbf{d}$ to help clustering.
+However, the constraint on translation cannot resolve problems from linear transformation, since any linear transformation will preserve the zero row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\).
+
+\subsubsection{Linear Transformation}
+To mimic PCA, one natural way is to make the loading orthogonal. Again, I'm considering two strategies: (1) make
+\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\); (2) make \( (\mathbf{C}'^{(1)},\cdots, \mathbf{C}'^{(J)})
+\begin{pmatrix}
+	\mathbf{C}^{(1)} \\
+	\vdots \\
+	\mathbf{C}^{(J)} \\
+\end{pmatrix} = \mathbf{I}_{p}\).
+Personally, I prefer the first one, for both computational and clustering reasons.
+Computationally, if I do things cluster-wise, I can easily use parallel when dimension becomes large.
+
+\textbf{However, this is not enough!} Consider \(\mathbf{x}_{t}^{*} = \mathbf{Ux}_{t}\), to
+preserve the orthogonality and block diagonal on \(\mathbf{C}\), \(\mathbf{U}\) must be
+block diagonal and each diagonal block \(\mathbf{U}^{(j)}\) is an orthogonal matrix (need to be the same for constraint (2)). More explicitly,
+
+\[\begin{pmatrix}
+	\mathbf{C}^{(1)} & \cdots & \mathbf{0} \\
+	\vdots & \ddots & \vdots \\
+	\mathbf{0} & \cdots & \mathbf{C}^{(J)} \\
+\end{pmatrix}\begin{pmatrix}
+	\mathbf{U}^{'(1)} & \cdots & \mathbf{0} \\
+	\vdots & \ddots & \vdots \\
+	\mathbf{0} & \cdots & \mathbf{U}^{'(J)} \\
+\end{pmatrix} = \begin{pmatrix}
+	\mathbf{C}^{(1)}\mathbf{U}^{'(J)} & \cdots & \mathbf{0} \\
+	\vdots & \ddots & \vdots \\
+	\mathbf{0} & \cdots & \mathbf{C}^{(1)}\mathbf{U}^{'(J)} \\
+\end{pmatrix}\]
+
+This transformation will also preserve the Frobenius norm of \(\mathbf{x}_{t}\),
+but the element value is not preserved. To ensure the unique solution,
+we need to add more constraints on linear dynamics, i.e. \(\mathbf{A}\) or/and
+\(\mathbf{Q}\). Simply making \(\mathbf{A}\) or \(\mathbf{Q}\) be block diagonal or even diagonal
+with same absolute value (e.g., \(\mathbf{I}_{Jp}\)) is still not enough. One solution is to force at least one of \(\mathbf{A}\) and \(\mathbf{Q}\) be diagonal.
+
+Take \(\mathbf{A}\) for example. Assume the true \(\mathbf{A}\) is diagonal and true
+\(\mathbf{C}^{\{ j\}}\) has orthogonal column. As shown previous, the linear transformation \(\mathbf{U}\) must be block diagonal and each diagonal block
+\(\mathbf{U}^{(j)}\) is an orthogonal matrix. Further, to preserve the diagonal
+structure of \(\mathbf{A}\) (the elements on the diagonal are generally not the
+same), we need \(\mathbf{U}^{(j)}\) to be diagonal. These two constraints force
+\(\mathbf{U}\) be diagonal with diagonal elements be \(\pm 1\). So the solution
+is unique up to sign. The same rationale for constraint on \(\mathbf{Q}\).
+
+\textbf{The above argument reminds me that I may need one more tiny constraint to solve the sign flipping issue. Do it later.}
+
+In summary, we need add three constraints: (1) translation of \(\mathbf{x}_{t}\),
+(2) orthogonality of loading and (3) diagonality of linear dynamics. In the following implementation, I tried 3 sets of constraints:
+
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
+	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), \(\mathbf{A}\) is full while \(\mathbf{Q}\) is
+	diagonal
+	\item
+	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
+	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), \(\mathbf{Q}\) is full while \(\mathbf{A}\) is
+	diagonal
+	\item
+	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
+	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), both \(\mathbf{A}\) and \(\mathbf{Q}\) are diagonal
+\end{enumerate}
+
+Constraint (1) and (2) allows interaction between clusters.
+
+Although, constraint (1) (diagonal $\mathbf{Q}$) is valid in theory, it doesn't work well in practice. Since elements $\mathbf{Q}$ is usually very small (~$10^{-3} ot 10^{-5}$).
+Therefore, I will discard this constraint later.
+
+\subsection{Constraint Implementation}
+
+\subsubsection{Translation}
+The constraint on translation is easy, just row-center the sample of \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) in each iteration.
+We can get the transformed full conditional distribution after some algebra.
+Let \(\mathbf{X} = \left( \mathbf{x}_{1},\ldots,\mathbf{x}_{T} \right) = \begin{pmatrix}
+	\mathbf{X}_{(1)}^{'} \\
+	\vdots \\
+	\mathbf{X}_{(Jp)}^{'} \\
+\end{pmatrix}\), where \(\mathbf{x}_{(k)} \in \mathbb{R}^{T}\) is the k-th row vector.
+Then the k-th row mean is
+\({\overline{\mathbf{x}}}_{(k)} = \frac{1}{T}\mathbf{x}_{(k)}^{'}\mathbf{1}_{T}\). The unconstraint full conditional distribution of \(vec(\mathbf{X})\) is derived in previous section. After
+normalization in each iteration, we are actually sampling from
+\(P(vec(\mathbf{X}) - \mathbf{1}_{T}\bigotimes\begin{pmatrix}
+	{\overline{\mathbf{x}}}_{(1)} \\
+	\vdots \\
+	{\overline{\mathbf{x}}}_{(Jp)} \\
+\end{pmatrix}|\mathbf{Y},\ldots)\). Notice that,
+\[\mathbf{1}_{T}\bigotimes\begin{pmatrix}
+	{\overline{\mathbf{x}}}_{(1)} \\
+	\vdots \\
+	{\overline{\mathbf{x}}}_{(Jp)} \\
+\end{pmatrix} = \mathbf{1}_{T}\bigotimes\frac{1}{T}\mathbf{X1}_{T} = \frac{1}{T}vec\left( \mathbf{1X}_{T}\mathbf{1}_{T}^{'} \right) = \frac{\mathbf{J}_{T}\bigotimes \mathbf{I}_{Jp}}{T}vec(\mathbf{X})\]
+, then we are just sampling from full conditional distribution of
+\((\mathbf{I}_{JpT} - \frac{\mathbf{J}_{T}\bigotimes \mathbf{I}_{Jp}}{T})vec(\mathbf{X})\)
+
+\subsubsection{Linear Transformation}
+The constraint on orthogonality is kind of tricky. Here, I just transformed \(\mathbf{C}^{(j)}\) be
+\(\mathbf{C}^{(j)} = \mathbf{K}^{(j)}\left( \mathbf{K}^{'(j)}\mathbf{K}^{(j)} \right)^{- 1/2}\). Therefore,
+we can just sample \(\mathbf{K}^{(j)}\) without any constrained. Since
+\(\mathbf{K}^{(j)}\) has the same structure as \(\mathbf{C}^{(j)}\), we can put some
+hierarchical prior to model \(\mathbf{C}^{(j)}\). Specially, when
+\(\mathbf{K}^{(j)} \sim MN_{n,p}(\mathbf{0},\bm{\Sigma}^{(j)},\mathbf{I})\) (\(MN_{n,p}\) means a matrix
+normal distribution), \(\mathbf{C}^{(j)} \sim MACG(\bm{\Sigma}^{(j)})\).
+\(MACG(\bm{\Sigma})\) means the matrix angular central Gaussian distribution,
+with density
+\(f_{\mathbf{X}}(\mathbf{X}) = |\bm{\Sigma}\left. \  \right|^{- p/2}{|\mathbf{X}'\Sigma^{- 1}\mathbf{X}|}^{- n/2}\).
+When \(\bm{\Sigma}^{(j)} = \mathbf{I}_{n}\) or \(n = p\), \(\mathbf{C}^{(j)}\) is just
+uniformly distributed on the Stiefel manifold
+\(S(p,n) = \left\{ \mathbf{C} \in \mathbb{R}^{n \times p} \right|\mathbf{C}'\mathbf{C} = \mathbf{I}_{p}\}\).
+
+\textbf{This reminds me to fix the problem when \(n < p\)}, which may happen when doing clustering. One easy solution is just use the constraint
+\( (\mathbf{C}'^{(1)},\cdots, \mathbf{C}'^{(J)})
+\begin{pmatrix}
+	\mathbf{C}^{(1)} \\
+	\vdots \\
+	\mathbf{C}^{(J)} \\
+\end{pmatrix} = \mathbf{I}_{p}\), or just merge clusters with \(n < p\) to other
+clusters. Fix that.
+
+Since now there's no closed form for full conditional, I just sample it by no U-turn sampler (NUTS).
+
+The diagonality in linear dynamics is trivial, except for the constraint with diagonal \(\mathbf{A}\) and full \(\mathbf{Q}\).
+This needs some algebra.
+Denote the diagonal \(\mathbf{A}\) as \(\mathbf{A} = diag(a_{1},\ldots,a_{Jp})\).
+Then,
+
+\[\mathbf{Y}_{BA} = \begin{pmatrix}
+	\mathbf{x}_{2}^{'} \\
+	\mathbf{x}_{3}^{'} \\
+	\vdots \\
+	\mathbf{x}_{T}^{'} \\
+\end{pmatrix} = \begin{pmatrix}
+	1 & x_{1,1} & \cdots & x_{1,Jp} \\
+	1 & x_{2,1} & \cdots & x_{2,Jp} \\
+	\vdots & \vdots & \vdots & \vdots \\
+	1 & x_{T - 1,1} & \cdots & x_{T - 1,Jp} \\
+\end{pmatrix}\begin{pmatrix}
+	b_{1} & \cdots & b_{Jp} \\
+	a_{1} & \cdots & 0 \\
+	\vdots & \ddots & \vdots \\
+	0 & \cdots & a_{Jp} \\
+\end{pmatrix} + \mathbf{E} = \mathbf{X}_{BA}\mathbf{F}_{BA} + \mathbf{E}\]
+
+, where
+\(\mathbf{E} = \left( \bm{\epsilon}_{1},\ldots,\bm{\epsilon}_{T - 1} \right)^{'}\)
+and \(\bm{\epsilon}_{t} \sim N(\mathbf{0},\mathbf{Q})\). Notice that
+\(\mathbf{x}_{t} = (x_{t, 1}, \cdots x_{t, Jp})'\).
+
+Equivalently, we can rewrite the model as:
+
+\[\mathbf{Y}_{ba} = vec\left(\mathbf{Y}_{BA} \right) = \begin{pmatrix}
+	1 & x_{1,1} & \cdots & 0 & 0 \\
+	\vdots & \vdots & \vdots & \vdots & \vdots \\
+	1 & x_{T - 1,1} & \cdots & 0 & 0 \\
+	\vdots & \vdots & \ddots & \vdots & \vdots \\
+	0 & 0 & \cdots & 1 & x_{1,Jp} \\
+	\vdots & \vdots & \vdots & \vdots & \vdots \\
+	0 & 0 & \cdots & 1 & x_{T - 1,Jp} \\
+\end{pmatrix}\begin{pmatrix}
+	b_{1} \\
+	a_{1} \\
+	\vdots \\
+	b_{Jp} \\
+	a_{Jp} \\
+\end{pmatrix} + vec(\mathbf{E}) = \mathbf{X}_{ba}\mathbf{f}_{ba} + vec(\mathbf{E})\]
+
+, where
+\(Cov\left(vec(\mathbf{E}) \right) = \mathbf{Q}\bigotimes \mathbf{I}_{T - 1}\).
+
+If we use the prior:
+\begin{align*}
+	\mathbf{Q} &\sim W^{- 1}(\Psi_{Q_{0}},\nu_{Q_{0}})\\
+	\mathbf{f}_{ba}|\mathbf{Q} &\sim N_{2Jp}(\mathbf{1}_{Jp}\bigotimes \mathbf{f}_{ba,0},\ \mathbf{Q}\bigotimes\bm{\Lambda}_{ba,0}^{- 1})
+\end{align*}
+
+, where \(\mathbf{f}_{ba,0} = (0, 1)'\) and \(\bm{\Lambda}_{ba,0} = \mathbf{I}_{2}\). Then the
+corresponding prior for \(vec(\mathbf{F}_{BA})\) is:
+
+\[vec\left( \mathbf{F}_{BA} \right)|\mathbf{Q} \sim N(vec\left( \mathbf{F}_{BA,0} \right),\ \mathbf{Q}\bigotimes\bm{\Lambda}_{BA,0}^{- 1})\]
+
+, where \(\mathbf{F}_{BA,0} = \begin{pmatrix}
+	\mathbf{0}_{Jp}^{'} \\
+	\mathbf{I}_{Jp} \\
+\end{pmatrix}\) and \(\bm{\Lambda}_{BA,0} = \mathbf{I}_{Jp + 1}\).
+
+Then the posteriors are:
+\begin{align*}
+	\mathbf{Q}|\mathbf{Y},\ldots &\sim W^{- 1}(\Psi_{Q},\nu_{Q})\\
+	\mathbf{f}_{ba}|\mathbf{Y},\mathbf{Q},\ldots &\sim N({\widehat{\mathbf{f}}}_{ba},\ (\mathbf{Q}\bigotimes \mathbf{I}_{2})\Lambda_{ba}^{- 1})
+\end{align*}
+
+, with parameters be:
+\begin{align*}
+	\bm{\Lambda}_{ba} &= \mathbf{X}_{ba}^{'}\mathbf{X}_{ba} + \mathbf{I}_{Jp}\bigotimes\bm{\Lambda}_{ba,0}\\
+	{\widehat{\mathbf{f}}}_{ba} &= \bm{\Lambda}_{ba}^{- 1}\left( \mathbf{X}_{ba}^{'}\mathbf{Y}_{ba} + \ \mathbf{1}_{Jp}\bigotimes\bm{\Lambda}_{ba,0}\mathbf{f}_{ba,0} \right)
+\end{align*}
+
+The corresponding estimate for \(\mathbf{F}_{BA}\) is denoted as
+\({\widehat{\mathbf{F}}}_{BA}\), which is just a transformation of \({\widehat{\mathbf{f}}}_{ba}\).
+\begin{align*}
+	\Psi_{Q} &= \Psi_{Q0} + \left( \mathbf{Y}_{BA} - \mathbf{X}_{BA}{\widehat{\mathbf{F}}}_{BA} \right)^{'}\left( \mathbf{Y}_{BA} - \mathbf{X}_{BA}{\widehat{\mathbf{F}}}_{BA} \right) + \left( {\widehat{\mathbf{F}}}_{BA} - \mathbf{F}_{BA,0} \right)^{'}\Lambda_{BA,0}^{- 1}\left( {\widehat{\mathbf{F}}}_{BA} - \mathbf{F}_{BA,0} \right)\\
+	\nu_{Q} &= \nu_{Q0} + T - 1
+\end{align*}
+
+\subsection{Simulation}
+The simulation example is as before, and all the code can be found in \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{this folder}.
+
+Since now we constrain the loading has orthogonal column for each cluster, just showing the Frobenius norm of $\mathbf{x}_t$ is not enough to check convergence.
+Here, I further check the value of $\mathbf{x}_{[T/2]}$.
+
+For \(\mathbf{C}^{(j)} = \mathbf{K}^{(j)}\left( \mathbf{K}^{'(j)}\mathbf{K}^{(j)} \right)^{- 1/2}\), I just put a simple hierarchical prior on $\mathbf{K}^{(j)}$ such that each element of $\mathbf{K}^{(j)}$ is i.i.d.
+$N(\mu_{k}^{(j)}, (\sigma^{(j)}_{k})^2)$.
+Well, maybe we can put some more careful priors later to reflect the covariance among neurons.
+
+Basically, I ran both MCMCs for 10,000 iterations. Here, I show 6 sets of plots:
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+	Full trace of Frobenius norm in $\mathbf{X} = (\mathbf{x}_{1},\ldots,\mathbf{x}_{T})$ and the value of $\mathbf{x}_{T/2} = \mathbf{x}_{500}$
+	\item
+	Full trace of 2-norm/ Frobenius norm in $\mathbf{d}$, $\mathbf{b}$, $\mathbf{A}$ and $\mathbf{Q}$ 
+	\item
+	partial trace plot of (1) from iteration 1 to 100
+	\item
+	partial trace plot of (2) from iteration 1 to 100
+	\item
+	average mean firing rate from iteration 5000 to 10,000.
+	\item
+	average $\mathbf{X}$ from iteration 5000 to 10,000.
+\end{enumerate}
+
+\subsubsection{\(\mathbf{Q}\) is full while \(\mathbf{A}\) is diagonal}
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{trace_X_full_diagA.png}}%
+		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{trace_dbAQ_full_diagA.png}}%
+	}
+	\caption{\(\mathbf{Q}\) is full while \(\mathbf{A}\) is diagonal, full trace}
+	\label{full trace, diag A}
+\end{figure}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{trace_X_part_diagA.png}}%
+		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{trace_dbAQ_part_diagA.png}}%
+	}
+	\caption{\(\mathbf{Q}\) is full while \(\mathbf{A}\) is diagonal, partial trace}
+	\label{partial trace, diag A}
+\end{figure}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[mean firing rate]{\label{fig:a}\includegraphics[width=0.75\textwidth]{FR_diagA.png}}%
+		\subfigure[latent vectors]{\label{fig:b}\includegraphics[width=0.75\textwidth]{latent_diagA.png}}%
+	}
+	\caption{\(\mathbf{Q}\) is full while \(\mathbf{A}\) is diagonal, mean FR and $\mathbf{x}_t$}
+	\label{FR and X, diag A}
+\end{figure}
+\clearpage
+
+\subsubsection{both \(\mathbf{A}\) and \(\mathbf{Q}\) are diagonal}
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{trace_X_full_diagAQ.png}}%
+		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{trace_dbAQ_full_diagAQ.png}}%
+	}
+	\caption{both \(\mathbf{A}\) and \(\mathbf{Q}\) are diagonal, full trace}
+	\label{full trace, diag AQ}
+\end{figure}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{trace_X_part_diagAQ.png}}%
+		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{trace_dbAQ_part_diagAQ.png}}%
+	}
+	\caption{both \(\mathbf{A}\) and \(\mathbf{Q}\) are diagonal, partial trace}
+	\label{partial trace, diag AQ}
+\end{figure}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[mean firing rate]{\label{fig:a}\includegraphics[width=0.75\textwidth]{FR_diagAQ.png}}%
+		\subfigure[latent vectors]{\label{fig:b}\includegraphics[width=0.75\textwidth]{latent_diagAQ.png}}%
+	}
+	\caption{both \(\mathbf{A}\) and \(\mathbf{Q}\) are diagonal, mean FR and $\mathbf{x}_t$}
+	\label{FR and X, diag AQ}
+\end{figure}
+
+We can see that either is fine, and the convergence is achieved at nearly 20 to 30 iterations.
+
+\textbf{In summary, to ensure convergence, we can put the following two constraints:}
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
+	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), \(\mathbf{Q}\) is full while \(\mathbf{A}\) is
+	diagonal
+	\item
+	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
+	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), both \(\mathbf{A}\) and \(\mathbf{Q}\) are diagonal
+\end{enumerate}
+
+If we really care about the interactions between clusters, just use constraint (1). Otherwise, either is fine.
+
+\subsection{Comments on Previous Models}
+When fitting the LDS or factor analysis (e.g.
+GPFA) model, I may only believe the temporal pattern or the change of the pattern.
+The pattern itself at a specific time point doesn't make a lot of sense, because of the affine transformation invariance.
+
+In previous work, they fit models by EM, variational Bayes or some other MAP/ MLE method.
+\textbf{This is quite dangerous} if they don't check the uniqueness (Thanks to the gold standard MCMC, otherwise I will never check this issue...).
+
+In the PLDS code that Ian shared me before, the default model has no constraint.
+They fit it by EM, and use ELBO as the convergence criteria.
+Although the parameters are not unique, the likelihood keeps the same, using ELBO is justified.
+However, since there's no randomness in EM, it's impossible to detect the non-uniqueness.
+Actually, I think their results are even not local optima, but just a single point in the optimal surface (I guess if they use something like stochastic EM, they will find the issue).
+
+The PLDS also provides some constraint options, but some of them are still not enough to ensure unique solution.
+BTW, this might be another reason they choose to use EM, since adding constraint in EM is trivial: just do unconstrained EM in each iteration and project the value to the constraint space (i.e.
+add constraint) at the end of each iteration.
+But this doesn't work for MCMC, since we are sampling the distribution but not just the single optimal value.
+
+\textbf{I guess similar problems may happen in other LDS/ GPFA research.}
+
+Even if they make enough constraints to ensure uniqueness, these research should be careful about the pattern they found.
+For example, in \href{https://papers.nips.cc/paper/2020/hash/aa1f5f73327ba40d47ebce155e785aaf-Abstract.html}{Joshua et al., 2020}, the changes/ different states of interaction make sense to me, but I'm quite skeptical about their results in the dynamic matrix $\mathbf{A}$ itself.
+Maybe because of block-diagonal structure in loading $\mathbf{C}$, the interaction between population make sense somewhat.
+But at least the relationship within population doesn't make any sense, because of transformation invariance.
 
@@
 Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
-	Do MH adaptively.
+	Improve the NUTS.
+	\item
+	Debug the clustering.
+	\item
+	Check and clear typo.
 %	\item

### 2021-09-27 LDS/blkDiag/traceCheck.m

@@ -1,0 +1,105 @@
+
+%%
+d_norm_fro = zeros(g,1);
+C_norm_fro = zeros(g, 1);
+b_norm = zeros(g, 1);
+A_norm_fro = zeros(g, 1);
+X_norm_fro = zeros(g, 1);
+K_norm_fro = zeros(g, 1);
+Q_norm_fro = zeros(g, 1);
+
+for k = 1:g
+    d_norm_fro(k) = norm(d_fit(:,:,k), 'fro');
+    C_norm_fro(k) = norm(C_fit(:,:,k), 'fro');
+    b_norm(k) = norm(b_fit(:,k));
+    A_norm_fro(k) = norm(A_fit(:,:,k), 'fro');
+    X_norm_fro(k) = norm(X_fit(:,:,k), 'fro');
+    K_norm_fro(k) = norm(K_fit(:,:,k), 'fro');
+    Q_norm_fro(k) = norm(Q_fit(:,:,k), 'fro');
+end
+
+plot(K_norm_fro)
+plot(muK_fit')
+plot(sig2K_fit')
+
+
+figure(1)
+subplot(2,1,1)
+plot(X_norm_fro)
+title("Frobenius norm of X, dim: " + nClus*p + "\times" + T)
+subplot(2,1,2)
+plot(squeeze(X_fit(:,round(T/2),1:g))')
+title("values of x_{500}, dim: " + size(X_fit, 1))
+
+figure(2)
+subplot(4,1,1)
+plot(d_norm_fro)
+title('norm of d')
+subplot(4,1,2)
+plot(b_norm)
+title('norm of b')
+subplot(4,1,3)
+plot(A_norm_fro)
+title('Frobenius norm of A')
+subplot(4,1,4)
+plot(Q_norm_fro)
+title('Frobenius norm of Q')
+
+
+figure(3)
+subplot(2,1,1)
+plot(X_norm_fro(1:100))
+title("Frobenius norm of X, dim: " + nClus*p + "\times" + T)
+subplot(2,1,2)
+plot(squeeze(X_fit(:,round(T/2),1:100))')
+title("values of x_{500}, dim: " + size(X_fit, 1))
+
+figure(4)
+subplot(4,1,1)
+plot(d_norm_fro(1:100))
+title('norm of d')
+subplot(4,1,2)
+plot(b_norm(1:100))
+title('norm of b')
+subplot(4,1,3)
[…]

### 2021-09-27 PLDS/core/LDS/LDSTransformParams.m

@@ -32,58 +32,62 @@ assignopts(who,varargin);
 xDim = size(params.model.A,1);
 
 switch TransformType
-
- case '0'
-  
-  % do nothing
-  
- case '1'
-  
-  [UC,SC,VC]    = svd(params.model.C,0);
-  [params seq]  = LDSApplyParamsTransformation(SC*VC',params,'seq',seq);
-  
-  params.model.Pi     = dlyap(params.model.A,params.model.Q);
-  if min(eig(params.model.Pi))<0
-    params.model.Pi = params.model.Q;
-  end
-  [UPi SPi VPi] = svd(params.model.Pi);
-  [params seq]  = LDSApplyParamsTransformation(UPi',params,'seq',seq);
-  
- case '2'  
-  
-  params.model.Pi = dlyap(params.model.A,params.model.Q);
-  if min(eig(params.model.Pi))<0
-    params.model.Pi = params.model.Q;
-  end
-  [UPi SPi VPi] = svd(params.model.Pi);
-  M    	      = diag(1./sqrt(diag(SPi)))*UPi';
-  [params seq]  = LDSApplyParamsTransformation(M,params,'seq',seq);    	
-  
-  [UC,SC,VC]    = svd(params.model.C,0);
-  [params seq]  = LDSApplyParamsTransformation(VC',params,'seq',seq);  	
-  
- case '3'
-
-  [params seq] = LDSApplyParamsTransformation(diag(sqrt(sum(params.model.C.^2,1))),params,'seq',seq);
-
- case '4'
-
-  Pi = dlyap(params.model.A,params.model.Q);
-  if min(eig(Pi))<0
-    Pi = params.model.Q;
-  end
-  M = diag(1./sqrt(diag(Pi)));
-  [params seq] = LDSApplyParamsTransformation(M,params,'seq',seq);
-  
- case '5'
-
-  [T B] = bdschur(params.model.A,inf);
-  [params seq] = LDSApplyParamsTransformation(pinv(T),params);
-  [params seq] = LDSApplyParamsTransformation(diag(sqrt(sum(params.model.C.^2,1))),params,'seq',seq);
-
-
- otherwise
-  
-  warning('Unknow paramter transformation type')
-  
+    
+    case '0'
+        
+        % do nothing
+        
+    case '1'
+        
+        [UC,SC,VC]    = svd(params.model.C,0);
+        disp('do SVD')
[…]

### 2021-10-03 LDS/constraint/LDS/AQ_full.m
transfer for documentation

@@ -1,0 +1,349 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(N, p*nClus);
+C = zeros(N, p);
+for k = 1:length(Lab)
+    C(k,:) = 2*repmat(sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)),1,2)+[-2 1];
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C(k,:);
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+% standardize the latent vectors
+
+% gtmp = -mean(X, 2);
+% M = inv(diag(max(abs(X),[], 2)));
+gtmp = -mean(X, 2);
[…]

### 2021-10-03 LDS/constraint/LDS/norm_AQ_full.m

@@ -1,0 +1,353 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(N, p*nClus);
+C = zeros(N, p);
+for k = 1:length(Lab)
+    C(k,:) = 2*repmat(sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)),1,2)+[-2 1];
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C(k,:);
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+% standardize the latent vectors
+
+% gtmp = -mean(X, 2);
+% M = inv(diag(max(abs(X),[], 2)));
+gtmp = -min(X,[], 2);
[…]

### 2021-10-03 LDS/constraint/LDS/norm_diagA.m

@@ -1,0 +1,371 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(N, p*nClus);
+C = zeros(N, p);
+for k = 1:length(Lab)
+    C(k,:) = 2*repmat(sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)),1,2)+[-2 1];
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C(k,:);
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+% standardize the latent vectors
+
+% gtmp = -mean(X, 2);
+% M = inv(diag(max(abs(X),[], 2)));
+gtmp = -min(X,[], 2);
[…]

### 2021-10-03 LDS/constraint/LDS/norm_diagAQ.m

@@ -1,0 +1,358 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(N, p*nClus);
+C = zeros(N, p);
+for k = 1:length(Lab)
+    C(k,:) = 2*repmat(sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)),1,2)+[-2 1];
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C(k,:);
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+% standardize the latent vectors
+
+% gtmp = -mean(X, 2);
+% M = inv(diag(max(abs(X),[], 2)));
+gtmp = -min(X,[], 2);
[…]

### 2021-10-03 LDS/constraint/LDS/norm_diagQ.m

@@ -1,0 +1,356 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(N, p*nClus);
+C = zeros(N, p);
+for k = 1:length(Lab)
+    C(k,:) = 2*repmat(sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)),1,2)+[-2 1];
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C(k,:);
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+% standardize the latent vectors
+
+% gtmp = -mean(X, 2);
+% M = inv(diag(max(abs(X),[], 2)));
+gtmp = -min(X,[], 2);
[…]

### 2021-10-03 LDS/constraint/LDS/traceCheck.m

@@ -24,49 +24,55 @@ end
 
 
 figure(1)
-subplot(2,1,1)
-plot(C_norm_fro)
-title("Frobenius norm of C, dim: " + nClus*p + "\times" + T)
-subplot(2,1,2)
-plot(squeeze(X_fit(:,round(T/2),1:g))')
-title("values of x_{500}, dim: " + size(X_fit, 1))
-
-figure(2)
-subplot(4,1,1)
+subplot(3,1,1)
 plot(d_norm_fro)
 title('norm of d')
-subplot(4,1,2)
+subplot(3,1,2)
+plot(C_norm_fro)
+title("Frobenius norm of C, dim: " + N + "\times" + p)
+subplot(3,1,3)
+plot(X_norm_fro)
+title("Frobenius norm of X, dim: " + nClus*p + "\times" + T)
+
+% subplot(2,1,2)
+% plot(squeeze(X_fit(:,round(T/2),1:g))')
+% title("values of x_{500}, dim: " + size(X_fit, 1))
+
+figure(2)
+subplot(3,1,1)
 plot(b_norm)
 title('norm of b')
-subplot(4,1,3)
+subplot(3,1,2)
 plot(A_norm_fro)
 title('Frobenius norm of A')
-subplot(4,1,4)
+subplot(3,1,3)
 plot(Q_norm_fro)
 title('Frobenius norm of Q')
 
-
+gTrunc = 1000;
 figure(3)
-subplot(2,1,1)
-plot(X_norm_fro(1:100))
+subplot(3,1,1)
+plot(d_norm_fro(1:gTrunc))
+title('norm of d')
+subplot(3,1,2)
+plot(C_norm_fro(1:gTrunc))
+title("Frobenius norm of C, dim: " + N + "\times" + p)
+subplot(3,1,3)
+plot(X_norm_fro(1:gTrunc))
 title("Frobenius norm of X, dim: " + nClus*p + "\times" + T)
-subplot(2,1,2)
-plot(squeeze(X_fit(:,round(T/2),1:100))')
-title("values of x_{500}, dim: " + size(X_fit, 1))
+
 
 figure(4)
-subplot(4,1,1)
-plot(d_norm_fro(1:100))
-title('norm of d')
-subplot(4,1,2)
-plot(b_norm(1:100))
+subplot(3,1,1)
+plot(b_norm(1:gTrunc))
 title('norm of b')
[…]

### 2021-10-03 documents/cluster_neural_population.tex
cluster_initial_done

@@
 
-%\section{Model Check [Problems for \(\mathbf{b}\) and \(\mathbf{A}\) partially solved]} \label{problems}
-%
-%\subsection{Estimation of \(\mathbf{d}\) and \(\mathbf{C}\) alone}
-%First, turn off all others except for loading (related) parameters and set them to be true values.
-There are 2 versions: (1) update priors as in subsection \ref{loading prior} and (2) no update of priors, and just set the prior for each as the standard normal.
-%For the estimation with prior update, the results are mean from iteration 50 to 100 (Figure \ref{loading alone}).
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image016.png}}%
-%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image017.png}}%
-%	}
-%	\caption{estimation of loading}
-%	\label{loading alone}
-%\end{figure}
-%
-%The mixing of convergence for loading priors (in subsection \ref{loading prior}) are fine.
-The update of priors don't influence results a lot, but this will help clustering a lot.
-It seems estimation of loading is fine.
-Let's see what happens when estimation of latents \(\mathbf{x}_t\) is on at the same time.
-%
-%\subsection{Estimation of \(\mathbf{d}\), \(\mathbf{C}\) and \(\mathbf{x}_{t}\)}
-%When the estimation of \(\mathbf{x}_t\) is on, we need more iterations. The results are mean from iteration 500 to 1000 (Figure \ref{loading and latent}).
-%
-%The estimation of loading:
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image018.png}}%
-%		\subfigure[]{\includegraphics[width=0.6\textwidth]{image020.png}}%
-%	}
-%	\caption{estimation of loading, with latents on}
-%	\label{loading and latent}
-%\end{figure}
-%
-%And the estimation of corresponding latents (Figure \ref{loading and latents}):
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[no update of prior]{\includegraphics[width=0.6\textwidth]{image019.png}}%
-%		\subfigure[with update of prior]{\includegraphics[width=0.6\textwidth]{image021.png}}%
-%	}
-%	\caption{estimation of loading}
-%	\label{loading and latents}
-%\end{figure}
-%
-%Things are still fine. The remaining parts may influence latent estimations are dynamics (\(\mathbf{b}\), \(\mathbf{A}\)) and prior process noise \(\mathbf{Q}\).
-%
-%\subsection{Estimation of \(\mathbf{b}\), \(\mathbf{A}\), \(\mathbf{Q}\) and \(\mathbf{x}_t\)}
-%I show the results of block-diagonal (Figure \ref{bAQX_blkDiag}) and full (Figure \ref{bAQX_full}) \(\mathbf{Q}\).
-These results are from 2000 MCMC samples.
-The latent and dynamics are from average from 500 to 2000 iterations.
-%
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_trace.png}}%
-%		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_latent.png}}%
-%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_blkDiag_A.png}}%
-%	}
-%	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, block-diagonal process noise}
-%	\label{bAQX_blkDiag}
-%\end{figure}
-%
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[trace]{\label{fig:a}\includegraphics[width=0.55\textwidth]{bAX_full_trace.png}}%
-%		\subfigure[latent]{\label{fig:b}\includegraphics[width=0.55\textwidth]{bAX_full_latent.png}}%
-%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{bAX_full_A.png}}%
-%	}
-%	\caption{sampling of $\mathbf{b}$, $\mathbf{A}$, $\mathbf{Q}$ and $\mathbf{x}_t$, full process noise}
-%	\label{bAQX_full}
-%\end{figure}
-%
-%Both block-diagonal and full looks fine, but the mixing of  full process noise version looks better.
-
 \section{Simulations}\label{sim}
@@
 \textbf{\textcolor{red}{All the simulation results are old and problematic. Update them later.}}
-
-%There are two set of simulation examples.
-The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics.
-In all following results, I fit things with both \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{block-diagonal} and \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/full}{full} \(\mathbf{Q}\).
-The fitting results for block-diagonal version is a bit better, because the underlying true \(\mathbf{Q}\) is diagonal.
-In the following part, I only show results from block-diagonal fitting for labeled data.
-For unlabeled data, i.e.
-clustering, all the results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
-%
-%\subsection{Simulation 1: Generate from LDS Directly}
-%In this simulation, there are 3 clusters with 10 neurons in each cluster.
-The dimension of latents in each cluster is 2.
-In the linear dynamics, the bias term \(\mathbf{b}\) is zero.
-There's no within-population interaction but has some weak between-population interactions.
-In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal.
-The details of simulation can be found in the first section of the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{simulation 1}.
-%
-%\subsubsection{Labeled Data: No Clustering}
-%The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{here}. 
-%
-%Here, I first compare 10k iterations for three methods (consider high dimension of \(\mathbf{x}_t\)): (1) both \(\mathbf{x}_t\) and loading (\(\mathbf{d}\) and \(\mathbf{C}\)) are updated by normal approximation.
-(2) update \(\mathbf{x}_t\) by normal approximation, but update loading by MH.
-(3) update both \(\mathbf{x}_t\) and loading by MH.
-%
-%The fitting results are average of iteration 5k to 10k.
-%
-%\textbf{Update \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by normal approximation}
-%
-%The trace plots are in Figure \ref{double N: trace}
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleN.png}}%
-%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleN.png}}%
-%	}
-%	\caption{both Normal Approx., trace plots}
-%	\label{double N: trace}
-%\end{figure}
-%
-%Well, the chain hasn't achieve the stationary distribution.
-So there's no need to see further results.
-But to show the improvement and for comparison, I still show the fitting of overall firing rate, latents and dynamics in Figure \ref{double N: fit}.
-%
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleN.png}}%
-%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleN.png}}%
-%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleN.png}}%
-%	}
-%	\caption{both Normal Approx., fit}
-%	\label{double N: fit}
-%\end{figure}
-%
-%\textbf{Update \(\mathbf{x}_t\) by normal approximation and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
-%
-%By using the normal approximated posterior variance as the proposal variance, the acceptance rates are around 0.45 (Since there are 30 neurons, there are 30 acceptance rates).
-%The trace plots are in Figure \ref{dc MH: trace}
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_dcMH.png}}%
-%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_dcMH.png}}%
-%	}
-%	\caption{latent Normal Approx. + loading MH, trace plots}
-%	\label{dc MH: trace}
-%\end{figure}
-%
-%Looks much better, although still not perfect yet. The fitting of overall firing rate, latents and dynamics are in Figure \ref{dc MH: fit}.
-%
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_dcMH.png}}%
-%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_dcMH.png}}%
-%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_dcMH.png}}%
-%	}
-%	\caption{latent Normal Approx. + loading MH, fit}
-%	\label{dc MH: fit}
-%\end{figure}
-%
-%See, nearly perfect (especially) for latent.
-%
-%\textbf{Update both \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}
-%
-%The acceptance rate (by using the optimal scalar) for latent is 0.6227, which is a bit high (but acceptable). Again, the acceptance rates for loading are around 0.45.
-%
-%The trace plots are in Figure \ref{double MH: trace}
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleMH.png}}%
-%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleMH.png}}%
-%	}
-%	\caption{both MH, trace plots}
-%	\label{double MH: trace}
-%\end{figure}
-%
-%The trace for dynamics is terrible... Maybe this is caused by high dimension of latent (\(d=6000\) in this case).
-%
-%The fitting of overall firing rate, latents and dynamics are in Figure \ref{double MH: fit}.
-%
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleMH.png}}%
-%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleMH.png}}%
-%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleMH.png}}%
-%	}
-%	\caption{both MH, fit}
-%	\label{double MH: fit}
-%\end{figure}
-%
-%OK, the fitting sucks...
-%
-%One natural idea is to combine MH and normal approximation in latent: at first few steps, use normal approximation at conditional posterior mode to bring things into somewhere close to mode quickly.
-Then use MH to stabilize everything.
-%
-%In the following, I still sample loading by MH. But for latent, the first half iteration is sampled by normal approximation, while the remainder is sampled from MH.
-%
-%The red line marks when I switch the sampling method for \(\mathbf{x}_t\).
-% 
-%The trace plots are in Figure \ref{half MH: trace}
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_halfMH.png}}%
-%		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_halfMH.png}}%
-%	}
-%	\caption{half MH for latent, trace plots}
-%	\label{half MH: trace}
-%\end{figure}
-%
-%The fitting of overall firing rate, latents and dynamics are in Figure \ref{half MH: fit}.
-%
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_halfMH.png}}%
-%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_halfMH.png}}%
-%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_halfMH.png}}%
-%	}
-%	\caption{half MH for latent, fit}
-%	\label{half MH: fit}
-%\end{figure}
-%
-%Well, everything is stable now. But this implementation is too naive. Maybe use some more advanced adaptive sampling method?
-
-%\begin{figure}[h!]
-%	\centering
-%	\includegraphics[width = .8\textwidth]{lds_blk_trace.png}
-%	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
-%	\label{lds norms}
-%\end{figure}\\
-%Some results are in Figure \ref{fig:LDS labeled}. These are means from iteration 1000 to 2000.
-%
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{lds_blk_FR.png}}%
-%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{lds_blk_latent.png}}%
-%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{lds_blk_A.png}}%
-%	}
-%	\caption{LDS sample with labels}
-%	\label{fig:LDS labeled}
-%\end{figure}
-%
-%Personally, I think the fitting is not bad and the mixing is fine (in previous version, we may need 10000 iterations to get a not bad result. But here 2000 iterations did even better than before).
-
-%I also fit the model by assuming one cluster.
-Again, the fitting for overall mean firing rate is perfect.
-This is why it's important to make the loading, \(\mathbf{d}\) and \(\mathbf{C}\), also be cluster-dependent (loading within clusters are correlated).
-%
-%If the loading only depends on neuron index and will not change for different clustering assignments, it's impossible to do clustering (at least in this case), since the loading is enough to capture all the patterns.
-%
-%
-%\subsubsection{Unlabeled Data: Clustering}
-%To give the full path of clustering, I show results in GIFs.
-All results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
-The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.
-%
-%Basically, in my current implementation, algorithms are good to merge clusters.
-However, generating new clusters is very hard...
-In other words, the newly generated cluster will usually not be sampled.
-%
-%\textbf{This is a big problem}. This makes DPMM loses its power. \textbf{Fix that later}.
-%
-%\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
-%In this simulation, there are again 3 clusters with 2 latents and 10 neurons in each.
-Now, the latents are generated directly without specifying the underlying linear dynamics of latents.
-However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal.
-The details of simulation can be found in the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/unspecifiedA_sample_blkDiag.m}{code}.
-%
-%\subsubsection{Labeled Data: No Clustering}
-%As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster.
-The code can be found \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/unspecifiedA_sample_blkDiag.m}{here}.
-%
-%
-%\textbf{I have deleted the old results to avoid confusion. I will post the new results here later.}
-
-%This time I ran 1000 iterations. The trace of the samples is shown in Figure \ref{noA norms}.
-%\begin{figure}[h!]
-%	\centering
-%	\includegraphics[width = .8\textwidth]{noA_blk_trace.png}
-%	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
-%	\label{noA norms}
-%\end{figure}\\
-%
-%Some results are in Figure \ref{fig:noA labeled}. These are means from iteration 500 to 1000.
-%
-%\begin{figure}[h!]
-%	\makebox[\linewidth][c]{%
-%		\centering
-%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{noA_blk_FR.png}}%
-%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{noA_blk_latent.png}}%
-%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{noA_blk_A.png}}%
-%	}
-%	\caption{LDS sample with labels}
-%	\label{fig:noA labeled}
-%\end{figure}
-%
-%This time the mixing of loading (\(\mathbf{d}\) and \(\mathbf{C}\)) is not fine. That reminds me that \textbf{I may need to make loading within cluster related by MV-(Poisson)-GLM directly.}
-
-%\subsubsection{Unlabeled Data: Clustering}
-%\textbf{The clustering results are old. Update when the MFM is implemented}
-%
-%Again, all results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}.
-The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.
-%
-%Besides fitting 3 clusters with 10 neurons each, I further fit a larger scale example (50 neurons for each cluster).
 
@@
 \subsection{Constraints}
-An affine transformation contains translation and linear transformation.
-
-\subsubsection{Translation}
-There are two strategies: (1) just ignore \(\mathbf{d}\) and let \(\mathbf{Cx}_{t}\)
-absorb all the intercepts; (2) center \(\mathbf{x}_t\) such that mean for each
-row of \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0.
-However, because of the block-diagonal constraints on $\mathbf{M}$, the strategy (1) may be problematic.
-Moreover, when $\mathbf{d}$ is non-zero, the \(\mathbf{x}_t\) will looks ugly.
-Another benefit for (2) is that we can use $\mathbf{d}$ to help clustering.
-However, the constraint on translation cannot resolve problems from linear transformation, since any linear transformation will preserve the zero row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\).
-
-\subsubsection{Linear Transformation}
-To mimic PCA, one natural way is to make the loading orthogonal. Again, I'm considering two strategies: (1) make
-\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\); (2) make \( (\mathbf{C}'^{(1)},\cdots, \mathbf{C}'^{(J)})
-\begin{pmatrix}
-	\mathbf{C}^{(1)} \\
-	\vdots \\
-	\mathbf{C}^{(J)} \\
-\end{pmatrix} = \mathbf{I}_{p}\).
-Personally, I prefer the first one, for both computational and clustering reasons.
-Computationally, if I do things cluster-wise, I can easily use parallel when dimension becomes large.
-
-\textbf{However, this is not enough!} Consider \(\mathbf{x}_{t}^{*} = \mathbf{Ux}_{t}\), to
-preserve the orthogonality and block diagonal on \(\mathbf{C}\), \(\mathbf{U}\) must be
-block diagonal and each diagonal block \(\mathbf{U}^{(j)}\) is an orthogonal matrix (need to be the same for constraint (2)). More explicitly,
-
-\[\begin{pmatrix}
-	\mathbf{C}^{(1)} & \cdots & \mathbf{0} \\
-	\vdots & \ddots & \vdots \\
-	\mathbf{0} & \cdots & \mathbf{C}^{(J)} \\
-\end{pmatrix}\begin{pmatrix}
-	\mathbf{U}^{'(1)} & \cdots & \mathbf{0} \\
-	\vdots & \ddots & \vdots \\
-	\mathbf{0} & \cdots & \mathbf{U}^{'(J)} \\
-\end{pmatrix} = \begin{pmatrix}
-	\mathbf{C}^{(1)}\mathbf{U}^{'(J)} & \cdots & \mathbf{0} \\
-	\vdots & \ddots & \vdots \\
-	\mathbf{0} & \cdots & \mathbf{C}^{(1)}\mathbf{U}^{'(J)} \\
-\end{pmatrix}\]
-
-This transformation will also preserve the Frobenius norm of \(\mathbf{x}_{t}\),
-but the element value is not preserved. To ensure the unique solution,
-we need to add more constraints on linear dynamics, i.e. \(\mathbf{A}\) or/and
-\(\mathbf{Q}\). Simply making \(\mathbf{A}\) or \(\mathbf{Q}\) be block diagonal or even diagonal
-with same absolute value (e.g., \(\mathbf{I}_{Jp}\)) is still not enough. One solution is to force at least one of \(\mathbf{A}\) and \(\mathbf{Q}\) be diagonal.
-
-Take \(\mathbf{A}\) for example. Assume the true \(\mathbf{A}\) is diagonal and true
-\(\mathbf{C}^{\{ j\}}\) has orthogonal column. As shown previous, the linear transformation \(\mathbf{U}\) must be block diagonal and each diagonal block
-\(\mathbf{U}^{(j)}\) is an orthogonal matrix. Further, to preserve the diagonal
-structure of \(\mathbf{A}\) (the elements on the diagonal are generally not the
-same), we need \(\mathbf{U}^{(j)}\) to be diagonal. These two constraints force
-\(\mathbf{U}\) be diagonal with diagonal elements be \(\pm 1\). So the solution
-is unique up to sign. The same rationale for constraint on \(\mathbf{Q}\).
-
-\textbf{The above argument reminds me that I may need one more tiny constraint to solve the sign flipping issue. Do it later.}
-
-In summary, we need add three constraints: (1) translation of \(\mathbf{x}_{t}\),
-(2) orthogonality of loading and (3) diagonality of linear dynamics. In the following implementation, I tried 3 sets of constraints:
-
-\begin{enumerate}
-	\def\labelenumi{(\arabic{enumi})}
-	\item
-	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
-	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), \(\mathbf{A}\) is full while \(\mathbf{Q}\) is
-	diagonal
-	\item
-	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
-	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), \(\mathbf{Q}\) is full while \(\mathbf{A}\) is
-	diagonal
-	\item
-	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
-	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), both \(\mathbf{A}\) and \(\mathbf{Q}\) are diagonal
-\end{enumerate}
-
-Constraint (1) and (2) allows interaction between clusters.
-
-Although, constraint (1) (diagonal $\mathbf{Q}$) is valid in theory, it doesn't work well in practice. Since elements $\mathbf{Q}$ is usually very small (~$10^{-3} ot 10^{-5}$).
-Therefore, I will discard this constraint later.
-
-\subsection{Constraint Implementation}
-
-\subsubsection{Translation}
-The constraint on translation is easy, just row-center the sample of \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) in each iteration.
-We can get the transformed full conditional distribution after some algebra.
+
+%\subsubsection{constraint on $\bm{x}_t$ only}
+\subsubsection{Constraint on latent vectors $\mathbf{x}_t$ only}
+The easiest way to constraint is to normalize or standardize each row of \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\).
 Let \(\mathbf{X} = \left( \mathbf{x}_{1},\ldots,\mathbf{x}_{T} \right) = \begin{pmatrix}
@@
 	\mathbf{X}_{(Jp)}^{'} \\
-\end{pmatrix}\), where \(\mathbf{x}_{(k)} \in \mathbb{R}^{T}\) is the k-th row vector.
-Then the k-th row mean is
+\end{pmatrix}\), where \(\mathbf{x}_{(k)} \in \mathbb{R}^{T}\) is the k-th row vector. Then,
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+	Normalization: \(\mathbf{x}_{(k)}^* = \frac{\mathbf{x}_{(k)}-\min \mathbf{x}_{(k)}}{\max \mathbf{x}_{(k)} - \min \mathbf{x}_{(k)}}\)
+	\item
+	Standardization: \(\mathbf{x}_{(k)}^* = \frac{\mathbf{x}_{(k)} - \bar{\mathbf{x}}_{(k)}}{S(\mathbf{x}_{(k)})}\)
+\end{enumerate}
+The purpose of this constraint is to let $\bm{d}$ capture the translation and $\bm{C}$ capture the scaling.
+
+After normalization/ standardization, the solution should be unique.
+
+\subsubsection{Joint constraint on $\mathbf{x}_t$, $\mathbf{C}$ and linear dynamics}
+
+\textbf{This is perfect when cluster is known} (more robust than constraints on $\bm{x}_t$ only), but when doing clusters, this is not very useful...
+Since the constraint depends on the number of elements in the cluster, but the number of elements will change when doing clustering.
+
+See details in the appendix.
+
+\subsection{Constraint Implementation}
+The constraint for normalization and standardization are fairly easy: just do transformation of the MCMC sample in each iteration.
+
+Normalization and standardization are both linear transformation, so after some algebra, we can even write the full conditional distribution explicitly.
+
+For example, let's write out the linear transformation for centering. Denote the k-th row mean as
 \({\overline{\mathbf{x}}}_{(k)} = \frac{1}{T}\mathbf{x}_{(k)}^{'}\mathbf{1}_{T}\). The unconstraint full conditional distribution of \(vec(\mathbf{X})\) is derived in previous section. After
@@
 
-\subsubsection{Linear Transformation}
+To help the convergence, we may further add constraint on linear dynamics.
+Most of them can be fairly implemented, by diagonal $\mathbf{A}$ with full $\mathbf{Q}$ need some algebra.
+See details in the appendix.
+
+\subsection{Simulation}
+Here, I ran both normalization and standardization for 10,000 iterations.
+Although pure normalization or standardization should be enough to ensure convergence, further constraints in linear dynamics may help convergence.
+Several constraints in linear dynamics (i.e. $\bm{A}$ and $\bm{Q}$) are also implemented:
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+	No constraints in $\bm{A}$ or $\bm{Q}$
+	\item
+	No constraints in $\bm{A}$, but $\bm{Q}$ is block-diagonal.
+	\item
+	No constraints in $\bm{A}$, but $\bm{Q}$ is diagonal.
+	\item
+	$\bm{A}$ is diagonal, but no constraints in $\bm{Q}$
+	(See details in appendix).
+	\item
+	Both $\bm{A}$ and $\bm{Q}$ are diagonal.
+\end{enumerate}
+
+By limited experiments, it seems when $\bm{A}$ is diagonal, the convergence \& mixing is best.
+If we put no constraint on $\bm{A}$, then putting no constraint on $\bm{Q}$ is the best.
+Here I show the 2-norm/ Frobenius norm of \(\left\{ \mathbf{d},\mathbf{C},\mathbf{X},\mathbf{b},\mathbf{A},\mathbf{Q} \right\}\) for 4 constraints combos:
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+	Normalization + both $\bm{A}$ and $\bm{Q}$ are diagonal.
+	\item
+	Normalization + $\bm{A}$ diagonal but $\bm{Q}$ full.
+	\item
+	Standardization + both $\bm{A}$ and $\bm{Q}$ are diagonal.
+	\item
+	Standardization + $\bm{A}$ diagonal but $\bm{Q}$ full.
+\end{enumerate}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[normalization + $\bm{A}$ and $\bm{Q}$ diagonal]{\label{fig:a}\includegraphics[width=0.75\textwidth]{norm_AQ_diag_dCX_trace_full.png}}%
+		\subfigure[normalization + $\bm{A}$ diagonal and $\bm{Q}$ full]{\label{fig:b}\includegraphics[width=0.75\textwidth]{norm_A_diag_dCX_trace_full.png}}%
+	}
+	\caption{normalization, full trace of $\bm{d}$, $\bm{C}$ and $\bm{X}$}
+	\label{full trace, normal, obs}
+\end{figure}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[standardization + $\bm{A}$ and $\bm{Q}$ diagonal]{\label{fig:a}\includegraphics[width=0.75\textwidth]{std_AQ_diag_dCX_trace_full.png}}%
+		\subfigure[standardization + $\bm{A}$ diagonal and $\bm{Q}$ full]{\label{fig:b}\includegraphics[width=0.75\textwidth]{std_A_diag_dCX_trace_full.png}}%
+	}
+	\caption{standardization, full trace of $\bm{d}$, $\bm{C}$ and $\bm{X}$}
+	\label{full trace, std, obs}
+\end{figure}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[normalization + $\bm{A}$ and $\bm{Q}$ diagonal]{\label{fig:a}\includegraphics[width=0.75\textwidth]{norm_AQ_diag_bAQ_trace_full.png}}%
+		\subfigure[normalization + $\bm{A}$ and $\bm{Q}$ full]{\label{fig:b}\includegraphics[width=0.75\textwidth]{norm_A_diag_bAQ_trace_full.png}}%
+	}
+	\caption{normalization, full trace of $\bm{b}$, $\bm{A}$ diagonal and $\bm{Q}$}
+	\label{full trace, normal, dynamics}
+\end{figure}
+
+\begin{figure}[h!]
+	\makebox[\linewidth][c]{%
+		\centering
+		\subfigure[standardization + $\bm{A}$ and $\bm{Q}$ diagonal]{\label{fig:a}\includegraphics[width=0.75\textwidth]{std_AQ_diag_bAQ_trace_full.png}}%
+		\subfigure[standardization + $\bm{A}$ and $\bm{Q}$ full]{\label{fig:b}\includegraphics[width=0.75\textwidth]{std_A_diag_bAQ_trace_full.png}}%
+	}
+	\caption{standardization, full trace of $\bm{b}$, $\bm{A}$ diagonal and $\bm{Q}$}
+	\label{full trace, std, dynamics}
+\end{figure}
+
+\subsection{Comments on Previous Models}
+When fitting the LDS or factor analysis (e.g.
+GPFA) model, I may only believe the temporal pattern or the change of the pattern.
+The pattern itself at a specific time point doesn't make a lot of sense, because of the affine transformation invariance.
+
+In previous work, they fit models by EM, variational Bayes or some other MAP/ MLE method.
+\textbf{This is quite dangerous} if they don't check the uniqueness (Thanks to the gold standard MCMC, otherwise I will never check this issue...).
+
+In the PLDS code that Ian shared me before, the default model has no constraint.
+They fit it by EM, and use ELBO as the convergence criteria.
+Although the parameters are not unique, the likelihood keeps the same, using ELBO is justified.
+However, since there's no randomness in EM, it's impossible to detect the non-uniqueness.
+Actually, I think their results are even not local optima, but just a single point in the optimal surface (I guess if they use something like stochastic EM, they will find the issue).
+
+The PLDS also provides some constraint options, but some of them are still not enough to ensure unique solution.
+BTW, this might be another reason they choose to use EM, since adding constraint in EM is trivial: just do unconstrained EM in each iteration and project the value to the constraint space (i.e.
+add constraint) at the end of each iteration.
+But this doesn't work for MCMC, since we are sampling the distribution but not just the single optimal value.
+
+\textbf{I guess similar problems may happen in other LDS/ GPFA research.}
+
+Even if they make enough constraints to ensure uniqueness, these research should be careful about the pattern they found.
+For example, in \href{https://papers.nips.cc/paper/2020/hash/aa1f5f73327ba40d47ebce155e785aaf-Abstract.html}{Joshua et al., 2020}, the changes/ different states of interaction make sense to me, but I'm quite skeptical about their results in the dynamic matrix $\mathbf{A}$ itself.
+Maybe because of block-diagonal structure in loading $\mathbf{C}$, the interaction between population make sense somewhat.
+But at least the relationship within population doesn't make any sense, because of transformation invariance.
+
+\clearpage
+
+\section{TODO}
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+%	Think about MV-(Poisson)-GLM.
+Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
+	Improve the NUTS.
+	\item
+	Debug the clustering.
+	\item
+	Check and clear typo.
+%	\item
+%	After resolving (1), see if the MCMC is still trapped in local optima sometimes.
+Now, the chain seems will get stuck in local mode.
+Maybe updating the loading as a whole will remedy the problem a bit.
+	\item
+	Implement the mixture of finite mixture (MFM) by Jeff Miller.
+	\item
+	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
+%	\item
+%	improve DPMM.
+In current brute force implementation, the number of potential clusters can even go beyond number of neurons (\(N\)).
+There are several improvements, e.g.
+\href{https://link.springer.com/article/10.1007/s11222-009-9150-y}{Kalli et al., 2011}, \href{http://proceedings.mlr.press/v37/gea15.html}{Ge et al., 2015} and \href{https://link.springer.com/article/10.1007/s11222-014-9471-3}{Hastie et al, 2015}.
+Check them later.
+	\item
+	After all of them are resolved, switch to GP version
+
+\end{enumerate}
+
+\clearpage
+\section{Appendix}
+
+\subsection{Joint constraint on $\mathbf{x}_t$, $\mathbf{C}$ and linear dynamics}
+For $\mathbf{x}_t$, we can row center it as previous. But instead of doing scaling on $\bm{x}_t$, we can put the scaling to the loading $\mathbf{C}$.
+
+To mimic PCA, one natural way is to make the loading orthogonal. Here, I simply make \(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\).
+
+\textbf{However, this is not enough!} Consider \(\mathbf{x}_{t}^{*} = \mathbf{Ux}_{t}\), to
+preserve the orthogonality and block diagonal on \(\mathbf{C}\), \(\mathbf{U}\) must be
+block diagonal and each diagonal block \(\mathbf{U}^{(j)}\) is an orthogonal matrix. More explicitly,
+
+\[\begin{pmatrix}
+	\mathbf{C}^{(1)} & \cdots & \mathbf{0} \\
+	\vdots & \ddots & \vdots \\
+	\mathbf{0} & \cdots & \mathbf{C}^{(J)} \\
+\end{pmatrix}\begin{pmatrix}
+	\mathbf{U}^{'(1)} & \cdots & \mathbf{0} \\
+	\vdots & \ddots & \vdots \\
+	\mathbf{0} & \cdots & \mathbf{U}^{'(J)} \\
+\end{pmatrix} = \begin{pmatrix}
+	\mathbf{C}^{(1)}\mathbf{U}^{'(J)} & \cdots & \mathbf{0} \\
+	\vdots & \ddots & \vdots \\
+	\mathbf{0} & \cdots & \mathbf{C}^{(1)}\mathbf{U}^{'(J)} \\
+\end{pmatrix}\]
+
+This transformation will also preserve the Frobenius norm of \(\mathbf{x}_{t}\),
+but the element value is not preserved. To ensure the unique solution,
+we need to add more constraints on linear dynamics, i.e. \(\mathbf{A}\) or/and
+\(\mathbf{Q}\). Simply making \(\mathbf{A}\) or \(\mathbf{Q}\) be block diagonal or even diagonal
+with same absolute value (e.g., \(\mathbf{I}_{Jp}\)) is still not enough. One solution is to force at least one of \(\mathbf{A}\) and \(\mathbf{Q}\) be diagonal.
+
+Take \(\mathbf{A}\) for example. Assume the true \(\mathbf{A}\) is diagonal and true
+\(\mathbf{C}^{\{ j\}}\) has orthogonal column. As shown previous, the linear transformation \(\mathbf{U}\) must be block diagonal and each diagonal block
+\(\mathbf{U}^{(j)}\) is an orthogonal matrix. Further, to preserve the diagonal
+structure of \(\mathbf{A}\) (the elements on the diagonal are generally not the
+same), we need \(\mathbf{U}^{(j)}\) to be diagonal. These two constraints force
+\(\mathbf{U}\) be diagonal with diagonal elements be \(\pm 1\). So the solution
+is unique up to sign. The same rationale for constraint on \(\mathbf{Q}\).
+
+In summary, we need add three constraints if do things jointly: (1) translation of \(\mathbf{x}_{t}\),
+(2) orthogonality of loading and (3) diagonality of linear dynamics. In the following implementation, I tried 3 sets of constraints:
+
+\begin{enumerate}
+	\def\labelenumi{(\arabic{enumi})}
+	\item
+	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
+	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), \(\mathbf{A}\) is full while \(\mathbf{Q}\) is
+	diagonal
+	\item
+	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
+	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), \(\mathbf{Q}\) is full while \(\mathbf{A}\) is
+	diagonal
+	\item
+	row mean in \((\mathbf{x}_{1},\ldots,\mathbf{x}_{T})\) is 0,
+	\(\mathbf{C}'^{(j)}\mathbf{C}^{(j)} = \mathbf{I}_{p}\), both \(\mathbf{A}\) and \(\mathbf{Q}\) are diagonal
+\end{enumerate}
+
+Constraint (1) and (2) allows interaction between clusters.
+
+Although, constraint (1) (diagonal $\mathbf{Q}$) is valid in theory, it doesn't work well in practice. Since elements $\mathbf{Q}$ is usually very small (~$10^{-3}$ to $10^{-5}$).
+Therefore, I will discard this constraint later.
+
+\subsection{sample the orthogonal loading}
 The constraint on orthogonality is kind of tricky. Here, I just transformed \(\mathbf{C}^{(j)}\) be
@@
 
-\textbf{This reminds me to fix the problem when \(n < p\)}, which may happen when doing clustering. One easy solution is just use the constraint
-\( (\mathbf{C}'^{(1)},\cdots, \mathbf{C}'^{(J)})
-\begin{pmatrix}
-	\mathbf{C}^{(1)} \\
-	\vdots \\
-	\mathbf{C}^{(J)} \\
-\end{pmatrix} = \mathbf{I}_{p}\), or just merge clusters with \(n < p\) to other
-clusters. Fix that.
-
 Since now there's no closed form for full conditional, I just sample it by no U-turn sampler (NUTS).
 
-The diagonality in linear dynamics is trivial, except for the constraint with diagonal \(\mathbf{A}\) and full \(\mathbf{Q}\).
-This needs some algebra.
-Denote the diagonal \(\mathbf{A}\) as \(\mathbf{A} = diag(a_{1},\ldots,a_{Jp})\).
-Then,
+\subsection{diagonal $\mathbf{A}$ and full $\mathbf{Q}$}
+Denote the diagonal \(\mathbf{A}\) as \(\mathbf{A} = diag(a_{1},\ldots,a_{Jp})\). Then,
 
@@
 
-\subsection{Simulation}
+\section{Simulation on joint constraint}
 The simulation example is as before, and all the code can be found in \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{this folder}.
@@
 	\item
-	Full trace of Frobenius norm in $\mathbf{X} = (\mathbf{x}_{1},\ldots,\mathbf{x}_{T})$ and the value of $\mathbf{x}_{T/2} = \mathbf{x}_{500}$
-	\item
-	Full trace of 2-norm/ Frobenius norm in $\mathbf{d}$, $\mathbf{b}$, $\mathbf{A}$ and $\mathbf{Q}$ 
-	\item
-	partial trace plot of (1) from iteration 1 to 100
-	\item
-	partial trace plot of (2) from iteration 1 to 100
+	Full trace of  2-norm/ Frobenius norm in $\bm{d}$, $\bm{C}$ and $\mathbf{X} = (\mathbf{x}_{1},\ldots,\mathbf{x}_{T})$
+	\item
+	Full trace of 2-norm/ Frobenius norm in $\mathbf{b}$, $\mathbf{A}$ and $\mathbf{Q}$ 
+	\item
+	partial trace plot of (1) from iteration 1 to 1000
+	\item
+	partial trace plot of (2) from iteration 1 to 1000
 	\item
@@
 		\centering
-		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{trace_X_full_diagA.png}}%
-		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{trace_dbAQ_full_diagA.png}}%
+		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{CCI_AQ_diag_dCX_trace_full.png}}%
+		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{CCI_AQ_diag_bAQ_trace_full.png}}%
 	}
@@
 		\centering
-		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{trace_X_part_diagA.png}}%
-		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{trace_dbAQ_part_diagA.png}}%
+		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{CCI_AQ_diag_dCX_trace_part.png}}%
+		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{CCI_AQ_diag_bAQ_trace_part.png}}%
 	}
@@
 		\centering
-		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{trace_X_full_diagAQ.png}}%
-		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{trace_dbAQ_full_diagAQ.png}}%
+		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{CCI_A_diag_dCX_trace_full.png}}%
+		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{CCI_A_diag_bAQ_trace_full.png}}%
 	}
@@
 		\centering
-		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{trace_X_part_diagAQ.png}}%
-		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{trace_dbAQ_part_diagAQ.png}}%
+		\subfigure[tracd of $\mathbf{X}$]{\label{fig:a}\includegraphics[width=0.75\textwidth]{CCI_A_diag_dCX_trace_part.png}}%
+		\subfigure[trace of loading and dynamics]{\label{fig:b}\includegraphics[width=0.75\textwidth]{CCI_A_diag_bAQ_trace_part.png}}%
 	}
@@
 
-\subsection{Comments on Previous Models}
-When fitting the LDS or factor analysis (e.g.
-GPFA) model, I may only believe the temporal pattern or the change of the pattern.
-The pattern itself at a specific time point doesn't make a lot of sense, because of the affine transformation invariance.
-
-In previous work, they fit models by EM, variational Bayes or some other MAP/ MLE method.
-\textbf{This is quite dangerous} if they don't check the uniqueness (Thanks to the gold standard MCMC, otherwise I will never check this issue...).
-
-In the PLDS code that Ian shared me before, the default model has no constraint.
-They fit it by EM, and use ELBO as the convergence criteria.
-Although the parameters are not unique, the likelihood keeps the same, using ELBO is justified.
-However, since there's no randomness in EM, it's impossible to detect the non-uniqueness.
-Actually, I think their results are even not local optima, but just a single point in the optimal surface (I guess if they use something like stochastic EM, they will find the issue).
-
-The PLDS also provides some constraint options, but some of them are still not enough to ensure unique solution.
-BTW, this might be another reason they choose to use EM, since adding constraint in EM is trivial: just do unconstrained EM in each iteration and project the value to the constraint space (i.e.
-add constraint) at the end of each iteration.
-But this doesn't work for MCMC, since we are sampling the distribution but not just the single optimal value.
-
-\textbf{I guess similar problems may happen in other LDS/ GPFA research.}
-
-Even if they make enough constraints to ensure uniqueness, these research should be careful about the pattern they found.
-For example, in \href{https://papers.nips.cc/paper/2020/hash/aa1f5f73327ba40d47ebce155e785aaf-Abstract.html}{Joshua et al., 2020}, the changes/ different states of interaction make sense to me, but I'm quite skeptical about their results in the dynamic matrix $\mathbf{A}$ itself.
-Maybe because of block-diagonal structure in loading $\mathbf{C}$, the interaction between population make sense somewhat.
-But at least the relationship within population doesn't make any sense, because of transformation invariance.
-
-\clearpage
-
-\section{TODO}
-\begin{enumerate}
-	\def\labelenumi{(\arabic{enumi})}
-	\item
-%	Think about MV-(Poisson)-GLM.
-Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
-	Improve the NUTS.
-	\item
-	Debug the clustering.
-	\item
-	Check and clear typo.
-%	\item
-%	After resolving (1), see if the MCMC is still trapped in local optima sometimes.
-Now, the chain seems will get stuck in local mode.
-Maybe updating the loading as a whole will remedy the problem a bit.
-	\item
-	Implement the mixture of finite mixture (MFM) by Jeff Miller.
-	\item
-	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
-%	\item
-%	improve DPMM.
-In current brute force implementation, the number of potential clusters can even go beyond number of neurons (\(N\)).
-There are several improvements, e.g.
-\href{https://link.springer.com/article/10.1007/s11222-009-9150-y}{Kalli et al., 2011}, \href{http://proceedings.mlr.press/v37/gea15.html}{Ge et al., 2015} and \href{https://link.springer.com/article/10.1007/s11222-014-9471-3}{Hastie et al, 2015}.
-Check them later.
-	\item
-	After all of them are resolved, switch to GP version
-
-\end{enumerate}
-
 \end{document}

### 2021-10-03 LDS/constraint/LDS/Q_blk.m

@@ -58,9 +58,13 @@ Y = poissrnd(exp(logLam));
 clusterPlot(Y, Lab)
 
 % standardize the latent vectors
-% center X around 0
-gtmp = -mean(X, 2);
-M = inv(diag(std(X, 0, 2)));
+
+% gtmp = -mean(X, 2);
+% M = inv(diag(max(abs(X),[], 2)));
+gtmp = -min(X,[], 2);
+M = inv(diag(range(X,2)));
+
+
 g = M*gtmp;
 
 X = M*X + g;
@@ -161,9 +165,10 @@ gradHess = @(vecX) gradHessX(vecX, d_tmp, C_fit(:,:,1), x0_fit(:,1), Q0,...
     Q_fit(:,:,1), A_fit(:,:,1), b_fit(:,1), Y);
 [muXvec,~,hess_tmp,~] = newtonGH(gradHess,X_tmp(:),1e-10,1000);
 X_fit(:,:,1) = reshape(muXvec, [], T);
-X_fit(:,:,1) = X_fit(:,:,1) - mean(X_fit(:,:,1), 2);
-X_fit(:,:,1) = (diag(std(X_fit(:,:,1), 0, 2)))\X_fit(:,:,1);
-
+% X_fit(:,:,1) = X_fit(:,:,1) - mean(X_fit(:,:,1), 2);
+% X_fit(:,:,1) = (diag(max(abs(X_fit(:,:,1)),[], 2)))\X_fit(:,:,1);
+X_fit(:,:,1) = X_fit(:,:,1) - min(X_fit(:,:,1),[], 2);
+X_fit(:,:,1) = (diag(range(X_fit(:,:,1),2)))\X_fit(:,:,1);
 
 %% MCMC
 optdc.M=1;
@@ -208,8 +213,11 @@ for g = 2:ng
     z = randn(length(muXvec), 1) + R'*muXvec;
     x_all = R'\z;
     X_fit(:,:,g) = reshape(x_all,[], T);
-    X_fit(:,:,g) = X_fit(:,:,g) - mean(X_fit(:,:,g), 2);
-    X_fit(:,:,g) = (diag(std(X_fit(:,:,g), 0, 2)))\X_fit(:,:,g);
+    X_fit(:,:,g) = X_fit(:,:,g) - min(X_fit(:,:,g),[], 2);
+    X_fit(:,:,g) = (diag(range(X_fit(:,:,g),2)))\X_fit(:,:,g);
+
[…]

### 2021-10-03 LDS/constraint/LDS_cluster/MM_Q_blk_norm.m

@@ -1,0 +1,206 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+% 
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+	A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+gtmp = -min(X,[], 2);
+M = inv(diag(range(X,2)));
+g = M*gtmp;
+
+X = M*X + g;
+x0 = M*x0 + g;
+d = d - (C_trans/M)*g;
+C_trans = C_trans/M;
+A = M*A/M;
+b = M*b + g - (M*A/M)*g;
[…]

### 2021-10-03 LDS/constraint/LDS_cluster/MM_full_norm.m

@@ -1,0 +1,193 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = ones(p,1)*0;
+b2 = ones(p,1)*0;
+b3 = ones(p,1)*0;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-3*eye(p);
+Q3 = 1e-3*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+% 
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+	A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    A = A+eye(size(Q,1))*0.92;
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+%%
+rng(2)
+ng = 30;
+kMM = 10;
+
+% pre-allocation
+Z_fit = zeros(N, ng);
+RHO_fit = zeros(kMM, ng);
+d_fit = zeros(N, kMM, ng);
+C_fit = zeros(N, p*kMM, ng);
[…]

### 2021-10-03 LDS/constraint/LDS_cluster/blockDiag_gibbsLoop_MM_norm.m

@@ -1,0 +1,300 @@
+function [XOut,x0Out,dOut,COut,...
+    mudcOut, SigdcOut,...
+    bOut,AOut,QOut, optdc, epsilon] =...
+    blockDiag_gibbsLoop_MM_norm(Y,X_tmp, Z_tmp, d_tmp, C_tmp,... % cluster-invariant
+    mudc_tmp, Sigdc_tmp,...
+    x0_tmp, b_tmp, A_tmp, Q_tmp, kMM,... % cluster-related
+    Q0, mux00, Sigx00, deltadc0, Taudc0,Psidc0,nudc0,...
+    BA0_all, Lamb0_f, Psi0,nu0, g, optdc, epsilon, burnIn) % priors
+
+% for development & debug...
+% Z_tmp = Z_fit(:,g);
+% X_tmp = X_fit(:,:,g-1);
+% d_tmp = d_fit(:,:,g-1);
+% C_tmp = C_fit(:,:,g-1);
+% mudc_tmp = mudc_fit(:,:,g-1);
+% Sigdc_tmp = Sigdc_fit(:,:,:,g-1);
+% x0_tmp = x0_fit(:,g-1);
+% b_tmp = b_fit(:,g-1);
+% A_tmp = A_fit(:,:,g-1);
+% Q_tmp = Q_fit(:,:,g-1);
+
+% output
+N = size(Y, 1);
+T = size(Y, 2);
+p = length(x0_tmp)/kMM;
+
+XOut = zeros(kMM*p, T);
+x0Out = zeros(kMM*p, 1);
+dOut = zeros(N, kMM);
+COut = zeros(N, p*kMM);
+mudcOut = zeros(p+1, kMM);
+SigdcOut = zeros(p+1, p+1, kMM);
+bOut = zeros(kMM*p, 1);
+AOut = zeros(kMM*p, kMM*p);
+QOut = zeros(kMM*p, kMM*p);
+
+[Zsort_tmp,idY] = sort(Z_tmp);
+uniZsort_tmp = unique(Zsort_tmp);
+nClus_tmp = length(uniZsort_tmp);
+latID = id2id(uniZsort_tmp, p);
+
+% labels without obs.: generate things by prior
+outLab = setdiff(1:kMM, uniZsort_tmp);
+if(~isempty(outLab))
+    
+    x0Out =  mvnrnd(mux00, Sigx00)';
+    for k =1:kMM
+        latID_tmp = id2id(k, p);
[…]

### 2021-10-03 LDS/constraint/LDS_cluster/full_gibbsLoop_MM_norm.m

@@ -1,0 +1,312 @@
+function [XOut,x0Out,dOut,COut,...
+    mudcOut, SigdcOut,...
+    bOut,AOut,QOut, optdc, epsilon] =...
+    full_gibbsLoop_MM_norm(Y,X_tmp, Z_tmp, d_tmp, C_tmp,... % cluster-invariant
+    mudc_tmp, Sigdc_tmp,...
+    x0_tmp, b_tmp, A_tmp, Q_tmp, kMM,... % cluster-related
+    Q0, mux00, Sigx00, deltadc0, Taudc0,Psidc0,nudc0,...
+    BA0_all, Lamb0_f, Psi0_f,nu0_f, g, optdc, epsilon, burnIn) % priors
+
+% for development & debug...
+% Z_tmp = Z_fit(:,g);
+% X_tmp = X_fit(:,:,g-1);
+% d_tmp = d_fit(:,:,g-1);
+% C_tmp = C_fit(:,:,g-1);
+% mudc_tmp = mudc_fit(:,:,g-1);
+% Sigdc_tmp = Sigdc_fit(:,:,:,g-1);
+% x0_tmp = x0_fit(:,g-1);
+% b_tmp = b_fit(:,g-1);
+% A_tmp = A_fit(:,:,g-1);
+% Q_tmp = Q_fit(:,:,g-1);
+
+% output
+N = size(Y, 1);
+T = size(Y, 2);
+p = length(x0_tmp)/kMM;
+
+XOut = zeros(kMM*p, T);
+x0Out = zeros(kMM*p, 1);
+dOut = zeros(N, kMM);
+COut = zeros(N, p*kMM);
+mudcOut = zeros(p+1, kMM);
+SigdcOut = zeros(p+1, p+1, kMM);
+bOut = zeros(kMM*p, 1);
+AOut = zeros(kMM*p, kMM*p);
+QOut = zeros(kMM*p, kMM*p);
+
+[Zsort_tmp,idY] = sort(Z_tmp);
+uniZsort_tmp = unique(Zsort_tmp);
+nClus_tmp = length(uniZsort_tmp);
+latID = id2id(uniZsort_tmp, p);
+
+% labels without obs.: generate things by prior
+outLab = setdiff(1:kMM, uniZsort_tmp);
+if(~isempty(outLab))
+    
+    x0Out =  mvnrnd(mux00, Sigx00)';
+    QOut = iwishrnd(Psi0_f(kMM), nu0_f(kMM));
+    
+    bOut = zeros(kMM*p, 1);
[…]

### 2021-10-03 LDS/constraint/LDS_cluster/unstable/MM_diagAQ_norm.m

@@ -54,96 +54,65 @@ end
 
 Y = poissrnd(exp(logLam));
 clusterPlot(Y, Lab)
+
+gtmp = -min(X,[], 2);
+M = inv(diag(range(X,2)));
+g = M*gtmp;
+
+X = M*X + g;
+x0 = M*x0 + g;
+d = d - (C_trans/M)*g;
+C_trans = C_trans/M;
+A = M*A/M;
+b = M*b + g - (M*A/M)*g;
+Q = M*Q*M';
+Q0 = M*Q0*M';
+
 %%
-rng(3)
+rng(6)
 ng = 50;
-kMM = 3;
+kMM = 10;
 
 % pre-allocation
 Z_fit = zeros(N, ng);
 RHO_fit = zeros(kMM, ng);
-
-X_fit = zeros(kMM*p, T, ng);
-x0_fit = zeros(kMM*p, ng);
-
 d_fit = zeros(N, kMM, ng);
-mud_fit = zeros(kMM, ng);
-sig2d_fit = zeros(kMM, ng);
-
-% do I need to update mean of K?
-K_fit = zeros(N, kMM*p, ng);
-muK_fit = zeros(kMM, ng);
-sig2K_fit = ones(kMM, ng);
-C_fit = zeros(N, kMM*p, ng);
-
-A_fit = zeros(kMM*p, kMM*p, ng);
+C_fit = zeros(N, p*kMM, ng);
+mudc_fit = zeros(p+1, kMM, ng);
+Sigdc_fit = zeros(p+1, p+1, kMM, ng);
 b_fit = zeros(kMM*p, ng);
+A_fit = zeros(kMM*p, kMM*p, ng);
 Q_fit = zeros(kMM*p, kMM*p, ng);
+x0_fit = zeros(kMM*p, ng);
+X_fit = zeros(kMM*p, T, ng);
 
-% priors...
+% priors
 delta0 = ones(1, kMM);
 Q0 = eye(kMM*p)*1e-2;
 
-% priors for initial (x0)
 mux00 = zeros(kMM*p, 1);
 Sigx00 = eye(kMM*p);
 
-% prior for K: C = K*(K'*K)^{-1/2}
-% vec(K) ~ N(mu_k*1, sig2K*I)
-% i.e. k_{ij} ~ (i.i.d.) N(mu_k, sig2K)
-nuK0 = 1;
-sig2K0 = 1;
-deltaK0 = 0;
-kK0 = 1;
[…]

### 2021-10-03 LDS/constraint/LDS_cluster/unstable/norm_AQ_diag_MM_MCMCLoop.m

@@ -1,0 +1,298 @@
+function [XOut,x0Out,dOut,COut,...
+    mudcOut, SigdcOut,...
+    bOut,AOut,QOut, optdc, epsilon] =...
+    norm_AQ_diag_MM_MCMCLoop(Y,X_tmp, Z_tmp, d_tmp, C_tmp,... % cluster-invariant
+    mudc_tmp, Sigdc_tmp,...
+    x0_tmp, b_tmp, A_tmp, Q_tmp, kMM,... % cluster-related
+    Q0, mux00, Sigx00, deltadc0, Taudc0,Psidc0,nudc0,...
+    BA0, Lamb0, Psi0,nu0, g, optdc, epsilon, burnIn) % priors
+
+% for development & debug...
+% Z_tmp = Z_fit(:,g);
+% X_tmp = X_fit(:,:,g-1);
+% d_tmp = d_fit(:,:,g-1);
+% C_tmp = C_fit(:,:,g-1);
+% mudc_tmp = mudc_fit(:,:,g-1);
+% Sigdc_tmp = Sigdc_fit(:,:,:,g-1);
+% x0_tmp = x0_fit(:,g-1);
+% b_tmp = b_fit(:,g-1);
+% A_tmp = A_fit(:,:,g-1);
+% Q_tmp = Q_fit(:,:,g-1);
+
+% output
+N = size(Y, 1);
+T = size(Y, 2);
+p = length(x0_tmp)/kMM;
+
+XOut = zeros(kMM*p, T);
+x0Out = zeros(kMM*p, 1);
+dOut = zeros(N, kMM);
+COut = zeros(N, p*kMM);
+mudcOut = zeros(p+1, kMM);
+SigdcOut = zeros(p+1, p+1, kMM);
+bOut = zeros(kMM*p, 1);
+AOut = zeros(kMM*p, kMM*p);
+QOut = zeros(kMM*p, kMM*p);
+
+[Zsort_tmp,idY] = sort(Z_tmp);
+uniZsort_tmp = unique(Zsort_tmp);
+nClus_tmp = length(uniZsort_tmp);
+latID = id2id(uniZsort_tmp, p);
+
+% labels without obs.: generate things by prior
+outLab = setdiff(1:kMM, uniZsort_tmp);
+if(~isempty(outLab))
+    
+    x0Out =  mvnrnd(mux00, Sigx00)';
+    for k =1:kMM
+        latID_tmp = id2id(k, p);
[…]

### 2021-10-04 documents/cluster_neural_population.tex
resolve clustering issues

@@
 
+\section{Some Issues in clustering}
+The first problem is how to update the unobserved components. In DP, this is the same as how to generate new components.
+
+Traditionally, the unobserved ones are just sampled from prior.
+However, this doesn't work in my case, since the dimension of $\bm{X}$ is super high ($6\times 1000$) in simulation.
+It's nearly impossible to accept the newly generated components.
+
+Here's my solution: other parameters are generated from prior, but $\bm{X}$ in new component is just the same  as, or just another sample for the largest cluster.
+
+In DP, we can do things even more aggressive: at the first few steps (e.g.
+pre-burn-in = 10), let all newly generated component parameters be the same (or samples from the same distribution) as the largest cluster.
+\textbf{This strategy encourages more clusters but keeps the clustering structure at the same time}.
+
+After pre-burn-in, we goes back to normal: all parameters except for $\bm{X}$ are generated from priors.
+This gives the algorithm some chances to hit a "good" one, but the probability is not as high as in pre-burn-in.
+
+The second problem comes when I let $\bm{Q}$ be full.
+For example, at one step, I have cluster index 1, 2, 4.
+But the sub-matrix of $\bm{Q}$ buying 1, 2, 4 columns and rows is not necessarily positive-definite.
+Here, I just do the easy thing: if the sub-matrix is not p.d., only keep the diagonal elements or block diagonal elements.
+
 \clearpage
@@
 	\item
-%	Think about MV-(Poisson)-GLM.
-Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
 	Improve the NUTS.
@@
 	\item
-	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
+	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected. (\textbf{resolved}?)
 %	\item

### 2021-10-04 LDS/constraint/LDS_cluster/DP_diagAQ_norm.m

@@ -1,0 +1,250 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-2 2];
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+figure(1)
+Y = poissrnd(exp(logLam));
+clusterPlot(Y, Lab)
+
+gtmp = -min(X,[], 2);
+M = inv(diag(range(X,2)));
+g = M*gtmp;
+
+X = M*X + g;
[…]

### 2021-10-04 LDS/constraint/LDS_cluster/MM_diagAQ_norm.m

@@ -15,7 +15,8 @@ pLab = repelem(1:nClus, p);
 d = randn(n*nClus,1)*0;
 C_trans = zeros(n*nClus, p*nClus);
 for k = 1:length(Lab)
-    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+1;
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-2 2];
 end
 
 X = zeros(p*nClus, T);
@@ -23,24 +24,25 @@ x0 = zeros(p*nClus, 1);
 Q0 = eye(nClus*p)*1e-2;
 X(:,1) = mvnrnd(x0, Q0)';
 
-b1 = ones(p,1)*0;
-b2 = ones(p,1)*0;
-b3 = ones(p,1)*0;
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
 b = [b1;b2;b3];
 
 Q1 = 1e-3*eye(p);
-Q2 = 1e-3*eye(p);
-Q3 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
 Q = blkdiag(Q1, Q2, Q3);
 
-% 
+%
 A = eye(size(Q,1));
 while any(imag(eig(A))==0)
     A= randn(size(Q));
     A = A-diag(diag(A));
-	A(squareform(pdist(pLab'))==0)=0;
+    A(squareform(pdist(pLab'))==0)=0;
     A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
-    A = A+eye(size(Q,1))*0.92;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
 end
 
 % let's generate lambda
@@ -52,6 +54,7 @@ for t=2:T
     logLam(:, t) = d + C_trans*X(:,t);
 end
 
+figure(1)
 Y = poissrnd(exp(logLam));
 clusterPlot(Y, Lab)
 
@@ -68,10 +71,14 @@ b = M*b + g - (M*A/M)*g;
[…]

### 2021-10-04 LDS/constraint/LDS_cluster/MM_full_norm.m

@@ -57,8 +57,8 @@ clusterPlot(Y, Lab)
 
 %%
 rng(2)
-ng = 30;
-kMM = 10;
+ng = 100;
+kMM = 30;
 
 % pre-allocation
 Z_fit = zeros(N, ng);

### 2021-10-04 LDS/constraint/LDS_cluster/norm_AQ_diag_DP_MCMCLoop_v2.m

@@ -1,38 +1,38 @@
 function [XOut,x0Out,dOut,COut,...
     mudcOut, SigdcOut,...
     bOut,AOut,QOut, optdc, epsilon] =...
-    full_gibbsLoop_MM_norm(Y,X_tmp, Z_tmp, d_tmp, C_tmp,... % cluster-invariant
+    norm_AQ_diag_DP_MCMCLoop_v2(Y,X_tmp, Z_tmp, d_tmp, C_tmp,...
     mudc_tmp, Sigdc_tmp,...
-    x0_tmp, b_tmp, A_tmp, Q_tmp, kMM,... % cluster-related
-    Q0, mux00, Sigx00, deltadc0, Taudc0,Psidc0,nudc0,...
-    BA0_all, Lamb0_f, Psi0_f,nu0_f, g, optdc, epsilon, burnIn) % priors
+    x0_tmp, b_tmp, A_tmp, Q_tmp, s_star,...
+    Q0_f, mux00_f, Sigx00_f, deltadc0, Taudc0,Psidc0,nudc0,...
+    BA0, Lamb0, Psi0,nu0, g, optdc, epsilon, burnIn)
 
 % for development & debug...
-% Z_tmp = Z_fit(:,g);
-% X_tmp = X_fit(:,:,g-1);
-% d_tmp = d_fit(:,:,g-1);
-% C_tmp = C_fit(:,:,g-1);
-% mudc_tmp = mudc_fit(:,:,g-1);
-% Sigdc_tmp = Sigdc_fit(:,:,:,g-1);
-% x0_tmp = x0_fit(:,g-1);
-% b_tmp = b_fit(:,g-1);
-% A_tmp = A_fit(:,:,g-1);
-% Q_tmp = Q_fit(:,:,g-1);
+% Z_tmp = Z_fit(:,g-1);
+% X_tmp = X_fit{g-1};
+% d_tmp = d_fit{g-1};
+% C_tmp = C_fit{g-1};
+% mudc_tmp = mudc_fit{g-1};
+% Sigdc_tmp = Sigdc_fit{g-1};
+% x0_tmp = x0_fit{g-1};
+% b_tmp = b_fit{g-1};
+% A_tmp = A_fit{g-1};
+% Q_tmp = Q_fit{g-1};
 
 % output
 N = size(Y, 1);
 T = size(Y, 2);
-p = length(x0_tmp)/kMM;
+p = 2;
 
[…]

### 2021-10-04 LDS/constraint/LDS_cluster/norm_AQ_diag_MM_MCMCLoop.m

@@ -44,13 +44,12 @@ outLab = setdiff(1:kMM, uniZsort_tmp);
 if(~isempty(outLab))
     
     x0Out =  mvnrnd(mux00, Sigx00)';
-    for k =1:kMM
-        latID_tmp = id2id(k, p);
-        QOut(latID_tmp,latID_tmp) = iwishrnd(Psi0,nu0);
+    for k = 1:(p*kMM)
+        QOut(k,k) = iwishrnd(Psi0,nu0);
+        bATmp = mvnrnd(BA0, kron(QOut(k,k), inv(Lamb0)));
+        bOut(k) = 0;% bATmp(1)
+        AOut(k,k) = 1; % bATmp(2)
     end
-    
-    bOut = zeros(kMM*p, 1);
-    AOut = eye(kMM*p);
     
     invQ0 = inv(sparse(Q0));
     R = chol(invQ0,'lower');
@@ -85,32 +84,16 @@ A_tmp2 = A_tmp(latID, latID);
 b_tmp2 = b_tmp(latID);
 Q_tmp2 = Q_tmp(latID, latID);
 
-% [muX,~,~] = ppasmoo_poissexp_v2(Y_tmp2,Csort_trans_tmp,...
-%     d_tmp2,x0_tmp2,Q0_tmp2,A_tmp2,b_tmp2,Q_tmp2);
-% hess_tmp = hessX(muX(:),d_tmp2,Csort_trans_tmp,Q0_tmp2,Q_tmp2,A_tmp2,Y_tmp2);
-
-% if use Newton-Raphson directly?
-% X_tmp = ppasmoo_poissexp_v2(Y_tmp2,Csort_trans_tmp,...
-%     d_tmp2,x0_tmp2,Q0_tmp2,A_tmp2,b_tmp2,Q_tmp2);
-% gradHess = @(vecX) gradHessX(vecX, d_tmp2, Csort_trans_tmp, x0_tmp2, Q0_tmp2, Q_tmp2, A_tmp2, b_tmp2, Y_tmp2);
-% [muXvec,~,hess_tmp,~] = newtonGH(gradHess,X_tmp(:),1e-6,1000);
-
-
+X_tmp2 = X_tmp(latID, :);
[…]

### 2021-10-04 LDS/constraint/LDS_cluster/norm_AQ_diag_MM_MCMCLoop_v2.m

@@ -1,11 +1,11 @@
 function [XOut,x0Out,dOut,COut,...
     mudcOut, SigdcOut,...
     bOut,AOut,QOut, optdc, epsilon] =...
-    full_gibbsLoop_MM_norm(Y,X_tmp, Z_tmp, d_tmp, C_tmp,... % cluster-invariant
+    norm_AQ_diag_MM_MCMCLoop_v2(Y,X_tmp, Z_tmp, d_tmp, C_tmp,... % cluster-invariant
     mudc_tmp, Sigdc_tmp,...
     x0_tmp, b_tmp, A_tmp, Q_tmp, kMM,... % cluster-related
     Q0, mux00, Sigx00, deltadc0, Taudc0,Psidc0,nudc0,...
-    BA0_all, Lamb0_f, Psi0_f,nu0_f, g, optdc, epsilon, burnIn) % priors
+    BA0, Lamb0, Psi0,nu0, g, optdc, epsilon, burnIn) % priors
 
 % for development & debug...
 % Z_tmp = Z_fit(:,g);
@@ -23,6 +23,17 @@ function [XOut,x0Out,dOut,COut,...
 N = size(Y, 1);
 T = size(Y, 2);
 p = length(x0_tmp)/kMM;
+
+% XOut = X_tmp;
+% x0Out = x0_tmp;
+% dOut = d_tmp;
+% COut = C_tmp;
+% mudcOut = mudc_tmp;
+% SigdcOut = Sigdc_tmp;
+% bOut = b_tmp;
+% AOut = A_tmp;
+% QOut = Q_tmp;
+
 
 XOut = zeros(kMM*p, T);
 x0Out = zeros(kMM*p, 1);
@@ -44,16 +55,19 @@ outLab = setdiff(1:kMM, uniZsort_tmp);
 if(~isempty(outLab))
     
     x0Out =  mvnrnd(mux00, Sigx00)';
-    QOut = iwishrnd(Psi0_f(kMM), nu0_f(kMM));
-    
-    bOut = zeros(kMM*p, 1);
-    AOut = eye(kMM*p);
+    for k = 1:(p*kMM)
+        QOut(k,k) = iwishrnd(Psi0,nu0);
+        bATmp = mvnrnd(BA0, kron(QOut(k,k), inv(Lamb0)));
+        bOut(k) = 0;% bATmp(1)
+        AOut(k,k) = 1; % bATmp(2)
+    end
     
     invQ0 = inv(sparse(Q0));
     R = chol(invQ0,'lower');
[…]

### 2021-10-04 LDS/constraint/wait_to_debug/CCI_all.m

@@ -1,348 +1,0 @@
-%%
-rng(123)
-n = 10;
-nClus = 3;
-N = n*nClus;
-p = 2;
-T = 1000;
-
-Lab = repelem(1:nClus, n);
-d = randn(n*nClus,1)/5;
-C_trans = zeros(N, p*nClus);
-C = zeros(N, p);
-for k = 1:length(Lab)
-    C(k,:) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)) + [-1 .5];
-    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C(k,:);
-end
-
-X = zeros(p*nClus, T);
-x0 = zeros(nClus*p,1);
-Q0 = eye(nClus*p)*1e-2;
-X(:,1) = mvnrnd(x0, Q0)';
-
-b1 = ones(p,1)*0.01;
-b2 = ones(p,1)*0;
-b3 = ones(p,1)*-0.03;
-b = [b1;b2;b3];
-b = b*0;
-
-Q1 = 1e-3*eye(p);
-Q2 = 1e-3*eye(p);
-Q3 = 1e-3*eye(p);
-Q = blkdiag(Q1, Q2, Q3);
-
-% Generate X offline (A unspecified)
-for i=1:size(Q,1)
-    k = ceil(rand()*20)+15;
-    X(i,:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
-end
-
-% let's generate lambda
-logLam = zeros(n*nClus, T);
-logLam(:,1) = d + C_trans*X(:,1);
-
-for t=2:T
-    logLam(:, t) = d + C_trans*X(:,t);
-end
-
-figure(1)
-Y = poissrnd(exp(logLam));
-clusterPlot(Y, Lab)
-
-[~, S, V] = svd(C, 0);
-Mcore = S*V';
-M = kron(eye(nClus), Mcore);
-
-X = M*X;
-x0 = M*x0;
-Q0 = M*Q0*M';
-C = C/Mcore;
-C_trans = C_trans/M;
-% b = M*b;
-% A = M*A/M;
-Q = M*Q*M';
-
-% center X around 0
-g = mean(X, 2);
-X = X - g;
-x0 = x0 - g;
-d = d - C_trans* g;
-% b = b + g - A*g;
-
-figure(2)
-figure(2)
-subplot(1,2,1)
-imagesc(exp(logLam))
[…]

### 2021-10-04 LDS/constraint/wait_to_debug/CCI_blk1.m

@@ -1,361 +1,0 @@
-%%
-rng(1)
-n = 10;
-nClus = 3;
-N = n*nClus;
-p = 2;
-T = 1000;
-
-Lab = repelem(1:nClus, n);
-d = randn(n*nClus,1)/5;
-C_trans = zeros(N, p*nClus);
-C = zeros(N, p);
-for k = 1:length(Lab)
-    C(k,:) = sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)) + [-1 .5];
-    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = C(k,:);
-end
-
-X = zeros(p*nClus, T);
-x0 = zeros(nClus*p,1);
-Q0 = eye(nClus*p)*1e-2;
-X(:,1) = mvnrnd(x0, Q0)';
-
-b1 = ones(p,1)*0.01;
-b2 = ones(p,1)*0;
-b3 = ones(p,1)*-0.03;
-b = [b1;b2;b3];
-b = b*0;
-
-Q1 = 1e-3*eye(p);
-Q2 = 1e-3*eye(p);
-Q3 = 1e-3*eye(p);
-Q = blkdiag(Q1, Q2, Q3);
-
-% Generate X offline (A unspecified)
-for i=1:size(Q,1)
-    k = ceil(rand()*20)+15;
-    X(i,:) = interp1(linspace(0,1,k),...
-        randn(k,1),linspace(0,1,T),'spline');
-end
-plot(X')
-
-
-% let's generate lambda
-logLam = zeros(n*nClus, T);
-logLam(:,1) = d + C_trans*X(:,1);
-
-for t=2:T
-    logLam(:, t) = d + C_trans*X(:,t);
-end
-
-figure(1)
-Y = poissrnd(exp(logLam));
-clusterPlot(Y, Lab)
-
-[~, S, V] = svd(C, 0);
-Mcore = S*V';
-M = kron(eye(nClus), Mcore);
-
-X = M*X;
-x0 = M*x0;
-Q0 = M*Q0*M';
-C = C/Mcore;
-C_trans = C_trans/M;
-% b = M*b;
-% A = M*A/M;
-Q = M*Q*M';
-
-% center X around 0
-g = mean(X, 2);
-X = X - g;
-x0 = x0 - g;
-d = d - C_trans* g;
-% b = b + g - A*g;
-
-figure(2)
-figure(2)
-subplot(1,2,1)
[…]

### 2021-10-04 LDS/constraint/wait_to_debug/CCI_blk2.m

@@ -1,258 +1,0 @@
-rng(3)
-ng = 10000;
-
-X_fit = zeros(nClus*p, T, ng);
-x0_fit = zeros(nClus*p, ng);
-
-d_fit = zeros(N, nClus, ng);
-mud_fit = zeros(nClus, ng);
-sig2d_fit = zeros(nClus, ng);
-
-% do I need to update mean of K?
-K_fit = zeros(N, nClus*p, ng);
-muK_fit = zeros(p, nClus, ng);
-SigK_fit = repmat(eye(p), 1, 1, nClus, ng);
-C_fit = zeros(N, nClus*p, ng);
-
-A_fit = zeros(nClus*p, nClus*p, ng);
-b_fit = zeros(nClus*p, ng);
-Q_fit = zeros(nClus*p, nClus*p, ng);
-
-% priors...
-Q0 = eye(nClus*p)*1e-2;
-
-% priors for initial (x0)
-mux00 = zeros(nClus*p, 1);
-Sigx00 = eye(nClus*p);
-
-% prior for K: C = K*(K'*K)^{-1/2}
-% (row) k_i iid ~ N(mu, Sigma)
-PsiK0 = eye(p);
-nuK0 = p+2;
-deltaK0 = zeros(p,1);
-kK0 = 1;
-
-
-% prior for d
-nud0 = 1;
-sig2d0 = 1e-2;
-deltad0 = 0;
-kd0 = 1;
-
-% prior for linear dyanmics (b, A, Q)
-BA0 =[0 1]';
-Lamb0 = eye(2);
-Psi0 = 1e-4;
-nu0 = 1+2;
-
-% initials...
-sig2d_fit(:,1) = ones(nClus,1)*1e-2;
-mud_fit(:,1) = zeros(nClus,1);
-d_fit(:,:,1) = zeros(N,nClus);
-
-muK_fit(:,:,1) = zeros(p, nClus);
-SigK_fit(:,:,:,1) = repmat(eye(p), 1, 1, nClus);
-
-K_raw = randn(N, p);
-d_tmp = zeros(N,1);
-for k = unique(Lab)
-    ladid_tmp = id2id(k, p);
-    K_tmp = K_raw(Lab == k, :);
-    K_fit(Lab == k, ladid_tmp, 1) = K_tmp;
-    C_fit(Lab == k, ladid_tmp, 1) = K_tmp/(sqrtm(K_tmp'*K_tmp));
[…]

### 2021-10-04 LDS/clusPlot.m
add DP plot

@@ -1,13 +1,14 @@
 addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
 %%
 clear all;close all;clc;
-load('C:\Users\gaw19004\Desktop\LDS_backup\new\noA_full_MM_below_4.mat')
+load('C:\Users\gaw19004\Desktop\LDS_backup\new3\cluster\DP_below_alpha5_norm_AQ_diag_rmRedun_1000.mat')
 
-plotFolder = "C:\Users\gaw19004\Desktop\LDS_backup\new\plot\noA_full_MM_3\";
-for k = 1:ng
+
+plotFolder = "C:\Users\gaw19004\Desktop\LDS_backup\new3\cluster\plot\DP_below\";
+for k = 1:10:500
     clus = figure;
     clusterPlot(Y, Z_fit(:,k)')
-    title("MM, Max Clust = " + kMM + ", k = " + k); % for MM
-%     title("DP, \alpha = " + alphaDP + ", k = " + k); % for DP
+%     title("MM, Max Clust = " + kMM + ", k = " + k); % for MM
+    title("DP, \alpha = " + alphaDP + ", burin =" + burnIn, ", k = " + k); % for DP
     saveas(clus, plotFolder+k+".png")
 end

### 2021-10-04 LDS/constraint/LDS_cluster/DP_diagAQ_norm.m

@@ -77,7 +77,7 @@ plot(X')
 %% MCMC setting
 rng(3)
 alphaDP = 5;
-ng = 500;
+ng = 1000;
 
 % pre-allocation
 Z_fit = zeros(N, ng);
@@ -101,9 +101,9 @@ nu0 = 1+2;
 
 % initials
 % single cluster
-Z_fit(:,1) = ones(N, 1);
+% Z_fit(:,1) = ones(N, 1);
 % N cluster
-% Z_fit(:,1) = 1:N;
+Z_fit(:,1) = 1:N;
 % Z_fit(:,1) = randsample(2, N, true);
 
 % reorder Y by labels
@@ -151,7 +151,7 @@ X_fit{1}(latID,:) = (diag(range(X_fit{1}(latID,:),2)))\X_fit{1}(latID,:);
 optdc.M=1;
 optdc.Madapt=0;
 epsilon = 0.01*ones(N,1);
-burnIn = 10;
+burnIn = 100;
 
 for g = 2:ng
     
@@ -204,46 +204,46 @@ for g = 2:ng
     [Z_fit(~isnan(clus_tmp(:,1)),g), ~] = find(clus_tmp(~isnan(clus_tmp(:,1)),:)');
     
     % delete the redundant...
-%     oldLab_uni = unique(Z_fit(:,g));
-%     nClus_new = length(oldLab_uni);
-%     
-%     X_new = zeros(nClus_new*p, T);
-%     x0_new = zeros(nClus_new*p, 1);
-%     d_new = zeros(N, nClus_new);
-%     C_new = zeros(N, p*nClus_new);
-%     mudc_new = zeros(p+1, nClus_new);
-%     Sigdc_new = zeros(p+1,p+1, nClus_new);
-%     b_new = zeros(nClus_new*p,1);
-%     A_new = eye(nClus_new*p);
-%     Q_new = eye(nClus_new*p);
-%     
-%     for k = 1:nClus_new
-%         Z_fit(Z_fit(:,g) == oldLab_uni(k), g) = k;
-%         
-%         latid_old = id2id(oldLab_uni(k), p);
-%         latid_new = id2id(k,p);
-%         
-%         X_new(latid_new, :) = X_fit{g}(latid_old,:);
-%         x0_new(latid_new) = x0_fit{g}(latid_old);
-%         d_new(:, k) = d_fit{g}(:,oldLab_uni(k));
-%         C_new(:,latid_new) = C_fit{g}(:,latid_old);
[…]

### 2021-10-04 LDS/constraint/LDS_cluster/MM_diagAQ_norm.m

@@ -76,9 +76,9 @@ plot(X')
 
 
 %%
-rng(2)
+rng(1)
 ng = 100;
-kMM = 4;
+kMM = 30;
 
 % pre-allocation
 Z_fit = zeros(N, ng);

### 2022-01-05 latexDraft/draft.tex
draft1 for latex

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2021

% ready for submission
\usepackage{neurips_2021}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2021}

% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2021}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2021}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{bm}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{subfigure}

\title{Clustering Neural Populations by Poisson Dynamic Factor Model}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
	David S.~Hippocampus\thanks{Use footnote for providing further information
		about author (webpage, alternative address)---\emph{not} for acknowledging
		funding agencies.} \\
	Department of Computer Science\\
	Cranberry-Lemon University\\
	Pittsburgh, PA 15213 \\
	\texttt{hippo@cs.cranberry-lemon.edu} \\
	% examples of more authors
	% \And
	% Coauthor \\
	% Affiliation \\
	% Address \\
	% \texttt{email} \\
	% \AND
	% Coauthor \\
	% Affiliation \\
	% Address \\
	% \texttt{email} \\
	% \And
	% Coauthor \\
	% Affiliation \\
	% Address \\
	% \texttt{email} \\
	% \And
	% Coauthor \\
	% Affiliation \\
	% Address \\
	% \texttt{email} \\
}

\begin{document}

	\maketitle

	\begin{abstract}
		\answerTODO{}
	\end{abstract}

	\section{Introduction}
	\label{intro}
	\answerTODO{}

	\section{Method}
	\label{method}

	\subsection{Poisson Dynmic Factor Model}
	Denote the observed spike-count of neuron $i \in \{ 1,\ldots,N\}$ at
	time bin $t \in \{ 1,\ldots,T\}$ as
	$y_{it} \in \mathbb{Z}_{\geq 0}$, and let
	$\bm{y}_{i} =  (y_{i1},\ldots,y_{iT})'$.
	Further, let $z_{i} = j$ denote the cluster indicator of neuron $i$. To facilitate clustering, we re-parametrize the regular Poisson linear dynamical system (PLDS) model to separate the mean log-firing-rate out. Assume neurons are independently Poisson distributed, conditional on the
	low-dimensional latent state $\bm{x}_{t}^{(j)} \in \mathbb{R}^{p}$ as follows:
	\begin{align*}
		y_{it} &\sim Poi(\lambda_{it})\\
		\log\lambda_{it} &= \mu_t^{(j)} + \bm{c}'_i\bm{x}^{(j)}_t
	\end{align*}
	, with $\bm{c}_i\sim N(\bm{0},\bm{I}_p)$. We further assume the intercept $\mu_t^{(j)}$ and the latent state $\bm{x}^{(j)}_t$ progress linearly with Gaussian noise as
	\begin{align*}
		\mu_1^{(j)} &\sim N(\mu_0, \Sigma_0)\\
		\mu_{t+1}^{(j)} &\sim N(f^{(j)}\mu_t^{(j)} + g^{(j)}, \Sigma^{(j)})\\
		\bm{x}_1^{(j)} &\sim N(\bm{x}_0, \bm{Q}_0)\\
		\bm{x}_{t+1}^{(j)} &\sim N(\bm{A}^{(j)}\bm{x}_t^{(j)} + \bm{b}^{(j)}, \bm{Q}^{(j)})
	\end{align*}
	If we denote $\bm{\lambda}_{i} =  (\lambda_{i1},\ldots,\lambda_{iT})'$, $\bm{\mu}^{(j)} = (\mu^{(j)}_1,\ldots,\mu^{(j)}_T)'$ and $\bm{X}^{(j)} = (\bm{x}^{(j)}_1,\ldots,\bm{x}^{(j)}_T)'$, the model can be equivalently written as regular Poisson factor model as
	\begin{align*}
		\bm{y}_i &\sim Poi(\bm{\lambda}_i)\\
		\log\bm{\lambda}_i &= \bm{\mu}^{(j)}+ \bm{X}^{(j)}\bm{c}_i 
	\end{align*}

	However, since the condition $T/N \rightarrow 0$ doesn’t hold, the latent state $bm{X}^{(j)}$ cannot be consistently estimated, and assuming linear dynamics of $\bm{X}^{(j)}$ resolves the problem. Note that when $p>1$, the model is not unique, since $\widetilde{\bm{X}}^{(j)} = \bm{X}^{(j)}\bm{U}$ also satisfies the equation for any orthogonal matrix $\bm{U}$ of order $p$. To ensure the model indefinability, we simply assume $\bm{A}^{(j)}$ and $\bm{Q}^{(j)}$
	are both diagonal for convenience. See more detailed discussions of the constraints in discussion. Overall, denote the cluster-related parameters of cluster $j$ as $\bm{\theta}^{(j)}= \{\bm{\mu}^{(j)}, \bm{X}^{(j)}, f^{(j)}, g^{(j)}, \Sigma^{(j)},\{\bm{A}^{(j)}, \{\bm{b}^{(j)}, \{\bm{Q}^{(j)}\}$ and the spike counts of neuron i is generated by Poisson dynamic factor model (PDFM) as $\bm{Y}_i\sim PDFM(\bm{\theta}^{(z_i)})$, with the prior of $\bm{\theta}^{(j)}$ as $\bm{H}$. The priors for $\{f^{(j)}, g^{(j)}, \Sigma^{(j)},\{\bm{A}^{(j)}, \{\bm{b}^{(j)}, \{\bm{Q}^{(j)}\}$ are regular  normal and inverse-gamma distribution (or multivariate normal and inverse-Wishart when using other non-diagonal constraints).

	The marginal likelihood of neuron i is

	$$M_{\bm{\theta}^{(j)}}(\bm{y}_i) = P(\bm{y}_i|\bm{\theta}^{(j)}) = \int P(\bm{y}_i|\bm{\theta}^{(j)}, \bm{c}_i)P(\bm{c}_i)\,d\bm{c}_i$$

	The marginal likelihood has no closed form and will be used for clustering. To help with fast clustering, instead of doing the Laplace approximation, we choose to make use of the Poisson-Gamma conjugacy [reference]. This leads to the closed form approximation. By the conditional independency assumption, $M_{\bm{\theta}^{(j)}}(\bm{y}_i)=\prod_{t=1}^{T}P(y_{it}|\bm{\theta}^{(j)})$. Since $\bm{c}_i\sim N(\bm{0},\bm{I}_p)$, $\lambda_{it} = \exp(\mu_t^{(j)} + \bm{c}'_i\bm{x}^{(j)}_t)\sim lognormal(\mu_t^{(j)}, \bm{x}'^{(j)}_t\bm{x}^{(j)}_t)$. Approximate the lognormal distribution by Gamma distribution, s.t. $lognormal(\mu_t^{(j)}, \bm{x}'^{(j)}_t\bm{x}^{(j)}_t) \approx Gamma(a_{it}, b_{it})$ with $a_{it} = (\bm{x}'^{(j)}_t\bm{x}^{(j)}_t)^{-1}$ and $b_{it} = \bm{x}'^{(j)}_t\bm{x}^{(j)}_t\cdot e^{\mu_t^{(j)}}$. Then by Poisson-Gamma conjugacy,

	$$
	P(y_{it}|\bm{\theta}^{(j)}) = \int P(y_{it}|\lambda_{it})P(\lambda_{it})\,d\lambda_{it}\approx NB(y_{it}|r_{it}, p_{it})
	$$

	, with $r_{it} = a_{it}$ and $p_{it} = 1/(1 + b_{it})$.

	Another more general idea is to approximate the log-likelihood by second-order polynomials, with coefficients determined by Chebyshev polynomial approximation [Refer to PAL]. However, the approximation doesn’t work well in practice, especially when the neural spike counts have a wide range. When doing the integration, we need to exponentiate the log-likelihood and this will exaggerate the approximation error.

	\subsection{Cluster by Mixture of Finite Mixtures Model}
	When the label is unknown, we cluster the neurons by mixture models. In practice, it’s usually impossible to know the number of neural population. One potential method is to do clustering by Dirichlet process mixtures (DPM) model. However, this is conceptually incorrect, since the number of neural populations is finite but unknown. Besides the conceptual incorrectness, using DPM is not easy to integrate the field knowledge about the number of neural populations. Here, we choose to use the mixture of finite mixtures (MFM) model as follows

	\begin{align*}
		K &\sim p_k &&\text{where}\, p_k \, \text{is a p.m.f. on} \{1,2,\ldots\}\\
		\bm{\pi}=(\pi_1,\ldots,\pi_k) &\sim Dir_k(\gamma, \ldots,\gamma) &&\text{given} K=k\\
		Z_1,\ldots,Z_n&\stackrel{i.i.d.}{\sim}\pi &&\text{given} \pi\\
		\bm{\theta}_1,\ldots,\bm{\theta}_k&\stackrel{i.i.d.}{\sim}\bm{H} &&\text{given} K=k\\
		\bm{Y}_i = (y_{i1},\ldots,y_{iT})' &\sim SSFM(\bm{\theta}^{(z_i)}) &&\text{independently for}\, i=1,\ldots,N, \text{given}\, \bm{\theta}_{1:K}\, \text{and}\, Z_{1:N}
	\end{align*}

	Besides the conceptual correctness, using MFM model allows us integrate the prior knowledge easily. Moreover, compared to DPM, MFM has some better properties for clustering, for example, MFM posterior on number of cluster is more concentrated and consistent, and MFM tend to give clusters size at the same order of magnitude while DPM may lead to a few large clusters and many small clusters. See [reference] for detailed discussion.

	\subsection{Inference}

	In this paper, we choose to do inference by MCMC. Because of the Poisson likelihood, the latent state $\bm{X}^{(j)}$has no closed full conditional distribution. We can sample the posterior by particle MCMC directly, but this can be slow. However, due to the Markovian structure of the model, the conditional log-posterior is concave and its Hessian is block-tridiagonal. Thus, we can do the global Laplace approximation efficiently in $\mathcal{O}(T)$.The cluster index and number of cluster are sampled by the analog of partition-based algorithm in DPM. See details of the MCMC in appendix.

	In practice, using variational Bayes (VB) instead of MCMC may be more favorable. The PLDS can be updated by variational EM. Using the stick-breaking representation of MFM model, we can do VB easily similar to [reference David Blei \& Michael I Jordan]. However, checking by the “gold standard” MCMC before doing VB is always a good choice.

	\section{Simulations}
	\label{sim}

	\subsection{Model Global Non-linearity by Clustering}
	There were a rich research results for single PLDS model, but it provides only a global linear model to represent the data in a lower dimensional subspace, which makes the application scope limited. When the input space is nonhomogeneous, a large dimension of latent state is needed and this may lead to the overfitting and poor performance. Mixture of PLDS/ PDFM models allow us to partition the input space into clusters and can therefore capture global nonlinearity by combining local linear models.

	We first simulated three neural populations, with 10 neurons in each. We set p=1 for each cluster and the recording length is T=1000. After checking the trace plots up to 10,000 iterations, the convergence achieved after several steps. The panel A and B in figure 1 show the posterior mean firing rate and fitted latent state, averaging from iteration 500 to 1000. The latent state is transformed into the commonly used PLDS model: $\log \lambda_i = \delta_i\bm{1}_T + \bm{G}^{(j)}\bm{d}_i$, where $d_i\in \mathbb{R}^q$. In PLDS parametrization, the latent state has one more dimension, i.e. $p=q+1$. The PLDS parametrization also doesn't have the unique solution. For comparison, we let $\bm{G}^{(j)}$ has mean zeros columns and is orthogonal.

	\begin{figure}[h!]
		\centering
		\subfigure[a]{\label{fig:1a}\includegraphics[width=0.5\textwidth]{image001.jpg}}%
		\subfigure[b]{\label{fig:1b}\includegraphics[width=0.5\textwidth]{image002.jpg}}%

		\subfigure[c]{\label{fig:1c}\includegraphics[width=0.5\textwidth]{image003.jpg}}%
		\caption{Model Global Non-linearity by Clustering}
	\end{figure}

	Then we held out $1/4$ and $1/2$ data as test set in a “speckled” pattern, i.e. randomly select subset of data for each neuron as held-out dataset. Then we fit the model with and without clusters, keeping the same latent dimension, i.e. (1) 3 clusters with $p=1$ for each and (2) 1 cluster with $p=2\cdot 3 - 1 = 5$. The procedure is replicated for 100 times for two proportions, and the difference of held-out likelihood per spike between 2 fittings are shown in panel C in figure 1. The difference between (1) and (2) are always positive, and this shows doing the mixture of PDFM performs better than single PDFM. Moreover, as the proportion of training decrease (less data), the benefit of clustering becomes more significant. This suggests that doing clustering is necessary.

	\subsubsection{Clustering}

	Then use the same setting, we remove the label and use the mixture model to do the clustering by MFM. The prior of the cluster number is $K\sim Geometric(0.2)$. The panel A of figure 2 shows the trace of label $z_i$ for each neuron for the first 100 iterations.

	\begin{figure}[h!]
		\centering
		\subfigure[a]{\label{fig:2a}\includegraphics[width=0.5\textwidth]{image004.jpg}}%
		\subfigure[b]{\label{fig:2b}\includegraphics[width=0.5\textwidth]{image005.jpg}}%

		\subfigure[c]{\label{fig:2c}\includegraphics[width=0.4\textwidth]{image006.jpg}}%
		\subfigure[c]{\label{fig:2d}\includegraphics[width=0.5\textwidth]{image007.jpg}}%
		\caption{Clustering}
	\end{figure}

	In this toy simulation example, the signal is strong and the spiking amplitudes of neurons in each population are similar. This might be too simplified for real situation. In practice, the neural activity is usually sparse and can be recorded in high resolution. Moreover, the spiking amplitude for neuron in one population vary a lot. Therefore, we simulate another more realistic example. In this simulation, there are 3 clusters and the dimension of latent state is $p=1$ for each. However, there are 20 neurons with wider range but smaller value of loading $c_i$. The simulated firing rate is shown in panel B in figure 2. The trace plot of number of cluster (panel C in figure 2) shows that the model tend to further split some clusters into sub-population, based on the spiking amplitude. The similarity matrix (panel D in figure 2) of posterior, averaging from iteration 500 to 1000, shows that the model tend to split cluster 1 and 3 into two sub-clusters.

	\section{Application}
	\label{app}

	\subsection{Neuropixels data}
	\answerTODO{}: brief introduction of Neuropixels dataset.

	The Neuropixels dataset contains recording of neural activities in different brain regions, when doing the drift-grating experiment. Here, we only consider neurons with SNR > 3 and the spiking counts > 1000. Then, we use the recording activity from Lateral posterior nucleus of the thalamus (LP, 20 neurons), anteromedial visual area (VISam, 12 neurons) and ventral posteromedial nucleus of the thalamus (VPM, 14 neurons) during the spontaneous period. The bin size is 0.1s and truncate the data up to 1000 steps (100s recording) for convenience.

	Panel A in figure 3 shows the spiking counts of these 46 neurons. Here we first fit model using all data, with $p=1$ and $K\sim Geometric(0.3)$. For the number of clusters, panel B in figure 2 shows the trace plot for the first 5000 iterations and panel C shows the histogram of iteration 2500 to 5000. These plots show that these neurons are quite non-homogenous and they tend to form many sub-populations.The posterior similarity matrix, averaging from iteration 2500 to 5000 (panel D in figure 3), shows that there are several sub-populations in each anotomical cluster, and there are some tiny confusions between VISam and VPM.

	\textcolor{red}{\textbf{I tried to held-out half of the data in a speckled pattern and fit the model with clustering on and off (single population). The dimension is selected by held-out likelihood, which is p=1 for clustering on and p=2 for single population analysis. The trace plots of the held-out log-likelihood per spike (starting from iteration 2) shows that doing clustering doesn’t improve things… Maybe because “turning clustering on” introduce too much variance… The benefit of clustering will pop out if we use the deterministic algorithm, such as VB mentioned.}}

	\begin{figure}[h!]
		\centering
		\subfigure[a]{\label{fig:3a}\includegraphics[width=0.5\textwidth]{image008.jpg}}%
		\subfigure[b]{\label{fig:3b}\includegraphics[width=0.4\textwidth]{image009.jpg}}%

		\subfigure[c]{\label{fig:3c}\includegraphics[width=0.4\textwidth]{image010.jpg}}%
		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.5\textwidth]{image011.jpg}}%

		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.5\textwidth]{image012.jpg}}%
		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.5\textwidth]{image013.jpg}}%
		\caption{Neuropixel}
	\end{figure}

	\subsection{Hippocampus data}

	\answerTODO{}

	\section{Discussion}

	As mentioned above, the factor model doesn’t have unique solution. Although this issue had been thoroughly discussed in statistics [reference], it has been ignored in some neuroscience research. Since the (P)LDS related model are usually fitted by deterministic algorithms such as variants of EM, the issue is kind of hard to detect. And because of the ignorance, there are some problematic comments on linear dynamics $\bm{A}$ and $\bm{Q}$. When using the variants of factor model, e.g. (P)LDS and GPFA, we should only focus on the “shape” of latent state but not the specific values of them.

	In this paper, to ensure the unique solution, we put the diagonal constraints in $\bm{A}^{(j)}$ and $\bm{Q}^{(j)}$ are used for convenience. However, only constraining on $p(p-1)/2$ s enough [reference]. Since we the label is unknown and will be switched when doing clustering, it’s inappropriate and convenient to put constraint on $\bm{c}_i$. Therefore, we instead treat $\bm{X}^{(j)}$ as the "loading" and can put constraints on it. In traditional Gaussian factor model, there are two equivalent types of constraints: (1) diagonal constraint: $\bm{X}'^{(j)}\bm{X}^{(j)}$ is diagonal; (2) block lower triangular constraint: The first $p$ rows of $\bm{X}^{(j)}$ is lower triagular and the diagonal elements are positive. Since we assume $\bm{X}^{(j)}$ has linear progression, using the diagonal $\bm{X}'^{(j)}\bm{X}^{(j)}$ constraint is more appropriate. This constraint can be easily achieved when conducting deterministic optimization algorithms (e.g. the VB mentioned in “method-inference” section). This can also be done in MCMC, by turning off the reflection of the projection (e.g. let the diagonal elements in the projection matrix be positive). Using diagonal $\bm{X}'^{(j)}\bm{X}^{(j)}$  leads to similar results.

	Moreover, the idea of doing clustering by mixture model can be extended beyond the Poisson distribution. We can further include other information, such as, the dispersion information by assuming negative-binomial distributed or even Conway-Maxwell-Poisson distributed neural spikes [reference to our paper in CMP]. This further information can be used for more detailed clustering, by expanding the state space.

	\section*{Acknowledgment}

	\section*{References}

	{
		\small

		[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
		connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
		(eds.), {\it Advances in Neural Information Processing Systems 7},
		pp.\ 609--616. Cambridge, MA: MIT Press.

		[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
			Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
		TELOS/Springer--Verlag.

		[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
		recall at excitatory recurrent synapses and cholinergic modulation in rat
		hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
	}

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

	\appendix

	\section{Appendix}

\end{document}

### 2022-01-10 latexDraft/draft.tex
latex update

@@
 \usepackage{subfigure}
+\bibliographystyle{unsrtnat}
 
@@
 	\begin{abstract}
-		\answerTODO{}
+	Modern recording techniques allow neuroscientists to study multiple neural populations over extended time periods in a large-scale, and the relationships within and between populations are summarized by low-dimensional latent vectors.
+When the neural activities are globally nonlinear, using single population analysis is inappropriate.
+However, defining the populations is usually difficult and  wrong cluster assignments will lead to bias in latent structure inferences.
+To tackle this challenge, we develop a clustering method based mixture of Poisson dynamic factor model.
+The number of cluster is treated as a parameter in mixture of finite mixtures (MFM) model, and the posteriors are sampled by a MCMC algorithm.
+To sample the posteriors efficiently, we approximate the full conditional distribution of latent state by Gaussian and approximate the marginal likelihood by making use of the Poisson-Gamma conjugacy.
+We further apply our method to neuropixel data and hippocampus data for illustration.
 	\end{abstract}
@@
 
-	However, since the condition $T/N \rightarrow 0$ doesn’t hold, the latent state $bm{X}^{(j)}$ cannot be consistently estimated, and assuming linear dynamics of $\bm{X}^{(j)}$ resolves the problem.
+	However, since the condition $T/N \rightarrow 0$ doesn’t hold \citep{bibid}, the latent state $\bm{X}^{(j)}$ cannot be consistently estimated, and assuming linear dynamics of $\bm{X}^{(j)}$ resolves the problem.
 Note that when $p>1$, the model is not unique, since $\widetilde{\bm{X}}^{(j)} = \bm{X}^{(j)}\bm{U}$ also satisfies the equation for any orthogonal matrix $\bm{U}$ of order $p$.
@@
 	The marginal likelihood has no closed form and will be used for clustering.
-To help with fast clustering, instead of doing the Laplace approximation, we choose to make use of the Poisson-Gamma conjugacy [reference].
-This leads to the closed form approximation.
+To help with fast clustering, instead of doing the Laplace approximation, we choose to make use of the Poisson-Gamma conjugacy.
+This approximation was originally used in \citet{El-Sayyad1973} to derive approximate posterior and the same method was applied to derive other approximations in \citet{Chan2009}.
+This approximation leads to the closed form approximation.
 By the conditional independency assumption, $M_{\bm{\theta}^{(j)}}(\bm{y}_i)=\prod_{t=1}^{T}P(y_{it}|\bm{\theta}^{(j)})$.
@@
 
-	Another more general idea is to approximate the log-likelihood by second-order polynomials, with coefficients determined by Chebyshev polynomial approximation [Refer to PAL].
+	Another more general idea is to approximate the log-likelihood by second-order polynomials, with coefficients determined by Chebyshev polynomial approximation \citet{Keeley2019}.
 However, the approximation doesn’t work well in practice, especially when the neural spike counts have a wide range.
@@
 Besides the conceptual incorrectness, using DPM is not easy to integrate the field knowledge about the number of neural populations.
-Here, we choose to use the mixture of finite mixtures (MFM) model as follows
+Here, we choose to use the mixture of finite mixtures (MFM, \citet{Miller2018}) model as follows
 
@@
 However, due to the Markovian structure of the model, the conditional log-posterior is concave and its Hessian is block-tridiagonal.
-Thus, we can do the global Laplace approximation efficiently in $\mathcal{O}(T)$.The cluster index and number of cluster are sampled by the analog of partition-based algorithm in DPM.
+Thus, we can do the global Laplace approximation efficiently in $\mathcal{O}(T)$ \citep{Paninski2010}.The cluster index and number of cluster are sampled by the analog of partition-based algorithm in DPM \citep{Neal2000}.
 See details of the MCMC in appendix.
@@
 The PLDS can be updated by variational EM.
-Using the stick-breaking representation of MFM model, we can do VB easily similar to [reference David Blei \& Michael I Jordan].
+Using the stick-breaking representation of MFM model, we can do VB easily similar to \citet{Blei2006}.
 However, checking by the “gold standard” MCMC before doing VB is always a good choice.
+
+	For the dimensionality $p$ of of the latent state, we can treat it as a parameter and sample the posterior by RJMCMC as in \citet{Lopes2004} or borrow the idea of adaptive Gibbs sampling with shrinkage prior \citep{Bhattacharya2011}.
+Here, we simply pre-set the $p$ or select the optimized value by the cross-validation, which can be easily conducted when switching to the deterministic algorithm in the future.
 
@@
 
-	\answerTODO{}
+	\textcolor{red}{\textbf{What kind of story do I need to tell here?
+This should be different from the story from Neuropixel data.
+Maybe focus on the potential improvement of held-out likelihood after clustering?
+But I guess similar problem will happen.
+When fitting data with 84 neurons, the number of component jumps around 10.}}
 
@@
 Therefore, we instead treat $\bm{X}^{(j)}$ as the "loading" and can put constraints on it.
-In traditional Gaussian factor model, there are two equivalent types of constraints: (1) diagonal constraint: $\bm{X}'^{(j)}\bm{X}^{(j)}$ is diagonal; (2) block lower triangular constraint: The first $p$ rows of $\bm{X}^{(j)}$ is lower triagular and the diagonal elements are positive.
+In traditional Gaussian factor model, there are two equivalent types of constraints: (1) diagonal constraint: $\bm{X}'^{(j)}\bm{X}^{(j)}$ is diagonal; (2) block lower triangular constraint \citep{Fokoue2003}: The first $p$ rows of $\bm{X}^{(j)}$ is lower triagular and the diagonal elements are positive.
 Since we assume $\bm{X}^{(j)}$ has linear progression, using the diagonal $\bm{X}'^{(j)}\bm{X}^{(j)}$ constraint is more appropriate.
@@
 
-	\section*{Acknowledgment}
+	\section*{Acknowledgments}
 
-	\section*{References}
+	\section*{Broader Impact}
 
+	\newpage
 	{
 		\small
-
-		[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
-		connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
-		(eds.), {\it Advances in Neural Information Processing Systems 7},
-		pp.\ 609--616. Cambridge, MA: MIT Press.
-
-		[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
-			Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
-		TELOS/Springer--Verlag.
-
-		[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
-		recall at excitatory recurrent synapses and cholinergic modulation in rat
-		hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
+		\bibliography{MyCollection}
 	}

### 2022-01-17 latexDraft/appendix/appendix.tex
ok, let's fix the bug...

\documentclass[]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx} % to insert pictures
\usepackage{bm}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}

\section{$x_t^{(j)}$ and $\mu_t^{(j)}$}

Let the $t^{\text{th}}$ column observation and firing rate of the $j^{\text{th}}$ cluster be $\widetilde{\bm{y}}_t = vec\left(\{y_{it}^{(j)}|z_i = j\}\right)$ and $\widetilde{\bm{\lambda}}_t^{(j)} = vec\left(\{\lambda_{it}|z_i = j\}\right)$. The number of neurons in cluster $j$ is $n_j=\left|\{i:z_i = j\}\right|$. The corresponding loading for these $n_j$ neurons is $\bm{C}^{(j)}\in\mathbb{R}^{n_j\times p}$, such that $\log \widetilde{\bm{\lambda}}_t^{(j)} = \mu_t^{(j)}\bm{1}_{n_j} + \bm{C}^{(j)}\bm{x}_t^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)\cdot \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$. Denote $\bm{\widetilde{C}}^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)$, $\bm{\widetilde{x}}_t^{(j)} = \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$, $\bm{\widetilde{x}}^{(j)} = \left(\bm{\widetilde{x}}_1^{'(j)},\ldots,\bm{\widetilde{x}}_T^{'(j)}\right)'$, $\widetilde{\bm{A}}^{(j)} = diag(f^{(j)}, \bm{A}^{(j)})$, $\widetilde{\bm{b}}^{(j)} = \left(g^{(j)}, \bm{b}^{'(j)}\right)'$ and $\widetilde{\bm{Q}}^{(j)} = diag(\Sigma^{(j)}, \bm{Q}^{(j)})$. The full conditional distribution $P(\bm{\widetilde{x}}^{(j)}|\ldots) = P(\bm{\widetilde{x}}^{(j)}|\{\widetilde{\bm{y}}_t\}_{t=1}^T, \bm{\widetilde{C}}^{(j)}, \widetilde{\bm{A}}^{(j)}, \widetilde{\bm{b}}^{(j)}, \widetilde{\bm{Q}}^{(j)})$ is approximated by a global Laplace approximation, i.e. $P(\bm{\widetilde{x}}^{(j)}|\ldots) \approx  N_{(p+1)T}(\bm{\widetilde{x}}^{(j)} | \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}, \bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}})$, with $\bm{\mu}_{\widetilde{\bm{x}}^{(j)}} = \argmax_{\widetilde{\bm{x}}^{(j)}}P(\bm{\widetilde{x}}^{(j)}|\ldots)$ and $\bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}}) = -\left(\nabla\nabla \log P(\bm{\widetilde{x}}^{(j)}|\ldots)\right |_{\bm{\widetilde{x}}^{(j)} = \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}} )^{-1}$. The log full conditional distribution $h(\bm{\widetilde{x}}^{(j)}) = \log P(\bm{\widetilde{x}}^{(j)}|\ldots)$ is given by:
$$
h = \text{const} + \sum_{t=1}^{T} \bm{\widetilde{C}}^{'(j)}\left(\widetilde{\bm{y}}_t - \widetilde{\bm{\lambda}}_t\right) - \frac{1}{2}(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}_0(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)}) - \sum_{t=2}^{T}\frac{1}{2}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)})
$$

\textbf{It should be $\bm{b}_t^{(j)}$, but not $\bm{b}^{(j)}$} ! Fix that and then edit the writing!

\end{document}

### 2022-01-17 demo/plot/plots.m
hc

@@ -427,6 +427,20 @@ set(fitFR,'PaperUnits','inches','PaperPosition',[0 0 4 3])
 saveas(fitFR, '16_fitFR.svg')
 saveas(fitFR, '16_fitFR.png')
 
+traceLLhd_all = figure;
+tracellhd = zeros(ng,1);
+for g = 1:ng
+    tracellhd(g) =...
+        nansum(log(poisspdf(Y,fitMFRTrace(:,:,g))), 'all')/nansum(Y, 'all');
+end
+
+traceLLhd_all = figure;
+plot(tracellhd)
+ylabel('llhd/spk')
+
+traceLLhd_part = figure;
+plot(tracellhd(1:100))
+ylabel('llhd/spk')
 
 %%
 traceLLhd = figure;

### 2022-01-17 demo/real/demo_hc_all.m

@@ -1,0 +1,228 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\data'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+load('ec016.272.mat')
+
+dt = 0.05;
+T = 1000;
+Tstart = 0;
+N = size(Tlist, 2);
+Y = zeros(N, T);
+for n = 1:N
+    for k = 1:T
+        Y(n,k) = sum((Tlist{n} > (dt*(k-1)) + Tstart) &...
+            (Tlist{n} < (dt*k + Tstart)));
+    end
+end
+
+imagesc(Y)
+colorbar()
+
+%%
+lAbsGam = @(x) log(abs(gamma(x)));
+rng(2)
+p=1;
+ng = 5000;
+t_max = N;
+
+% this is the DP setting, replace to MFM later...
+% DPMM = true;
+% alpha_random = true;
+% sigma_alpha = 0.1; % scale for MH proposals in alpha move
+% alphaDP = 1;
+% log_v = (1:t_max+1)*log(alphaDP) - lAbsGam(alphaDP+N) + lAbsGam(alphaDP);
+% a = 1;
+% b = 0;
+
+
+% MFM:
+DPMM = false;
+alpha_random = false;
+MFMgamma = 1;
+% K ~ Geometric(r)
+r = 0.3;
+log_pk = @(k) log(r) + (k-1)*log(1-r);
+
+a = MFMgamma;
+b = MFMgamma;
+log_v = MFMcoeff(log_pk, MFMgamma, N, t_max + 1);
+
+logNb = log((1:N) + b);
+
+% priors...
+% theta = (d x)
+prior.theta0 = zeros(1+p,1);
+prior.Q0 = eye(1+p);
+
+prior.muC0 = zeros(p,1);
+prior.SigC0 = eye(p);
+
+prior.BA0 =[0 1]';
+prior.Lamb0 = eye(2);
+prior.Psi0 = 1e-4;
+prior.nu0 = 1+2;
+
+% pre-allocation
+t_fit = zeros(1, ng);
+Z_fit = zeros(N, ng);
+numClus_fit = zeros(t_max + 3, ng);
+
+t_fit(1) = 1;
+Z_fit(:,1) = ones(N, 1);
+numClus_fit(1,1) = N;
[…]

### 2022-01-17 latexDraft/draft.tex

@@
 
+	To show the connection between the proposed model and PLDS parametrization, we simulate the data using the PLDS model, although these two are equivalent after some algebra.
+Denote the common PLDS model as $\log \lambda_i = \delta_i\bm{1}_T + \bm{G}^{(j)}\bm{d}_i$, where $d_i\in \mathbb{R}^q$.
+In PLDS parametrization, the latent state has one more dimension, i.e.
+$p=q+1$.
+
 	We first simulated three neural populations, with 10 neurons in each.
-We set p=1 for each cluster and the recording length is T=1000.
+We set $q=2$ for each cluster and the recording length is T=1000.
 After checking the trace plots up to 10,000 iterations, the convergence achieved after several steps.
 The panel A and B in figure 1 show the posterior mean firing rate and fitted latent state, averaging from iteration 500 to 1000.
-The latent state is transformed into the commonly used PLDS model: $\log \lambda_i = \delta_i\bm{1}_T + \bm{G}^{(j)}\bm{d}_i$, where $d_i\in \mathbb{R}^q$.
-In PLDS parametrization, the latent state has one more dimension, i.e.
-$p=q+1$.
-The PLDS parametrization also doesn't have the unique solution.
-For comparison, we let $\bm{G}^{(j)}$ has mean zeros columns and is orthogonal.
+The results are transformed to PLDS parametrization for comparison.
+Notice that the PLDS parametrization also doesn't have the unique solution, we let $\bm{G}^{(j)}$ has mean zeros columns and is orthogonal.\\
+	\textbf{Still has the sign flipping issue. Fix later.}
 
@@
 		\centering
-		\subfigure[a]{\label{fig:3a}\includegraphics[width=0.5\textwidth]{image008.jpg}}%
-		\subfigure[b]{\label{fig:3b}\includegraphics[width=0.4\textwidth]{image009.jpg}}%
+		\subfigure[a]{\label{fig:3a}\includegraphics[width=0.4\textwidth]{image008.jpg}}%
+		\subfigure[b]{\label{fig:3b}\includegraphics[width=0.3\textwidth]{image009.jpg}}%
 
-		\subfigure[c]{\label{fig:3c}\includegraphics[width=0.4\textwidth]{image010.jpg}}%
-		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.5\textwidth]{image011.jpg}}%
+		\subfigure[c]{\label{fig:3c}\includegraphics[width=0.3\textwidth]{image010.jpg}}%
+		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image011.jpg}}%
 
-		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.5\textwidth]{image012.jpg}}%
-		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.5\textwidth]{image013.jpg}}%
+		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image012.jpg}}%
+		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image013.jpg}}%
+
+		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image001.png}}%
+		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image002.png}}%
 		\caption{Neuropixel}

### 2022-01-17 demo/bt/gradHessX_bt.m
do b_t

@@ -1,0 +1,45 @@
+function gradHess = gradHessX_bt(vecX, d_tmp, C_trans_tmp, x0_tmp, Q0_tmp, Q_tmp, A_tmp, b_tmp, Y)
+
+% for debugging...
+% dX = [theta_a.d;theta_a.X];
+% vecX = dX(:);
+% d_tmp = zeros(N_tmp,1);
+% C_trans_tmp = [ones(N_tmp,1) theta_a.C(obsIdx,:)];
+% x0_tmp = prior.theta0;
+% Q0_tmp = prior.Q0;
+% Q_tmp = theta_a.Q;
+% A_tmp = theta_a.A;
+% b_tmp = theta_a.b;
+% Y = Y_tmp;
+
+T = size(Y,2);
+X_all = reshape(vecX, [], T);
+lam_tmp = exp(C_trans_tmp*X_all + d_tmp);
+
+hessup = repmat((Q_tmp\A_tmp)', 1, 1, T-1);
+hessub = repmat(Q_tmp\A_tmp, 1, 1, T-1);
+hessmed = repmat(zeros(size(X_all, 1)),1,1,T);
+for t = 1:T
+    if (t==1)
+        hessmed(:,:,t) = -C_trans_tmp'*diag(lam_tmp(:,t))*C_trans_tmp-...
+            inv(Q0_tmp)- A_tmp'*(Q_tmp\A_tmp);
+    elseif (t == T)
+        hessmed(:,:,t) = -C_trans_tmp'*diag(lam_tmp(:,t))*C_trans_tmp -inv(Q_tmp);
+    else
+        hessmed(:,:,t) = -C_trans_tmp'*diag(lam_tmp(:,t))*C_trans_tmp -...
+            inv(Q_tmp) - A_tmp'*(Q_tmp\A_tmp);
+    end
+end
+
+gradHess{1} = C_trans_tmp'*(Y - lam_tmp) + [-Q0_tmp\(X_all(:,1) - x0_tmp)+...
+    A_tmp'*(Q_tmp\(X_all(:,2) - A_tmp*X_all(:,1)-b_tmp(:,1))),...
+    -Q_tmp\(X_all(:,2:(T-1)) - A_tmp*X_all(:,1:(T-2))-b_tmp(:, 1:(T-2)))+...
[…]

### 2022-01-17 demo/bt/ppasmoo_poissexp_bt.m

@@ -1,0 +1,79 @@
+function [x,W,lam,lamPred] = ppasmoo_poissexp_bt(n,C,d,x0,W0,A,b,Q)
+
+% to debug
+% n = Y_train;
+% C = C_fit(:,:,1);
+% d = d_tmp;
+% x0 = x0_fit(:,1);
+% W0 = Q0;
+% A = A_fit(:,:,1);
+% b = b_fit(:,1);
+% Q = Q_fit(:,:,1);
+
+
+warning('off');
+lastwarn('')
+nCell = size(n, 1);
+T = size(n, 2);
+
+% Preallocate
+x   = zeros(length(x0),T);
+W   = zeros([size(W0) T]);
+lamPred = zeros(nCell, T);
+
+% Initialize
+x(:,1)   = x0;
+W(:,:,1) = W0;
+lamPred(:,1)   = exp(C*x0 + d);
+
+xpred = x;
+Wpred = W;
+lam = lamPred;
+
+I = eye(size(W0));
+
+% Forward-Pass (Filtering)
+for i=2:size(n,2)
+    xpred(:,i) = A*x(:,i-1) + b(:,i-1);
+    lamPred(:,i) = exp(C*xpred(:,i) + d);
+    Wpred(:,:,i) = A*W(:,:,i-1)*A' + Q;
+    
+    INFO = zeros(size(W0));
+    SCORE = zeros(size(x0));
+    
+    for k=1:nCell
+        if(~isnan(n(k,i)))
+            INFO = INFO + C(k,:)'*(lamPred(k,i))*C(k,:);
+            SCORE = SCORE + C(k,:)'*(n(k,i) - lamPred(k, i));
+        end
+    end
+    Wpostinv = inv(Wpred(:,:,i)) + INFO;
+    W(:,:,i) = inv(Wpostinv);
+    W(:,:,i) = (W(:,:,i) + W(:,:,i)')/2;
+    
+    x(:,i)  = xpred(:,i) + W(:,:,i)*SCORE;
+    
+    lam(:,i) = exp(C*x(:,i) + d);
+    
+    [~, msgid] = lastwarn;
+    if strcmp(msgid,'MATLAB:illConditionedMatrix')
+        error('Error: singular matrix');
+    end
+end
+% lastwarn('')
+
+
+% Backward-Pass (RTS)
+for i=(T-1):-1:1
[…]

### 2022-01-17 demo/bt/raw_bt.m

@@ -1,0 +1,197 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+d(1:10) = d(1:10) + 1;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-1 2];
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p);
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
[…]

### 2022-01-17 demo/bt/sample_prior_bt.m

@@ -1,0 +1,66 @@
+function theta =...
+    sample_prior_new(prior, N, T, p, sampleDynamics, maxCount)
+
+flag = 1;
+while flag
+    % lodaing: C
+    theta.C = randn(N, p);
+    theta.muC = theta.C;
+    
+    % linear dynamics: b, A & Q
+    theta.A = eye(p+1);
+    theta.b = zeros(p+1,T-1);
+    theta.Q = eye(p+1)*prior.Psi0;
+    
+    % mean & latent: d, X
+    dX = ones(1+p,T)*Inf;
+    dX(:,1) = mvnrnd(prior.theta0, prior.Q0);
+    
+    if sampleDynamics
+        
+        for k = 1:(p+1)
+            theta.Q(k,k)= iwishrnd(prior.Psi0, prior.nu0);
+            theta.A(k,k) = mvnrnd(prior.A0, kron(theta.Q(k,k), inv(prior.Lamb0)))';
+            for t = 1:(T-1)
+                theta.b(k, t) = normrnd(prior.b0,prior.sig2b0);
+            end
+        end
+    end
+    
+    for t= 2:T
+        dX(:, t) = mvnrnd(theta.b(:,t-1) + theta.A*dX(:, t-1), theta.Q);
+    end
+    
+    theta.d = dX(1,:);
+    theta.X = dX(2:end,:);
+    
+    % test
+%     theta.X(abs(theta.X) > 1) = sign(theta.X(abs(theta.X) > 1));
+    
+    
+%     dX = zeros(p+1, T);
+%     for i=1:(p+1)
+%         k = ceil(rand()*20)+10;
+%         if i == 1
+%             pt = randn(k,1);
+%         else
+%             pt = randn(k,1)*.25;
+%         end
+%         dX(i,:) = interp1(linspace(0,1,k),pt,linspace(0,1,T),'spline');
+%     end
+% 
+%     theta.d = dX(1,:);
+%     theta.X = dX(2:end,:);
+    
+    
+    
+    
+    lamTmp = exp([ones(N,1) theta.C]*[theta.d ;theta.X]);
+    if(sum(isinf(lamTmp(:))) ==0 && max(lamTmp(:)) <= maxCount)
+       flag = 0; 
+    end
+    
+end
+
+
+end

### 2022-01-17 demo/bt/unLabeled/demo_new_gamma.m

@@ -1,0 +1,328 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+d(1:10) = d(1:10) + 1;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-1 2];
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p);
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
[…]

### 2022-01-17 demo/bt/update_cluster_bt.m

@@ -1,0 +1,161 @@
+function [theta_b, epsilonOut, log_pdf] =...
+    update_cluster_new(Y_tmp,theta_a,theta_b,...
+    prior, N, T, p, obsIdx, active, density, OPTDC_tmp)
+
+% for debug
+% c=j;
+% obsIdx = obsIdx;
+% Y_tmp = Y(obsIdx,:);
+% theta_a = THETA{g-1}(c);
+% theta_b = THETA{g}(c);
+% active = true;
+% density = false;
+% OPTDC_tmp = OPTDC(obsIdx);
+
+N_tmp = size(Y_tmp, 1);
+log_pdf = NaN;
+
+% (1) update X_fit
+
+gradHess = @(vecdX) gradHessX_bt(vecdX, zeros(N_tmp,1),...
+    [ones(N_tmp,1) theta_a.C(obsIdx,:)],...
+    prior.theta0, prior.Q0, theta_a.Q, theta_a.A, theta_a.b, Y_tmp);
+
+dX = [theta_a.d;theta_a.X];
+[mudXvec,~,hess_tmp,~, eoi] = newtonGH(gradHess,dX(:),1e-6,1000);
+
+if((sum(isnan(mudXvec)) ~= 0) || eoi == 1000)
+    disp('use adaptive smoother initial')
+    try
+        dX_tmp = ppasmoo_poissexp_bt(Y_tmp,[ones(N_tmp,1) theta_a.C(obsIdx,:)],...
+            zeros(N_tmp,1),prior.theta0,prior.Q0,theta_a.A,theta_a.b,theta_a.Q);
+    catch
+        dX_tmp = dX*0;
+    end
+    
+    [mudXvec,~,hess_tmp,~] = newtonGH(gradHess,dX_tmp(:),1e-6,1000);
+end
+
+% use Cholesky decomposition to sample efficiently
+if active
+    try
+        RdX = chol(-hess_tmp,'lower'); % sparse
+        zdX = randn(length(mudXvec), 1) + RdX'*mudXvec;
+        dXsamp = RdX'\zdX;
+    catch
+        dXsamp = mudXvec;
+    end
+    dXsamp_trans = reshape(dXsamp,[], T);
+    
+    theta_b.d = dXsamp_trans(1,:);
+    theta_b.X = dXsamp_trans(2:end,:);
+end
+
+if density
+    log_pdf = 0;
+%     log_pdf = log_pdf + mvnlpdf(theta_b.Xori(:), muXvec, -hess_tmp);
+end
+
+
[…]

### 2022-01-18 latexDraft/draft.tex
add appendix

@@
 \bibliographystyle{unsrtnat}
+
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 	Further, let $z_{i} = j$ denote the cluster indicator of neuron $i$.
-To facilitate clustering, we re-parametrize the regular Poisson linear dynamical system (PLDS) model to separate the mean log-firing-rate out.
+To facilitate clustering, we re-parametrize the regular Poisson linear dynamical system (PLDS) model \citep{Macke2011} to separate the mean log-firing-rate out.
 Assume neurons are independently Poisson distributed, conditional on the
@@
 		\mu_1^{(j)} &\sim N(\mu_0, \Sigma_0)\\
-		\mu_{t+1}^{(j)} &\sim N(f^{(j)}\mu_t^{(j)} + g^{(j)}, \Sigma^{(j)})\\
+		\mu_{t+1}^{(j)} &\sim N(f^{(j)}\mu_t^{(j)} + g^{(j)}, \sigma^{2(j)})\\
 		\bm{x}_1^{(j)} &\sim N(\bm{x}_0, \bm{Q}_0)\\
@@
 	\end{align*}
+
+	The bias term ($\bm{b}^{(j)}$) in the latent state is constant across time, unlike ind PLDS, which is  $\bm{b}_t^{(j)}$.
+Allowing the time-varying bias term will lead to infinite solutions of linear dynamics.
+If the $\bm{A}^{(j)}$ is one solution, then $\bm{A}^{*(j)} = \bm{A}^{(j)} + \bm{M}$ and $\bm{b}_t^{*(j)} = \bm{b}_t^{(j)} - \bm{M}\bm{x}^{(j)}_t$ will be another set of solution, for any $\bm{M} \in \mathbb{R}^{p\times p}$.
+The same reason for using time constant $g^{(j)}$.
+
 	If we denote $\bm{\lambda}_{i} =  (\lambda_{i1},\ldots,\lambda_{iT})'$, $\bm{\mu}^{(j)} = (\mu^{(j)}_1,\ldots,\mu^{(j)}_T)'$ and $\bm{X}^{(j)} = (\bm{x}^{(j)}_1,\ldots,\bm{x}^{(j)}_T)'$, the model can be equivalently written as regular Poisson factor model as
@@
 
-	However, since the condition $T/N \rightarrow 0$ doesn’t hold \citep{bibid}, the latent state $\bm{X}^{(j)}$ cannot be consistently estimated, and assuming linear dynamics of $\bm{X}^{(j)}$ resolves the problem.
-Note that when $p>1$, the model is not unique, since $\widetilde{\bm{X}}^{(j)} = \bm{X}^{(j)}\bm{U}$ also satisfies the equation for any orthogonal matrix $\bm{U}$ of order $p$.
+	However, since the condition $T/N \rightarrow 0$ doesn’t hold \citep{Johnstone2009}, the latent state $\bm{X}^{(j)}$ cannot be consistently estimated, and assuming linear dynamics of $\bm{X}^{(j)}$ resolves the problem.
+Note that when $p>1$, the model is not unique, since $\bm{X}^{*(j)} = \bm{X}^{(j)}\bm{U}$ also satisfies the equation for any orthogonal matrix $\bm{U}$ of order $p$.
 To ensure the model indefinability, we simply assume $\bm{A}^{(j)}$ and $\bm{Q}^{(j)}$
@@
 See more detailed discussions of the constraints in discussion.
-Overall, denote the cluster-related parameters of cluster $j$ as $\bm{\theta}^{(j)}= \{\bm{\mu}^{(j)}, \bm{X}^{(j)}, f^{(j)}, g^{(j)}, \Sigma^{(j)},\{\bm{A}^{(j)}, \{\bm{b}^{(j)}, \{\bm{Q}^{(j)}\}$ and the spike counts of neuron i is generated by Poisson dynamic factor model (PDFM) as $\bm{Y}_i\sim PDFM(\bm{\theta}^{(z_i)})$, with the prior of $\bm{\theta}^{(j)}$ as $\bm{H}$.
-The priors for $\{f^{(j)}, g^{(j)}, \Sigma^{(j)},\{\bm{A}^{(j)}, \{\bm{b}^{(j)}, \{\bm{Q}^{(j)}\}$ are regular  normal and inverse-gamma distribution (or multivariate normal and inverse-Wishart when using other non-diagonal constraints).
+Overall, denote the cluster-related parameters of cluster $j$ as $\bm{\theta}^{(j)}= \{\bm{\mu}^{(j)}, \bm{X}^{(j)}, f^{(j)}, g^{(j)}, \sigma^{2(j)},\bm{A}^{(j)}, \bm{b}^{(j)}, \bm{Q}^{(j)}\}$ and the spike counts of neuron i is generated by Poisson dynamic factor model (PDFM) as $\bm{Y}_i\sim PDFM(\bm{\theta}^{(z_i)})$, with the prior of $\bm{\theta}^{(j)}$ as $\bm{H}$.
+The priors for $\{f^{(j)}, g^{(j)}, \sigma^{2(j)},\bm{A}^{(j)}, \bm{b}^{(j)}, \bm{Q}^{(j)}\}$ are regular  normal and inverse-gamma distribution (or multivariate normal and inverse-Wishart when using other non-diagonal constraints).
 
@@
 
+	\subsection{Full Conditional Distributions}
+
+	\underline{\textit{$x_t^{(j)}$ and $\mu_t^{(j)}$}}:
+	Let the $t^{\text{th}}$ column observation and firing rate of the $j^{\text{th}}$ cluster be $\widetilde{\bm{y}}_t = vec\left(\{y_{it}^{(j)}|z_i = j\}\right)$ and $\widetilde{\bm{\lambda}}_t^{(j)} = vec\left(\{\lambda_{it}|z_i = j\}\right)$.
+The number of neurons in cluster $j$ is $n_j=\left|\{i:z_i = j\}\right|$.
+The corresponding loading for these $n_j$ neurons is $\bm{C}^{(j)}\in\mathbb{R}^{n_j\times p}$, such that $\log \widetilde{\bm{\lambda}}_t^{(j)} = \mu_t^{(j)}\bm{1}_{n_j} + \bm{C}^{(j)}\bm{x}_t^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)\cdot \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$.
+Denote $\bm{\widetilde{C}}^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)$, $\bm{\widetilde{x}}_t^{(j)} = \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$, $\bm{\widetilde{x}}^{(j)} = \left(\bm{\widetilde{x}}_1^{'(j)},\ldots,\bm{\widetilde{x}}_T^{'(j)}\right)'$, $\widetilde{\bm{A}}^{(j)} = diag(f^{(j)}, \bm{A}^{(j)})$, $\widetilde{\bm{b}}^{(j)} = \left(g^{(j)}, \bm{b}^{'(j)}\right)'$ and $\widetilde{\bm{Q}}^{(j)} = diag(\sigma^{2(j)}, \bm{Q}^{(j)})$.
+The full conditional distribution $P(\bm{\widetilde{x}}^{(j)}|\ldots) = P(\bm{\widetilde{x}}^{(j)}|\{\widetilde{\bm{y}}_t\}_{t=1}^T, \bm{\widetilde{C}}^{(j)}, \widetilde{\bm{A}}^{(j)}, \widetilde{\bm{b}}^{(j)}, \widetilde{\bm{Q}}^{(j)})$ is approximated by a global Laplace approximation, i.e.
+$P(\bm{\widetilde{x}}^{(j)}|\ldots) \approx  N_{(p+1)T}(\bm{\widetilde{x}}^{(j)} | \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}, \bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}})$, with $\bm{\mu}_{\widetilde{\bm{x}}^{(j)}} = \argmax_{\widetilde{\bm{x}}^{(j)}}P(\bm{\widetilde{x}}^{(j)}|\ldots)$ and $\bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}}) = -\left(\nabla\nabla \log P(\bm{\widetilde{x}}^{(j)}|\ldots)\right |_{\bm{\widetilde{x}}^{(j)} = \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}} )^{-1}$.
+The log full conditional distribution $h(\bm{\widetilde{x}}^{(j)}) = \log P(\bm{\widetilde{x}}^{(j)}|\ldots)$ is given by:
+	\begin{align*}
+		h &= \text{const} + \sum_{t=1}^{T} \widetilde{\bm{y}}'_t\bm{\widetilde{C}}^{(j)}\bm{\widetilde{x}}_t^{(j)} - \widetilde{\bm{\lambda}}_t
+		- \frac{1}{2}(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}_0(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})\\ 
+		&-\sum_{t=2}^{T}\frac{1}{2}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)} - \widetilde{\bm{b}}^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)} - \widetilde{\bm{b}}^{(j)})
+	\end{align*}
+
+	The $h(\bm{\widetilde{x}}^{(j)})$ is concave and hence unimodal, and the Markovian structure of the latent dynamics, and hence the tri-block-diagonal Hessian, makes it possible to compute a Newton update in $\mathcal{O}(T)$ \citep{Paninski2010}.
+
+	To facilitate convergence, we use a smoothing estimate with local Gaussian approximation as a "warm start".
+The forward filtering for a dynamic Poisson model has been previously described in \cite{Eden2004}.
+Because the approximated filtering estimates are Gaussian distributed, we can further find the smoothing estimates using a backward pass as in \cite{RAUCH1965}.
+
+	\underline{\textit{$\bm{c}_i$}}: 
+	When writing the observation mapping in matrix form,
+	$\log\bm{\lambda}_i = \bm{\mu}^{(j)}+ \bm{X}^{(j)}\bm{c}_i$, the update of $\bm{c}_i$ reduces to a regular Poisson regression problem, given $\bm{\mu}^{(j)}$ and $\bm{X}^{(j)}$ are known.
+Here, we sample the posterior by no a No U-Turn Sampler (NUTS, \cite{Hoffman2011}).
+
+	\underline{\textit{Linear dynamics of latent state}}: The parameters for linear dynamics are $f^{(j)}$, $g^{(j)}$, $\sigma^{2(j)}$, $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$.
+To make the model identifiable, we simply assume $\bm{A}^{(j)} = diag(a_1^{(j)},\ldots, a_p^{(j)})$ and $\bm{Q}^{(j)} = diag(q_1^{(j)},\ldots, q_p^{(j)})$.
+Therefore, we can update $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$ dimension-by-dimension, as the update in $f^{(j)}$, $g^{(j)}$ and $\sigma^{2(j)}$.
+Use the priors $\sigma^{2(j)} \sim IG\left(\nu_0/2, \nu_0\sigma_0^2/2\right)$ and  $\left( g^{(j)}, f^{(j)}\right)'\sim N(\bm{\mu}_0, \sigma^{2(j)}\bm{\Lambda}_0^{-1})$.
+The prior $\mu_0$ is specified as $(0,1)'$.
+Denote $\bm{y}_{\mu^{(j)}} = \left(\mu_2^{(j)}, \ldots, \mu_T^{(j)}\right)'$ and $\bm{X}_{\mu^{(j)}} = \left(\bm{1}_{T-1}, \bm{\mu}_{1:(T-1)}^{(j)}\right)$, with $\bm{\mu}_{1:(T-1)}^{(j)} = \left(\mu_1^{(j)}, \ldots, \mu_{T-1}^{(j)}\right)'$.
+The full conditional distributions are $\sigma^{2(j)}|\ldots \sim IG\left(\frac{\nu_0 + T-1}{2}, \frac{\nu_0\sigma^2_0 + \bm{y}'_{\mu^{(j)}}\bm{y}_{\mu^{(j)}} + \bm{\mu}'_0\bm{\Lambda}_0\bm{\mu}_0 - \bm{\mu}'_n\bm{\Lambda}_n\bm{\mu}_n }{2}\right)$ and $\left( g^{(j)}, f^{(j)}\right)'|\ldots \sim N(\bm{\mu}_n, \sigma^{2(j)}\bm{\Lambda}_n^{-1})$, with $\bm{\Lambda}_n = \bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}} + \bm{\Lambda}_0$ and $\bm{\mu}_n = \bm{\Lambda}_n^{-1}\left(\bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}}\left(\bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}}\right)^{-1}\bm{X}'_{\mu^{(j)}}\bm{y}_{\mu^{(j)}} + \bm{\Lambda}_0\bm{\mu}_0\right)$.
+The updates of $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$ are similar, with diagonal assumption.
+
 \end{document}

### 2022-01-20 demo/plot/plots.m
transfer for plotting

@@ -195,7 +195,6 @@ set(FR,'PaperUnits','inches','PaperPosition',[0 0 6 3])
 saveas(FR, '4_FR.svg')
 saveas(FR, '4_FR.png')
 
-%%
 %% cluster plot...
 
 sim10 = figure;
@@ -227,8 +226,8 @@ set(gca,'FontSize',10, 'LineWidth', 1.5,'TickDir','out')
 box off
 
 set(sim10,'PaperUnits','inches','PaperPosition',[0 0 5 3])
-saveas(sim10, '3_sim10.svg')
-saveas(sim10, '3_sim10.png')
+saveas(sim10, '5_sim10.svg')
+saveas(sim10, '5_sim10.png')
 
 
 sim30 = figure;
@@ -260,14 +259,14 @@ set(gca,'FontSize',10, 'LineWidth', 1.5,'TickDir','out')
 box off
 
 set(sim30,'PaperUnits','inches','PaperPosition',[0 0 5 3])
-saveas(sim30, '4_sim30.svg')
-saveas(sim30, '4_sim30.png')
+saveas(sim30, '6_sim30.svg')
+saveas(sim30, '6_sim30.png')
 
 clusLast = figure;
 clusterPlot(Y, Z_fit(:,ng)')
 set(clusLast,'PaperUnits','inches','PaperPosition',[0 0 5 3])
-saveas(clusLast, '5_sim30Last.svg')
-saveas(clusLast, '5_sim30Last.png')
+saveas(clusLast, '7_sim30Last.svg')
+saveas(clusLast, '7_sim30Last.png')
 
 
 
@@ -284,8 +283,8 @@ ylabel('neuron')
 xlabel('T')
 
 set(FR20,'PaperUnits','inches','PaperPosition',[0 0 4 3])
-saveas(FR20, '6_FR20.svg')
-saveas(FR20, '6_FR20.png')
+saveas(FR20, '8_FR20.svg')
+saveas(FR20, '8_FR20.png')
 
 
 sim20 = figure;
@@ -302,8 +301,8 @@ colormap(flipud(hot))
 colorbar()
 
 set(sim20,'PaperUnits','inches','PaperPosition',[0 0 4 3])
-saveas(sim20, '7_sim20.svg')
-saveas(sim20, '7_sim20.png')
+saveas(sim20, '9_sim20.svg')
+saveas(sim20, '9_sim20.png')
 
 hist20 = figure;
[…]

### 2022-01-20 demo/real/demo_pixel_all.m

@@ -43,10 +43,10 @@ lab_all_str = string(tab.ecephys_structure_acronym);
 
 % spontaneous
 % Tstart = 29.8301073815904; Tend = 89.89682738;
-Tstart =  1001.89177167499; Tend = 1290.88309738159;
+% Tstart =  1001.89177167499; Tend = 1290.88309738159;
 
 % drifting_gratings 
-% Tstart = 1591.13385738159; Tend = 2190.63454310612;
+Tstart = 1591.13385738159; Tend = 2190.63454310612;
 
 dt = 0.1;
 T = min(floor((Tend-Tstart)/dt), 1000);

### 2022-01-20 latexDraft/draft.tex
draft 1 update

@@
 To sample the posteriors efficiently, we approximate the full conditional distribution of latent state by Gaussian and approximate the marginal likelihood by making use of the Poisson-Gamma conjugacy.
-We further apply our method to neuropixel data and hippocampus data for illustration.
+We further apply our method to neuropixel data (and hippocampus data) for illustration.
 	\end{abstract}
@@
 	The bias term ($\bm{b}^{(j)}$) in the latent state is constant across time, unlike ind PLDS, which is  $\bm{b}_t^{(j)}$.
-Allowing the time-varying bias term will lead to infinite solutions of linear dynamics.
+When the bias term changes across the time, the update order of $\bm{A}^{(j)}$ and $\bm{b}_t^{(j)}$ will influence the results (usually, the $\bm{A}^{(j)}$ is updated first).
+When ignoring the updating order, the linear dynamics will have infinite solutions.
 If the $\bm{A}^{(j)}$ is one solution, then $\bm{A}^{*(j)} = \bm{A}^{(j)} + \bm{M}$ and $\bm{b}_t^{*(j)} = \bm{b}_t^{(j)} - \bm{M}\bm{x}^{(j)}_t$ will be another set of solution, for any $\bm{M} \in \mathbb{R}^{p\times p}$.
-The same reason for using time constant $g^{(j)}$.
+This causes some inconvenience for inference.
+The same reason for using the time-constant $g^{(j)}$.
 
@@
 	\begin{align*}
-		K &\sim p_k &&\text{where}\, p_k \, \text{is a p.m.f. on} \{1,2,\ldots\}\\
+		K &\sim p_K &&\text{where}\, p_K \, \text{is a p.m.f. on} \{1,2,\ldots\}\\
 		\bm{\pi}=(\pi_1,\ldots,\pi_k) &\sim Dir_k(\gamma, \ldots,\gamma) &&\text{given} K=k\\
@@
 Moreover, compared to DPM, MFM has some better properties for clustering, for example, MFM posterior on number of cluster is more concentrated and consistent, and MFM tend to give clusters size at the same order of magnitude while DPM may lead to a few large clusters and many small clusters.
-See [reference] for detailed discussion.
+See \cite{Miller2018} for detailed discussion.
 
@@
 Thus, we can do the global Laplace approximation efficiently in $\mathcal{O}(T)$ \citep{Paninski2010}.The cluster index and number of cluster are sampled by the analog of partition-based algorithm in DPM \citep{Neal2000}.
-See details of the MCMC in appendix.
+See details of the MCMC update in the Appendix.
 
@@
 We set $q=2$ for each cluster and the recording length is T=1000.
-After checking the trace plots up to 10,000 iterations, the convergence achieved after several steps.
-The panel A and B in figure 1 show the posterior mean firing rate and fitted latent state, averaging from iteration 500 to 1000.
-The results are transformed to PLDS parametrization for comparison.
-Notice that the PLDS parametrization also doesn't have the unique solution, we let $\bm{G}^{(j)}$ has mean zeros columns and is orthogonal.\\
-	\textbf{Still has the sign flipping issue. Fix later.}
+The $\delta_i = 1$ for $i=1,\ldots, 10$ and is $0$ for the remaining neurons.
+The $\bm{G}^{(j)}$ is $0.1(1,\ldots,10)'\otimes\bm{1}'_2 + \bm{1}_{10}\otimes(-1, 2)$ to set up a gradient for the loading in each cluster and to differ the contribution of two latent vectors.
+We allow $\bm{A}^{(j)}$ be non-diagonal and possible interactions between clusters, which is used in single population PLDS model.
+The diagonal element is generated from $N(0.92, 0.1^2)$.
+The bias terms are $\bm{b}^{(j)}$ is $10^{-3}\bm{1}_2$ for the first 2 clusters and $10^{-4}\bm{1}_2$ for the last cluster.
+The $\bm{Q}^{(j)}$ are $10^{-3}\bm{I}_2$, $10^{-4}\bm{I}_2$ and $10^{-5}\bm{I}_2$.
 
-	\begin{figure}[h!]
-		\centering
-		\subfigure[a]{\label{fig:1a}\includegraphics[width=0.5\textwidth]{image001.jpg}}%
-		\subfigure[b]{\label{fig:1b}\includegraphics[width=0.5\textwidth]{image002.jpg}}%
-
-		\subfigure[c]{\label{fig:1c}\includegraphics[width=0.5\textwidth]{image003.jpg}}%
-		\caption{Model Global Non-linearity by Clustering}
-	\end{figure}
+	The Figure \ref{fig:1a} shows the trace plot for the log-likelihood per spike, up to 10,000 iterations.
+The plot shows (Figure \ref{fig:1b}) that the convergence achieved after several steps.
+Therefore, we show the results by averaging iteration 500 to 1000.
+Figure \ref{fig:1c} shows the posterior mean of the mean firing rate, while Figure \ref{fig:1d} shows the fitted latent state.
+The latent vectors are transformed to PLDS parametrization for comparison.
+Notice that the PLDS parametrization also doesn't have the unique solution, we let $\bm{G}^{(j)}$ has mean zeros columns and is orthogonal.
 
@@
 randomly select subset of data for each neuron as held-out dataset.
-Then we fit the model with and without clusters, keeping the same latent dimension, i.e.
+The "speckled" cross-validation was previously used in \cite{Williams2020}.
+We then fit the model with and without clusters, keeping the same latent dimension, i.e.
 (1) 3 clusters with $p=1$ for each and (2) 1 cluster with $p=2\cdot 3 - 1 = 5$.
-The procedure is replicated for 100 times for two proportions, and the difference of held-out likelihood per spike between 2 fittings are shown in panel C in figure 1.
+The procedure is replicated for 100 times for two proportions, and the difference of held-out likelihood per spike between 2 fittings are shown in Figure \ref{fig:1e}.
 The difference between (1) and (2) are always positive, and this shows doing the mixture of PDFM performs better than single PDFM.
@@
 
+	\begin{figure}[h!]
+		\centering
+		\subfigure[Trace plot of  log-likelihood per spike]{\label{fig:1a}\includegraphics[width=0.5\textwidth]{2_llhd1.png}}%
+		\subfigure[Trace plot of  log-likelihood per spike, truncated (1000)]{\label{fig:1b}\includegraphics[width=0.5\textwidth]{3_llhd2.png}}%
+
+		\subfigure[true vs. fitted firing rate]{\label{fig:1c}\includegraphics[width=0.5\textwidth]{image001.jpg}}%
+		\subfigure[true vs. fitted latent vectors]{\label{fig:1d}\includegraphics[width=0.5\textwidth]{LF.png}}%
+
+		\subfigure[cross-validation: clustered vs. single population]{\label{fig:1e}\includegraphics[width=0.5\textwidth]{image003.jpg}}%
+		\subfigure[trace plot for non-labeled case]{\label{fig:1f}\includegraphics[width=0.5\textwidth]{image004.jpg}}%
+		\caption{Simple case}
+	\end{figure}
+
 	\subsubsection{Clustering}
-
 	Then use the same setting, we remove the label and use the mixture model to do the clustering by MFM.
 The prior of the cluster number is $K\sim Geometric(0.2)$.
-The panel A of figure 2 shows the trace of label $z_i$ for each neuron for the first 100 iterations.
+Figure \ref{fig:1f} shows the trace of label $z_i$ for each neuron for the first 100 iterations.
 
@@
 		\centering
-		\subfigure[a]{\label{fig:2a}\includegraphics[width=0.5\textwidth]{image004.jpg}}%
-		\subfigure[b]{\label{fig:2b}\includegraphics[width=0.5\textwidth]{image005.jpg}}%
-
-		\subfigure[c]{\label{fig:2c}\includegraphics[width=0.4\textwidth]{image006.jpg}}%
-		\subfigure[c]{\label{fig:2d}\includegraphics[width=0.5\textwidth]{image007.jpg}}%
-		\caption{Clustering}
+		\subfigure[Firing rate]{\label{fig:2a}\includegraphics[width=0.4\textwidth]{image005.jpg}}%
+		\subfigure[trace plot for cluster number]{\label{fig:2b}\includegraphics[width=0.4\textwidth]{image006.jpg}}%
+		\subfigure[similarity matrix]{\label{fig:2c}\includegraphics[width=0.4\textwidth]{image007.jpg}}%
+		\caption{More complicated case: more variety of amplitude \& weaker signal}
 	\end{figure}
@@
 In practice, the neural activity is usually sparse and can be recorded in high resolution.
-Moreover, the spiking amplitude for neuron in one population vary a lot.
-Therefore, we simulate another more realistic example.
+Moreover, the spiking amplitudes for neuron in one nominated population vary a lot, and hence may form a few subpopulations.
+Therefore, we then simulate another more realistic example.
 In this simulation, there are 3 clusters and the dimension of latent state is $p=1$ for each.
-However, there are 20 neurons with wider range but smaller value of loading $c_i$.
-The simulated firing rate is shown in panel B in figure 2.
-The trace plot of number of cluster (panel C in figure 2) shows that the model tend to further split some clusters into sub-population, based on the spiking amplitude.
-The similarity matrix (panel D in figure 2) of posterior, averaging from iteration 500 to 1000, shows that the model tend to split cluster 1 and 3 into two sub-clusters.
+However, there are 20 neurons with wider range but smaller value of loading $c_i$, corresponding to potential subpopulations and weak signal.
+The simulated firing rate is shown in Figure \ref{fig:2a}.
+The trace plot of number of cluster (Figure \ref{fig:2b}) shows that the model tend to further split some clusters into sub-population, based on the spiking amplitude.
+This is more clearly shown in the posterior similarity matrix ((Figure \ref{fig:2c}), averaging from iteration 500 to 1000.
+In similarity matrix, the entry $(i,j)$ is the posterior probability that data points $i$ and $j$ belong to the same cluster.
+The rows and columns are ordered according to the ground truth.
+The similarity matrix shows that the model tend to split cluster 1 and 3 into two sub-clusters.
 
@@
 
-	The Neuropixels dataset contains recording of neural activities in different brain regions, when doing the drift-grating experiment.
+	The Neuropixels dataset contains recording of neural activities in different brain regions of mouse.
 Here, we only consider neurons with SNR > 3 and the spiking counts > 1000.
-Then, we use the recording activity from Lateral posterior nucleus of the thalamus (LP, 20 neurons), anteromedial visual area (VISam, 12 neurons) and ventral posteromedial nucleus of the thalamus (VPM, 14 neurons) during the spontaneous period.
+Then, we first use the recording activity from Lateral posterior nucleus of the thalamus (LP, 20 neurons), anteromedial visual area (VISam, 12 neurons) and ventral posteromedial nucleus of the thalamus (VPM, 14 neurons) during the spontaneous period.
 The bin size is 0.1s and truncate the data up to 1000 steps (100s recording) for convenience.
 
-	Panel A in figure 3 shows the spiking counts of these 46 neurons.
+	Figure \ref{fig:3a} shows the spiking counts of these 46 neurons.
 Here we first fit model using all data, with $p=1$ and $K\sim Geometric(0.3)$.
-For the number of clusters, panel B in figure 2 shows the trace plot for the first 5000 iterations and panel C shows the histogram of iteration 2500 to 5000.
-These plots show that these neurons are quite non-homogenous and they tend to form many sub-populations.The posterior similarity matrix, averaging from iteration 2500 to 5000 (panel D in figure 3), shows that there are several sub-populations in each anotomical cluster, and there are some tiny confusions between VISam and VPM.
+For the number of clusters, Figure \ref{fig:3b} shows the trace plot for the first 5000 iterations.
+All the following results are from averaging samples from iteration 2500 to 5000.
+The Figure \ref{fig:3c} shows the histogram of the posterior samples.
+These plots show that these neurons are quite non-homogeneous and they tend to form many sub-populations.
+These 3 nominated populations by anatomy tend to form around 10 (sub)-clusters.
+
+	The Figure \ref{fig:3d} and Figure \ref{fig:3e} show the trace of log-likelihood per spike.
+The fitted mean firing rate is shown in Figure \ref{fig:3f}.
+When investigating the similarity matrix in Figure \ref{fig:3g} (sorted according to the anatomical population, such that 1-20 are LP, 21-32 are VISam and 33-46 are VPM), there are several subpopulations in each anatomical cluster.
+Generally, the clustering results are consistent with the anatomical assignment, although there are some tiny confusions between VISam and VPM.
+
+	However, this is not the case when the mouse is in the experiment.
+We choose the same set of neurons but use recordings when the mouse are watching drift gratings.
+The Figure \ref{fig:3h} shows the trace plot of cluster number.
+The Figure \ref{fig:3i} and Figure \ref{fig:3j} show the fitted firing rate and similarity matrix, again averaging from iteration 2500 to 5000.
+
+	When comparing the observed spiking activities, it's fairly hard to see differences when the mouse is exposed to the stimuli and not.
+However, the posterior similarity matrix in the experiment shows that there's more confusion between LP and VISam, compared to the spontaneous response.
+This suggests that the functional connection pattern between neurons in different areas when the stimuli are changed.
 
@@
 The dimension is selected by held-out likelihood, which is p=1 for clustering on and p=2 for single population analysis.
-The trace plots of the held-out log-likelihood per spike (starting from iteration 2) shows that doing clustering doesn’t improve things… Maybe because “turning clustering on” introduce too much variance… The benefit of clustering will pop out if we use the deterministic algorithm, such as VB mentioned.}}
+The trace plots of the held-out log-likelihood per spike (starting from iteration 2) shows that doing clustering doesn’t improve things (although they are very close)… Maybe because “turning clustering on” introduce too much variance… The benefit of clustering will pop out if we use the deterministic algorithm, such as VB mentioned.}}
 
@@
 		\centering
-		\subfigure[a]{\label{fig:3a}\includegraphics[width=0.4\textwidth]{image008.jpg}}%
-		\subfigure[b]{\label{fig:3b}\includegraphics[width=0.3\textwidth]{image009.jpg}}%
+		\subfigure[spontaneous: spike counts]{\label{fig:3a}\includegraphics[width=0.5\textwidth]{image008.jpg}}%
+		\subfigure[spontaneous: cluster number trace]{\label{fig:3b}\includegraphics[width=0.35\textwidth]{image009.jpg}}%
+		\subfigure[spontaneous: cluster number histogram]{\label{fig:3c}\includegraphics[width=0.35\textwidth]{image010.jpg}}%
 
-		\subfigure[c]{\label{fig:3c}\includegraphics[width=0.3\textwidth]{image010.jpg}}%
-		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image011.jpg}}%
+		\subfigure[spontaneous: llhd/spk trace]{\label{fig:3d}\includegraphics[width=0.3\textwidth]{image001.png}}%
+		\subfigure[spontaneous: llhd/spk trace, truncated]{\label{fig:3e}\includegraphics[width=0.3\textwidth]{image002.png}}%
+		\subfigure[spontaneous: true vs. fitted firing rate]{\label{fig:3f}\includegraphics[width=0.3\textwidth]{image011.jpg}}%
+		\subfigure[spontaneous: similarity matrix]{\label{fig:3g}\includegraphics[width=0.3\textwidth]{image012.jpg}}%
 
-		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image012.jpg}}%
-		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image013.jpg}}%
-
-		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image001.png}}%
-		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image002.png}}%
+		\subfigure[drift-grating: cluster number trace]{\label{fig:3h}\includegraphics[width=0.4\textwidth]{12_numTrace.png}}%
+		\subfigure[drift-grating: true vs. fitted firing rate]{\label{fig:3i}\includegraphics[width=0.4\textwidth]{16_fitFR.png}}%
+		\subfigure[drift-grating: similarity matrix]{\label{fig:3j}\includegraphics[width=0.4\textwidth]{14_simPixel.png}}%
 		\caption{Neuropixel}
@@
 
-	\textcolor{red}{\textbf{What kind of story do I need to tell here?
-This should be different from the story from Neuropixel data.
-Maybe focus on the potential improvement of held-out likelihood after clustering?
-But I guess similar problem will happen.
-When fitting data with 84 neurons, the number of component jumps around 10.}}
+	\textcolor{red}{\textbf{I prefer just give one application.} Because 1) there's a page limit (9 pages excluding ref \& appendix in 2021, and now it seems we will exceed the limit), and 2) the V1 data is enough to illustrate the idea If we really need to add the second example, we should tell the story in another angle}.
 
@@
 
-	\subsection{Full Conditional Distributions}
+	\subsection{MCMC updates}
+	The posteriors are sampled by a Gibbs sampler, the full conditional distributions, if available, are shown as follows:
 
-	\underline{\textit{$x_t^{(j)}$ and $\mu_t^{(j)}$}}:
+	\underline{\textit{Update $x_t^{(j)}$ and $\mu_t^{(j)}$}}:
 	Let the $t^{\text{th}}$ column observation and firing rate of the $j^{\text{th}}$ cluster be $\widetilde{\bm{y}}_t = vec\left(\{y_{it}^{(j)}|z_i = j\}\right)$ and $\widetilde{\bm{\lambda}}_t^{(j)} = vec\left(\{\lambda_{it}|z_i = j\}\right)$.
@@
 	\begin{align*}
-		h &= \text{const} + \sum_{t=1}^{T} \widetilde{\bm{y}}'_t\bm{\widetilde{C}}^{(j)}\bm{\widetilde{x}}_t^{(j)} - \widetilde{\bm{\lambda}}_t
+		h &= \text{const} + \sum_{t=1}^{T} \left(\widetilde{\bm{y}}'_t\bm{\widetilde{C}}^{(j)}\bm{\widetilde{x}}_t^{(j)} - \widetilde{\bm{\lambda}}_t\right)
 		- \frac{1}{2}(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}_0(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})\\ 
@@
 	\end{align*}
-
+	, where "const" represents the constant.
 The $h(\bm{\widetilde{x}}^{(j)})$ is concave and hence unimodal, and the Markovian structure of the latent dynamics, and hence the tri-block-diagonal Hessian, makes it possible to compute a Newton update in $\mathcal{O}(T)$ \citep{Paninski2010}.
+The same technique is used in the E-step for PLDS model \citep{Macke2011}.
 
@@
 
-	\underline{\textit{$\bm{c}_i$}}: 
+	\underline{\textit{Update $\bm{c}_i$}}: 
 	When writing the observation mapping in matrix form,
 	$\log\bm{\lambda}_i = \bm{\mu}^{(j)}+ \bm{X}^{(j)}\bm{c}_i$, the update of $\bm{c}_i$ reduces to a regular Poisson regression problem, given $\bm{\mu}^{(j)}$ and $\bm{X}^{(j)}$ are known.
-Here, we sample the posterior by no a No U-Turn Sampler (NUTS, \cite{Hoffman2011}).
+Here, we sample the posterior by no a No U-Turn Sampler (NUTS, \cite{Hoffman2011}), within the Gibbs sampler.
 
-	\underline{\textit{Linear dynamics of latent state}}: The parameters for linear dynamics are $f^{(j)}$, $g^{(j)}$, $\sigma^{2(j)}$, $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$.
+	\underline{\textit{Update Linear dynamics of latent state}}: The parameters for linear dynamics are $f^{(j)}$, $g^{(j)}$, $\sigma^{2(j)}$, $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$.
 To make the model identifiable, we simply assume $\bm{A}^{(j)} = diag(a_1^{(j)},\ldots, a_p^{(j)})$ and $\bm{Q}^{(j)} = diag(q_1^{(j)},\ldots, q_p^{(j)})$.
@@
 
+	\underline{\textit{Update $z_i$}}: To update the cluster assignments for each neuron $i$, we modify a partition based algorithm in DPM for MFM.
+Because of the non-conjugacy, the marginal likelihood in terms of all cluster parameters $\theta_{z_i}$ can not be easily computed.
+This issue can be solved by introducing auxiliary variable, as in "Algorithm 8" in \cite{Neal2000} for DPM.
+We can use the same idea for inference in the MFM by two substitutions: 1) replace $|c_i|$ by $|c_i| + \gamma$ and 2) replace $\alpha$ by $\gamma V_n(t+1)/ V_n(t)$.
+Here, $t$ is the number of partition obtained by removing the neuron $i$.
+The $V_n(t) = \sum_{k=1}^{\infty}\frac{k_{(t)}}{(\gamma k)^{(n)}}p_K(k)$, with $x^{(m)} = x(x+1)\cdots(x+m-1)$, $x_{(m)} = x(x-1)\cdots(x-m+1)$, $x^{(0)} = 1$ and $x_{(0)} = 1$.
+See details in \cite{Miller2018}
+
 \end{document}

### 2022-01-26 core/lds/newtonGH.m
new plots

@@ -1,11 +1,17 @@
 function [x,fx,dfdx,xx,eoi] = newtonGH(fdf,x0,TolX,MaxIter)
 
+% to debug
+% fdf = gradHess;
+% x0 = dX_tmp(:);
+% TolX = 1e-4;
+% MaxIter = 2;
+% 
+
 TolFun=eps;
 xx(:,1) = x0;
 dh = feval(fdf, x0);
 fx = dh{1};
 % disp(norm(fx))
-warning('off');
 for k = 1:MaxIter
 %     disp(k)
     dfdx = dh{2};

### 2022-01-26 core/lds/update_cluster_new.m

@@ -1,6 +1,6 @@
 function [theta_b, epsilonOut, log_pdf] =...
     update_cluster_new(Y_tmp,theta_a,theta_b,...
-    prior, N, T, p, obsIdx, active, density, OPTDC_tmp, Y)
+    prior, N, T, p, obsIdx, active, density, OPTDC_tmp)
 
 % for debug
 % c=j;
@@ -22,18 +22,23 @@ gradHess = @(vecdX) gradHessX(vecdX, zeros(N_tmp,1),...
     prior.theta0, prior.Q0, theta_a.Q, theta_a.A, theta_a.b, Y_tmp);
 
 dX = [theta_a.d;theta_a.X];
+warning('off');
 [mudXvec,~,hess_tmp,~, eoi] = newtonGH(gradHess,dX(:),1e-6,1000);
+warning('on');
 
 if((sum(isnan(mudXvec)) ~= 0) || eoi == 1000)
-    disp('use adaptive smoother initial')
+    
     try
-        dX_tmp = ppasmoo_poissexp_v2(Y_tmp,[ones(N_tmp,1) theta_a.C(obsIdx,:)],...
+        disp('use adaptive smoother initial')
+        dX_tmp = ppasmoo_poissexp_na(Y_tmp,[ones(N_tmp,1) theta_a.C(obsIdx,:)],...
             zeros(N_tmp,1),prior.theta0,prior.Q0,theta_a.A,theta_a.b,theta_a.Q);
     catch
+        disp('use 0')
         dX_tmp = dX*0;
     end
-    
-    [mudXvec,~,hess_tmp,~] = newtonGH(gradHess,dX_tmp(:),1e-6,1000);
+    warning('off');
+    [mudXvec,~,hess_tmp,~] = newtonGH(gradHess,dX_tmp(:),1e-4,1000);
+    warning('on');
 end
 
 % use Cholesky decomposition to sample efficiently
@@ -117,29 +122,43 @@ end
 
 dX = [theta_b.d;theta_b.X];
 for k = 1:(p+1)
-    Y_BA = dX(k,2:T)';
-    X_BA = [ones(T-1,1) dX(k,1:(T-1))'];
-    
-    BAn = (X_BA'*X_BA + prior.Lamb0)\(X_BA'*Y_BA + prior.Lamb0*prior.BA0);
-    PsiQ = prior.Psi0 + (Y_BA - X_BA*BAn)'*(Y_BA - X_BA*BAn) +...
[…]

### 2022-01-26 demo/globalNonlinear/demo_labeled.m

@@ -1,0 +1,288 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(123)
+n = 5;
+nClus = 10;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+
+d = randn(n*nClus,1)/5;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = (sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)) + [-1.5 0])*0.5;
+end
+
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus,1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b = zeros(p*nClus,1);
+Q = eye(nClus*p)*1e-3;
+
+% Generate X offline (A unspecified)
+for i=1:size(Q,1)
+    k = ceil(rand()*20)+10;
+    X(i,:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+% figure(1)
+Y = poissrnd(exp(logLam));
+% clusterPlot(Y, Lab)
+
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
+    [QX, RX] = mgson(X(latidTmp,:)');
+    Mdiag{l} = inv(RX');
+end
+M = sparse(blkdiag(Mdiag{:}));
+g = M*gtmp;
+
+X = M*X + g;
+x0 = M*x0 + g;
+d = d - (C_trans/M)*g;
+C_trans = C_trans/M;
+Q = M*Q*M';
+Q0 = M*Q0*M';
+
+
+
+%% MCMC settings
+p=1;
+rng(1)
+ng = 1000;
+
+% theta = (d x)
+prior.theta0 = zeros(1+p,1);
+prior.Q0 = eye(1+p);
+
+prior.muC0 = zeros(p,1);
[…]

### 2022-01-26 demo/globalNonlinear/demo_labeled_compare.m

@@ -1,0 +1,262 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+rng(123)
+n = 5;
+nClus = 10;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+
+d = randn(n*nClus,1)/5;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = (sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)) + [-1.5 0])*0.5;
+end
+
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus,1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b = zeros(p*nClus,1);
+Q = eye(nClus*p)*1e-3;
+
+% Generate X offline (A unspecified)
+for i=1:size(Q,1)
+    k = ceil(rand()*20)+10;
+    X(i,:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+% figure(1)
+Y = poissrnd(exp(logLam));
+% clusterPlot(Y, Lab)
+
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
+    [QX, RX] = mgson(X(latidTmp,:)');
+    Mdiag{l} = inv(RX');
+end
+M = sparse(blkdiag(Mdiag{:}));
+g = M*gtmp;
+
+X = M*X + g;
+x0 = M*x0 + g;
+d = d - (C_trans/M)*g;
+C_trans = C_trans/M;
+Q = M*Q*M';
+Q0 = M*Q0*M';
+
+
+%%
+nRep = 50;
+propTrain = [3/4 1/2];
+ng = 50;
+idx = round(ng/2):ng;
+
+Ytest_spk = zeros(nRep, 2);
+llhd_spk_test_3clus = zeros(nRep, 2);
[…]

### 2022-01-26 demo/globalNonlinear/demo_unLabeled.m

@@ -1,0 +1,314 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+close all;
+rng(123)
+n = 5;
+nClus = 10;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+
+d = randn(n*nClus,1)/5;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = (sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)) + [-1.5 0])*0.5;
+end
+
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus,1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b = zeros(p*nClus,1);
+Q = eye(nClus*p)*1e-3;
+
+% Generate X offline (A unspecified)
+for i=1:size(Q,1)
+    k = ceil(rand()*20)+10;
+    X(i,:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
+end
+
+for kk = [3 5 8 9]
+    idTmp = id2id(kk,p);
+    k = ceil(rand()*25)+10;
+    for j = 1:p
+        X(idTmp(j),:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
+    end
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+% figure(1)
+Y = poissrnd(exp(logLam));
+% clusterPlot(Y, Lab)
+
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
+    [QX, RX] = mgson(X(latidTmp,:)');
+    Mdiag{l} = inv(RX');
+end
+M = sparse(blkdiag(Mdiag{:}));
+g = M*gtmp;
+
+X = M*X + g;
+x0 = M*x0 + g;
[…]

### 2022-01-26 demo/globalNonlinear/demo_unLabeled_compare.m

@@ -1,0 +1,503 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(123)
+n = 5;
+nClus = 10;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+
+d = randn(n*nClus,1)/5;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = (sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)) + [-1.5 0])*0.5;
+end
+
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus,1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b = zeros(p*nClus,1);
+Q = eye(nClus*p)*1e-3;
+
+% Generate X offline (A unspecified)
+for i=1:size(Q,1)
+    k = ceil(rand()*20)+10;
+    X(i,:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+% figure(1)
+Y = poissrnd(exp(logLam));
+% clusterPlot(Y, Lab)
+
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
+    [QX, RX] = mgson(X(latidTmp,:)');
+    Mdiag{l} = inv(RX');
+end
+M = sparse(blkdiag(Mdiag{:}));
+g = M*gtmp;
+
+X = M*X + g;
+x0 = M*x0 + g;
+d = d - (C_trans/M)*g;
+C_trans = C_trans/M;
+Q = M*Q*M';
+Q0 = M*Q0*M';
+
+imagesc(exp(C_trans*X + d))
+colorbar()
+
+%% split train & test
+rng(1)
+propTrain = 1/2;
+nTrain = round(T*propTrain);
[…]

### 2022-01-26 demo/labeled/raw_new_cvSpkeckle_all.m

@@ -369,6 +369,10 @@ plot(llhd_spk_test(1:2,50:end)')
 plot(llhd_spk_train(3:4,50:end)')
 plot(llhd_spk_test(3:4,50:end)')
 
+%% let's further select the number of latent...
+
+
+
 
 
 

### 2022-01-26 demo/labeled/raw_new_cvSpkeckle_all2.m

@@ -1,0 +1,344 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+d(1:10) = d(1:10) + 1;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        (sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-1 2])*0.5;
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p);
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
[…]

### 2022-01-26 demo/labeled/try_globalNonlinear.m

@@ -1,0 +1,242 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%%
+rng(123)
+n = 5;
+nClus = 10;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+
+d = randn(n*nClus,1)/5;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) = (sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k)) + [-1.5 0])*0.5;
+end
+
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus,1);
+Q0 = eye(nClus*p)*1e-2;
+X(:,1) = mvnrnd(x0, Q0)';
+
+b = zeros(p*nClus,1);
+Q = eye(nClus*p)*1e-3;
+
+% Generate X offline (A unspecified)
+for i=1:size(Q,1)
+    k = ceil(rand()*20)+10;
+    X(i,:) = interp1(linspace(0,1,k),randn(k,1),linspace(0,1,T),'spline');
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+% figure(1)
+Y = poissrnd(exp(logLam));
+% clusterPlot(Y, Lab)
+
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
+    [QX, RX] = mgson(X(latidTmp,:)');
+    Mdiag{l} = inv(RX');
+end
+M = sparse(blkdiag(Mdiag{:}));
+g = M*gtmp;
+
+X = M*X + g;
+x0 = M*x0 + g;
+d = d - (C_trans/M)*g;
+C_trans = C_trans/M;
+Q = M*Q*M';
+Q0 = M*Q0*M';
+
+figure(1)
+subplot(1,2,1)
+imagesc(Y)
+colorbar()
+subplot(1,2,2)
+imagesc(exp(C_trans*X + d))
+colorbar()
+
+
+
[…]

### 2022-01-26 demo/plot/plots.m

@@ -1,8 +1,43 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
 
-plotFolder = 'C:\Users\gaw19004\Documents\GitHub\state-space-clustering\plots';
+%% true
+plotFolder = 'C:\Users\gaw19004\Documents\GitHub\state-space-clustering\demo\globalNonlinear\plots';
 cd(plotFolder)
 
-%%
+dxSum3 = zeros(p+1,T);
+dxSum5 = zeros(p+1,T);
+
+c = 0;
+for g = round(ng/2):ng
+    c= c+1;
+    dxSum3 = dxSum3 + [THETA{g}(3).d;THETA{g}(3).X];
+    dxSum5 = dxSum5 + [THETA{g}(5).d;THETA{g}(5).X];
+end
+
+dxMean3 = dxSum3/c;
+dxMean3 = dxMean3 - mean(dxMean3, 2);
+[dxMean3, ~] = mgson(dxMean3');
+
+dxMean5 = dxSum5/c;
+dxMean5 = dxMean5 - mean(dxMean5, 2);
+[dxMean5, ~] = mgson(dxMean5');
+
+hold on
+plot(-X(5,:), 'k', 'LineWidth', 1)
+plot(dxMean3(:,1), 'r', 'LineWidth', 1)
+plot(dxMean5(:,1), 'b', 'LineWidth', 1)
+hold off
+
+
+
+
+
+%% no cluster
+clear all;close all;clc;
+% plotFolder = 'C:\Users\gaw19004\Documents\GitHub\state-space-clustering\plots\noCluster';
+plotFolder = 'C:\Users\gaw19004\Documents\GitHub\state-space-clustering\demo\globalNonlinear\plots\noCluster';
+cd(plotFolder)
+
 dSum1 = zeros(1,T);
 dSum2 = zeros(1,T);
 dSum3 = zeros(1,T);
@@ -46,94 +81,69 @@ dxMean2 = dxMean2 - mean(dxMean2, 2);
 dxMean3 = dxMean3 - mean(dxMean3, 2);
 [dxMean3, ~] = mgson(dxMean3');
 
-%% LF
+% LF
+
 LF = figure;
-subplot(3,4,1)
-plot(X(id2id(1,2),:)')
-y1 = ylim;
-title('dX-trans: true')
-subplot(3,4,2)
-plot(dSum1'/c)
-title('d:fit')
-subplot(3,4,3)
-plot(xSum1'/c)
[…]

### 2022-01-26 demo/real/demo_hc_all.m

@@ -1,228 +1,0 @@
-addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
-addpath(genpath('C:\Users\gaw19004\Documents\GitHub\data'));
-% addpath(genpath('D:\github\state-space-clustering'));
-
-%%
-load('ec016.272.mat')
-
-dt = 0.05;
-T = 1000;
-Tstart = 0;
-N = size(Tlist, 2);
-Y = zeros(N, T);
-for n = 1:N
-    for k = 1:T
-        Y(n,k) = sum((Tlist{n} > (dt*(k-1)) + Tstart) &...
-            (Tlist{n} < (dt*k + Tstart)));
-    end
-end
-
-imagesc(Y)
-colorbar()
-
-%%
-lAbsGam = @(x) log(abs(gamma(x)));
-rng(2)
-p=1;
-ng = 5000;
-t_max = N;
-
-% this is the DP setting, replace to MFM later...
-% DPMM = true;
-% alpha_random = true;
-% sigma_alpha = 0.1; % scale for MH proposals in alpha move
-% alphaDP = 1;
-% log_v = (1:t_max+1)*log(alphaDP) - lAbsGam(alphaDP+N) + lAbsGam(alphaDP);
-% a = 1;
-% b = 0;
-
-
-% MFM:
-DPMM = false;
-alpha_random = false;
-MFMgamma = 1;
-% K ~ Geometric(r)
-r = 0.3;
-log_pk = @(k) log(r) + (k-1)*log(1-r);
-
-a = MFMgamma;
-b = MFMgamma;
-log_v = MFMcoeff(log_pk, MFMgamma, N, t_max + 1);
-
-logNb = log((1:N) + b);
-
-% priors...
-% theta = (d x)
-prior.theta0 = zeros(1+p,1);
-prior.Q0 = eye(1+p);
-
-prior.muC0 = zeros(p,1);
-prior.SigC0 = eye(p);
-
-prior.BA0 =[0 1]';
-prior.Lamb0 = eye(2);
-prior.Psi0 = 1e-4;
-prior.nu0 = 1+2;
-
-% pre-allocation
-t_fit = zeros(1, ng);
-Z_fit = zeros(N, ng);
-numClus_fit = zeros(t_max + 3, ng);
-
-t_fit(1) = 1;
-Z_fit(:,1) = ones(N, 1);
-numClus_fit(1,1) = N;
[…]

### 2022-01-26 demo/real/demo_pixel.m

@@ -9,7 +9,7 @@ tab = readtable('719161530_units.csv'); % meta-data
 % tab.ecephys_structure_acronym % names of anatomical structures for each single-unit
 subset = find(tab.snr>3 & cellfun(@length,Tlist')>1000);
 idx50 = find(ismember(string(tab.ecephys_structure_acronym),...
-    ["VISam" "LP" "VPM"]));
+    ["LGd" "VISp" "CA1" "LP"]));
 
 % "CA1" "LGd" "POL" "VISp" "VISl"
 % VISam
@@ -42,15 +42,39 @@ lab_all_str = string(tab.ecephys_structure_acronym);
 [lab_num_sub2, clusIdx] = findgroups(lab_all_str(subset2));
 
 % spontaneous
-% Tstart = 29.8301073815904; Tend = 89.89682738;
-Tstart =  1001.89177167499; Tend = 1290.88309738159;
+% Tstart = 29.83010738; Tend = 89.89682738; % 1
+Tstart = 1001.891772; Tend = 1290.883097; % 2
+% Tstart = 1589.382401; Tend = 1591.133857; % 3
+% Tstart = 2190.634543; Tend = 2221.660447; % 4
+% Tstart = 2822.161967; Tend = 2852.187037; % 5
+% Tstart = 3152.437777; Tend = 3182.462857; % 6
+% Tstart = 3781.963503; Tend = 4083.215117; % 7
+% Tstart = 4683.716567; Tend = 4713.741627; % 8
+% Tstart = 5397.312443; Tend = 5398.313257; % 9
+% Tstart = 5878.714467; Tend = 5908.739537; % 10
+% Tstart = 6389.157337; Tend = 6689.408117; % 11
+% Tstart = 7169.809297; Tend = 7199.834317; % 12
+% Tstart = 7680.268867; Tend = 7710.293937; % 13
+% Tstart = 8010.544677; Tend = 8040.569727; % 14
+% Tstart = 8568.51062;  Tend = 8611.046137; % 15
 
-% drifting_gratings 
-% Tstart = 1591.13385738159; Tend = 2190.63454310612;
+% natural scences
+% Tstart = 5908.739537; Tend = 6389.157337;
+% Tstart = 6689.408117; Tend = 7169.809297;
+% Tstart = 8040.569727; Tend = 8568.51062;
+
+% drifting gratings
+% Tstart = 1591.133857; Tend = 2190.634543;
[…]

### 2022-01-26 demo/real/demo_pixel_all.m

@@ -9,7 +9,7 @@ tab = readtable('719161530_units.csv'); % meta-data
 % tab.ecephys_structure_acronym % names of anatomical structures for each single-unit
 subset = find(tab.snr>3 & cellfun(@length,Tlist')>1000);
 idx50 = find(ismember(string(tab.ecephys_structure_acronym),...
-    ["VISam" "LP" "VPM"]));
+    ["LGd" "VISp" "CA1" "LP"]));
 
 % "CA1" "LGd" "POL" "VISp" "VISl"
 % VISam
@@ -42,15 +42,39 @@ lab_all_str = string(tab.ecephys_structure_acronym);
 [lab_num_sub2, clusIdx] = findgroups(lab_all_str(subset2));
 
 % spontaneous
-% Tstart = 29.8301073815904; Tend = 89.89682738;
-% Tstart =  1001.89177167499; Tend = 1290.88309738159;
+% Tstart = 29.83010738; Tend = 89.89682738; % 1
+Tstart = 1001.891772; Tend = 1290.883097; % 2
+% Tstart = 1589.382401; Tend = 1591.133857; % 3
+% Tstart = 2190.634543; Tend = 2221.660447; % 4
+% Tstart = 2822.161967; Tend = 2852.187037; % 5
+% Tstart = 3152.437777; Tend = 3182.462857; % 6
+% Tstart = 3781.963503; Tend = 4083.215117; % 7
+% Tstart = 4683.716567; Tend = 4713.741627; % 8
+% Tstart = 5397.312443; Tend = 5398.313257; % 9
+% Tstart = 5878.714467; Tend = 5908.739537; % 10
+% Tstart = 6389.157337; Tend = 6689.408117; % 11
+% Tstart = 7169.809297; Tend = 7199.834317; % 12
+% Tstart = 7680.268867; Tend = 7710.293937; % 13
+% Tstart = 8010.544677; Tend = 8040.569727; % 14
+% Tstart = 8568.51062;  Tend = 8611.046137; % 15
 
-% drifting_gratings 
-Tstart = 1591.13385738159; Tend = 2190.63454310612;
+% natural scences
+% Tstart = 5908.739537; Tend = 6389.157337;
+% Tstart = 6689.408117; Tend = 7169.809297;
+% Tstart = 8040.569727; Tend = 8568.51062;
+
+% drifting gratings
[…]

### 2022-01-26 demo/unLabeled/demo_new_gamma.m

@@ -137,7 +137,7 @@ actList = zeros(t_max+3,1); actList(1) = 1;
 c_next = 2;
 
 for k = 1:t_fit(1)
-    THETA{1}(k) = sample_prior_new(prior, N, T, p, true, Inf);
+    THETA{1}(k) = sample_prior_new(prior, N, T, p, false, Inf);
 end
 
 %% MCMC
@@ -150,6 +150,11 @@ end
 burnIn = round(ng/10);
 epsilon = 0.01*ones(N,1);
 
+simMat = zeros(N,N);
+for k = 1:size(simMat, 1)
+    simMat(k,:) = simMat(k,:) + (Z_fit(k, 1) == Z_fit(:, 1))';
+end
+fitMFRTrace = zeros(N, T, ng);
 for g = 2:ng
     
     if(g < burnIn);disp("iter " + g + ", changing"); % change epsilon
@@ -166,10 +171,7 @@ for g = 2:ng
         
         [THETA{g}(c), epsilon(obsIdx), log_pdf] =...
             update_cluster_new(Y(obsIdx,:),THETA{g-1}(c),THETA{g}(c),...
-            prior, N, T, p, obsIdx, true, false, OPTDC(obsIdx), Y);
-%         [THETA{g}(c), epsilon(obsIdx), log_pdf] =...
-%             update_cluster_new_aug(Y(obsIdx,:),THETA{g-1}(c),THETA{g}(c),...
-%             prior, N, T, p, obsIdx, true, false, OPTDC(obsIdx), Y);
+            prior, N, T, p, obsIdx, true, false, OPTDC(obsIdx));
     end
     
     if(g == burnIn)
@@ -205,7 +207,7 @@ for g = 2:ng
         numClus_fit(c,g) = numClus_fit(c,g) - 1;
         if(numClus_fit(c,g) > 0)
             c_prop = c_next;
-            THETA{g}(c_prop) = sample_prior_new(prior, N, T, p, true, Inf);
+            THETA{g}(c_prop) = sample_prior_new(prior, N, T, p, false, Inf);
         else
             c_prop = c;
             actList = ordered_remove(c, actList, t_fit(g));
@@ -241,13 +243,22 @@ for g = 2:ng
         numClus_fit(c,g) = numClus_fit(c,g) + 1;
     end
     
-    figure(2)
+    figure(1)
     clusterPlot(Y, Z_fit(:,g)')
+    
[…]

### 2022-01-26 demo/unLabeled/demo_new_gamma2.m

@@ -1,0 +1,340 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+d(1:10) = d(1:10) + 1;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-1 2];
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p);
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
[…]

### 2022-01-26 demo/unLabeled/demo_new_gamma3.m

@@ -1,0 +1,339 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+d(1:10) = d(1:10) + 1;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-1 2];
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p);
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
[…]

### 2022-01-26 demo/unLabeled/demo_new_gamma4.m

@@ -1,0 +1,349 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+d(1:10) = d(1:10) + 1;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-1 2];
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p);
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
[…]

### 2022-01-26 demo/unLabeled/demo_new_gamma_compare.m

@@ -1,0 +1,349 @@
+%% split train & test
+rng(0)
+propTrain = 1/4;
+nTrain = round(T*propTrain);
+Y_train = nan*Y;
+Y_test = nan*Y;
+for k = 1:N
+    trIdx = randsample(T, nTrain);
+    teIdx = setdiff(1:T, trIdx);
+    Y_train(k,trIdx) = Y(k,trIdx);
+    Y_test(k, teIdx) = Y(k,teIdx);
+end
+
+
+
+%% cluter-on
+rng(1)
+p=1;
+ng = 1000;
+t_max = N;
+
+% MFM:
+DPMM = false;
+alpha_random = false;
+MFMgamma = 1;
+% K ~ Geometric(r)
+r = 0.2;
+log_pk = @(k) log(r) + (k-1)*log(1-r);
+
+a = MFMgamma;
+b = MFMgamma;
+log_v = MFMcoeff(log_pk, MFMgamma, N, t_max + 1);
+
+logNb = log((1:N) + b);
+
+% priors...
+% theta = (d x)
+prior.theta0 = zeros(1+p,1);
+prior.Q0 = eye(1+p);
+
+prior.muC0 = zeros(p,1);
+prior.SigC0 = eye(p);
+
+prior.BA0 =[0 1]';
+prior.Lamb0 = eye(2);
+prior.Psi0 = 1e-3;
+prior.nu0 = 1+2;
+
+% pre-allocation
+t_fit = zeros(1, ng);
+Z_fit = zeros(N, ng);
+numClus_fit = zeros(t_max + 3, ng);
+
+t_fit(1) = 1;
+Z_fit(:,1) = ones(N, 1);
+numClus_fit(1,1) = N;
+actList = zeros(t_max+3,1); actList(1) = 1;
+c_next = 2;
+
+for k = 1:t_fit(1)
+    THETA{1}(k) = sample_prior_new(prior, N, T, p, false, Inf);
+end
+
+%% MCMC
+
+burnIn = round(ng/10);
+epsilon = 0.01*ones(N,1);
+for k = 1:N
+    optdc.M=1;
+    optdc.Madapt=0;
+    OPTDC{k} = optdc;
+%     OPTDC{k}.epsilon = epsilon(k);
+end
+
+fitMFRTrace = zeros(N, T, ng);
+llhd_spk_train = zeros(ng,1);
+llhd_spk_test = zeros(ng,1);
+
+for g = 2:ng
+    
[…]

### 2022-01-26 demo/unLabeled/try/augmentation/demo.m

@@ -103,7 +103,7 @@ DPMM = false;
 alpha_random = false;
 MFMgamma = 1;
 % K ~ Geometric(r)
-r = 0.2;
+r = 0.3;
 log_pk = @(k) log(r) + (k-1)*log(1-r);
 
 a = MFMgamma;
@@ -141,25 +141,32 @@ for k = 1:t_fit(1)
 end
 
 %% MCMC
-for k = 1:N
-    optdc.M=1;
-    optdc.Madapt=0;
-    OPTDC{k} = optdc;
-end
 
 burnIn = round(ng/10);
 epsilon = 0.01*ones(N,1);
 
+for i = 1:N
+    optdc.M=1;
+    optdc.Madapt=0;
+    OPTDC{i} = optdc;
+    OPTDC{i}.epsilon = epsilon(i); % no tune of epsilon
+end
+
+simMat = zeros(N,N);
+for k = 1:size(simMat, 1)
+    simMat(k,:) = simMat(k,:) + (Z_fit(k, 1) == Z_fit(:, 1))';
+end
+fitMFRTrace = zeros(N, T, ng);
 for g = 2:ng
     
     if(g < burnIn);disp("iter " + g + ", changing"); % change epsilon
     elseif(g == burnIn) % tune epsilon
-        for(k = 1:N);OPTDC{k}.Madapt=50;end
+        %         for(k = 1:N);OPTDC{k}.Madapt=10;end
         disp("iter " + g + ", tuning");
     else;disp("iter " + g + ", tuned");
     end % fix epsilon
     
-    % (1) impute Z_fit...
+    % (1) update cluster index z_i
     Z_fit(:,g) = Z_fit(:,g-1);
     numClus_fit(:,g) = numClus_fit(:,g-1);
     t_fit(g) = t_fit(g-1);
@@ -180,12 +187,12 @@ for g = 2:ng
     
     for ii = 1:N
         
-        % (a) remove point i from its cluster
+        % (a) remove point ii from its cluster
         c = Z_fit(ii, g);
         numClus_fit(c,g) = numClus_fit(c,g) - 1;
         if(numClus_fit(c,g) > 0)
             c_prop = c_next;
-            THETA{g-1}(c_prop) = sample_prior_new(prior, N, T, p, true, Inf);
+            THETA{g-1}(c_prop) = sample_prior_new(prior, N, T, p, false, Inf);
         else
             c_prop = c;
[…]

### 2022-01-26 demo/unLabeled/try/augmentation/demo2.m

@@ -1,0 +1,330 @@
+addpath(genpath('C:\Users\gaw19004\Documents\GitHub\state-space-clustering'));
+% addpath(genpath('D:\github\state-space-clustering'));
+
+%% simulation
+rng(1)
+n = 10;
+nClus = 3;
+N = n*nClus;
+p = 2;
+T = 1000;
+
+Lab = repelem(1:nClus, n);
+pLab = repelem(1:nClus, p);
+
+d = randn(n*nClus,1)*0;
+d(1:10) = d(1:10) + 1;
+C_trans = zeros(n*nClus, p*nClus);
+for k = 1:length(Lab)
+    C_trans(k, ((Lab(k)-1)*p+1):(Lab(k)*p)) =...
+        sum(Lab(1:k)==Lab(k))/sum(Lab==Lab(k))+[-1 2];
+end
+
+X = zeros(p*nClus, T);
+x0 = zeros(p*nClus, 1);
+Q0 = eye(nClus*p);
+X(:,1) = mvnrnd(x0, Q0)';
+
+b1 = randn(p, 1)*1e-3;
+b2 = randn(p, 1)*1e-3;
+b3 = randn(p, 1)*1e-4;
+b = [b1;b2;b3];
+
+Q1 = 1e-3*eye(p);
+Q2 = 1e-4*eye(p);
+Q3 = 1e-5*eye(p);
+Q = blkdiag(Q1, Q2, Q3);
+
+%
+A = eye(size(Q,1));
+while any(imag(eig(A))==0)
+    A= randn(size(Q));
+    A = A-diag(diag(A));
+    A(squareform(pdist(pLab'))==0)=0;
+    A = A./sqrt(sum((A-diag(diag(A))).^2,2))*0.1;
+    %     A = A+eye(size(Q,1))*0.92;
+    A = A + diag(randn(nClus*p, 1)*1e-1 + 0.92);
+end
+
+% let's generate lambda
+logLam = zeros(n*nClus, T);
+logLam(:,1) = d + C_trans*X(:,1);
+
+for t=2:T
+    X(:, t) = mvnrnd(A*X(:, t-1) + b, Q)';
+    logLam(:, t) = d + C_trans*X(:,t);
+end
+
+Y = poissrnd(exp(logLam));
+
+gtmp = -mean(X, 2);
+Mdiag = {};
+for l = unique(Lab)
+    latidTmp = id2id(l,p);
[…]

### 2022-01-26 demo/unLabeled/try/augmentation/note.m

@@ -1,0 +1,67 @@
+
+
+
+% calculate Chebychev parameters
+% chebMax = 10*max(Y(:));
+% Cheb = zeros(1+chebMax, 2+1);
+% for k = 0:chebMax % k = min(Y(:)):max(Y(:))
+%     if k ==0
+%         cehbTmp = compute_cheb([log(0.1)-2 log(0.1)+2]);
+%         %         cehbTmp = compute_cheb([log(0.1) log(1)]);
+%     else
+%         cehbTmp = compute_cheb([log(k)-2 log(k)+2]);
+%         %         cehbTmp = compute_cheb([log(k) log(k+1)]);
+%     end
+%     Cheb(k+1,:) = cehbTmp;
+% end
+
+
+
+%             cheb_a = zeros(T,1);
+%             cheb_b = zeros(T,1);
+%             for t = 1:T
+%                 idx = min(round(exp(THETA{g}(cc).d(t))) + 1, chebMax + 1);
+%                 cheb_a(t) = Cheb(idx,3);
+%                 cheb_b(t) = Cheb(idx,2);
+%             end
+%             logLike = poiLogMarg_PAL(Y(ii,:)', THETA{g}(cc).X', THETA{g}(cc).d', prior,...
+%                 cheb_a, cheb_b);
+
+
+
+
+
+%         cheb_a = zeros(T,1);
+%         cheb_b = zeros(T,1);
+%         for t = 1:T
+%             idx = min(round(exp(THETA{g}(c_prop).d(t))) + 1, chebMax + 1);
+%             cheb_a(t) = Cheb(idx,3);
+%             cheb_b(t) = Cheb(idx,2);
+%         end
+%         logLike = poiLogMarg_PAL(Y(ii,:)', THETA{g}(c_prop).X', THETA{g}(c_prop).d', prior,...
+%             cheb_a, cheb_b);
+
+
+
+
+%% MH_notes...
+
+
+
+% ues MH
+%         derc = @(c) X_tmpC'*(Y(i,:)' - lamC(c)) - prior.SigC0\(c - prior.muC0);
+%         hessc = @(c) -X_tmpC'*diag(lamC(c))*X_tmpC - inv(prior.SigC0);
+%         c0 = C_fit(i,:,g-1)';
+%
+%         [muc,~,niSigc,~] = newton(derc,hessc,c0,1e-8,1000);
+%         if(sum(isnan(muc)) ~= 0)
+%             disp('use 0')
[…]

### 2022-01-26 demo/unLabeled/try/augmentation/sample_prior_new_aug2.m

@@ -1,0 +1,36 @@
+function theta =...
+    sample_prior_new_aug2(prior, N, T, p, sampleDynamics)
+
+flag = 1;
+while flag
+    
+    % linear dynamics: b, A & Q
+    theta.A = eye(p+1);
+    theta.b = zeros(p+1,1);
+    theta.Q = eye(p+1)*prior.Psi0;
+    
+    % mean & latent: d, X
+    dX = ones(1+p,T)*Inf;
+    dX(:,1) = mvnrnd(prior.theta0, prior.Q0);
+    
+    if sampleDynamics
+        for k = 1:p+1
+            theta.Q(k,k)= iwishrnd(prior.Psi0, prior.nu0);
+            baSamp = mvnrnd(prior.BA0, kron(theta.Q(k,k), inv(prior.Lamb0)))';
+            theta.b(k) = baSamp(1);
+            theta.A(k,k) = baSamp(2);
+        end
+    end
+    
+    for t= 2:T
+        dX(:, t) = mvnrnd(theta.b + theta.A*dX(:, t-1), theta.Q);
+    end
+    
+    theta.d = dX(1,:);
+    theta.X = dX(2:end,:);
+    flag = 0; 
+    
+end
+
+
+end

### 2022-01-26 demo/unLabeled/try/augmentation/update_cluster_new_aug.m

@@ -1,148 +1,75 @@
-function [theta_b, epsilonOut, log_pdf] =...
-    update_cluster_new_aug(Y_tmp,theta_a,theta_b,...
-    prior, N, T, p, obsIdx, active, density, OPTDC_tmp, Y)
+function theta_b =...
+    update_cluster_new_aug(Y_tmp,theta_a,theta_b,prior, T, p, obsIdx)
 
 % for debug
-% c=j; c = actList(j);
-% obsIdx = obsIdx;
+% c=j; 
+% c = actList(j);
 % Y_tmp = Y(obsIdx,:);
 % theta_a = THETA{g-1}(c);
 % theta_b = THETA{g}(c);
-% active = true;
-% density = false;
-% OPTDC_tmp = OPTDC(obsIdx);
 
 N_tmp = size(Y_tmp, 1);
-log_pdf = NaN;
 
-% (0) update C
-% the transition kernel for NUTS is symmetric --> no need for calculation
-epsilonOut = ones(N_tmp, 1)*0.01;
-
-if active
-    for i = 1:N_tmp
-        X_tmpC = theta_a.X';
-        
-        lamC = @(c) exp(theta_a.d' + X_tmpC*c);
-        
-        % use NUTS
-        lpdf = @(c) sum(log(poisspdf(Y_tmp(i,:)', lamC(c)))) +...
-            log(mvnpdf(c, prior.muC0, prior.SigC0));
-        glpdf = @(c) X_tmpC'*(Y_tmp(i,:)' - lamC(c)) - prior.SigC0\(c - prior.muC0);
-        
-        fg=@(dc_r) deal(lpdf(dc_r'), glpdf(dc_r')'); % log density and gradient
-        c0 = theta_a.C(obsIdx(i),:)';
-        
-        [c_NUTS, ~, diagn]=hmc_nuts(fg, c0',OPTDC_tmp{i});
-        epsilonOut(i) = diagn.opt.epsilonbar;
-        theta_b.C(obsIdx(i),:) = c_NUTS(end,:);
-        
-        % ues MH
-%         derc = @(c) X_tmpC'*(Y_tmp(i,:)' - lamC(c)) - prior.SigC0\(c - prior.muC0);
-%         hessc = @(c) -X_tmpC'*diag(lamC(c))*X_tmpC - inv(prior.SigC0);
-%         c0 = theta_a.C(obsIdx(i),:)';
-%         
[…]

### 2022-01-26 demo/unLabeled/try/augmentation/update_cluster_new_aug2.m

@@ -1,0 +1,77 @@
+function theta_b =...
+    update_cluster_new_aug2(Y_tmp,theta_a,theta_b,prior, T, p, obsIdx, Ctmp)
+
+% for debug
+% c=j; 
+% c = actList(j);
+% Y_tmp = Y(obsIdx,:);
+% theta_a = THETA{g-1}(c);
+% theta_b = THETA{g}(c);
+
+N_tmp = size(Y_tmp, 1);
+
+% (1) update X...
+gradHess = @(vecdX) gradHessX(vecdX, zeros(N_tmp,1),...
+    [ones(N_tmp,1) Ctmp(obsIdx,:)],...
+    prior.theta0, prior.Q0, theta_a.Q, theta_a.A, theta_a.b, Y_tmp);
+
+dX = [theta_a.d;theta_a.X];
+warning('off');
+[mudXvec,~,hess_tmp,~, eoi] = newtonGH(gradHess,dX(:),1e-4,1000);
+warning('on');
+
+if((sum(isnan(mudXvec)) ~= 0) || eoi == 1000)
+    try
+        disp('use adaptive smoother initial')
+        dX_tmp = ppasmoo_poissexp_na(Y_tmp,[ones(N_tmp,1) Ctmp(obsIdx,:)],...
+            zeros(N_tmp,1),prior.theta0,prior.Q0,theta_a.A,theta_a.b,theta_a.Q);
+    catch
+        disp('use 0')
+        dX_tmp = dX*0;
+    end
+    warning('off');
+    [mudXvec,~,hess_tmp,~] = newtonGH(gradHess,dX_tmp(:),1e-4,1000);
+    warning('on');
+end
+
+% use Cholesky decomposition to sample efficiently
+try
+    RdX = chol(-hess_tmp,'lower'); % sparse
+    zdX = randn(length(mudXvec), 1) + RdX'*mudXvec;
+    dXsamp = RdX'\zdX;
+catch
+    dXsamp = mudXvec;
+end
+dXsamp_trans = reshape(dXsamp,[], T);
+
+theta_b.d = dXsamp_trans(1,:);
+theta_b.X = dXsamp_trans(2:end,:);
+
+% (2) update linear dynamics...
+dX = [theta_b.d;theta_b.X];
+for k = 1:(p+1)
+    try
+        Y_BA = dX(k,2:T)';
+        X_BA = [ones(T-1,1) dX(k,1:(T-1))'];
+        
[…]

### 2022-01-26 demo/unLabeled/update_cluster_new2.m

@@ -1,0 +1,133 @@
+function [theta_b, epsilonOut, log_pdf] =...
+    update_cluster_new2(Y_tmp,theta_a,theta_b,...
+    prior, N, T, p, obsIdx, active, density, OPTDC_tmp)
+
+% for debug
+% c=j;
+% obsIdx = obsIdx;
+% Y_tmp = Y(obsIdx,:);
+% theta_a = THETA{g-1}(c);
+% theta_b = THETA{g}(c);
+% active = true;
+% density = false;
+% OPTDC_tmp = OPTDC(obsIdx);
+
+N_tmp = size(Y_tmp, 1);
+log_pdf = NaN;
+
+
+% (2) update loading: C
+% the transition kernel for NUTS is symmetric --> no need for calculation
+epsilonOut = ones(N_tmp, 1)*0.01;
+
+if active
+    for i = 1:N_tmp
+        X_tmpC = theta_a.X';
+        
+        lamC = @(c) exp(theta_a.d' + X_tmpC*c);
+        
+        % use NUTS
+        lpdf = @(c) sum(log(poisspdf(Y_tmp(i,:)', lamC(c)))) +...
+            log(mvnpdf(c, prior.muC0, prior.SigC0));
+        glpdf = @(c) X_tmpC'*(Y_tmp(i,:)' - lamC(c)) - prior.SigC0\(c - prior.muC0);
+        
+        fg=@(dc_r) deal(lpdf(dc_r'), glpdf(dc_r')'); % log density and gradient
+        c0 = theta_a.C(obsIdx(i),:)';
+        
+        [c_NUTS, ~, diagn]=hmc_nuts(fg, c0',OPTDC_tmp{i});
+        epsilonOut(i) = diagn.opt.epsilonbar;
+        theta_b.C(obsIdx(i),:) = c_NUTS(end,:); 
+    end
+end
+
+
+
+
+% (1) update X_fit
+gradHess = @(vecdX) gradHessX(vecdX, zeros(N_tmp,1),...
+    [ones(N_tmp,1) theta_b.C(obsIdx,:)],...
+    prior.theta0, prior.Q0, theta_a.Q, theta_a.A, theta_a.b, Y_tmp);
+
+dX = [theta_a.d;theta_a.X];
+warning('off');
+[mudXvec,~,hess_tmp,~, eoi] = newtonGH(gradHess,dX(:),1e-6,1000);
+warning('on');
+
+if((sum(isnan(mudXvec)) ~= 0) || eoi == 1000)
+    
+    try
[…]

### 2022-01-26 demo/unLabeled/update_cluster_new3.m

@@ -1,0 +1,107 @@
+function [theta_b, log_pdf] =...
+    update_cluster_new3(Y_tmp,theta_a,theta_b,...
+    prior, N, T, p, obsIdx, active, density, OPTDC_tmp)
+
+% for debug
+% c=j;
+% obsIdx = obsIdx;
+% Y_tmp = Y(obsIdx,:);
+% theta_a = THETA{g-1}(c);
+% theta_b = THETA{g}(c);
+% active = true;
+% density = false;
+% OPTDC_tmp = OPTDC(obsIdx);
+
+N_tmp = size(Y_tmp, 1);
+log_pdf = NaN;
+
+% (1) update X_fit
+
+gradHess = @(vecdX) gradHessX(vecdX, zeros(N_tmp,1),...
+    [ones(N_tmp,1) theta_a.C(obsIdx,:)],...
+    prior.theta0, prior.Q0, theta_a.Q, theta_a.A, theta_a.b, Y_tmp);
+
+dX = [theta_a.d;theta_a.X];
+warning('off');
+[mudXvec,~,hess_tmp,~, eoi] = newtonGH(gradHess,dX(:),1e-6,1000);
+warning('on');
+
+if((sum(isnan(mudXvec)) ~= 0) || eoi == 1000)
+    
+    try
+        disp('use adaptive smoother initial')
+        dX_tmp = ppasmoo_poissexp_na(Y_tmp,[ones(N_tmp,1) theta_a.C(obsIdx,:)],...
+            zeros(N_tmp,1),prior.theta0,prior.Q0,theta_a.A,theta_a.b,theta_a.Q);
+    catch
+        disp('use 0')
+        dX_tmp = dX*0;
+    end
+    warning('off');
+    [mudXvec,~,hess_tmp,~] = newtonGH(gradHess,dX_tmp(:),1e-4,1000);
+    warning('on');
+end
+
+% use Cholesky decomposition to sample efficiently
+if active
+    try
+        RdX = chol(-hess_tmp,'lower'); % sparse
+        zdX = randn(length(mudXvec), 1) + RdX'*mudXvec;
+        dXsamp = RdX'\zdX;
+    catch
+        dXsamp = mudXvec;
+    end
+    dXsamp_trans = reshape(dXsamp,[], T);
+    
+    theta_b.d = dXsamp_trans(1,:);
+    theta_b.X = dXsamp_trans(2:end,:);
+end
+
+if density
+    log_pdf = 0;
[…]

### 2022-01-26 demo/unLabeled/update_cluster_new4.m

@@ -1,0 +1,45 @@
+function [theta_b, epsilonOut, log_pdf] =...
+    update_cluster_new4(Y_tmp,theta_a,theta_b,...
+    prior, N, T, p, obsIdx, active, density, OPTDC_tmp)
+
+% for debug
+% c=j;
+% obsIdx = obsIdx;
+% Y_tmp = Y(obsIdx,:);
+% theta_a = THETA{g-1}(c);
+% theta_b = THETA{g}(c);
+% active = true;
+% density = false;
+% OPTDC_tmp = OPTDC(obsIdx);
+
+N_tmp = size(Y_tmp, 1);
+log_pdf = NaN;
+
+
+% (2) update loading: C
+% the transition kernel for NUTS is symmetric --> no need for calculation
+epsilonOut = ones(N_tmp, 1)*0.01;
+
+if active
+    for i = 1:N_tmp
+        X_tmpC = theta_a.X';
+        
+        lamC = @(c) exp(theta_a.d' + X_tmpC*c);
+        
+        % use NUTS
+        lpdf = @(c) sum(log(poisspdf(Y_tmp(i,:)', lamC(c)))) +...
+            log(mvnpdf(c, prior.muC0, prior.SigC0));
+        glpdf = @(c) X_tmpC'*(Y_tmp(i,:)' - lamC(c)) - prior.SigC0\(c - prior.muC0);
+        
+        fg=@(dc_r) deal(lpdf(dc_r'), glpdf(dc_r')'); % log density and gradient
+        c0 = theta_a.C(obsIdx(i),:)';
+        
+        [c_NUTS, ~, diagn]=hmc_nuts(fg, c0',OPTDC_tmp{i});
+        epsilonOut(i) = diagn.opt.epsilonbar;
+        theta_b.C(obsIdx(i),:) = c_NUTS(end,:);
+    end
+end
+
+
+
+end

### 2022-01-26 latexDraft/draft.tex
update plots

@@
 % ready for submission
-\usepackage{neurips_2021}
+%\usepackage{neurips_2021}
 
@@
 % [preprint] option:
-%     \usepackage[preprint]{neurips_2021}
+\usepackage[preprint]{neurips_2021}
 
@@
 
-\title{Clustering Neural Populations by Poisson Dynamic Factor Model}
+\title{Clustering Neural Populations with a Poisson Dynamic Factor Model}
 
@@
 \author{%
-	David S.~Hippocampus\thanks{Use footnote for providing further information
-		about author (webpage, alternative address)---\emph{not} for acknowledging
-		funding agencies.} \\
-	Department of Computer Science\\
-	Cranberry-Lemon University\\
-	Pittsburgh, PA 15213 \\
-	\texttt{hippo@cs.cranberry-lemon.edu} \\
-	% examples of more authors
-	% \And
-	% Coauthor \\
-	% Affiliation \\
-	% Address \\
-	% \texttt{email} \\
-	% \AND
-	% Coauthor \\
-	% Affiliation \\
-	% Address \\
-	% \texttt{email} \\
+	Ganchao Wei\\
+	Department of Statistics\\
+	University of Connecticut\\
+	\texttt{ganchao.wei@uconn.edu}\\
+	\And
+	Ian. H. Stevenson\\
+	Department of Psychological Sciences\\
+	University of Connecticut\\
+	\texttt{ian.stevenson@uconn.edu} \\
+	\And
+	Xiaojing Wang\\
+	Department of Statistics\\
+	University of Connecticut\\
+	\texttt{xiaojing.wang@uconn.edu} \\
 	% \And
@@
 	\begin{abstract}
-	Modern recording techniques allow neuroscientists to study multiple neural populations over extended time periods in a large-scale, and the relationships within and between populations are summarized by low-dimensional latent vectors.
+	Modern neural recording techniques allow neuroscientists to observe the spiking activity of many neurons simultaneously.
+Although previous work has illustrated how activity within and between known populations of neurons can be summarized by low-dimensional latent vectors, in many cases what determines a discrete population may be unclear.
+Neurons differ in their anatomical location, but, also, in their cell types and response properties.
 When the neural activities are globally nonlinear, using single population analysis is inappropriate.
 However, defining the populations is usually difficult and  wrong cluster assignments will lead to bias in latent structure inferences.
-To tackle this challenge, we develop a clustering method based mixture of Poisson dynamic factor model.
-The number of cluster is treated as a parameter in mixture of finite mixtures (MFM) model, and the posteriors are sampled by a MCMC algorithm.
-To sample the posteriors efficiently, we approximate the full conditional distribution of latent state by Gaussian and approximate the marginal likelihood by making use of the Poisson-Gamma conjugacy.
-We further apply our method to neuropixel data (and hippocampus data) for illustration.
+To tackle this challenge, here we develop a clustering method based on a mixture of Poisson dynamic factor model with the number of cluster is treated as a parameter in mixture of finite mixtures (MFM) model.
+To sample efficiently from the posteriors, we approximate the full conditional distribution of latent state by Gaussian and approximate the marginal likelihood by making use of the Poisson-Gamma conjugacy.
+We apply our method to multi-region recordings, neuropixels recordings for illustration.
+This approach may be a useful tool for clustering neurons based on function.
 	\end{abstract}
@@
 
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=1\textwidth]{figure1.png}
+		\caption{\textbf{Overview of the model. A.} Briefly describe the neuroscience motivation. \textbf{B.} The graphical model}
+	\end{figure}
+
 	\section{Method}
@@
 
-	\subsection{Poisson Dynmic Factor Model}
+	\subsection{Poisson Dynamic Factor Model}
 	Denote the observed spike-count of neuron $i \in \{ 1,\ldots,N\}$ at
@@
 To facilitate clustering, we re-parametrize the regular Poisson linear dynamical system (PLDS) model \citep{Macke2011} to separate the mean log-firing-rate out.
-Assume neurons are independently Poisson distributed, conditional on the
+Assume neural responses are independently Poisson distributed, conditional on the
 	low-dimensional latent state $\bm{x}_{t}^{(j)} \in \mathbb{R}^{p}$ as follows:
@@
 	\end{align*}
-	, with $\bm{c}_i\sim N(\bm{0},\bm{I}_p)$. We further assume the intercept $\mu_t^{(j)}$ and the latent state $\bm{x}^{(j)}_t$ progress linearly with Gaussian noise as
+	, with $\bm{c}_i\sim N(\bm{0},\bm{I}_p)$. We further assume the intercept $\mu_t^{(j)}$ and the latent state $\bm{x}^{(j)}_t$ evolve linearly over time with Gaussian noise as
 	\begin{align*}
@@
 
-	The bias term ($\bm{b}^{(j)}$) in the latent state is constant across time, unlike ind PLDS, which is  $\bm{b}_t^{(j)}$.
+	The bias term $\bm{b}^{(j)}$ in the latent state is constant across time, unlike ind PLDS, which is  $\bm{b}_t^{(j)}$.
 When the bias term changes across the time, the update order of $\bm{A}^{(j)}$ and $\bm{b}_t^{(j)}$ will influence the results (usually, the $\bm{A}^{(j)}$ is updated first).
@@
 	\subsection{Model Global Non-linearity by Clustering}
-	There were a rich research results for single PLDS model, but it provides only a global linear model to represent the data in a lower dimensional subspace, which makes the application scope limited.
+
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=1\textwidth]{figure2.jpg}
+		\caption{\textbf{Global non-linearity by multiple clusters} \textcolor{red}{(The labels are known now.
+The purpose is to check MCMC do fit well.)}Simulate 10 clusters without specifying linear dynamics, with 5 neurons each.
+The dimension in each cluster is $p=1 (q=2)$.
+All the fitted posteriors are averages from iteration 500 to 1000.
+\textbf{A.} The trace plot of log-likelihood per spike.
+\textbf{B.} The fitted mean firing rate.
+\textbf{C.} The fitted latent vectors in the first 3 clusters.\textbf{D.} and \textbf{E.} Select 3/4 and 1/2 of the data as training in a speckled pattern, and replicate the sampling 50 times in each case.
+With each training sample, fit two models: 1) 10-cluster model: fit every single model for each, with $p=1$ (true value) in each and 2) 1-cluster model: fit a single model for all neurons, with $p$ selected by cross-validation.
+\textbf{D.} The held-out log-likelihood per spike decrease for less training samples.
+With the same training sample, 10-cluster model performs better than 1-cluster model.
+The selected $p$ for 1-cluster model is shown at the top.
+\textbf{E.} The differences of the held-out log-likelihood per spike between 10-cluster model and 1-cluster model.
+The less the training data, the more significant the improvement by fitting model cluster-wise.}
+	\end{figure}
+
+%	There were a rich research results for single PLDS model, but it provides only a global linear model to represent the data in a lower dimensional subspace, which makes the application scope limited.
 When the input space is nonhomogeneous, a large dimension of latent state is needed and this may lead to the overfitting and poor performance.
 Mixture of PLDS/ PDFM models allow us to partition the input space into clusters and can therefore capture global nonlinearity by combining local linear models.
-
-	To show the connection between the proposed model and PLDS parametrization, we simulate the data using the PLDS model, although these two are equivalent after some algebra.
+%	
+%	To show the connection between the proposed model and PLDS parametrization, we simulate the data using the PLDS model, although these two are equivalent after some algebra.
 Denote the common PLDS model as $\log \lambda_i = \delta_i\bm{1}_T + \bm{G}^{(j)}\bm{d}_i$, where $d_i\in \mathbb{R}^q$.
@@
 $p=q+1$.
-
-	We first simulated three neural populations, with 10 neurons in each.
+%	
+%	We first simulated three neural populations, with 10 neurons in each.
 We set $q=2$ for each cluster and the recording length is T=1000.
@@
 The $\bm{Q}^{(j)}$ are $10^{-3}\bm{I}_2$, $10^{-4}\bm{I}_2$ and $10^{-5}\bm{I}_2$.
-
-	The Figure \ref{fig:1a} shows the trace plot for the log-likelihood per spike, up to 10,000 iterations.
+%	
+%	The Figure \ref{fig:1a} shows the trace plot for the log-likelihood per spike, up to 10,000 iterations.
 The plot shows (Figure \ref{fig:1b}) that the convergence achieved after several steps.
@@
 Notice that the PLDS parametrization also doesn't have the unique solution, we let $\bm{G}^{(j)}$ has mean zeros columns and is orthogonal.
-
-	Then we held out $1/4$ and $1/2$ data as test set in a “speckled” pattern, i.e.
+%	
+%	Then we held out $1/4$ and $1/2$ data as test set in a “speckled” pattern, i.e.
 randomly select subset of data for each neuron as held-out dataset.
@@
 This suggests that doing clustering is necessary.
+%	
+%	\begin{figure}[h!]
+%		\centering
+%		\subfigure[Trace plot of  log-likelihood per spike]{\label{fig:1a}\includegraphics[width=0.5\textwidth]{2_llhd1.png}}%
+%		\subfigure[Trace plot of  log-likelihood per spike, truncated (1000)]{\label{fig:1b}\includegraphics[width=0.5\textwidth]{3_llhd2.png}}%
+%		
+%		\subfigure[true vs. fitted firing rate]{\label{fig:1c}\includegraphics[width=0.5\textwidth]{image001.jpg}}%
+%		\subfigure[true vs. fitted latent vectors]{\label{fig:1d}\includegraphics[width=0.5\textwidth]{LF.png}}%
+%		
+%		\subfigure[cross-validation: clustered vs. single population]{\label{fig:1e}\includegraphics[width=0.5\textwidth]{image003.jpg}}%
+%		\subfigure[trace plot for non-labeled case]{\label{fig:1f}\includegraphics[width=0.5\textwidth]{image004.jpg}}%
+%		\caption{Simple case}
+%	\end{figure}
+
+	\subsubsection{Clustering}
 
@@
 		\centering
-		\subfigure[Trace plot of  log-likelihood per spike]{\label{fig:1a}\includegraphics[width=0.5\textwidth]{2_llhd1.png}}%
-		\subfigure[Trace plot of  log-likelihood per spike, truncated (1000)]{\label{fig:1b}\includegraphics[width=0.5\textwidth]{3_llhd2.png}}%
-
-		\subfigure[true vs. fitted firing rate]{\label{fig:1c}\includegraphics[width=0.5\textwidth]{image001.jpg}}%
-		\subfigure[true vs. fitted latent vectors]{\label{fig:1d}\includegraphics[width=0.5\textwidth]{LF.png}}%
-
-		\subfigure[cross-validation: clustered vs. single population]{\label{fig:1e}\includegraphics[width=0.5\textwidth]{image003.jpg}}%
-		\subfigure[trace plot for non-labeled case]{\label{fig:1f}\includegraphics[width=0.5\textwidth]{image004.jpg}}%
-		\caption{Simple case}
+		\includegraphics[width=1\textwidth]{figure3.jpg}
+		\caption{\textbf{Clustering} The same simulation setting as in Figure 2, but remove the labels.
+\textcolor{red}{(The cluster 3 and 5 have nearly the same spiking patterns, because of the "bad luck"...
+May resample with another seed)}.
+The fitting is conducted using all and half data as training.
+\textbf{A.} The trace plot of training log-likelihood per spike and number of cluster.
+All the posterior results are averages from iteration 500 to 1000.
+\textbf{B.} The fitted firing rate.
+\textcolor{red}{(This is different from figure 2, because the labels are removed)} \textbf{C.} The similarity matrix.
+\textbf{D.} and \textbf{E.} Compare the held-out (1/2 data) likelihood for 3 fittings: 1) 10-cluster model: use the true labels and fit the model in each cluster with $p=1$; 2) cluster-on model: unknown labels but turn on the clustering, with $p=1$ (CV) and 3) 1-cluster model: single population model, with $p=8$ (CV).
+\textbf{D.} The trace plot of held-out log-likelihood per spike.
+The traces for 10-cluster and cluster-on are close, but more variation (occasional "drop-down") in cluster-on due to estimation of unknown labels.
+\textbf{E.} The histograms of posterior held-out log-likelihood per spike for 3 models.
+10-cluster and cluster-on models are close, and they perform better than 1-cluster model.}
 	\end{figure}
 
-	\subsubsection{Clustering}
-	Then use the same setting, we remove the label and use the mixture model to do the clustering by MFM.
+%	Then use the same setting, we remove the label and use the mixture model to do the clustering by MFM.
 The prior of the cluster number is $K\sim Geometric(0.2)$.
 Figure \ref{fig:1f} shows the trace of label $z_i$ for each neuron for the first 100 iterations.
-
-	\begin{figure}[h!]
-		\centering
-		\subfigure[Firing rate]{\label{fig:2a}\includegraphics[width=0.4\textwidth]{image005.jpg}}%
-		\subfigure[trace plot for cluster number]{\label{fig:2b}\includegraphics[width=0.4\textwidth]{image006.jpg}}%
-		\subfigure[similarity matrix]{\label{fig:2c}\includegraphics[width=0.4\textwidth]{image007.jpg}}%
-		\caption{More complicated case: more variety of amplitude \& weaker signal}
-	\end{figure}
-
-	In this toy simulation example, the signal is strong and the spiking amplitudes of neurons in each population are similar.
+%	
+%	\begin{figure}[h!]
+%		\centering
+%		\subfigure[Firing rate]{\label{fig:2a}\includegraphics[width=0.4\textwidth]{image005.jpg}}%
+%		\subfigure[trace plot for cluster number]{\label{fig:2b}\includegraphics[width=0.4\textwidth]{image006.jpg}}%
+%		\subfigure[similarity matrix]{\label{fig:2c}\includegraphics[width=0.4\textwidth]{image007.jpg}}%
+%		\caption{More complicated case: more variety of amplitude \& weaker signal}
+%	\end{figure}
+%	
+%	In this toy simulation example, the signal is strong and the spiking amplitudes of neurons in each population are similar.
 This might be too simplified for real situation.
@@
 
-	\section{Application}
+	\section{Application: Neuropixels data}
 	\label{app}
 
-	\subsection{Neuropixels data}
-	\answerTODO{}: brief introduction of Neuropixels dataset.
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=1\textwidth]{figure4.jpg}
+		\caption{\textbf{Application in Neuropixels data.} We first choose a one minute recordings from spontaneous, with bin size in $0.1s$ \textcolor{red}{(29.8 to 89.8)}.\textbf{A.} The trace plots of log-likelihood per spike and number of cluster \textcolor{red}{(Maybe again overlaid by 1/2-training results?)}.
+All the following results are averages from iteration 500 to 1000.
+\textbf{B} The fitted firing rate.
+\textbf{C.} The histogram of posterior cluster number.
+\textbf{D.} We then hold out 1/2 of data in the speckled pattern and fit 3 models: 1) cluster-on model: turn on clustering with $p=1$ (CV); 2) 1-cluster: single population with $p=2$ (CV) and 3) anatomy-cluster: fit models each based on 4 anatomical labels, with $p=(1,1,1,3)$ (CV).
+Doing clustering is the best.
+Anatomical labels give the "wrong" labels, in terms of the spiking pattern, which make things even worse.
+\textbf{E.} The similarity matrices when fitting four 1-min epochs: 1) spontaneous 1: 29.8s to 89.8s; 2) spontaneous 2: 1001.9s to 1061.9s; 3) natural movie: 2221.7s to 2281.7s and 4) drift grating: 1890.9s to 1950.9s.
+All are sorted by mode of posterior cluster index within each anatomical region.
+The patterns found in spontaneous are robust and there are a large confusion between VISp and CA1.
+The confusion between VISp and CA1 also exists in natural movie, but decrease a lot.
+However, the overall confusions become larger.
+When exposing to the drift grating movie, the confusions among all four regions decrease even more.}
+	\end{figure}
 
-	The Neuropixels dataset contains recording of neural activities in different brain regions of mouse.
+%	\answerTODO{}: brief introduction of Neuropixels dataset.
+%	
+%	The Neuropixels dataset contains recording of neural activities in different brain regions of mouse.
 Here, we only consider neurons with SNR > 3 and the spiking counts > 1000.
@@
 The bin size is 0.1s and truncate the data up to 1000 steps (100s recording) for convenience.
-
-	Figure \ref{fig:3a} shows the spiking counts of these 46 neurons.
+%	
+%	Figure \ref{fig:3a} shows the spiking counts of these 46 neurons.
 Here we first fit model using all data, with $p=1$ and $K\sim Geometric(0.3)$.
@@
 These 3 nominated populations by anatomy tend to form around 10 (sub)-clusters.
-
-	The Figure \ref{fig:3d} and Figure \ref{fig:3e} show the trace of log-likelihood per spike.
+%	
+%	The Figure \ref{fig:3d} and Figure \ref{fig:3e} show the trace of log-likelihood per spike.
 The fitted mean firing rate is shown in Figure \ref{fig:3f}.
@@
 Generally, the clustering results are consistent with the anatomical assignment, although there are some tiny confusions between VISam and VPM.
-
-	However, this is not the case when the mouse is in the experiment.
+%	
+%	However, this is not the case when the mouse is in the experiment.
 We choose the same set of neurons but use recordings when the mouse are watching drift gratings.
@@
 The Figure \ref{fig:3i} and Figure \ref{fig:3j} show the fitted firing rate and similarity matrix, again averaging from iteration 2500 to 5000.
-
-	When comparing the observed spiking activities, it's fairly hard to see differences when the mouse is exposed to the stimuli and not.
+%	
+%	When comparing the observed spiking activities, it's fairly hard to see differences when the mouse is exposed to the stimuli and not.
 However, the posterior similarity matrix in the experiment shows that there's more confusion between LP and VISam, compared to the spontaneous response.
 This suggests that the functional connection pattern between neurons in different areas when the stimuli are changed.
-
-	\textcolor{red}{\textbf{I tried to held-out half of the data in a speckled pattern and fit the model with clustering on and off (single population).
+%	
+%	\textcolor{red}{\textbf{I tried to held-out half of the data in a speckled pattern and fit the model with clustering on and off (single population).
 The dimension is selected by held-out likelihood, which is p=1 for clustering on and p=2 for single population analysis.
 The trace plots of the held-out log-likelihood per spike (starting from iteration 2) shows that doing clustering doesn’t improve things (although they are very close)… Maybe because “turning clustering on” introduce too much variance… The benefit of clustering will pop out if we use the deterministic algorithm, such as VB mentioned.}}
-
-	\begin{figure}[h!]
-		\centering
-		\subfigure[spontaneous: spike counts]{\label{fig:3a}\includegraphics[width=0.5\textwidth]{image008.jpg}}%
-		\subfigure[spontaneous: cluster number trace]{\label{fig:3b}\includegraphics[width=0.35\textwidth]{image009.jpg}}%
-		\subfigure[spontaneous: cluster number histogram]{\label{fig:3c}\includegraphics[width=0.35\textwidth]{image010.jpg}}%
-
-		\subfigure[spontaneous: llhd/spk trace]{\label{fig:3d}\includegraphics[width=0.3\textwidth]{image001.png}}%
-		\subfigure[spontaneous: llhd/spk trace, truncated]{\label{fig:3e}\includegraphics[width=0.3\textwidth]{image002.png}}%
-		\subfigure[spontaneous: true vs. fitted firing rate]{\label{fig:3f}\includegraphics[width=0.3\textwidth]{image011.jpg}}%
-		\subfigure[spontaneous: similarity matrix]{\label{fig:3g}\includegraphics[width=0.3\textwidth]{image012.jpg}}%
-
-		\subfigure[drift-grating: cluster number trace]{\label{fig:3h}\includegraphics[width=0.4\textwidth]{12_numTrace.png}}%
-		\subfigure[drift-grating: true vs. fitted firing rate]{\label{fig:3i}\includegraphics[width=0.4\textwidth]{16_fitFR.png}}%
-		\subfigure[drift-grating: similarity matrix]{\label{fig:3j}\includegraphics[width=0.4\textwidth]{14_simPixel.png}}%
-		\caption{Neuropixel}
-	\end{figure}
-
-	\subsection{Hippocampus data}
-
-	\textcolor{red}{\textbf{I prefer just give one application.} Because 1) there's a page limit (9 pages excluding ref \& appendix in 2021, and now it seems we will exceed the limit), and 2) the V1 data is enough to illustrate the idea If we really need to add the second example, we should tell the story in another angle}.
+%	
+%	\begin{figure}[h!]
+%		\centering
+%		\subfigure[spontaneous: spike counts]{\label{fig:3a}\includegraphics[width=0.5\textwidth]{image008.jpg}}%
+%		\subfigure[spontaneous: cluster number trace]{\label{fig:3b}\includegraphics[width=0.35\textwidth]{image009.jpg}}%
+%		\subfigure[spontaneous: cluster number histogram]{\label{fig:3c}\includegraphics[width=0.35\textwidth]{image010.jpg}}%
+%		
+%		\subfigure[spontaneous: llhd/spk trace]{\label{fig:3d}\includegraphics[width=0.3\textwidth]{image001.png}}%
+%		\subfigure[spontaneous: llhd/spk trace, truncated]{\label{fig:3e}\includegraphics[width=0.3\textwidth]{image002.png}}%
+%		\subfigure[spontaneous: true vs. fitted firing rate]{\label{fig:3f}\includegraphics[width=0.3\textwidth]{image011.jpg}}%
+%		\subfigure[spontaneous: similarity matrix]{\label{fig:3g}\includegraphics[width=0.3\textwidth]{image012.jpg}}%
+%		
+%		\subfigure[drift-grating: cluster number trace]{\label{fig:3h}\includegraphics[width=0.4\textwidth]{12_numTrace.png}}%
+%		\subfigure[drift-grating: true vs. fitted firing rate]{\label{fig:3i}\includegraphics[width=0.4\textwidth]{16_fitFR.png}}%
+%		\subfigure[drift-grating: similarity matrix]{\label{fig:3j}\includegraphics[width=0.4\textwidth]{14_simPixel.png}}%
+%		\caption{Neuropixel}
+%	\end{figure}
 
